### 2017-03-28 paper.tex
Update on Overleaf.

\documentclass[a4paper,12pt]{article}
\usepackage[french]{babel}
\usepackage{amsfonts}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{latexsym}

\usepackage[margin=1.5cm]{geometry}

%%%%%
% variable to include comments or not in the compilation ; set to 1 to include
\def \draft {1}
%\def \draft {0}

% writing utilities

% comments and responses
%  -> use this comment to ask questions on what other wrote/answer questions with optional arguments (up to 4 answers)
\usepackage{xparse}
\usepackage{ifthen}
\DeclareDocumentCommand{\comment}{m o o o o}
{\ifthenelse{\draft=1}{
    \textcolor{red}{\textbf{C : }#1}
    \IfValueT{#2}{\textcolor{blue}{\textbf{A1 : }#2}}
    \IfValueT{#3}{\textcolor{ForestGreen}{\textbf{A2 : }#3}}
    \IfValueT{#4}{\textcolor{red!50!blue}{\textbf{A3 : }#4}}
    \IfValueT{#5}{\textcolor{Aquamarine}{\textbf{A4 : }#5}}
 }{}
}

% todo
\newcommand{\todo}[1]{
\ifthenelse{\draft=1}{\textcolor{red!50!blue}{\textbf{TODO : \textit{#1}}}}{}
}

\begin{document}

\title{
% title propositions
%Space matters
For a Systematic Sensitivity Analysis of Spatial Meta-parameters in Models of Simulation
}

\date{}

\maketitle

\section{Space does matter in simulation modeling}

generalities about simulation modeling of social facts and initial conditions

the choice of working (as an illustration) on Schelling segregation model

\subsection{Literature review on why we believe space actually does matter}

Pourquoi est-ce qu'on pense que forme grille spatiale peut influencer Schelling ?

Spatio-temporal model of a social fact, so spatio-temporal patterns and dynamics are important. Spatial grid as an input variable whose impact on the model behaviour and output should be systematically tested while looking for general conclusions : e.g. effect of the size of the neighbourhood on the decision to move.

clementine \& marion already started to write something up on that topic

\subsection{With a simple Schelling model}

simple illustration based on Netlogo Segregation model (Florent)
For now on: one agent per cell, some cells cannot host agents, Moore neighbourhood. The idea is to illustrate how this simple variation in the population density of a cell (0 or 1) will impact their neighbourhood and therefore the probability to host blue or red agents.

\subsection{In reality...}

(if enough material) : what type of cities actually produced more / less segregation?
See if there is some empirical findings the link between the shape of a city and its capability to produce segregated residential patterns.

end of section : we got frustrated and wanted to systematically study the influence of initial spatial configuration on how models (here Schelling) does behave

this is why...

%%%%%%%%%%%%%%%
\section{Research framework}
%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%
\subsection{But in reality space is more complicated than basic Schelling model}

We want to consider the fact that there are different morphologies of cities (so different topological relations between 'neighbouring' cells) and various level of densities at the block$/$residential building level (so for the same definition of the scope of a neighbourhood, not the same number of agents nor the same potentiality to host a new agent).

we worked with Schelling model extended with heterogeneous cell capacities: find the references of previous work where spatial heterogeneity and densities have been studied.

basic description of the model that we use, of the parameters and behaviour of agents that we simulate and finally of the output indicators.

Our output indicators: the level of segregation at equilibrium. Global indicator and$/$or also Local Moran's (number of significant clusters)?

What we are interested in is how the relation between the input variables and the output indicators may vary according to the spatial grid we're using. Not sure we have to mention it here, but just as a reminder for us.

We have to be explicit about the relation between density and neighbourhood: if neighbourhood is describe as a number of agents, or by a particular spatial scope that we vary. In monocentric cities with highly populated cells, large neighbourhoods won't make a difference as most of the cells won't have the capacity to host migrants. Etc.

Of course, we have to define what we call an equilibrium: static, dynamic (agents are still moving as there are still some unhappy ones but the segregation patterns$/$ values are not evolving anymore $\rightarrow$ we fix this time scope as being $n$ time steps).

%%%%%%%%%%%%%%%
\subsection{And studying influence of space requires appropriate tools}

(which is maybe why scholars tend not to do that : it's hard)
our research framework : many combinations of parameters and many urban forms, plus random components (asynchronous decision making by agents which requires to order the agents...) : OpenMole to the rescue!

%%%%%%%%%%%%%%%
\subsection{Generation of various urban forms}

Juste's density generation model and comparison with Florent's categories : 
now everything is on stage for the results

%%%%%%%%%%%%%%%
\section{How space matters}

Again we have two points of analysis: the range and variability of segregation levels per city types that we obtained and how the shape of cities impact on the model behaviour, i.e. the relation between input and output variables.

%%%%%%%%%%%%%%%
\subsection{Shape of cities and segregation capability}

We can imagine a figure where we would have one example of city of each type and two sizes of neighbourhood scope for instance (everything else being equal), and we see the resulting distribution of green and blue agents at equilibrium.

Plus descriptive statistics (factor analysis? regression?)

%%%%%%%%%%%%%%%
\subsection{Systematic analysis}

\paragraph{Systematic Exploration}

The relation between these values and the value of the input parameters $\rightarrow$ phase diagram better than a regression analysis for instance as we suppose the relations are not continuous. We want to see if there are (and if yes, where in the parameter space) the ruptures and clusters of same behaviours.

z(x,y) en fonction des formes de grilles initiales, z étant l'indice de segregation (Moran global ou nombre de clusters significatifs obtenus avec un local moran?) et (x,y) nos inputs (proportion of similar people wanted and size of the neighbourhood)

\paragraph{Comparing Phase Diagrams}

Problem: we have as many phase diagrams than we have spatial grids, so we cannot define the phases qualitatively. We need quantitative procedures.

Potential methods to explore: anisotropic spatial segregation index (number of clusters found and where in the parameter spaces? but may be weak...), earth Movers Distance (i.e. distance between probability distributions), comparison of transition matrices of the dynamic associated to the potential...
See also: http://mck.riks.nl/ to get some ideas.

\paragraph{Results}

Et on met le mod{\`e}le plus sioux avec toutes les réplications dans plein de configurations de villes d'une part : diagramme de phase par rapport aux meta-param{\`e}tres(=indicateurs de forme, dans un plan principal = 80\% variance expliqu{\'e}e) de la variabilit{\'e} des diagrammes de phase selon critères de comparaison quantitatifs par rapport à une r{\'e}f{\'e}rence.

%%%%%%%%%%%%%%%
\subsection{Application: the Sugarscape model}

To illustrate the generic nature of our methodology, we apply it here to an other model implying a spatialized distribution of resources for which agents compete: the ``Sugarscape'' model, introduced in~\cite{epstein1996growing}. We extend the implementation with agents wealth distribution of~\cite{li2009netlogo}.

%%%%%%%%%%%%%%%
\subsection{Application: SimpopNet}

More complicated initial conditions such as a synthetic transportation network also fully enter our framework, even these may be more difficult to generate or explore. We illustrate this other type of application by exploring the sensitivity of the \emph{SimpopNet} model to the initial network. This model introduced in \cite{schmitt2014modelisation} aims at simulating the co-evolution of cities and transportation networks. The implementation of the model for which the behavior is known corresponds to a fixed stylized initial network. We modulate it here in a basic way, within a range of plausibility given by network indicators.

%%%%%%%%%%%%%%%
\section{Discussion}
%%%%%%%%%%%%%%%

Puis discussion : space matters (les valeurs prises par les indicateurs segrégations sont pas les mêmes) et ensuite en ouverture implications pour planning ?

%%%%%%%%%%%%%%%
% Biblio
%%%%%%%%%%%%%%%

\bibliographystyle{apalike}
\bibliography{biblio}

\end{document}

### 2017-04-03 paper.tex
.

@@
 %%%%%%%%%%%%%%%
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/fixed_phase_diagram}
+\caption{Phase diagram of the sugarscape model (final smoothed Gini indicator), for a basic bicentric spatial setup, with varying parameters population $N$, minimal sugar reload $s_{min}$ and maximal sugar reload $s_{max}$.}
+\label{fig:sugarscape-phasediagram}
+\end{figure}
+%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%
+\begin{figure}
+\centering
+%\includegraphics[width=\textwidth]{}
+\caption{Phase diagrams varying across the container morphological space.
+Around 10000 realistic ressources distribution were generated, and used as container landscape (patches sugar capacity) for the sugarscape model.
+Spatial configurations were classified by reducing the dimension from four morphological indicators to two principal components ($\simeq$80\% of variance), for which the resulting space was gridded into 10x10 bins.
+We show here for each cell the average phase diagram along $(N,s_{min})$, taken at $s_{max}=200$.}
+\label{fig:sugarscape-meta}
+\end{figure}
+%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%
+\begin{figure}
+\centering
+%
+\caption{Meta phase diagram, i.e.
+phase diagram of phase diagram behavior, along parameters of spatial configuration generation ($\alpha$,$n_D$,$\beta$,$N_G$) \todo{described these somewhere}, aggregated for each as an Earth Mover Distance to a fixed reference (phase diagram of Fig.~\ref{fig:sugarscape-phasediagram}).}
+\label{fig:sugarscape-meta}
+\end{figure}
+%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%
 \subsection{Application: SimpopNet}

### 2017-04-03 paper.tex
Update on Overleaf.

@@
 \usepackage[margin=1.5cm]{geometry}
+
+\usepackage{graphicx}
 
@@
 \caption{Meta phase diagram, i.e.
-phase diagram of phase diagram behavior, along parameters of spatial configuration generation ($\alpha$,$n_D$,$\beta$,$N_G$) \todo{described these somewhere}, aggregated for each as an Earth Mover Distance to a fixed reference (phase diagram of Fig.~\ref{fig:sugarscape-phasediagram}).}
+phase diagram of phase diagram behavior, along parameters of spatial configuration generation ($\alpha$,$n_D$,$\beta$,$N_G$) (described these somewhere), aggregated for each as an Earth Mover Distance to a fixed reference (phase diagram of Fig.~\ref{fig:sugarscape-phasediagram}).}
 \label{fig:sugarscape-meta}

### 2017-04-03 paper.tex
.

--- 2017-03-28 paper.tex
@@
 
+Basic parameters if the sugarscape model on which phase diagrams are computed are: agent population $N$, agent minimal ressource $s_{min}$ and agent maximal ressource $s_{max}$.
+
+%%%%%%%%%%%%%%%
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/fixed_phase_diagram}
+\caption{Phase diagram of the sugarscape model (final smoothed Gini indicator), for a basic bicentric spatial setup, with varying parameters population $N$, minimal sugar reload $s_{min}$ and maximal sugar reload $s_{max}$.
+For each parameter point, average is taken over 50 repetitions.}
+\label{fig:sugarscape-phasediagram}
+\end{figure}
+%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%
+\begin{figure}
+\centering
+\includegraphics[width=\textwidth]{figures/phasediagrams_pca_maxSugar200}
+\caption{Phase diagrams varying across the container morphological space.
+50 realistic ressources distributions were generated, and used as container landscape (patches sugar capacity) for the sugarscape model, and phase diagrams were computed as in Fig.~\ref{fig:sugarscape-phasediagram} for each one.
+Spatial configurations were classified by reducing the dimension from four morphological indicators to two principal components ($\simeq$80\% of variance), for which the resulting space was gridded into 5x5 bins.
+We show here for each cell the average phase diagram along $(N,s_{min})$, taken at $s_{max}=200$.
+Other values of $s_{max}$ provide similar qualitative variations.
+The two missing cells correspond to non-feasible cells in the morphological space for the given generating model and parameters.}
+\label{fig:sugarscape-meta}
+\end{figure}
+%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%
+\begin{figure}
+\centering
+%
+\caption{Meta phase diagram, i.e.
+phase diagram of phase diagram behavior, along parameters of spatial configuration generation ($\alpha$,$n_D$,$\beta$,$N_G$) (described these somewhere), aggregated for each as an Earth Mover Distance to a fixed reference (phase diagram of Fig.~\ref{fig:sugarscape-phasediagram}).}
+\label{fig:sugarscape-meta}
+\end{figure}
+%%%%%%%%%%%%%%%
+
 %%%%%%%%%%%%%%%

### 2017-04-07 Raimbault_Cottineau_LeTexier_LeNechet_Reuillon.tex
.

\documentclass[11pt]{article}
\date{}
\renewcommand\abstractname{\fontsize{14pt}{0}\textbf{Abstract}\selectfont}

\usepackage[left=25mm, right=25mm, top=25mm, bottom=25mm, includehead=false, includefoot=false]{geometry}

\usepackage{graphicx}
\usepackage{url}
\usepackage[round,semicolon]{natbib}  % Citation styles https://www.sharelatex.com/learn/Natbib_citation_styles
\bibliographystyle{humannat}
\renewcommand{\bibsection}{}
\renewcommand{\bibhang}{\setlength{-1px}}

\usepackage{authblk} % For author lists
\renewcommand\Authfont{\fontsize{11}{1}\selectfont}
\renewcommand\Affilfont{\fontsize{9}{1}\selectfont}

\renewcommand*\footnoterule{}

\usepackage[table]{xcolor}
\usepackage[parfill]{parskip} % Line between paragraphs
\usepackage{amsmath}
\pagenumbering{arabic}

\usepackage{sectsty}
\allsectionsfont{\sffamily}

\usepackage[pdftex]{hyperref} 
\hypersetup{pdfborder={0 0 0} }

% **************  TITLE AND AUTHOR INFORMATION **************

\title{\sffamily\fontsize{16}{0}\textbf{Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?}}

\author[1]{Cl{\' e}mentine Cottineau\thanks{}}
\author[2,3]{Juste Raimbault}
\author[3]{Florent Le N{\' e}chet}
\author[4]{Marion Le Texier}
\author[2,5]{Romain Reuillon}
\affil[1]{Centre for Advanced Spatial Analysis, University College London, UK}
\affil[2]{UMR 8504 G{\' e}ographie-cit{\' e}s, Paris, France}
\affil[3]{Laboratoire Ville Mobilit{\' e} Transport, Universit{\' e} Paris-Est, France}
\affil[4]{UMR 6266 IDEES, Universit{\' e} de Rouen, France}
\affil[5]{Institut des Syst{\` e}mes Complexes Paris Ile-de-France, France}

\affil[*]{\texttt{Email: c.cottineau@ucl.ac.uk }}

\begin{document}

\maketitle

% **************  ABSTRACT  **************

\begin{abstract}
\noindent
\setlength{\parindent}{0pt}
Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust. Sensitivity analysis usually include the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes. However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interaction of agents and of their interactions with the environment. In this contribution, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameter to create density grids used in spatial simulation models. We show, with the example of two very classical agent-based models (Schelling's models of segregation and Sugarscape) that the effect of space in simulation is significant, and sometimes even larger than parameters themselves. We do so using high performance computing in a very simple and straightforward open-source workflow.
$ $ \\ {\bf Keywords:} Space, Initial conditions, Sensitivity, ABM.

\end{abstract}

% **************  MAIN BODY OF THE PAPER **************

\section{Introduction}

Simulation has been recognised and increasingly used by geographers to explore various geographical processes and problems within virtual laboratories. It appears as a very fruitful way to overcome the difficulty of analytic resolution of many spatial models developed in the past, and to explore the possible trajectories of social and ecological systems in space and time. The specificity of geographical models compared with models from other social sciences is usually regarded as the way geographers consider space and spatial interactions, driven by an explicit interest in the way space influences the outcomes of the model. Geographers are indeed concerned about understanding and modelling how space plays a role in social interactions and environmental processes, and whether its action is placed-based or place-neutral. We think simulation can become a very good tool to achieve this, provided that models include relevant spatial description and modelling, behavioural rules that take space into account, and provided model evaluation  stresses the sensitivity of output variations to the way space is modelled. This communication focuses on a way to do  so: by using a sensitivity method to test the model's outcomes to initial spatial conditions.

% **************

\section{Why spatial patterns are expected to impact social simulations}

\subsection{Because actual cities are not regular grids}
Although this may seem obvious, cities are not regular grids of isotropic densities. Modelling may be about abstracting features to highlight processes in a way which is simpler than reality. However, the uniform grid which represents space in most simulation models is potentially not enough to represent urban processes because density and accessibility have environmental, economic and social consequences. An intermediate and more meaningful way of abstracting space might thus be to consider, not the peculiarities of every city, but their broad density structures. In Europe for example, one can find broad types of density distributions (\cite{LeNechet2015}). At initialisation, we then expect agents to be distributed in different patterns, which influence results in the long run. A first approach could be to evaluate the effects of the same set of social mechanisms, but modelled in different types of cities, for example monocentric, polycentric and discontinuous.

\subsection{Because actual agents are not random walkers}
We expect spatial patterns to influence model outcomes also because the agents' rule of action itself might depend on the spatial structure of the environment. Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density. For example, agents tend to create buffer zones if they are allowed network-based rather than isotropic movements (\cite{Banos2012}), while the vision / sensing mechanism is sensitive to the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}). Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.

% **************

\section{Method}

In this paper, we suggest tackling sensitivity to spatial conditions by generating a variety of density grids to feed into simulation models at initialisation, and exploring the sensitivity of the model outcomes to the variation of meta-parameters (i.e. parameters used to generate initial spatial conditions). The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters within a typical category of space (a monocentric case for example) and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric vs. polycentric for example) on the results of a given model (cf. figure \ref{fig:method}).

\begin{figure}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/SchemaMeta_1.png}
} \caption{General workflow of our method} \label{fig:method} \end{center} \end{figure} %

\subsection{Constructing a spatial generator}

Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}). In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$. This first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$. Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed. To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment of 100,000 units (cf. figure \ref{fig:spatialGen}).

\begin{figure}[htbp] \begin{center} 
\resizebox{0.9\textwidth}{!}{ 
	\includegraphics{figures/spatialGen.png}
} \caption{Four examples of grids produced by the spatial generator} \label{fig:spatialGen} \end{center} \end{figure} %

\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}

In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions: their concentration index, hierarchy index, centrality index and continuity index (cf. \cite{LeNechet2015}). A stochastic exploration of a Latin Hypercube Sampling of 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n$, $N$} gives a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions. These are further clustered into three classes of morphology: compact (e.g. Vienna), polycentric (Liege) and discontinuous (Augsburg) in order to evaluate the non-trivial effects of urban form on simulation results. We select 15 grids of each type to work with.

% ************** 
\section{Application cases}

We use our set of initial density grids in two spatial simulation models: Schelling and Sugarscape.

\subsection{Schelling's model of residential segregation}

Schelling's model consists in a abstract urban housing market where agents of different nature sense their environment, evaluate their satisfaction in terms of neighbourhood composition, and relocate if unsatisfied. It has been shown by \cite{Schelling1969} that even tolerant agents tended to produce segregated patterns because of the complexity of their local interactions. The main parameters of this model are the tolerance level (\% of agents similar to {\it ego}), the scope of sensing and the percentage of vacant spaces in the housing market. In addition, we are interested in testing the impact of the spatial distribution of housing capacity in this project, using the generated grids.

\subsection{Sugarscape model of resource extraction and population settlement}

Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population (\cite{EpsteinAxtell1996}). Agents of different vision scopes and different metabolisms harvest a self-regenerating resource available heterogeneously in the initial landscape, they settle and collect this resource, which leads some of them to survive and others to perish. The main parameters of this model are the number of agents, their minimal and maximal resource. In addition, we are interested in testing the impact of the spatial distribution of the resource in this project, using the generated grids. The outcome of the model is measured as a phase diagram of an index of inequality for ressource distribution (Gini index).

% ************** 
\section{Results}

We proceed to 4,500,000 simulation runs of the Schelling model (1000 parameter combinations x 45 density grids x 100 replications), using OpenMOLE to distribute the computation, and apply segregation measures to characterise the results. We find that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index. For Sugarscape, 2,500,000 simulations (1000 parameter points x 50 density grids x 50 replications) allow us to show that the model is more sensitive to space than to its parameter, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.

% **************

% \hfill \break
% \itshape{This is a sub, subheading}\normalfont

% \hfill\break

%
%
%\begin{table}[htp]
%
%\begin{center}
%\begin{tabular}{c c c c}
%\arrayrulecolor{black}
%\hline 
%This & Is & A & Table\\
%\arrayrulecolor{lightgray}
%\hline 
%\arrayrulecolor{black}
%Label & 0.1 & 0.2 & 0.3\\
%Label & 1.0 & 2.0 & 3.0\\
%\hline
%\end{tabular}
%\end{center}
%\label{first_table}
%\caption{This is a table caption}
%\end{table}%
%
%
%
%\begin{equation}
%a^2 + b^2 = c^2
%\tag*{Equation 1}
%\end{equation}

\section{Acknowledgements}

The authors acknowledge the funding of their institutions and the EPSRC project number EP/M023583/1.

\section{References}

\bibliography{spacematters}

\end{document}

### 2017-05-11 paper.tex
Update on Overleaf.

--- 2017-04-03 paper.tex
@@
-\documentclass[a4paper,12pt]{article}
-\usepackage[french]{babel}
-\usepackage{amsfonts}
-\usepackage{amsmath,amssymb,amsthm}
-\usepackage{latexsym}
+\documentclass[11pt]{article}
+\input{header.tex}
 
-\usepackage[margin=1.5cm]{geometry}
+\date{}
+\renewcommand\abstractname{\fontsize{14pt}{0}\textbf{Abstract}\selectfont}
+
+\usepackage[left=25mm, right=25mm, top=25mm, bottom=25mm, includehead=false, includefoot=false]{geometry}
 
 \usepackage{graphicx}
+\usepackage{url}
+\usepackage[round,semicolon]{natbib}  % Citation styles https://www.sharelatex.com/learn/Natbib_citation_styles
+\bibliographystyle{humannat}
+\renewcommand{\bibsection}{}
+\renewcommand{\bibhang}{\setlength{-1px}}
 
-%%%%%
-% variable to include comments or not in the compilation ; set to 1 to include
-\def \draft {1}
-%\def \draft {0}
+\usepackage{authblk} % For author lists
+\renewcommand\Authfont{\fontsize{11}{1}\selectfont}
+\renewcommand\Affilfont{\fontsize{9}{1}\selectfont}
 
-% writing utilities
+\renewcommand*\footnoterule{}
 
-% comments and responses
-%  -> use this comment to ask questions on what other wrote/answer questions with optional arguments (up to 4 answers)
-\usepackage{xparse}
-\usepackage{ifthen}
-\DeclareDocumentCommand{\comment}{m o o o o}
-{\ifthenelse{\draft=1}{
-    \textcolor{red}{\textbf{C : }#1}
-    \IfValueT{#2}{\textcolor{blue}{\textbf{A1 : }#2}}
-    \IfValueT{#3}{\textcolor{ForestGreen}{\textbf{A2 : }#3}}
-    \IfValueT{#4}{\textcolor{red!50!blue}{\textbf{A3 : }#4}}
-    \IfValueT{#5}{\textcolor{Aquamarine}{\textbf{A4 : }#5}}
- }{}
-}
+\usepackage[table]{xcolor}
+\usepackage[parfill]{parskip} % Line between paragraphs
+\usepackage{amsmath}
+\pagenumbering{arabic}
 
-% todo
-\newcommand{\todo}[1]{
-\ifthenelse{\draft=1}{\textcolor{red!50!blue}{\textbf{TODO : \textit{#1}}}}{}
-}
+\usepackage{sectsty}
+\allsectionsfont{\sffamily}
+
+\usepackage[pdftex]{hyperref} 
+\hypersetup{pdfborder={0 0 0} }
+
+% **************  TITLE AND AUTHOR INFORMATION **************
+
+\title{\sffamily\fontsize{16}{0}\textbf{Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?}}
+
+\author[1]{Cl{\' e}mentine Cottineau\thanks{}}
+\author[2,4]{Juste Raimbault}
+\author[3]{Marion Le Texier}
+\author[4]{Florent Le N{\' e}chet}
+\author[2,5]{Romain Reuillon}
+\affil[1]{Centre for Advanced Spatial Analysis, University College London, UK}
+\affil[2]{UMR 8504 G{\' e}ographie-cit{\' e}s, Paris, France}
+\affil[3]{UMR 6266 IDEES, Universit{\' e} de Rouen Normandie, France}
+\affil[4]{Laboratoire Ville Mobilit{\' e} Transport, Universit{\' e} Paris-Est, France}
+\affil[5]{Instit, ut des Syst{\` e}mes Complexes Paris Ile-de-France, France}
+
+\affil[*]{\texttt{, Email: c.cottineau@ucl.ac.uk }}
 
@@
 
-\title{
-% title propositions
-%Space matters
-For a Systematic Sensitivity Analysis of Spatial Meta-parameters in Models of Simulation
-}
-
-\date{}
-
 \maketitle
 
-\section{Space does matter in simulation modeling}
+% **************  ABSTRACT  **************
 
-generalities about simulation modeling of social facts and initial conditions
+\begin{abstract}
+\noindent
+\setlength{\parindent}{0pt}
+Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust.
+Sensitivity analysis usually include the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes.
+However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interaction of agents and of their interactions with the environment.
+In this contribution, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameter to create density grids used in spatial simulation models.
+We show, with the example of two very classical agent-based models (Schelling's models of segregation and Sugarscape) that the effect of space in simulation is significant, and sometimes even larger than parameters themselves.
+We do so using high performance computing in a very simple and straightforward open-source workflow.
+$ $ \\ {\bf Keywords:} Space, Initial conditions, Sensitivity, ABM.
 
-the choice of working (as an illustration) on Schelling segregation model
+\end{abstract},
 
-\subsection{Literature review on why we believe space actually does matter}
+% **************  MAIN BODY OF THE PAPER **************
 
-Pourquoi est-ce qu'on pense que forme grille spatiale peut influencer Schelling ?
+\section{Introduction}
 
-Spatio-temporal model of a social fact, so spatio-temporal patterns and dynamics are important.
-Spatial grid as an input variable whose impact on the model behaviour and output should be systematically tested while looking for general conclusions : e.g.
-effect of the size of the neighbourhood on the decision to move.
+Simulation has been recognised and increasingly used by geographers to explore various geographical processes and problems within virtual laboratories.
+It appears as a very fruitful way to overcome the difficulty of analytic resolution of many spatial models developed in the past, and to explore the possible trajectories of social and ecological systems in space and time.
+The specificity of geographical models compared with models from other social sciences is usually regarded as the way geographers consider space and spatial interactions, driven by an explicit interest in the way space influences the outcomes of the model.
+Geographers are indeed concerned about understanding and modelling how space plays a role in social interactions and environmental processes, and whether its action is placed-based or place-neutral.
+We think simulation can become a very good tool to achieve this, provided that models include relevant spatial description and modelling, behavioural rules that take space into account, and provided model evaluation  stresses the sensitivity of output variations to the way space is modelled.
+This communication focuses on a way to do  so: by using a sensitivity method to test the model's outcomes to initial spatial conditions.
 
-clementine \& marion already started to write something up on that topic
+\subsection{Why spatial patterns are expected to impact social simulations}
 
-\subsection{With a simple Schelling model}
+\paragraph{Because actual cities are not regular grids}
+Although this may seem obvious, cities are not regular grids of isotropic densities.
+Modelling may be about abstracting features to highlight processes in a way which is simpler than reality.
+However, the uniform grid which represents space in most simulation models is potentially not enough to represent urban processes because density and accessibility have environmental, economic and social consequences.
+An intermediate and more meaningful way of abstracting space might thus be to consider, not the peculiarities of every city, but their broad density structures.
+In Europe for example, one can find broad types of density distributions (\cite{LeNechet2015}).
+At initialisation, we then expect agents to be distributed in different patterns, which influence results in the long run \cite{Castellanoetal2009}.
+A first approach could be to evaluate the effects of the same set of social mechanisms, but modelled in different types of cities, for example monocentric, polycentric and discontinuous.
 
-simple illustration based on Netlogo Segregation model (Florent)
-For now on: one agent per cell, some cells cannot host agents, Moore neighbourhood.
-The idea is to illustrate how this simple variation in the population density of a cell (0 or 1) will impact their neighbourhood and therefore the probability to host blue or red agents.
+\paragraph{Because actual agents are not random walkers}
+We expect spatial patterns to influence model outcomes also because the agents' rule of action itself might depend on the spatial structure of the environment.
+Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.
+For example, agents tend to create buffer zones if they are allowed network-based rather than isotropic movements (\cite{Banos2012}), while the vision / sensing mechanism is sensitive to the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}).
+Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.
 
-\subsection{In reality...}
+\paragraph{Due to spatio-temporal path dependencies}
 
-(if enough material) : what type of cities actually produced more / less segregation?
-See if there is some empirical findings the link between the shape of a city and its capability to produce segregated residential patterns.
+\subsection{Objective}
 
-end of section : we got frustrated and wanted to systematically study the influence of initial spatial configuration on how models (here Schelling) does behave
+In this paper, we suggest tackling sensitivity to spatial conditions by generating a variety of density grids to feed into simulation models at initialisation, and exploring the sensitivity of the model outcomes to the variation of meta-parameters (i.e.
+parameters used to generate initial spatial conditions).
+The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters within a typical category of space (a monocentric case for example) and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric vs.
+polycentric for example) on the results of a given model.
 
-this is why...
+% **************
 
-%%%%%%%%%%%%%%%
-\section{Research framework}
-%%%%%%%%%%%%%%%
+\section{Previous attempts in the literature}
 
-%%%%%%%%%%%%%%%
-\subsection{But in reality space is more complicated than basic Schelling model}
+\subsection{In general}
 
-We want to consider the fact that there are different morphologies of cities (so different topological relations between 'neighbouring' cells) and various level of densities at the block$/$residential building level (so for the same definition of the scope of a neighbourhood, not the same number of agents nor the same potentiality to host a new agent).
+Paragraph not as detailed as the following one, but just to illustrate the fact that the issue is an important concern in different types of spatial simulation modelling approaches?
 
-we worked with Schelling model extended with heterogeneous cell capacities: find the references of previous work where spatial heterogeneity and densities have been studied.
+Eg. : effects of delineation and scale of cities in LUTI models \citep{Thomasetal2017}, etc.
 
-basic description of the model that we use, of the parameters and behaviour of agents that we simulate and finally of the output indicators.
+\subsection{For Schelling model in particular}
 
-Our output indicators: the level of segregation at equilibrium. Global indicator and$/$or also Local Moran's (number of significant clusters)?
+Recent attempts of running Schelling-like models beyond the chessboard grid have proved that migration decisions are not only dependent of what positions agents occupy relative to each other, but also of where they occupy these positions.
+\citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents are highest in a rectangular grid and lowest in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents.
+\citet{StaufferSolomon2007} introduce asymetric interactions and empty residences on a large and regular lattice.
+They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns.
+\citet{Singhetal2009} show that the segregation patterns obtained by Schelling for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.
+\citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents.
+They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile.
+\citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours.
+Interestingly, \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents).
+In this paper, we build on the contributions of these previous studies and go further by systematically measuring the impact of the city lattice on Schelling model behaviour so far only assessed through a small set of illustrative spatial structures.
 
-What we are interested in is how the relation between the input variables and the output indicators may vary according to the spatial grid we're using.
-Not sure we have to mention it here, but just as a reminder for us.
+\subsection{And for SugarScape?}
 
-We have to be explicit about the relation between density and neighbourhood: if neighbourhood is describe as a number of agents, or by a particular spatial scope that we vary.
-In monocentric cities with highly populated cells, large neighbourhoods won't make a difference as most of the cells won't have the capacity to host migrants.
-Etc.
+% **************
 
-Of course, we have to define what we call an equilibrium: static, dynamic (agents are still moving as there are still some unhappy ones but the segregation patterns$/$ values are not evolving anymore $\rightarrow$ we fix this time scope as being $n$ time steps).
+\section{Method}
 
-%%%%%%%%%%%%%%%
-\subsection{And studying influence of space requires appropriate tools}
+In this section, we detail the method developped to tackle simulation models sensitivity to initial spatial conditions. The general workflow of the method is illustrated in Figure \ref{fig:method}.
 
-(which is maybe why scholars tend not to do that : it's hard)
-our research framework : many combinations of parameters and many urban forms, plus random components (asynchronous decision making by agents which requires to order the agents...) : OpenMole to the rescue!
+\begin{figure}[htbp] \begin{center} 
+\resizebox{0.9\textwidth}{!}{ 
+	\includegraphics{figures/SchemaMeta_1.png}
+} \caption{General workflow of our method} \label{fig:method} \end{center} \end{figure} %
 
-%%%%%%%%%%%%%%%
-\subsection{Generation of various urban forms}
+\comment[MLT]{If we want to keep this figure, I propose that we link it more clearly to the following subsections, by numbering the different parts of the figure?
+By graphically showing on which parts of the figure the different subsections focus?
+Besides, we need to change it as we know want to go further than analysing the outputs qualitatively.}
 
-Juste's density generation model and comparison with Florent's categories : 
-now everything is on stage for the results
+\subsection{Constructing a spatial generator}
 
-%%%%%%%%%%%%%%%
-\section{How space matters}
+Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}).
+In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$.
+This first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$.
+Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed.
+To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment of 100,000 units (cf.
+figure \ref{fig:spatialGen}).
 
-Again we have two points of analysis: the range and variability of segregation levels per city types that we obtained and how the shape of cities impact on the model behaviour, i.e.
-the relation between input and output variables.
+\begin{figure}[htbp] \begin{center} 
+\resizebox{0.9\textwidth}{!}{ 
+	\includegraphics{figures/spatialGen.png}
+} \caption{Four examples of grids produced by the spatial generator} \label{fig:spatialGen} \end{center} \end{figure} %
 
-%%%%%%%%%%%%%%%
-\subsection{Shape of cities and segregation capability}
+\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}
 
-We can imagine a figure where we would have one example of city of each type and two sizes of neighbourhood scope for instance (everything else being equal), and we see the resulting distribution of green and blue agents at equilibrium.
+In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions: their concentration index, hierarchy index, centrality index and continuity index (cf.
+\cite{LeNechet2015}).
+A stochastic exploration of a Latin Hypercube Sampling of 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n$, $N$} gives a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions.
+These are further clustered into three classes of morphology: compact (e.g.
+Vienna), polycentric (Liege) and discontinuous (Augsburg) in order to evaluate the non-trivial effects of urban form on simulation results.
+We select 15 grids of each type to work with.
 
-Plus descriptive statistics (factor analysis? regression?)
+%cite{Gauvinetal2009}
 
-%%%%%%%%%%%%%%%
-\subsection{Systematic analysis}
-
-\paragraph{Systematic Exploration}
-
-The relation between these values and the value of the input parameters $\rightarrow$ phase diagram better than a regression analysis for instance as we suppose the relations are not continuous.
-We want to see if there are (and if yes, where in the parameter space) the ruptures and clusters of same behaviours.
-
-z(x,y) en fonction des formes de grilles initiales, z étant l'indice de segregation (Moran global ou nombre de clusters significatifs obtenus avec un local moran?) et (x,y) nos inputs (proportion of similar people wanted and size of the neighbourhood)
-
-\paragraph{Comparing Phase Diagrams}
+\subsection{Comparing Phase Diagrams}
 
@@
 
-\paragraph{Results}
+A simple measure of the distance between two phase diagrams, relative to their internal variability, is given by
 
-Et on met le mod{\`e}le plus sioux avec toutes les réplications dans plein de configurations de villes d'une part : diagramme de phase par rapport aux meta-param{\`e}tres(=indicateurs de forme, dans un plan principal = 80\% variance expliqu{\'e}e) de la variabilit{\'e} des diagrammes de phase selon critères de comparaison quantitatifs par rapport à une r{\'e}f{\'e}rence.
+\[
+d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
+\]
 
-%%%%%%%%%%%%%%%
-\subsection{Application: the Sugarscape model}
+where $f_{\vec{alpha}}\left[\vec{x}\right]$ the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a functional distance that can be taken for example as basic L2 distance or the Earth's Mover Distance.
 
-To illustrate the generic nature of our methodology, we apply it here to an other model implying a spatialized distribution of resources for which agents compete: the ``Sugarscape'' model, introduced in~\cite{epstein1996growing}.
+% ************** 
+\section{Application cases}
+
+We use our set of initial density grids in two spatial simulation models: Schelling and Sugarscape.
+
+\subsection{Schelling's model of residential segregation}
+
+Schelling's model consists in a abstract urban housing market where agents of different nature sense their environment, evaluate their satisfaction in terms of neighbourhood composition, and relocate if unsatisfied.
+It has been shown by \cite{Schelling1969} that even tolerant agents tended to produce segregated patterns because of the complexity of their local interactions.
+The main parameters of this model are the tolerance level (\% of agents similar to {\it ego}), the scope of sensing and the percentage of vacant spaces in the housing market.
+In addition, we are interested in testing the impact of the spatial distribution of housing capacity in this project, using the generated grids.
+
+\subsection{Sugarscape model of resource extraction and population settlement}
+
+Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population (\cite{EpsteinAxtell1996}).
+Agents of different vision scopes and different metabolisms harvest a self-regenerating resource available heterogeneously in the initial landscape, they settle and collect this resource, which leads some of them to survive and others to perish.
+The main parameters of this model are the number of agents, their minimal and maximal resource.
+In addition, we are interested in testing the impact of the spatial distribution of the resource in this project, using the generated grids.
+The outcome of the model is measured as a phase diagram of an index of inequality for ressource distribution (Gini index).
 We extend the implementation with agents wealth distribution of~\cite{li2009netlogo}.
 
-%%%%%%%%%%%%%%%
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{figures/fixed_phase_diagram}
-\caption{Phase diagram of the sugarscape model (final smoothed Gini indicator), for a basic bicentric spatial setup, with varying parameters population $N$, minimal sugar reload $s_{min}$ and maximal sugar reload $s_{max}$.}
-\label{fig:sugarscape-phasediagram}
-\end{figure}
-%%%%%%%%%%%%%%%
+% ************** 
+\section{Results}
 
-%%%%%%%%%%%%%%%
-\begin{figure}
-\centering
-%\includegraphics[width=\textwidth]{}
-\caption{Phase diagrams varying across the container morphological space.
-Around 10000 realistic ressources distribution were generated, and used as container landscape (patches sugar capacity) for the sugarscape model.
-Spatial configurations were classified by reducing the dimension from four morphological indicators to two principal components ($\simeq$80\% of variance), for which the resulting space was gridded into 10x10 bins.
-We show here for each cell the average phase diagram along $(N,s_{min})$, taken at $s_{max}=200$.}
-\label{fig:sugarscape-meta}
-\end{figure}
-%%%%%%%%%%%%%%%
+\comment[JR]{for the plan of the results section, I think the best is to organize by application case.}
 
-%%%%%%%%%%%%%%%
-\begin{figure}
-\centering
+We proceed to 4,500,000 simulation runs of the Schelling model (1000 parameter combinations x 45 density grids x 100 replications), using OpenMOLE to distribute the computation, and apply segregation measures to characterise the results.
+We find that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
+For Sugarscape, 2,500,000 simulations (1000 parameter points x 50 density grids x 50 replications) allow us to show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.
+
+% ************** 
+\section{Discussion}
+
+% **************
+
+% \hfill \break
+% \itshape{This is a sub, subheading}\normalfont
+
+% \hfill\break
+
 %
-\caption{Meta phase diagram, i.e.
-phase diagram of phase diagram behavior, along parameters of spatial configuration generation ($\alpha$,$n_D$,$\beta$,$N_G$) (described these somewhere), aggregated for each as an Earth Mover Distance to a fixed reference (phase diagram of Fig.~\ref{fig:sugarscape-phasediagram}).}
-\label{fig:sugarscape-meta}
-\end{figure}
-%%%%%%%%%%%%%%%
+%
+%\begin{table}[htp]
+%
+%\begin{center}
+%\begin{tabular}{c c c c}
+%\arrayrulecolor{black}
+%\hline 
+%This & Is & A & Table\\
+%\arrayrulecolor{lightgray}
+%\hline 
+%\arrayrulecolor{black}
+%Label & 0.1 & 0.2 & 0.3\\
+%Label & 1.0 & 2.0 & 3.0\\
+%\hline
+%\end{tabular}
+%\end{center}
+%\label{first_table}
+%\caption{This is a table caption}
+%\end{table}%
+%
+%
+%
+%\begin{equation}
+%a^2 + b^2 = c^2
+%\tag*{Equation 1}
+%\end{equation}
 
-%%%%%%%%%%%%%%%
-\subsection{Application: SimpopNet}
+\section{Acknowledgements}
 
-More complicated initial conditions such as a synthetic transportation network also fully enter our framework, even these may be more difficult to generate or explore.
-We illustrate this other type of application by exploring the sensitivity of the \emph{SimpopNet} model to the initial network.
-This model introduced in \cite{schmitt2014modelisation} aims at simulating the co-evolution of cities and transportation networks.
-The implementation of the model for which the behavior is known corresponds to a fixed stylized initial network.
-We modulate it here in a basic way, within a range of plausibility given by network indicators.
+The authors acknowledge the funding of their institutions and the EPSRC project number EP/M023583/1.
+Results obtained in this paper were computed on the vo.complex-system.eu virtual organization of the European Grid Infrastructure ( http://www.egi.eu ).
+We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.
 
-%%%%%%%%%%%%%%%
-\section{Discussion}
-%%%%%%%%%%%%%%%
+\section{References}
 
-Puis discussion : space matters (les valeurs prises par les indicateurs segrégations sont pas les mêmes) et ensuite en ouverture implications pour planning ?
-
-%%%%%%%%%%%%%%%
-% Biblio
-%%%%%%%%%%%%%%%
-
-\bibliographystyle{apalike}
-\bibliography{biblio}
+\bibliography{spacematters}
 

### 2017-05-16 paper.tex
Update on Overleaf.

@@
-\documentclass[11pt]{article}
+\documentclass[Afour,sageh,times]{sagej}
 \input{header.tex}
 
-\date{}
-\renewcommand\abstractname{\fontsize{14pt}{0}\textbf{Abstract}\selectfont}
-
-\usepackage[left=25mm, right=25mm, top=25mm, bottom=25mm, includehead=false, includefoot=false]{geometry}
-
-\usepackage{graphicx}
-\usepackage{url}
-\usepackage[round,semicolon]{natbib}  % Citation styles https://www.sharelatex.com/learn/Natbib_citation_styles
-\bibliographystyle{humannat}
-\renewcommand{\bibsection}{}
-\renewcommand{\bibhang}{\setlength{-1px}}
-
-\usepackage{authblk} % For author lists
-\renewcommand\Authfont{\fontsize{11}{1}\selectfont}
-\renewcommand\Affilfont{\fontsize{9}{1}\selectfont}
-
-\renewcommand*\footnoterule{}
-
-\usepackage[table]{xcolor}
-\usepackage[parfill]{parskip} % Line between paragraphs
-\usepackage{amsmath}
-\pagenumbering{arabic}
-
-\usepackage{sectsty}
-\allsectionsfont{\sffamily}
-
-\usepackage[pdftex]{hyperref} 
-\hypersetup{pdfborder={0 0 0} }
+\begin{document}
 
@@
 
-\title{\sffamily\fontsize{16}{0}\textbf{Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?}}
+\runninghead{Cottineau \& al.}
 
-\author[1]{Cl{\' e}mentine Cottineau\thanks{}}
-\author[2,4]{Juste Raimbault}
-\author[3]{Marion Le Texier}
-\author[4]{Florent Le N{\' e}chet}
-\author[2,5]{Romain Reuillon}
-\affil[1]{Centre for Advanced Spatial Analysis, University College London, UK}
-\affil[2]{UMR 8504 G{\' e}ographie-cit{\' e}s, Paris, France}
-\affil[3]{UMR 6266 IDEES, Universit{\' e} de Rouen Normandie, France}
-\affil[4]{Laboratoire Ville Mobilit{\' e} Transport, Universit{\' e} Paris-Est, France}
-\affil[5]{Instit, ut des Syst{\` e}mes Complexes Paris Ile-de-France, France}
+\title{Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?}
 
-\affil[*]{\texttt{, Email: c.cottineau@ucl.ac.uk }}
+\author{Cl{\' e}mentine Cottineau\affilnum{1}, Juste Raimbault\affilnum{2,4}, Marion Le Texier\affilnum{3}, Florent Le N{\' e}chet\affilnum{4}, Romain Reuillon\affilnum{2,5}
+}
 
-\begin{document}
+\affiliation{\affilnum{1}Centre for Advanced Spatial Analysis, University College London, UK\\
+\affilnum{2}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
+\affilnum{3}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
+\affilnum{4}Laboratoire Ville Mobilit{\'e} Transport, Universit{\'e} Paris-Est, France\\
+\affilnum{5}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}
 
-\maketitle
+\corrauth{Corresponding author
+address.}
+
+\email{c.cottineau@ucl.ac.uk}
 
@@
 \begin{abstract}
-\noindent
-\setlength{\parindent}{0pt}
+%\noindent
+%\setlength{\parindent}{0pt}
 Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust.
@@
 We do so using high performance computing in a very simple and straightforward open-source workflow.
-$ $ \\ {\bf Keywords:} Space, Initial conditions, Sensitivity, ABM.
+\end{abstract}
 
-\end{abstract},
+\keywords{Space, Initial conditions, Sensitivity, ABM}
+
+\maketitle
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Why spatial patterns are expected to impact social simulations}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Objective}
@@
 
-% **************
+%%%%%%%%%%%%%%%%%%%%%%
+\section{Previous attempts in the literature}
+%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Previous attempts in the literature}
-
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{In general}
 
-Paragraph not as detailed as the following one, but just to illustrate the fact that the issue is an important concern in different types of spatial simulation modelling approaches?
+The effect of the spatial configuration on area-based attributes of human behaviours has been largely discussed in geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984,FotheringhamWong1991}.
+Recently, \citet{Kwan2012} claims for a careful examination of what she coins the uncertain geographic context problem (UGCoP), that is of the spatial configuration of geographical units even if the size and delineation of the area are the same.
+On the contrary, the scarcity of these considerations in the geographic simulation model literature questions the generalisation of their results, as it has for instance been showed in the case of LUTI models \citep{Thomasetal2017}, of diffusion processes using ABM \citep{LeTexierCaruso2017}.
 
-Eg. : effects of delineation and scale of cities in LUTI models \citep{Thomasetal2017}, etc.
-
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{For Schelling model in particular}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{And for SugarScape?}
 
-% **************
-
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Method}
+%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\begin{figure}[htbp] \begin{center} 
+%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure*}[htbp] \begin{center} 
 \resizebox{0.9\textwidth}{!}{ 
 	\includegraphics{figures/SchemaMeta_1.png}
-} \caption{General workflow of our method} \label{fig:method} \end{center} \end{figure} %
+} \caption{General workflow of our method\comment[MLT]{If we want to keep this figure, I propose that we link it more clearly to the following subsections, by numbering the different parts of the figure?
+By graphically showing on which parts of the figure the different subsections focus?
+Besides, we need to change it as we know want to go further than analysing the outputs qualitatively.}} \label{fig:method} \end{center} \end{figure*} %
+%%%%%%%%%%%%%%%%%%%%%%
 
-\comment[MLT]{If we want to keep this figure, I propose that we link it more clearly to the following subsections, by numbering the different parts of the figure?
-By graphically showing on which parts of the figure the different subsections focus?
-Besides, we need to change it as we know want to go further than analysing the outputs qualitatively.}
-
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Constructing a spatial generator}
@@
 
-\begin{figure}[htbp] \begin{center} 
+%%%%%%%%%%%%%%%%%%%%%%
+\begin{figure*}[htbp] \begin{center} 
 \resizebox{0.9\textwidth}{!}{ 
 	\includegraphics{figures/spatialGen.png}
-} \caption{Four examples of grids produced by the spatial generator} \label{fig:spatialGen} \end{center} \end{figure} %
+} \caption{Four examples of grids produced by the spatial generator} \label{fig:spatialGen} \end{center} \end{figure*} %
+%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Comparing Phase Diagrams}
@@
 
-where $f_{\vec{alpha}}\left[\vec{x}\right]$ the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a functional distance that can be taken for example as basic L2 distance or the Earth's Mover Distance.
+where $f_{\vec{\alpha}}\left[\vec{x}\right]$ the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a functional distance that can be taken for example as basic L2 distance or the Earth's Mover Distance.
 
-% ************** 
+\todo[JR]{explain this in clear words}
+
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Application cases}
+%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Schelling's model of residential segregation}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Sugarscape model of resource extraction and population settlement}
@@
 
-% ************** 
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Results}
+%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Schelling's model}
+
 We proceed to 4,500,000 simulation runs of the Schelling model (1000 parameter combinations x 45 density grids x 100 replications), using OpenMOLE to distribute the computation, and apply segregation measures to characterise the results.
 We find that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Sugarscape model}
+
 For Sugarscape, 2,500,000 simulations (1000 parameter points x 50 density grids x 50 replications) allow us to show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.
 
-% ************** 
+%%%%%%%%%%%%%%%%%%%%%%
 \section{Discussion}
-
-% **************
+%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section{Acknowledgements}
+%%%%%%%%%%%%%
+% Acknowledgements
 
+\begin{acks}
 The authors acknowledge the funding of their institutions and the EPSRC project number EP/M023583/1.
@@
 We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.
+\end{acks}
 
-\section{References}
+%%%%%%%%%%%%
+%% References
+%%%%%%%%%%%%
 
-\bibliography{spacematters}
+\bibliographystyle{SageH}
+\bibliography{spacematters,biblio}
 

### 2017-05-31 paper.tex
writing

@@
 We think simulation can become a very good tool to achieve this, provided that models include relevant spatial description and modelling, behavioural rules that take space into account, and provided model evaluation  stresses the sensitivity of output variations to the way space is modelled.
-This communication focuses on a way to do  so: by using a sensitivity method to test the model's outcomes to initial spatial conditions.
+This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of model's outcomes to initial spatial conditions.
 
@@
 \paragraph{Due to spatio-temporal path dependencies}
+A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
+Similar to what \todo{cit Quarck and Jaguar} calls \emph{frozen accidents} in any complex system, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system: the example of transportation networks is a good illustration, as their spatial shape and hierarchy is strongly influenced by past investment decisions, technical choices, or political decisions sometimes not rational~(\cite{zembri2010new}).
+Some aggregated indicators will not take into account positions and trajectories of each agent (such as segregation in the Schelling model) but others, as in the case of spatial patterns of accessibility in a system of cities, fully capture the path-dependency and may therefore be highly dependent of the initial spatial configuration.
+It is not clear for example what shifted the economical and political capital of France from Lyon to Paris in the early Middle Age, some assumptions being the reconfiguration of trade patterns from South to North of Europe and thus an increased centrality for Paris due to its spatial position: the bifurcation induced by socio-economic and political factors took a deep significance with worldwide repercussions until today when magnified by the spatial configuration.
+\comment[JR]{quelqu'un sait si d'une part c'est vrai et d'autre part si des references backupent ca ?
+- je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\section{Method}
+\section{Methods}
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-Problem: we have as many phase diagrams than we have spatial grids, so we cannot define the phases qualitatively. We need quantitative procedures.
+In order to test for the influence of spatial initial conditions, we need a systematic method to compare phase diagrams.
+Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic.
+A solution is to use systematic quantitative procedures.
+Several potential methods could be used: for example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
+The use of metrics comparing spatial distributions, such as the Earth Movers Distance, or the comparison of transition matrices of the dynamic associated to the potential described by each distribution.
 
-Potential methods to explore: anisotropic spatial segregation index (number of clusters found and where in the parameter spaces?
-but may be weak...), earth Movers Distance (i.e.
-distance between probability distributions), comparison of transition matrices of the dynamic associated to the potential...
-See also: http://mck.riks.nl/ to get some ideas.
+\comment[JR]{just realized there are a bunch of metrics in physical sciences (hydrology, meteorology, etc.) to compare spatial fields : cf http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0178165 - with spatio-temporal empirical orthogonal functions we can have a 3D parameter space for example ?}
 
-A simple measure of the distance between two phase diagrams, relative to their internal variability, is given by
+\comment[JR]{See also: http://mck.riks.nl/ to get some ideas.}
+
+To keep it simple and as such methodological considerations are auxiliary to the main purpose of this paper, we propose an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability.
+More formally, the distance is given by
 
@@
 
-where $f_{\vec{\alpha}}\left[\vec{x}\right]$ the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a functional distance that can be taken for example as basic L2 distance or the Earth's Mover Distance.
+where $\alpha \mapsto \left[\vec{x} \mapsto f_{\vec{\alpha}}\left(\vec{x}\right)\right]$ is the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a distance between probability distributions that can be taken for example as basic L2 distance or the Earth's Mover Distance.
+For each values $\vec{\alpha_i}$, the phase diagram is seen as a random spatial field, facilitating the definition of variances and distance.
 
-\todo[JR]{explain this in clear words}
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Model Exploration Workflows as an (meta-)sensitivity analysis method}
+
+\comment[JR]{ca serait interessant d'inserer un paragraphe specifique sur Openmole, le multimodeling, les workflows modulables etc, comme outil très puissant pour ce qu'on veut faire : ca permet de rendre transparent les analyses de sensitivite mais egalement les meta-analyses de sensitivite (d'ou le titre).}
 

### 2017-06-12 paper.tex
Update on Overleaf.

@@
 
-\author{Cl{\' e}mentine Cottineau\affilnum{1}, Juste Raimbault\affilnum{2,4}, Marion Le Texier\affilnum{3}, Florent Le N{\' e}chet\affilnum{4}, Romain Reuillon\affilnum{2,5}
+\author{Juste Raimbault\affilnum{1,4}, 
+Cl{\' e}mentine Cottineau\affilnum{2},
+Marion Le Texier\affilnum{3}, 
+Florent Le N{\' e}chet\affilnum{4}, 
+Romain Reuillon\affilnum{1,5}
 }
 
-\affiliation{\affilnum{1}Centre for Advanced Spatial Analysis, University College London, UK\\
-\affilnum{2}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
+\affiliation{\affilnum{1}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
+\affilnum{2}Centre for Advanced Spatial Analysis, University College London, UK\\
 \affilnum{3}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
@@
 This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of model's outcomes to initial spatial conditions.
+To demonstrate the genericity of our approach, we develop two applications from simulation models which are commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996}: Schelling's model of segregation and Epstein and Axtell's Sugarscape model.
+Both models are defined at the intraurban level.
 
@@
 \paragraph{Due to spatio-temporal path dependencies}
+
 A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
-Similar to what \todo{cit Quarck and Jaguar} calls \emph{frozen accidents} in any complex system, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system: the example of transportation networks is a good illustration, as their spatial shape and hierarchy is strongly influenced by past investment decisions, technical choices, or political decisions sometimes not rational~(\cite{zembri2010new}).
+Similar to what \todo{cit Quarck and Jaguar} calls \emph{frozen accidents} in any complex system, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
+Temporal and cumulative effects have been considered in various geographical subfields and at various geographical scales, including transportation and urban economics, urban geography, and interregional migrations \citep[for e.g.]{White1977,White1978, AllenSanglier1979,Wilson1981,Pumainetal1989,AllenSanglier1981,WeidlichHaag1988,Portugali2000,Wilson2002,Batty2007,AzizAlaouiBertelle2009}.
+Less studied is the impact of the spatial setting on models dynamics and potential bifurcations.
+
+The example of transportation networks is a good illustration, as their spatial shape and hierarchy is strongly influenced by past investment decisions, technical choices, or political decisions sometimes not rational~(\cite{zembri2010new}).
 Some aggregated indicators will not take into account positions and trajectories of each agent (such as segregation in the Schelling model) but others, as in the case of spatial patterns of accessibility in a system of cities, fully capture the path-dependency and may therefore be highly dependent of the initial spatial configuration.
@@
 \comment[JR]{quelqu'un sait si d'une part c'est vrai et d'autre part si des references backupent ca ?
-- je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.}
+- je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.} \comment[MLT]{L'idée de donner un exemple me semble bonne mais je ne suis pas sûre qu'on l'amène de la bonne manière?}
 
@@
 
-In this section, we detail the method developped to tackle simulation models sensitivity to initial spatial conditions. The general workflow of the method is illustrated in Figure \ref{fig:method}.
+In this section, we detail the method developed to analyse the sensitivity of simulation models to initial spatial conditions.
+The general method workflow is illustrated in Figure \ref{fig:method}.
+In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters and relating these variations of values to the variations in the simulation results, we here introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by parameters and produces sets of spatial initial conditions.
+Initial spatial conditions are clustered to represent types of spaces ex-ante (for example: moonocentric or polycentric density grids), and the sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types.
+It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.
 

### 2017-06-28 paper.tex
.

@@
 A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
-Similar to what \todo{cit Quarck and Jaguar} calls \emph{frozen accidents} in any complex system, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
+Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
 Temporal and cumulative effects have been considered in various geographical subfields and at various geographical scales, including transportation and urban economics, urban geography, and interregional migrations \citep[for e.g.]{White1977,White1978, AllenSanglier1979,Wilson1981,Pumainetal1989,AllenSanglier1981,WeidlichHaag1988,Portugali2000,Wilson2002,Batty2007,AzizAlaouiBertelle2009}.
@@
 Several potential methods could be used: for example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
-The use of metrics comparing spatial distributions, such as the Earth Movers Distance, or the comparison of transition matrices of the dynamic associated to the potential described by each distribution.
-
-\comment[JR]{just realized there are a bunch of metrics in physical sciences (hydrology, meteorology, etc.) to compare spatial fields : cf http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0178165 - with spatio-temporal empirical orthogonal functions we can have a 3D parameter space for example ?}
-
-\comment[JR]{See also: http://mck.riks.nl/ to get some ideas.}
-
+The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
+Map comparison methods, popular in environmental sciences, provide numeral tools to compare two dimensional fields~\cite{visser2006map}.
+To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}.
 To keep it simple and as such methodological considerations are auxiliary to the main purpose of this paper, we propose an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability.
@@
 
-\comment[JR]{for the plan of the results section, I think the best is to organize by application case.}
-
 %%%%%%%%%%%%%%%%%%%%%%
@@
 For Sugarscape, 2,500,000 simulations (1000 parameter points x 50 density grids x 50 replications) allow us to show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.
+
+More precisely, we explore the basic parameter space of the model, which is the population of agents $P$, the minimal ressource capacity
+
+% summary stats
+%    Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
+%0.009408 0.084490 1.266000 0.930400 1.574000 1.858000
+
+% pca of morphological space
+% "","PC1","PC2","PC3","PC4"
+%"distance",0.762358566609464,-0.260991693298744,0.200656405132039,0.557162237616392
+%"entropy",0.601306167355116,0.181706245959277,0.0958379422351422,-0.772158547261002
+%"moran",0.0311129390452153,0.912155429075071,0.30114271129527,0.276256268103684
+%"slope",0.237217819823539,0.258531718397015,-0.927289147645628,0.130475642169329
+
+%relativedistance_metaparams.pdf
+%relativedistance_morphspace.pdf
+
+%phasediagram_id0_maxSugar110.png
+%phasediagram_id27_maxSugar110.png
+
+% phase diagrams
+%          spAlpha spDiffsteps spDiffusion spGrowth spPopulation
+% id=27 : 0.7913103    2.376837   0.1440293 157.4147 4852.746
+% id=0 : 2.562398    3.753032   0.1316788 128.4632 4753.983
+% maxSugar = 110
 

### 2017-06-28 paper.tex
.

@@
 \comment[JR]{quelqu'un sait si d'une part c'est vrai et d'autre part si des references backupent ca ?
-- je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.} \comment[MLT]{L'idée de donner un exemple me semble bonne mais je ne suis pas sûre qu'on l'amène de la bonne manière?}
+- je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.}[(MLT)L'idée de donner un exemple me semble bonne mais je ne suis pas sûre qu'on l'amène de la bonne manière?]
 
@@
 
-\[
-d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
-\]
+\begin{equation}\label{eq:phase-distance}
+d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
+\end{equation}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Model Exploration Workflows as an (meta-)sensitivity analysis method}
+\subsection{Model Exploration Workflows as an meta-sensitivity analysis method}
 
-\comment[JR]{ca serait interessant d'inserer un paragraphe specifique sur Openmole, le multimodeling, les workflows modulables etc, comme outil très puissant pour ce qu'on veut faire : ca permet de rendre transparent les analyses de sensitivite mais egalement les meta-analyses de sensitivite (d'ou le titre).}
+The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows.
+The ideas of multi-modeling and extensive model exploration are nothing from new as Openshaw already advocated for ``model-crunching'' in~\cite{openshaw1983data}, but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities: \cite{cottineau2016back} claims for a renewed approach on multi-modeling.
+Coupling models as we do answers to similar questions.
+In that stream of research, the model exploration platform OpenMole~\cite{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
+In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file.
 
@@
 
-More precisely, we explore the basic parameter space of the model, which is the population of agents $P$, the minimal ressource capacity
+More precisely, we explore a grid of a basic parameter space of the model, which three dimensions are the population of agents $P\in \left[10;510\right]$, the minimal initial agent ressource $s_{-}\in \left[10;100\right]$ and the maximal initial agent ressource $s_{+}\in \left[110;200\right]$.
+Each parameter is binned into 10 values, giving 1000 parameter points.
+We run 50 repetitions for each configuration, what yield reasonable convergence properties.
+The initial spatial configuration varies across 50 different grids, generated by sampling meta-parameters for the generator in a LHS.
+We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model.
+We mesure the distance of all 3-dimensional phase diagrams to the reference phase diagram computed on the default model setup (see Fig.~\ref{fig:sugarscape-distance} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability.
+Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each.
+Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.
 
@@
 %   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-%0.009408 0.084490 1.266000 0.930400 1.574000 1.858000
+% 0.08909 0.19790 1.52200 1.29600 2.16400 2.98100
+
+We obtain a very strong sensitivity to initial conditions, as the distribution of the relative distance to reference across grids ranges from 0.09 to 2.98 with a median of 1.52 and an average of 1.30.
+It means that in average, the model is more sensitive to meta-parameters than to parameters, and the relation variation can reach a factor of 3.
+We plot in Fig.~\ref{fig:sugarscape-distance} their distribution in a morphological space.
+The reduced morphological space is obtained by computing 4 raw indicators of urban form, namely Moran index, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and contextualization), and by reducing the dimension with a principal component analysis for which we keep the first two components (92\% of cumulated variance).
+The first measures a ``level of sprawl'' and of scattering, whereas the second measures aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation.
+It is confirmed by the behavior as a function of meta-parameters, as high values of $\alpha$ also yield high distance.
+In terms of model processes, it shows that congestion mechanisms induce rapidly higher levels of inequality.
 
@@
 
-%relativedistance_metaparams.pdf
-%relativedistance_morphspace.pdf
+%%%%%%%%%%%%%
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{figures/relativedistance_metaparams}\\
+\includegraphics[width=0.5\textwidth]{figures/relativedistance_morphspace}
+\caption{\textbf{Relative distances of phase diagrams to the reference across grids.} (Top) Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and diffusion ($\beta$, strength of diffusion process).
+(Bottom) Relative distance as a function of two first principal components of the morphological space (see text).
+Red point correspond to the reference spatial configuration.
+Green frame and blue frame give respectively the first and second particular phase diagrams shown in Fig.~\ref{fig:sugarscape-phasediagrams}.}
+\label{fig:sugarscape-distance}
+\end{figure}
+%%%%%%%%%%%%%
 
-%phasediagram_id0_maxSugar110.png
-%phasediagram_id27_maxSugar110.png
-
-% phase diagrams
+% phase diagrams -> ok well different qualitatively
 %          spAlpha spDiffsteps spDiffusion spGrowth spPopulation
@@
 % maxSugar = 110
+
+%%%%%%%%%%%%%
+\begin{figure}
+\centering
+\includegraphics[width=0.24\textwidth]{figures/phasediagram_id27_maxSugar110}
+\includegraphics[width=0.24\textwidth]{figures/phasediagram_id0_maxSugar110}
+\caption{\textbf{Examples of phase diagrams.} We show two dimensional phase diagrams on $(P,s_-)$, both at fixed $s_+ = 110$.
+(Left) Green frame, obtained with $\alpha = 0.79$, $n=2$, $\beta = 0.14$, $N=157$; (Right) Blue frame, obtained with $\alpha = 2.56$, $n=3$, $\beta = 0.13$, $N=128$.}
+\label{fig:sugarscape-phasediagrams}
+\end{figure}
+%%%%%%%%%%%%%
+
+We now check the sensitivity in terms of qualitative behavior of phase diagrams.
+We show the phase diagrams for two very opposite morphologies in term of sprawling, but controlling for aggregation with the same $PC2$ value.
+These correspond to the green and blue frames in Fig.~\ref{fig:sugarscape-distance}.
+The behaviors are rather stable for varying $s_+$, what means that the poorest agents have a determinant role in trajectories.
+The two examples have not only a very distant baseline inequality (the ceil of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values.
+The process is thus completely inverted, what would have significant impacts if one tried to schematize policies from this model.
+This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
 

### 2017-07-05 paper.tex
Update on Overleaf.

@@
 
-\runninghead{Cottineau \& al.}
+\runninghead{Raimbault \& al.}
 
@@
 
-Simulation has been recognised and increasingly used by geographers to explore various geographical processes and problems within virtual laboratories.
+Simulation has been recognised and increasingly used by geographers to explore various geographical processes and problems within virtual laboratories \citep{Quesneletal2009} \comment{Batty 1978?
+James Doran 1970's : lire thèse de Seb ;) }.
 It appears as a very fruitful way to overcome the difficulty of analytic resolution of many spatial models developed in the past, and to explore the possible trajectories of social and ecological systems in space and time.
@@
 
+\subsection{Objective}
+
+In this paper, we suggest tackling sensitivity to spatial conditions by generating a variety of density grids to feed into simulation models at initialisation, and exploring the sensitivity of the model outcomes to the variation of meta-parameters (i.e.
+parameters used to generate initial spatial conditions).
+The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters within a typical category of space (a monocentric case for example) and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric vs.
+polycentric for example) on the results of a given model.
+
 %%%%%%%%%%%%%%%%%%%%%%
 \subsection{Why spatial patterns are expected to impact social simulations}
+
+\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.}
+\paragraph{Because agents are not random walkers}
+
+We expect spatial patterns to influence model outcomes also because the agents' rule of action itself might depend on the spatial structure of the environment.
+Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.
+For example, agents tend to create buffer zones if they are allowed network-based rather than isotropic movements (\cite{Banos2012}), while the vision / sensing mechanism is sensitive to the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}).
+Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.
+\comment[MLT]{Du coup, je mettrais ce paragraphe à la suite du suivant.} \comment[FL]{je suis d'accord}
+
+\paragraph{Because the evolution of cities is path dependent}
+
+The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolve irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009} \comment[MLT]{Ca fait très name dropping mais je voulais souligner que l'on connaissait la biblio hors Denise hihi.}.
+In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to be (partially) driven by the initial spatial setting, making it difficult to interprete the resulting global structures.
+A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
+Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
 
@@
 
-\paragraph{Because actual agents are not random walkers}
-We expect spatial patterns to influence model outcomes also because the agents' rule of action itself might depend on the spatial structure of the environment.
-Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.
-For example, agents tend to create buffer zones if they are allowed network-based rather than isotropic movements (\cite{Banos2012}), while the vision / sensing mechanism is sensitive to the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}).
-Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.
+\paragraph{Considerations of urban form in simulation models}
 
-\paragraph{Due to spatio-temporal path dependencies}
+The effect of the spatial configuration on area-based attributes of human behaviours has been largely discussed in geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984,FotheringhamWong1991}.
+Recently, \citet{Kwan2012} claims for a careful examination of what she coins the uncertain geographic context problem (UGCoP), that is of the spatial configuration of geographical units even if the size and delineation of the area are the same.
+On the contrary, the scarcity of these considerations in the geographic simulation model literature questions the generalisation of their results, as it has for instance been showed in the case of LUTI models \citep{Thomasetal2017} and of diffusion processes using ABM \citep{LeTexierCaruso2017}.
+\comment[FL]{Je pense que cette partie est vraiment utile, et devrait être étoffée, en procédant peut être à une typologie plus systématique des "initial condition" qui nous intéressent.
+Il y a déjà une différence entre l'hétérogénéité de l'espace (densités réparties mono ou poly, etc) et les différentes façons de représenter l'espace (grilles carrées, hexagones).
+Par ailleurs aussi des discussions sur l'incertitude de l'information (est ce que par exemple les emplois downscalés au carreau INSEE par Pivano 2016 c'est vraiment fiable ou très ?) et la précision de la grille}
 
-A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
-Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
-Temporal and cumulative effects have been considered in various geographical subfields and at various geographical scales, including transportation and urban economics, urban geography, and interregional migrations \citep[for e.g.]{White1977,White1978, AllenSanglier1979,Wilson1981,Pumainetal1989,AllenSanglier1981,WeidlichHaag1988,Portugali2000,Wilson2002,Batty2007,AzizAlaouiBertelle2009}.
-Less studied is the impact of the spatial setting on models dynamics and potential bifurcations.
-
-The example of transportation networks is a good illustration, as their spatial shape and hierarchy is strongly influenced by past investment decisions, technical choices, or political decisions sometimes not rational~(\cite{zembri2010new}).
+%The example of transportation networks is a good illustration, as their spatial shape and hierarchy is strongly influenced by past investment decisions, technical choices, or political decisions sometimes not rational~(\cite{zembri2010new}).
 Some aggregated indicators will not take into account positions and trajectories of each agent (such as segregation in the Schelling model) but others, as in the case of spatial patterns of accessibility in a system of cities, fully capture the path-dependency and may therefore be highly dependent of the initial spatial configuration.
@@
 \comment[JR]{quelqu'un sait si d'une part c'est vrai et d'autre part si des references backupent ca ?
-- je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.}[(MLT)L'idée de donner un exemple me semble bonne mais je ne suis pas sûre qu'on l'amène de la bonne manière?]
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Objective}
-
-In this paper, we suggest tackling sensitivity to spatial conditions by generating a variety of density grids to feed into simulation models at initialisation, and exploring the sensitivity of the model outcomes to the variation of meta-parameters (i.e.
-parameters used to generate initial spatial conditions).
-The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters within a typical category of space (a monocentric case for example) and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric vs.
-polycentric for example) on the results of a given model.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\section{Previous attempts in the literature}
-%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{In general}
-
-The effect of the spatial configuration on area-based attributes of human behaviours has been largely discussed in geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984,FotheringhamWong1991}.
-Recently, \citet{Kwan2012} claims for a careful examination of what she coins the uncertain geographic context problem (UGCoP), that is of the spatial configuration of geographical units even if the size and delineation of the area are the same.
-On the contrary, the scarcity of these considerations in the geographic simulation model literature questions the generalisation of their results, as it has for instance been showed in the case of LUTI models \citep{Thomasetal2017}, of diffusion processes using ABM \citep{LeTexierCaruso2017}.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{For Schelling model in particular}
-
-Recent attempts of running Schelling-like models beyond the chessboard grid have proved that migration decisions are not only dependent of what positions agents occupy relative to each other, but also of where they occupy these positions.
-\citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents are highest in a rectangular grid and lowest in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents.
-\citet{StaufferSolomon2007} introduce asymetric interactions and empty residences on a large and regular lattice.
-They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns.
-\citet{Singhetal2009} show that the segregation patterns obtained by Schelling for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.
-\citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents.
-They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile.
-\citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours.
-Interestingly, \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents).
-In this paper, we build on the contributions of these previous studies and go further by systematically measuring the impact of the city lattice on Schelling model behaviour so far only assessed through a small set of illustrative spatial structures.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{And for SugarScape?}
+- je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.}[(MLT)L'idée de donner un exemple me semble bonne mais je ne suis pas sûre qu'on l'amène de la bonne manière?
+--> en fait je crois que l'on a pas besoin de ce paragraphe illustratif, ça complique le papier.
+Qu'en pensez-vous?]  [(FL) : illustrer par les transports c'est une bonne idée mais pas besoin de parler du moyen âge!
+la faible croissance de Dijon depuis 30 ans peut être une illustration tarte à la crême.
+mais pas sûr effectivement qu'on ait absolument besoin d'illustration, en tous cas il n'y en a pas dans le § d'avant (bientôt après ;o)]
 
@@
 The general method workflow is illustrated in Figure \ref{fig:method}.
-In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters and relating these variations of values to the variations in the simulation results, we here introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by parameters and produces sets of spatial initial conditions.
+In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters and relating these variations of values to the variations in the simulation results, we here introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions.
 Initial spatial conditions are clustered to represent types of spaces ex-ante (for example: moonocentric or polycentric density grids), and the sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types.
@@
 	\includegraphics{figures/SchemaMeta_1.png}
-} \caption{General workflow of our method\comment[MLT]{If we want to keep this figure, I propose that we link it more clearly to the following subsections, by numbering the different parts of the figure?
-By graphically showing on which parts of the figure the different subsections focus?
-Besides, we need to change it as we know want to go further than analysing the outputs qualitatively.}} \label{fig:method} \end{center} \end{figure*} %
+} \caption{General workflow of our method} \label{fig:method} \end{center} \end{figure*} %
 %%%%%%%%%%%%%%%%%%%%%%
@@
 Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}).
-In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$.
+\comment[FL]{il faut que les sens de tous les paramètres soient explicités avec les équations ?} In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$.
 This first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$.
@@
 figure \ref{fig:spatialGen}).
+\comment{A detailler en incluant les equations de morphogenese}
 
@@
 	\includegraphics{figures/spatialGen.png}
-} \caption{Four examples of grids produced by the spatial generator} \label{fig:spatialGen} \end{center} \end{figure*} %
+} \caption{Four examples of grids produced by the spatial generator. The lighter the red, the denser the area.} \label{fig:spatialGen} \end{center} \end{figure*} %
 %%%%%%%%%%%%%%%%%%%%%%
@@
 A solution is to use systematic quantitative procedures.
-Several potential methods could be used: for example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
+Several potential methods could be used \comment[MLT]{Ici je me demande si l'on ne devrait pas simplement dire que l'on a pas trouvé de méthodes de comparaisons systématiques de diagrammes de phase dans la littérature SMA ou geographic modelling, et que du coup on propose de comparer la variance intra à la variance inter-grille selon la formule ...
+Et je garderais la précision sur les autres méthodes que l'on aurait pu choisir pour la section de discussion}: for example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
 The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
-Map comparison methods, popular in environmental sciences, provide numeral tools to compare two dimensional fields~\cite{visser2006map}.
+Map comparison methods, popular in environmental sciences \comment[MLT]{Et pas en géo?
+Qu'est-ce que ces méthodes ont de particulier?}, provide numeral tools to compare two dimensional fields~\cite{visser2006map}.
 To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}.
@@
 
-where $\alpha \mapsto \left[\vec{x} \mapsto f_{\vec{\alpha}}\left(\vec{x}\right)\right]$ is the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a distance between probability distributions that can be taken for example as basic L2 distance or the Earth's Mover Distance.
+where $\alpha \mapsto \left[\vec{x} \mapsto f_{\vec{\alpha}}\left(\vec{x}\right)\right]$ is the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a distance between probability distributions that can be taken for example as basic L2 distance \comment[MLT]{Euclidean?
+Je crois qu'il faut adapter le vocabulaire utilisé aux lecteurs d'EPB, qu'en pensez-vous?} or the Earth's Mover Distance \comment[MLT]{Si l'on mentionne les deux, ne va-t-on pas nous demander de tester l'effet du choix de l'une ou de l'autre sur nos résultats?}.
 For each values $\vec{\alpha_i}$, the phase diagram is seen as a random spatial field, facilitating the definition of variances and distance.
+\comment[FL]{je pense que c'est un peu compliqué effectivement cette partie.}
 
@@
 The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows.
-The ideas of multi-modeling and extensive model exploration are nothing from new as Openshaw already advocated for ``model-crunching'' in~\cite{openshaw1983data}, but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities: \cite{cottineau2016back} claims for a renewed approach on multi-modeling.
-Coupling models as we do answers to similar questions.
-In that stream of research, the model exploration platform OpenMole~\cite{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
+The ideas of multi-modeling and extensive model exploration are nothing from new as Openshaw already advocated for ``model-crunching'' in~\cite{openshaw1983data}, but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities.
+The model exploration platform OpenMole~\cite{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
 In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file.
+\comment[MLT]{Je me suis permis de bouger la partie sur le multi-modeling en fin de paragraphe, mais ai essayé d'en garder l'esprit tel quel!} In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modeling claimed by \cite{cottineau2016back}.
 
@@
 In addition, we are interested in testing the impact of the spatial distribution of housing capacity in this project, using the generated grids.
+\comment[MLT]{Dans le paragraphe sur Sugarscape on parle du diagramme de phase que l'on étudie en sortie, pas ici.
+On harmonise dans quel sens?}
+
+%Recent attempts of running Schelling-like models beyond the chessboard grid have proved that migration decisions are not only dependent of what positions agents occupy relative to each other, but also of where they occupy these positions.
+\citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents are highest in a rectangular grid and lowest in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents.
+\citet{StaufferSolomon2007} introduce asymetric interactions and empty residences on a large and regular lattice.
+They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns.
+\citet{Singhetal2009} show that the segregation patterns obtained by Schelling for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.
+\citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents.
+They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile.
+\citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours.
+Interestingly, \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents).
+In this paper, we build on the contributions of these previous studies and go further by systematically measuring the impact of the city lattice on Schelling model behaviour so far only assessed through a small set of illustrative spatial structures.
+\comment[MLT]{Le remettre en moins détaillé dans l'intro en lien avec les catégories de prise en compte de la forme de la grille.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
+
+\comment[FL]{il faudrait peut être davantage renforcer l'analyse a propos de figure 1 de si c'est plutôt les mu parameters ou les spatial parameters qui ont une grosse influence sur les résultats.}
 
@@
 
+\comment{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}
+
 % \hfill \break

### 2017-07-05 paper.tex
.

--- 2017-06-28 paper.tex
@@
 In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$.
-This first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$.
+More precisely, each added unit has a probability proportional to $P_i/\sum P_k$ to be added to patch $i$ with population $P_i$.
+At the end of each time step, this first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches).
+To avoid bord effects such as a reflexion on the border of the world, border patches diffuse to the outside.
+The model stops when a fixed total population is reached.
 Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed.
@@
 figure \ref{fig:spatialGen}).
+Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
 

### 2017-07-05 paper.tex

--- 2017-07-05 paper.tex
@@
 Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}).
-\comment[FL]{il faut que les sens de tous les paramètres soient explicités avec les équations ?} In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$.
-This first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$.
+In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$.
+More precisely, each added unit has a probability proportional to $P_i/\sum P_k$ to be added to patch $i$ with population $P_i$.
+At the end of each time step, this first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches).
+To avoid bord effects such as a reflexion on the border of the world, border patches diffuse to the outside.
+The model stops when a fixed total population is reached.
 Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed.
@@
 figure \ref{fig:spatialGen}).
+Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
 \comment{A detailler en incluant les equations de morphogenese}

### 2017-07-05 paper.tex
Update on Overleaf.

--- 2017-07-05 paper.tex
@@
 
-\author{Juste Raimbault\affilnum{1,4}, 
-Cl{\' e}mentine Cottineau\affilnum{2},
-Marion Le Texier\affilnum{3}, 
-Florent Le N{\' e}chet\affilnum{4}, 
+\author{Juste Raimbault\affilnum{1,2}, 
+Cl{\' e}mentine Cottineau\affilnum{3},
+Marion Le Texier\affilnum{4}, 
+Florent Le N{\' e}chet\affilnum{2}, 
 Romain Reuillon\affilnum{1,5}
@@
 \affiliation{\affilnum{1}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
-\affilnum{2}Centre for Advanced Spatial Analysis, University College London, UK\\
-\affilnum{3}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
-\affilnum{4}Laboratoire Ville Mobilit{\'e} Transport, Universit{\'e} Paris-Est, France\\
+\affilnum{2}Laboratoire Ville Mobilit{\'e} Transport, Universit{\'e} Paris-Est, France\\
+\affilnum{3}Centre for Advanced Spatial Analysis, University College London, UK\\
+\affilnum{4}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
 \affilnum{5}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}

### 2017-07-18 paper.tex
Update on Overleaf.

@@
 
-\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.}
-\paragraph{Because agents are not random walkers}
+\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.} \comment[MLT]{Du coup, ça je le garderais plutôt pour la discussion, à moins que l'on avance des références explicites mais je crois que la sous-section d'après explicite assez bien ce besoin.}
 
-We expect spatial patterns to influence model outcomes also because the agents' rule of action itself might depend on the spatial structure of the environment.
+\comment[MLT]{Je propose de reformuler les 3 paragraphes ci-dessous de la manière suivante : The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009}.
+In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to be (partially) driven by the initial spatial setting, making it difficult to interprete the resulting global structures.
+At initialisation, we then expect the spatial distribution of agents to influence results in the long run \cite{Castellanoetal2009}.
+Although this may seem obvious, cities are not regular grids of isotropic densities.
+Modelling may be about abstracting features to highlight processes in a way which is simpler than reality.
+However, the uniform grid which represents space in most simulation models is potentially not enough to represent urban processes because density and accessibility have environmental, economic and social consequences.
+An intermediate and more meaningful way of abstracting space might thus be to consider, not the peculiarities of every city, but their broad density structures.
+In Europe for example, one can find broad types of density distributions (\cite{LeNechet2015}).
+Secondly we expect spatial patterns to influence model outcomes because the agents' rule of action itself might depend on the spatial structure of the environment.
 Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.
@@
 Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.
-\comment[MLT]{Du coup, je mettrais ce paragraphe à la suite du suivant.} \comment[FL]{je suis d'accord}
-
-\paragraph{Because the evolution of cities is path dependent}
-
-The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolve irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009} \comment[MLT]{Ca fait très name dropping mais je voulais souligner que l'on connaissait la biblio hors Denise hihi.}.
-In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to be (partially) driven by the initial spatial setting, making it difficult to interprete the resulting global structures.
-A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
-Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
+A first approach could be to evaluate the effects of the same set of social mechanisms, but modelled in different types of cities, for example monocentric, polycentric and discontinuous.}
 
@@
 In Europe for example, one can find broad types of density distributions (\cite{LeNechet2015}).
-At initialisation, we then expect agents to be distributed in different patterns, which influence results in the long run \cite{Castellanoetal2009}.
+At initialisation, we then expect agents to be distributed in different patterns, which influence results in the long run \citep{Castellanoetal2009}.
 A first approach could be to evaluate the effects of the same set of social mechanisms, but modelled in different types of cities, for example monocentric, polycentric and discontinuous.
+
+\paragraph{Because the evolution of cities is path dependent}
+
+The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009}.
+In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to be (partially) driven by the initial spatial setting, making it difficult to interprete the resulting global structures.
+A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
+Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
+
+\paragraph{Because agents are not random walkers}
+
+We expect spatial patterns to influence model outcomes also because the agents' rule of action itself might depend on the spatial structure of the environment.
+Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.
+For example, agents tend to create buffer zones if they are allowed network-based rather than isotropic movements (\cite{Banos2012}), while the vision / sensing mechanism is sensitive to the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}).
+Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.
 
@@
 
+\comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}
+
 % \hfill \break

### 2017-07-20 paper.tex
.

@@
 In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters and relating these variations of values to the variations in the simulation results, we here introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions.
-Initial spatial conditions are clustered to represent types of spaces ex-ante (for example: moonocentric or polycentric density grids), and the sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types.
+Initial spatial conditions are clustered to represent types of spaces ex-ante (for example: monocentric or polycentric density grids), and the sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types.
 It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.
@@
 In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$.
-More precisely, each added unit has a probability proportional to $P_i/\sum P_k$ to be added to patch $i$ with population $P_i$.
+More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
 At the end of each time step, this first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches).
@@
 Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
-\comment{A detailler en incluant les equations de morphogenese}
+\comment{A detailler en incluant les equations de morphogenese}[(JR) j'ai détaillé le processus, c'est suffisant ?]
 
@@
 	\includegraphics{figures/spatialGen.png}
-} \caption{Four examples of grids produced by the spatial generator. The lighter the red, the denser the area.} \label{fig:spatialGen} \end{center} \end{figure*} %
+} \caption{Four examples of grids produced by the spatial generator.
+The lighter the red, the denser the area.
+Changing the growth rate $N$ allows to have more or less chaotic shapes corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$.} \label{fig:spatialGen} \end{center} \end{figure*} %
 %%%%%%%%%%%%%%%%%%%%%%
@@
 A solution is to use systematic quantitative procedures.
-Several potential methods could be used \comment[MLT]{Ici je me demande si l'on ne devrait pas simplement dire que l'on a pas trouvé de méthodes de comparaisons systématiques de diagrammes de phase dans la littérature SMA ou geographic modelling, et que du coup on propose de comparer la variance intra à la variance inter-grille selon la formule ...
-Et je garderais la précision sur les autres méthodes que l'on aurait pu choisir pour la section de discussion}: for example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
-The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
-Map comparison methods, popular in environmental sciences \comment[MLT]{Et pas en géo?
-Qu'est-ce que ces méthodes ont de particulier?}, provide numeral tools to compare two dimensional fields~\cite{visser2006map}.
-To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}.
-To keep it simple and as such methodological considerations are auxiliary to the main purpose of this paper, we propose an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability.
+To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature.
+Several potential methods from other fields such as environmental science could be used, but we keep it simple and such methodological considerations are furthermore auxiliary to the main purpose of this paper.
+we propose therefore an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability.
 More formally, the distance is given by
@@
 
-where $\alpha \mapsto \left[\vec{x} \mapsto f_{\vec{\alpha}}\left(\vec{x}\right)\right]$ is the operator giving phase diagrams with $\vec{x}$ parameters and $\vec{\alpha}$ meta-parameters, and $d$ is a distance between probability distributions that can be taken for example as basic L2 distance \comment[MLT]{Euclidean?
-Je crois qu'il faut adapter le vocabulaire utilisé aux lecteurs d'EPB, qu'en pensez-vous?} or the Earth's Mover Distance \comment[MLT]{Si l'on mentionne les deux, ne va-t-on pas nous demander de tester l'effet du choix de l'une ou de l'autre sur nos résultats?}.
-For each values $\vec{\alpha_i}$, the phase diagram is seen as a random spatial field, facilitating the definition of variances and distance.
-\comment[FL]{je pense que c'est un peu compliqué effectivement cette partie.}
+where $f_{\vec{\alpha}}(\vec{x})$ is the phase diagrams on parameters $\vec{x}$ at given meta-parameters $\vec{\alpha}$.
+$d$ is a functional distance that we take simply as Euclidian distance.
+The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$.
+
+\comment[JR]{j'ai simplifié au max}
 
@@
 In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file.
-\comment[MLT]{Je me suis permis de bouger la partie sur le multi-modeling en fin de paragraphe, mais ai essayé d'en garder l'esprit tel quel!} In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modeling claimed by \cite{cottineau2016back}.
+In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modeling claimed by \cite{cottineau2016back}.
 
@@
 
-We use our set of initial density grids in two spatial simulation models: Schelling and Sugarscape.
+We use our set of initial density grids in two ``classical'' spatial simulation models: Schelling and Sugarscape.
 
@@
 
-\comment[FL]{il faudrait peut être davantage renforcer l'analyse a propos de figure 1 de si c'est plutôt les mu parameters ou les spatial parameters qui ont une grosse influence sur les résultats.}
+\comment[FL]{il faudrait peut être davantage renforcer l'analyse a propos de figure 1 de si c'est plutôt les mu parameters ou les spatial parameters qui ont une grosse influence sur les résultats.}[(JR) assez explicite pour sugarscape, j'ai rajouté un peu tout de même.]
 
@@
 In terms of model processes, it shows that congestion mechanisms induce rapidly higher levels of inequality.
+To put these results in perspective of our workflow given in Fig.~\ref{fig:method}, we have a sensitivity to spatial parameters in average greater than the sensitivity to model parameters.
 
@@
 
-\comment{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}
+\comment[FL]{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}[(JR) bonne idée, y'a pas mal d'exemples ou il faut pas mal contourner l'usage ``naturel'' de Netlogo pour s'en sortir - sur les représentations raster/vecteur c'est souvent un dilemne et ça peut induire des biais d'implémentation - je vais essayer de trouver un exemple simple (ceux que j'ai là trop usine à gaz, exemple Lutecia)]
 
 \comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}
+
+\paragraph{Comparing phase diagrams}
+
+Comparing phase diagrams is as we saw not straightforward, and further developments of our method imply testing alternative methods for this particular point.
+For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
+The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
+Map comparison methods, popular in environmental sciences \comment[MLT]{Et pas en géo?
+Qu'est-ce que ces méthodes ont de particulier?}, provide numeral tools to compare two dimensional fields~\cite{visser2006map}.
+To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}.
+Note that with the intuitive index we used, more subtle distances such as the Earth's Mover Distance can be used.
 

### 2017-07-20 paper.tex
Update on Overleaf.

@@
 
-\comment[MLT]{Je propose de reformuler les 3 paragraphes ci-dessous de la manière suivante : The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009}.
+\paragraph{Because actual cities are not regular grids}
+
+\comment[MLT]{Je propose de reformuler les 3 paragraphes ci-dessous plus ou moins de la manière suivante :} The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009}.
 In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to be (partially) driven by the initial spatial setting, making it difficult to interprete the resulting global structures.
@@
 Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.
-A first approach could be to evaluate the effects of the same set of social mechanisms, but modelled in different types of cities, for example monocentric, polycentric and discontinuous.}
 
-\paragraph{Because actual cities are not regular grids}
-Although this may seem obvious, cities are not regular grids of isotropic densities.
+%\paragraph{Because actual cities are not regular grids}
+%Although this may seem obvious, cities are not regular grids of isotropic densities.
 Modelling may be about abstracting features to highlight processes in a way which is simpler than reality.
@@
 
-\paragraph{Because the evolution of cities is path dependent}
-
-The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009}.
+%\paragraph{Because the evolution of cities is path dependent}
+%The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009}.
 In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to be (partially) driven by the initial spatial setting, making it difficult to interprete the resulting global structures.
@@
 
-\paragraph{Because agents are not random walkers}
-
-We expect spatial patterns to influence model outcomes also because the agents' rule of action itself might depend on the spatial structure of the environment.
+%\paragraph{Because agents are not random walkers}
+% We expect spatial patterns to influence model outcomes also because the agents' rule of action itself might depend on the spatial structure of the environment.
 Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.
@@
 
+%\comment[FL]{Je pense que cette partie est vraiment utile, et devrait être étoffée, en procédant peut être à une typologie plus systématique des "initial condition" qui nous intéressent.
+Il y a déjà une différence entre l'hétérogénéité de l'espace (densités réparties mono ou poly, etc) et les différentes façons de représenter l'espace (grilles carrées, hexagones).
+Par ailleurs aussi des discussions sur l'incertitude de l'information (est ce que par exemple les emplois downscalés au carreau INSEE par Pivano 2016 c'est vraiment fiable ou très ?) et la précision de la grille}
+
 The effect of the spatial configuration on area-based attributes of human behaviours has been largely discussed in geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984,FotheringhamWong1991}.
 Recently, \citet{Kwan2012} claims for a careful examination of what she coins the uncertain geographic context problem (UGCoP), that is of the spatial configuration of geographical units even if the size and delineation of the area are the same.
-On the contrary, the scarcity of these considerations in the geographic simulation model literature questions the generalisation of their results, as it has for instance been showed in the case of LUTI models \citep{Thomasetal2017} and of diffusion processes using ABM \citep{LeTexierCaruso2017}.
-\comment[FL]{Je pense que cette partie est vraiment utile, et devrait être étoffée, en procédant peut être à une typologie plus systématique des "initial condition" qui nous intéressent.
-Il y a déjà une différence entre l'hétérogénéité de l'espace (densités réparties mono ou poly, etc) et les différentes façons de représenter l'espace (grilles carrées, hexagones).
-Par ailleurs aussi des discussions sur l'incertitude de l'information (est ce que par exemple les emplois downscalés au carreau INSEE par Pivano 2016 c'est vraiment fiable ou très ?) et la précision de la grille}
+Considerations of such issues in the geographic simulation model literature are rather scarce, but they have been some noticeable examples of papers attempting to evaluation the effects of the same set of mechanisms runned from different initial spatial conditions.
+Mainly, these approaches have been focusing on the impact of geo-localised input data accuracy, of the spatial system shape, precision, and boundaries, as well as of the impact of spatial heterogeneity.
+
+\textbf{Geo-localised input data accuracy:} \citep{Thomasetal2017} show that data selection in LUTI model is interelated to the delineation of the spatial sytem boundaries and scale of analysis.
+They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs help avoiding measurement errors of model behaviour and outcomes.
+
+\textbf{Spatial system shape, precision, and boundaries:} \citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are highest in a rectangular grid and lowest in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents.
+\citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours.
+\citet{LeTexierCaruso2017} simulate an individual-based dynamic model where agents move and exchange coins across regions.
+Using a set of different theoretical spatial systems, they demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics.
+
+\textbf{Spatial heterogeneity:} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
+They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns.
+\citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents.
+They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile.
+\citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents).
+
+\textbf{Combination:} \citet{Singhetal2009} show that the segregation patterns obtained by Schelling for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.
+
+In this paper, we build on the contributions of these previous studies and go further by systematically measuring the impact of the city lattice on model behaviour so far only assessed through a small set of illustrative spatial structures.
+We illustrate the potential generalization of our approach by running two distinctive agent-based models: Schelling's model of residential segregation and Sugarscape.
 
@@
 On harmonise dans quel sens?}
-
-%Recent attempts of running Schelling-like models beyond the chessboard grid have proved that migration decisions are not only dependent of what positions agents occupy relative to each other, but also of where they occupy these positions.
-\citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents are highest in a rectangular grid and lowest in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents.
-\citet{StaufferSolomon2007} introduce asymetric interactions and empty residences on a large and regular lattice.
-They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns.
-\citet{Singhetal2009} show that the segregation patterns obtained by Schelling for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.
-\citet{Gauvinetal2010} run Schelling's segregation process in an open city-lattice to study how the variations in tolerance levels, vacancy rates and city attractiveness may create lines of vacancy lots between clusters of agents.
-They conclude on the functional role of vacancies, which allow weakly tolerant agents to live and be satisfied in a city environment they nevertheless perceive as hostile.
-\citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours.
-Interestingly, \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents).
-In this paper, we build on the contributions of these previous studies and go further by systematically measuring the impact of the city lattice on Schelling model behaviour so far only assessed through a small set of illustrative spatial structures.
-\comment[MLT]{Le remettre en moins détaillé dans l'intro en lien avec les catégories de prise en compte de la forme de la grille.}
 

### 2017-07-20 paper.tex
.

--- 2017-07-20 paper.tex
@@
 
-\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.} \comment[MLT]{Du coup, ça je le garderais plutôt pour la discussion, à moins que l'on avance des références explicites mais je crois que la sous-section d'après explicite assez bien ce besoin.}
-
-\comment[MLT]{Je propose de reformuler les 3 paragraphes ci-dessous de la manière suivante : The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009}.
-In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to be (partially) driven by the initial spatial setting, making it difficult to interprete the resulting global structures.
-At initialisation, we then expect the spatial distribution of agents to influence results in the long run \cite{Castellanoetal2009}.
+An urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects: self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002,Batty2007}.
+In that frame, complexity is partially due to bifurcations \citep{Wilson1981} which are determinant in spatial systems~\citep{Wilson2002}.
+\comment[JR]{je suis pas fan des train de citations ; peut-on détailler pour certaines refs au moins ce qu'elles apportent ?
+- j'ai formulé une proposition.}
+ In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interprete the resulting global structures.
+\comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}
+ % A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
+Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
+ At initialisation, we then expect the spatial distribution of agents to influence results in the long run \citep{Castellanoetal2009}.
 Although this may seem obvious, cities are not regular grids of isotropic densities.
@@
 Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.
-For example, agents tend to create buffer zones if they are allowed network-based rather than isotropic movements (\cite{Banos2012}), while the vision / sensing mechanism is sensitive to the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}).
+For example, agents tend to create buffer zones if they are allowed network-based rather than isotropic movements (\cite{Banos2012}), while the vision and sensing mechanism are sensitive to the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}).
 Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.
-A first approach could be to evaluate the effects of the same set of social mechanisms, but modelled in different types of cities, for example monocentric, polycentric and discontinuous.}
-
-\paragraph{Because actual cities are not regular grids}
-Although this may seem obvious, cities are not regular grids of isotropic densities.
-Modelling may be about abstracting features to highlight processes in a way which is simpler than reality.
-However, the uniform grid which represents space in most simulation models is potentially not enough to represent urban processes because density and accessibility have environmental, economic and social consequences.
-An intermediate and more meaningful way of abstracting space might thus be to consider, not the peculiarities of every city, but their broad density structures.
-In Europe for example, one can find broad types of density distributions (\cite{LeNechet2015}).
-At initialisation, we then expect agents to be distributed in different patterns, which influence results in the long run \citep{Castellanoetal2009}.
 A first approach could be to evaluate the effects of the same set of social mechanisms, but modelled in different types of cities, for example monocentric, polycentric and discontinuous.
 
-\paragraph{Because the evolution of cities is path dependent}
-
-The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009}.
-In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to be (partially) driven by the initial spatial setting, making it difficult to interprete the resulting global structures.
-A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
-Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
-
-\paragraph{Because agents are not random walkers}
-
-We expect spatial patterns to influence model outcomes also because the agents' rule of action itself might depend on the spatial structure of the environment.
-Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.
-For example, agents tend to create buffer zones if they are allowed network-based rather than isotropic movements (\cite{Banos2012}), while the vision / sensing mechanism is sensitive to the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}).
-Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.
-
-\paragraph{Considerations of urban form in simulation models}
+\subsection{Considerations of urban form in simulation models}
 
@@
 The lighter the red, the denser the area.
-Changing the growth rate $N$ allows to have more or less chaotic shapes corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$.} \label{fig:spatialGen} \end{center} \end{figure*} %
+Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$.} \label{fig:spatialGen} \end{center} \end{figure*} %
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
+\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.} \comment[MLT]{Du coup, ça je le garderais plutôt pour la discussion, à moins que l'on avance des références explicites mais je crois que la sous-section d'après explicite assez bien ce besoin.}
+
 \paragraph{Comparing phase diagrams}

### 2017-07-20 paper.tex

--- 2017-07-20 paper.tex
@@
 
-\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.} \comment[MLT]{Du coup, ça je le garderais plutôt pour la discussion, à moins que l'on avance des références explicites mais je crois que la sous-section d'après explicite assez bien ce besoin.}
-
-\paragraph{Because actual cities are not regular grids}
-
-\comment[MLT]{Je propose de reformuler les 3 paragraphes ci-dessous plus ou moins de la manière suivante :} The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009}.
-In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to be (partially) driven by the initial spatial setting, making it difficult to interprete the resulting global structures.
+An urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects: self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002,Batty2007}.
+In that frame, complexity is partially due to bifurcations \citep{Wilson1981} which are determinant in spatial systems~\citep{Wilson2002}.
+\comment[JR]{je suis pas fan des train de citations ; peut-on détailler pour certaines refs au moins ce qu'elles apportent ?
+- j'ai formulé une proposition.}
+ In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interprete the resulting global structures.
+\comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}
+ % A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
+Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
 At initialisation, we then expect the spatial distribution of agents to influence results in the long run \cite{Castellanoetal2009}.
@@
 Secondly we expect spatial patterns to influence model outcomes because the agents' rule of action itself might depend on the spatial structure of the environment.
-Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.
-For example, agents tend to create buffer zones if they are allowed network-based rather than isotropic movements (\cite{Banos2012}), while the vision / sensing mechanism is sensitive to the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}).
-Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.
-
-%\paragraph{Because actual cities are not regular grids}
-%Although this may seem obvious, cities are not regular grids of isotropic densities.
-Modelling may be about abstracting features to highlight processes in a way which is simpler than reality.
-However, the uniform grid which represents space in most simulation models is potentially not enough to represent urban processes because density and accessibility have environmental, economic and social consequences.
-An intermediate and more meaningful way of abstracting space might thus be to consider, not the peculiarities of every city, but their broad density structures.
-In Europe for example, one can find broad types of density distributions (\cite{LeNechet2015}).
-At initialisation, we then expect agents to be distributed in different patterns, which influence results in the long run \citep{Castellanoetal2009}.
-A first approach could be to evaluate the effects of the same set of social mechanisms, but modelled in different types of cities, for example monocentric, polycentric and discontinuous.
-
-%\paragraph{Because the evolution of cities is path dependent}
-%The urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects \citep{AllenSanglier1981,Portugali2000,Wilson1981,Wilson2002,Wu2002,Batty2007,AzizAlaouiBertelle2009}.
-In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to be (partially) driven by the initial spatial setting, making it difficult to interprete the resulting global structures.
-A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
-Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
-
-%\paragraph{Because agents are not random walkers}
-% We expect spatial patterns to influence model outcomes also because the agents' rule of action itself might depend on the spatial structure of the environment.
 Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.

### 2017-07-20 paper.tex
.
Update on Overleaf.

@@
 
-An urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects: self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002,Batty2007}.
+An urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002,Batty2007}.
 In that frame, complexity is partially due to bifurcations \citep{Wilson1981} which are determinant in spatial systems~\citep{Wilson2002}.
-\comment[JR]{je suis pas fan des train de citations ; peut-on détailler pour certaines refs au moins ce qu'elles apportent ?
-- j'ai formulé une proposition.}
 In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interprete the resulting global structures.
-\comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}
+\comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}\comment[MLT]{Non, mais je pense que la seconde phrase pourrait être reformulée/développée pour mieux raccrocher avec le reste, qu'en penses-tu?}
  % A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
 Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
+
 At initialisation, we then expect the spatial distribution of agents to influence results in the long run \cite{Castellanoetal2009}.

### 2017-07-20 paper.tex
Update on Overleaf.

@@
 \subsection{Why spatial patterns are expected to impact social simulations}
-
+\comment[MLT]{Je me demande si tout ce qui vient après jusqu'à la littérature review sur les attempts de prise en compte de l'effet de la configuration spatiale sur le comportement et les résultats des modèles ne serait pas mieux en intro?
+Ca ferait une longue intro mais ça éviterait de trop se répéter?}
 An urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects.
@@
 In that frame, complexity is partially due to bifurcations \citep{Wilson1981} which are determinant in spatial systems~\citep{Wilson2002}.
+
 In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interprete the resulting global structures.
-\comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}\comment[MLT]{Non, mais je pense que la seconde phrase pourrait être reformulée/développée pour mieux raccrocher avec le reste, qu'en penses-tu?}
- % A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
+\comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}\comment[MLT]{Non, mais je pense que la seconde phrase pourrait être reformulée/développée pour mieux raccrocher avec le reste, qu'en penses-tu?} A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
 Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
 
-At initialisation, we then expect the spatial distribution of agents to influence results in the long run \cite{Castellanoetal2009}.
+At initialisation, we then expect the spatial distribution of agents to influence results in the long run \cite{Castellanoetal2009}, as agents' rule of action itself might depend on the spatial structure of the environment: households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), mechanisms of surrounding sensing are impacted by different distributions of density (\cite{Banos2012}) or the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}), etc.
+The way modellers abstract space is therefore a central element of any spatially explicite simulation model.
 Although this may seem obvious, cities are not regular grids of isotropic densities.
-Modelling may be about abstracting features to highlight processes in a way which is simpler than reality.
 However, the uniform grid which represents space in most simulation models is potentially not enough to represent urban processes because density and accessibility have environmental, economic and social consequences.
@@
 In Europe for example, one can find broad types of density distributions (\cite{LeNechet2015}).
-Secondly we expect spatial patterns to influence model outcomes because the agents' rule of action itself might depend on the spatial structure of the environment.
-Indeed, mechanisms of surrounding sensing will be impacted by different distributions of density.
-For example, agents tend to create buffer zones if they are allowed network-based rather than isotropic movements (\cite{Banos2012}), while the vision / sensing mechanism is sensitive to the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}).
-Finally, households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), thus creating a sensitivity to the initial form of the city modelled.
 
-\paragraph{Considerations of urban form in simulation models}
+\paragraph{Previous considerations of urban form in simulation models}
 
@@
 Recently, \citet{Kwan2012} claims for a careful examination of what she coins the uncertain geographic context problem (UGCoP), that is of the spatial configuration of geographical units even if the size and delineation of the area are the same.
-Considerations of such issues in the geographic simulation model literature are rather scarce, but they have been some noticeable examples of papers attempting to evaluation the effects of the same set of mechanisms runned from different initial spatial conditions.
+Considerations of such issues in the geographic simulation model literature are rather scarce, but they have been some noticeable examples of papers attempting to evaluate the effects of the same set of mechanisms runned from different initial spatial conditions.
 Mainly, these approaches have been focusing on the impact of geo-localised input data accuracy, of the spatial system shape, precision, and boundaries, as well as of the impact of spatial heterogeneity.
@@
 Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}).
-In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step, allocating it through preferential attachment characterised by its strength of attraction $\alpha$.
+In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment characterised by its strength of attraction $\alpha$.
 More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
 At the end of each time step, this first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches).
-To avoid bord effects such as a reflexion on the border of the world, border patches diffuse to the outside.
-The model stops when a fixed total population is reached.
+To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside.
+The model stops when a fixed total population \comment{n?} is reached.
 Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed.
@@
 Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
-\comment{A detailler en incluant les equations de morphogenese}[(JR) j'ai détaillé le processus, c'est suffisant ?]
+\comment{A detailler en incluant les equations de morphogenese}[(JR) j'ai détaillé le processus, c'est suffisant ?]\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n).
+Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$?
+au pas de temps $t$?
+Pourquoi ne pas indexer alpha?
+Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.}
 
@@
 Several potential methods from other fields such as environmental science could be used, but we keep it simple and such methodological considerations are furthermore auxiliary to the main purpose of this paper.
-we propose therefore an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability.
+We propose therefore an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability.
 More formally, the distance is given by
@@
 
-\comment[JR]{j'ai simplifié au max}
+\comment[JR]{j'ai simplifié au max}\comment[mLT]{Attention, on utilise alpha plus haut.}
 
@@
 
-We use our set of initial density grids in two ``classical'' spatial simulation models: Schelling and Sugarscape.
+In this section, we briefly recall the main components of the two ``classical'' agent-based simulation models used to test how spatial density variations may impact simulation models behaviour and results, and how general the method proposed is.
 
@@
 
-Schelling's model consists in a abstract urban housing market where agents of different nature sense their environment, evaluate their satisfaction in terms of neighbourhood composition, and relocate if unsatisfied.
+Schelling's model consists in an abstract urban housing market where agents of different nature sense their environment, evaluate their satisfaction in terms of neighbourhood composition, and relocate if unsatisfied.
 It has been shown by \cite{Schelling1969} that even tolerant agents tended to produce segregated patterns because of the complexity of their local interactions.
@@
 In addition, we are interested in testing the impact of the spatial distribution of housing capacity in this project, using the generated grids.
-\comment[MLT]{Dans le paragraphe sur Sugarscape on parle du diagramme de phase que l'on étudie en sortie, pas ici.
-On harmonise dans quel sens?}
+\comment{We extend the implementation of...
+The outcome of the model is measured as a phase diagram of segregation level (Dissimilarity, Moran's I, Entropy, Exposure, all four of them?).
+.}
 
@@
 In addition, we are interested in testing the impact of the spatial distribution of the resource in this project, using the generated grids.
+We extend the implementation with agents wealth distribution of~\cite{li2009netlogo}.
 The outcome of the model is measured as a phase diagram of an index of inequality for ressource distribution (Gini index).
-We extend the implementation with agents wealth distribution of~\cite{li2009netlogo}.
 

### 2017-07-20 paper.tex
.

--- 2017-07-20 paper.tex
@@
 In that frame, complexity is partially due to bifurcations \citep{Wilson1981} which are determinant in spatial systems~\citep{Wilson2002}.
-In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interprete the resulting global structures.
+In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures.
 \comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}\comment[MLT]{Non, mais je pense que la seconde phrase pourrait être reformulée/développée pour mieux raccrocher avec le reste, qu'en penses-tu?}

### 2017-07-21 paper.tex
Update on Overleaf.

--- 2017-07-20 paper.tex
@@
 
-\subsection{Objective}
-
-In this paper, we suggest tackling sensitivity to spatial conditions by generating a variety of density grids to feed into simulation models at initialisation, and exploring the sensitivity of the model outcomes to the variation of meta-parameters (i.e.
-parameters used to generate initial spatial conditions).
-The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters within a typical category of space (a monocentric case for example) and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric vs.
-polycentric for example) on the results of a given model.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Why spatial patterns are expected to impact social simulations}
-\comment[MLT]{Je me demande si tout ce qui vient après jusqu'à la littérature review sur les attempts de prise en compte de l'effet de la configuration spatiale sur le comportement et les résultats des modèles ne serait pas mieux en intro?
-Ca ferait une longue intro mais ça éviterait de trop se répéter?}
 An urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects.
@@
 
+\subsection{Objective}
+
+In this paper, we suggest tackling sensitivity to spatial conditions by generating a variety of density grids to feed into simulation models at initialisation, and exploring the sensitivity of the model outcomes to the variation of meta-parameters (i.e.
+parameters used to generate initial spatial conditions).
+The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters within a typical category of space (a monocentric case for example\comment[FL]{je ne trouve pas cela facile à comprendre car on n'a pas encore parlé des différentes formes de villes}) and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric vs.
+polycentric for example) on the results of a given model.
+
+%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Why spatial patterns are expected to impact social simulations}
+
+\comment[MLT]{Je me demande si tout ce qui vient après jusqu'à la littérature review sur les attempts de prise en compte de l'effet de la configuration spatiale sur le comportement et les résultats des modèles ne serait pas mieux en intro?
+Ca ferait une longue intro mais ça éviterait de trop se répéter?}
+
 \paragraph{Previous considerations of urban form in simulation models}
+\begin{itemize}
+\item spatial context
+\begin{itemize}
+\item extent
+\item objects
+\item shape
+\end{itemize}
 
+\item spatial "encoding"
+\begin{itemize}
+\item heterogeneity
+\item algorithm of desagregation
+\item uncertainty
+\end{itemize}
+
+\item spatial effects
+
+\begin{itemize}
+\item topology
+\item interaction (??)
+\end{itemize}
+\item spatial outputs
+\begin{itemize}
+\item agregation
+\item use of models
+\end{itemize}
+\end{itemize}
 %\comment[FL]{Je pense que cette partie est vraiment utile, et devrait être étoffée, en procédant peut être à une typologie plus systématique des "initial condition" qui nous intéressent.

### 2017-07-21 paper.tex
Update on Overleaf.

@@
 parameters used to generate initial spatial conditions).
-The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters within a typical category of space (a monocentric case for example\comment[FL]{je ne trouve pas cela facile à comprendre car on n'a pas encore parlé des différentes formes de villes}) and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric vs.
-polycentric for example) on the results of a given model.
+The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters within a typical category of space (a monocentric case for example\comment[FL]{je ne trouve pas cela facile à comprendre car on n'a pas encore parlé des différentes formes de villes}) and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
 
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Why spatial patterns are expected to impact social simulations}
+\subsection{Previous considerations of urban form in simulation models}
 
-\comment[MLT]{Je me demande si tout ce qui vient après jusqu'à la littérature review sur les attempts de prise en compte de l'effet de la configuration spatiale sur le comportement et les résultats des modèles ne serait pas mieux en intro?
-Ca ferait une longue intro mais ça éviterait de trop se répéter?}
-
-\paragraph{Previous considerations of urban form in simulation models}
 \begin{itemize}
@@
 \item heterogeneity
-\item algorithm of desagregation
+\item algorithm of disaggregation
 \item uncertainty
@@
 \item spatial effects
-
 \begin{itemize}
@@
 \end{itemize}
+
 \item spatial outputs
 \begin{itemize}
-\item agregation
+\item aggregation
 \item use of models
@@
 Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}).
-In a nutshell, grids are generated through an iterative process which adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment characterised by its strength of attraction $\alpha$.
+In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment \comment[FL]{à quoi?} characterised by its strength of attraction $\alpha$.
 More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
-At the end of each time step, this first growth process is then smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches).
-To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside.
+At the end of each time step, this growth process is smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches).
+To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside\comment[FL]{comment est-ce qu'on contrôle la perte de population induite?}.
 The model stops when a fixed total population \comment{n?} is reached.
@@
 Pourquoi ne pas indexer alpha?
-Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.}
+Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.} \comment[FL]{je crois que N c'est l'incrément de pop et n le nombre de fois où c'est smoothé mais c'est quoi la pop totale?}
 
@@
 
-In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions: their concentration index, hierarchy index, centrality index and continuity index (cf.
+In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure : their concentration index, hierarchy index, centrality index and continuity index (cf.
 \cite{LeNechet2015}).
@@
 
-In order to test for the influence of spatial initial conditions, we need a systematic method to compare phase diagrams.
+In order to test for the influence of spatial initial conditions on model outputs, we need a systematic method to compare phase diagrams.
 Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic.
@@
 The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows.
-The ideas of multi-modeling and extensive model exploration are nothing from new as Openshaw already advocated for ``model-crunching'' in~\cite{openshaw1983data}, but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities.
-The model exploration platform OpenMole~\cite{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
+The ideas of multi-modeling and extensive model exploration are nothing from new as Openshaw already advocated for ``model-crunching'' in, but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities.
+The model exploration platform OpenMole~\citep{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
 In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file.
@@
 
+\comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
+
+\comment[MLT]{Mise à disposition des grilles générées? D'un échantillon?}
+
 \paragraph{Comparing phase diagrams}

### 2017-07-21 paper.tex
Update on Overleaf.

@@
 
+\comment{Ajouter un paragraphe sur le lien entre forme de grille et comportement du modèle/résultats.}
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.} \comment[MLT]{Du coup, ça je le garderais plutôt pour la discussion, à moins que l'on avance des références explicites mais je crois que la sous-section d'après explicite assez bien ce besoin.}
+\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.}
 

### 2017-08-03 paper.tex
Update on Overleaf.

@@
 
-\corrauth{Corresponding author
-address.}
-
-\email{c.cottineau@ucl.ac.uk}
+\corrauth{\email{juste.raimbault@polytechnique.edu}}
 
@@
 %\setlength{\parindent}{0pt}
-Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust.
-Sensitivity analysis usually include the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes.
-However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interaction of agents and of their interactions with the environment.
-In this contribution, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameter to create density grids used in spatial simulation models.
-We show, with the example of two very classical agent-based models (Schelling's models of segregation and Sugarscape) that the effect of space in simulation is significant, and sometimes even larger than parameters themselves.
-We do so using high performance computing in a very simple and straightforward open-source workflow.
+Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust and representative of 'real-world' conditions.
+Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes.
+However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment.
+In this paper, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create the density grids with which spatial simulation models are initialised.
+We show, with the example of two very classical agent-based models (Schelling's models of segregation and Sugarscape's model of resource harvesting) and a straightforward open-source work-flow using high performance computing, that the effect of space is significant in the two types of simulation.
+Furthermore, this effect is sometimes larger than the effect of parameters' value change.
 \end{abstract}
@@
 % **************  MAIN BODY OF THE PAPER **************
+\newpage
 
@@
 
-Simulation has been recognised and increasingly used by geographers to explore various geographical processes and problems within virtual laboratories \citep{Quesneletal2009} \comment{Batty 1978?
-James Doran 1970's : lire thèse de Seb ;) }.
-It appears as a very fruitful way to overcome the difficulty of analytic resolution of many spatial models developed in the past, and to explore the possible trajectories of social and ecological systems in space and time.
-The specificity of geographical models compared with models from other social sciences is usually regarded as the way geographers consider space and spatial interactions, driven by an explicit interest in the way space influences the outcomes of the model.
-Geographers are indeed concerned about understanding and modelling how space plays a role in social interactions and environmental processes, and whether its action is placed-based or place-neutral.
-We think simulation can become a very good tool to achieve this, provided that models include relevant spatial description and modelling, behavioural rules that take space into account, and provided model evaluation stresses the sensitivity of output variations to the way space is modelled.
-This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of model's outcomes to initial spatial conditions.
-To demonstrate the genericity of our approach, we develop two applications from simulation models which are commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996}: Schelling's model of segregation and Epstein and Axtell's Sugarscape model.
-Both models are defined at the intraurban level.
+Simulation within computer models has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}.
+\comment{+ Seb \& James Doran in the 1970s??} Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of social and ecological systems in space and time.
+The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model.
+Indeed, they usually seek to understand and model the role of space in social interactions and environmental processes, in a city for example.
+We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial description and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled.
+Unfortunately, the first condition is not always met, and the second is seldom even mentioned.
+This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions.
+To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model, both defined at the intra-urban level.\\
 
-An urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002,Batty2007}.
-In that frame, complexity is partially due to bifurcations \citep{Wilson1981} which are determinant in spatial systems~\citep{Wilson2002}.
+An urban system can be defined as a large number of diverse social agents interacting with one another, within a limited portion of space.
+Social agents thus constitute the microscopic level of urban systems, which are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.
+Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures.
+\\
+Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity~\citep{pumain2012urban}, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average.
+On the contrary, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time provide the use of ergodic models.
+Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
+\\
 
-In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interprete the resulting global structures.
-\comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}\comment[MLT]{Non, mais je pense que la seconde phrase pourrait être reformulée/développée pour mieux raccrocher avec le reste, qu'en penses-tu?} A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories.
-Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
-
-At initialisation, we then expect the spatial distribution of agents to influence results in the long run \cite{Castellanoetal2009}, as agents' rule of action itself might depend on the spatial structure of the environment: households can have different preferences with respect to the built-environment (\cite{SpielmanHarrison2014}), mechanisms of surrounding sensing are impacted by different distributions of density (\cite{Banos2012}) or the scale of modelled environments (\cite{LauriJaggi2003, FossettDietrich2009}), etc.
-The way modellers abstract space is therefore a central element of any spatially explicite simulation model.
-Although this may seem obvious, cities are not regular grids of isotropic densities.
-However, the uniform grid which represents space in most simulation models is potentially not enough to represent urban processes because density and accessibility have environmental, economic and social consequences.
-An intermediate and more meaningful way of abstracting space might thus be to consider, not the peculiarities of every city, but their broad density structures.
-In Europe for example, one can find broad types of density distributions (\cite{LeNechet2015}).
+Although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) in cities is far from isotropic, even in the most sprawled cities.
+On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities.
+In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions.
+However, most geographical models, especially cellular automata, represent cities as uniform grids of isotropic densities.
+Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}.
+The isotropic simplification is potentially harmful to the representation of the urban processes under inquiry because density and accessibility have environmental, economic and social consequences.
+Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment.
+For example, households can have different preferences with respect to the built-environment they want to live in \citep{SpielmanHarrison2014}, or agents can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}.
+The way modellers abstract space is therefore a central element of geographical simulation models.
+A meaningful way of abstracting space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures.
 
@@
 
-In this paper, we suggest tackling sensitivity to spatial conditions by generating a variety of density grids to feed into simulation models at initialisation, and exploring the sensitivity of the model outcomes to the variation of meta-parameters (i.e.
-parameters used to generate initial spatial conditions).
-The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters within a typical category of space (a monocentric case for example\comment[FL]{je ne trouve pas cela facile à comprendre car on n'a pas encore parlé des différentes formes de villes}) and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
+In this paper, we suggest a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation.
+The generator being controlled by meta-parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations.
+The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
+Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which has been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies, which tend to focus on single aspects of the debate.
+In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.
 
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Previous considerations of urban form in simulation models}
+\subsection{Previous considerations on the effects of the urban form in simulation models}
 
-\begin{itemize}
-\item spatial context
-\begin{itemize}
-\item extent
-\item objects
-\item shape
-\end{itemize}
-
-\item spatial "encoding"
-\begin{itemize}
-\item heterogeneity
-\item algorithm of disaggregation
-\item uncertainty
-\end{itemize}
-
-\item spatial effects
-\begin{itemize}
-\item topology
-\item interaction (??)
-\end{itemize}
-
-\item spatial outputs
-\begin{itemize}
-\item aggregation
-\item use of models
-\end{itemize}
-\end{itemize}
 %\comment[FL]{Je pense que cette partie est vraiment utile, et devrait être étoffée, en procédant peut être à une typologie plus systématique des "initial condition" qui nous intéressent.
@@
 
-The effect of the spatial configuration on area-based attributes of human behaviours has been largely discussed in geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984,FotheringhamWong1991}.
-Recently, \citet{Kwan2012} claims for a careful examination of what she coins the uncertain geographic context problem (UGCoP), that is of the spatial configuration of geographical units even if the size and delineation of the area are the same.
-Considerations of such issues in the geographic simulation model literature are rather scarce, but they have been some noticeable examples of papers attempting to evaluate the effects of the same set of mechanisms runned from different initial spatial conditions.
-Mainly, these approaches have been focusing on the impact of geo-localised input data accuracy, of the spatial system shape, precision, and boundaries, as well as of the impact of spatial heterogeneity.
+Different spatial configurations produce different distributions of income, carbon emissions, educational outcomes, etc.
+For example, \citet{wheeler2006urban} shows that, in the US, sprawling cities tend to be more unequal than their compact counterparts with respects to income.
+Dynamically, sprawl consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}.
+Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that{\it "deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking"}.
+The discussion of these effects has also been extensive in the field of geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984, FotheringhamWong1991}.
+Recently, \citet{Kwan2012} has argued for a careful examination of what she coins the uncertain geographic context problem (UGCoP), i.e.
+of the spatial configuration of geographical units even when the size and delineation of the area are the same.\\
 
-\textbf{Geo-localised input data accuracy:} \citep{Thomasetal2017} show that data selection in LUTI model is interelated to the delineation of the spatial sytem boundaries and scale of analysis.
-They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs help avoiding measurement errors of model behaviour and outcomes.
+Considerations of such issues in the geographical modelling literature are rather scarce, although there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions.
+Mainly, they have focused on the impact of:
+\begin{itemize}
+\item The accuraty of geo-localised input data;
+\item The spatial system shape, precision and boundaries;
+\item Spatial heterogeneity.
+\end{itemize}
+We detail these approaches in the paragraphs below.
+Other approaches, not included in the scope of this article, include the effects of the topology of space on the processes (cf.
+\citet{moreno2009integrating} who explore the role of the topology of possible interactions in a Schelling model).
+Generally, sensitivity analyses focus either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely both at the same time.
 
-\textbf{Spatial system shape, precision, and boundaries:} \citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are highest in a rectangular grid and lowest in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents.
+\paragraph{Geo-localised input data accuracy:} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and scale of analysis.
+They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes.
+In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.
+
+\paragraph{Spatial system shape, precision, and boundaries:} \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e.
+made equivalent to) to Axelrod Culture Model.
+\citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are higher in a rectangular grid and lower in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents.
 \citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours.
-\citet{LeTexierCaruso2017} simulate an individual-based dynamic model where agents move and exchange coins across regions.
-Using a set of different theoretical spatial systems, they demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics.
+\citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins.
+\comment[FL]{je dois compléter\\
+\cite{horritt2001effects} identified the effects of grid cell size on the behaviour of a raster flood model
+\cite{vazquez2002effect} ont exploré la meilleure granularité pour la précision d'un modèle d'écoulement hydrologie et le temps de calcul (à reprendre)}
 
-\textbf{Spatial heterogeneity:} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
+\paragraph{Spatial heterogeneity:} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
 They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns.
@@
 
-\textbf{Combination:} \citet{Singhetal2009} show that the segregation patterns obtained by Schelling for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.
+\paragraph{Combination:} \citet{Singhetal2009} show that the segregation patterns obtained by Schelling for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.
 
-In this paper, we build on the contributions of these previous studies and go further by systematically measuring the impact of the city lattice on model behaviour so far only assessed through a small set of illustrative spatial structures.
-We illustrate the potential generalization of our approach by running two distinctive agent-based models: Schelling's model of residential segregation and Sugarscape.
+\comment[clem]{Ce serait cool d'avoir aussi des exemples de simu hors Schelling pour les deux dernieres sous-parties}
+
+In this paper, we build on the contributions of these previous studies and go further by systematically measuring the impact of the city lattice on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures.
+We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model of residential segregation and Sugarscape.
 
@@
 The general method workflow is illustrated in Figure \ref{fig:method}.
-In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters and relating these variations of values to the variations in the simulation results, we here introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions.
-Initial spatial conditions are clustered to represent types of spaces ex-ante (for example: monocentric or polycentric density grids), and the sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types.
-It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.
-
+In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters, we introduced the lower branch in figure \ref{fig:method}.
+We wanted to emphasize that spatial effects do not only derive from grid size or shape effects, but indeed from the heterogeneity of relevant geographic entities (people, housing, networks, etc.) through geographical space.
+In models, the initial configuration of space can be either flat (most economic models\comment[FL]{c'est peut être un peu brutal?}), or heterogeneous,  based on external databases and internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
+In our approach, we have introduced a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions. 
+To allow for the exploration of the relative effects of model parameters and spatial configuration, we clustered initial spatial conditions  to represent types of spaces ex-ante (for example: monocentric or polycentric density grids).
+The sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types.
+It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.
+
+In order to test for the influence of spatial initial conditions on model outputs, we use a systematic method to compare phase diagrams.
+We define a phase diagram as a vector of  outputs\comment[FL]{Juste, peut-être des choses qui sont mal dites?
+: par exemple, dans diagramme de phase est-ce qu'on entend pas une "réification" de catégories - des clusters par ex?
+Et du coup est-ce que cela ne manque pas là?}.
+
+Comparing phase diagrams is useful because it allows us to compare the 
+Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic\comment[FL]{donner un ordre de grandeur du nombre de simulations?}.
+A solution is to use systematic quantitative procedures.
+To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature.
+We used a distance between two phase diagram corresponding to the share of between-diagrams variability relative to their internal variability\comment[FL]{je ne comprends pas ce choix : pourquoi faut il considérer que des classes ayant une forte cohérence interne seraient plus éloignées?
+ou alors la question ne se pose pas en ces termes...
+Par ailleurs est ce que la formule est bien homogène du point de vue des dimensions?
+En tous cas je pense qu'on peut se passer de la formule à ce stade, éventuellement à mettre dans l'affichage des résultats si on en a besoin.}
+
+%Several potential methods from other fields such as environmental science could be used, but we keep it simple and such methodological considerations are furthermore auxiliary to the main purpose of this paper.
+We propose therefore an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability.
+More formally, the distance is given by
+
+%\begin{equation}\label{eq:phase-distance}
+%d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}%{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
+%\end{equation}
+
+%where $f_{\vec{\alpha}}(\vec{x})$ is the phase diagrams on parameters $\vec{x}$ at given meta-parameters $\vec{\alpha}$.
+$d$ is a functional distance that we take simply as Euclidian distance.
+The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$.
+
+\comment[JR]{j'ai simplifié au max}\comment[mLT]{Attention, on utilise alpha plus haut.}
+
+The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows.
+The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities.
+The model exploration platform OpenMole~\citep{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
+In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file.
+In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modelling claimed by \citet{cottineau2015modular}.
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructing a spatial generator}
+\subsection{Constructing a spatial generator of density grids}
 
@@
 At the end of each time step, this growth process is smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches).
-To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside\comment[FL]{comment est-ce qu'on contrôle la perte de population induite?}.
-The model stops when a fixed total population \comment{n?} is reached.
+To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside.
+The procedure stops when a fixed number of steps $k$ is reached.
+The grid then has a population of $k \times N$ (the population lost due to diffusion process to the outside is reallowed through a normalization procedure at the end of the steps).
 Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed.
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}
+\begin{figure*}[htbp] \begin{center} 
+\resizebox{0.9\textwidth}{!}{ 
+	\includegraphics{figures/GridPresentation.png}
+} \caption{Correspondence between European urban density structures and grids produced with the spatial generator.} 
+\label{fig:densityTypes} \end{center} \end{figure*} %
+%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%
+%\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}
 
@@
 \cite{LeNechet2015}).
-A stochastic exploration of a Latin Hypercube Sampling of 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n$, $N$} gives a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions.
+A stochastic exploration of a Latin Hypercube Sampling\comment[FL]{ce n'est pas explicite} of 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n$, $N$} gives a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions.
 These are further clustered into three classes of morphology: compact (e.g.
 Vienna), polycentric (Liege) and discontinuous (Augsburg) in order to evaluate the non-trivial effects of urban form on simulation results.
-We select 15 grids of each type to work with.
+We select 15 grids of each type to work with in the computation of sensitivity analysis.
+\comment[FL]{pourquoi 15?} \comment[FL]{dire explicitement ce qu'on va faire avec ces catégories dans la partie résultats?}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Comparing Phase Diagrams}
-
-In order to test for the influence of spatial initial conditions on model outputs, we need a systematic method to compare phase diagrams.
-Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic.
-A solution is to use systematic quantitative procedures.
-To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature.
-Several potential methods from other fields such as environmental science could be used, but we keep it simple and such methodological considerations are furthermore auxiliary to the main purpose of this paper.
-We propose therefore an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability.
-More formally, the distance is given by
-
-\begin{equation}\label{eq:phase-distance}
-d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
-\end{equation}
-
-where $f_{\vec{\alpha}}(\vec{x})$ is the phase diagrams on parameters $\vec{x}$ at given meta-parameters $\vec{\alpha}$.
-$d$ is a functional distance that we take simply as Euclidian distance.
-The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$.
-
-\comment[JR]{j'ai simplifié au max}\comment[mLT]{Attention, on utilise alpha plus haut.}
+%\subsection{Comparing Phase Diagrams}
+\comment[FL]{j'ai proposé de mettre cette partie au début de method et de la simplifier}
 
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Model Exploration Workflows as an meta-sensitivity analysis method}
+%\subsection{Model Exploration Workflows as an meta-sensitivity analysis method}
+\comment[FL]{item}
 
-The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows.
-The ideas of multi-modeling and extensive model exploration are nothing from new as Openshaw already advocated for ``model-crunching'' in, but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities.
-The model exploration platform OpenMole~\citep{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
-In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file.
-In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modeling claimed by \cite{cottineau2016back}.
+In the following section, we briefly recall the main components of the two ``classical'' agent-based simulation models used to test how spatial density variations may impact simulation models behaviour and results, and how general the method proposed is.
 
-%%%%%%%%%%%%%%%%%%%%%%
-\section{Application cases}
-%%%%%%%%%%%%%%%%%%%%%%
-
-In this section, we briefly recall the main components of the two ``classical'' agent-based simulation models used to test how spatial density variations may impact simulation models behaviour and results, and how general the method proposed is.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Schelling's model of residential segregation}
+\subsection{Case study models}
 
@@
 The outcome of the model is measured as a phase diagram of segregation level (Dissimilarity, Moran's I, Entropy, Exposure, all four of them?).
-.}
+.}\\
 
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sugarscape model of resource extraction and population settlement}
-
-Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population (\cite{EpsteinAxtell1996}).
+Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \citep{EpsteinAxtell1996}.
+Although it {\it "is designed to study the interaction of many plausible social mechanisms"} \citep[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where {\it "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its toroidal array of cells"}.
 Agents of different vision scopes and different metabolisms harvest a self-regenerating resource available heterogeneously in the initial landscape, they settle and collect this resource, which leads some of them to survive and others to perish.
@@
 In addition, we are interested in testing the impact of the spatial distribution of the resource in this project, using the generated grids.
-We extend the implementation with agents wealth distribution of~\cite{li2009netlogo}.
-The outcome of the model is measured as a phase diagram of an index of inequality for ressource distribution (Gini index).
+We extend the implementation with agents wealth distribution of~\citet{li2009netlogo}.
+The outcome of the model is measured as a phase diagram of an index of inequality for resource distribution (Gini index).
+
+\subsection{Experiment design}
+For Schelling's model, we explore a grid of a basic parameter space of the model, which is composed of two dimensions: the minimum proportion of similar agents required in the neighbourhood for the agent to be satisfied (or intolerance level) $S\in \left[0;1\right]$ and the vacancy rate of the city $V\in \left[0;1\right]$, which then defines the number of green and red agents, present in equal proportion for this experiment.
+Each parameter is binned into 10 values, giving 1000 parameter points.
+We run 100 repetitions for each configuration, what yield reasonable convergence properties.
+The initial spatial configuration varies across 45 different grids, chosen among the ones which are most representative of the three types of urban morphology: 15 compact grids (like Vienna, cf figure \ref{fig:densityTypes}), 15 polycentric grids (like Liege) and 15 discontinuous grids (like Augsburg).
+The full experiment thus equates to 4,500,000 simulations (1000 parameter combinations x 45 density grids x 100 replications).
+We use OpenMOLE to distribute the computation, and apply segregation measures to characterise the results.
+
+For Sugarscape, we explore a grid of a basic parameter space of the model, which is composed of three dimensions: the population of agents $P\in \left[10;510\right]$, the minimal initial agent resource $s_{-}\in \left[10;100\right]$ and the maximal initial agent resource $s_{+}\in \left[110;200\right]$.
+Each parameter is binned into 10 values, giving 1000 parameter points.
+We run 50 repetitions for each configuration, what yield reasonable convergence properties.
+The initial spatial configuration varies across 50 different grids, generated by sampling meta-parameters for the generator in a LHS.
+We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model.
+The full experiment thus equates to 2,500,000 simulations (1000 parameter combinations x 50 density grids x 50 replications).
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Schelling's model}
+\subsection{Quantitative variation across grids}
+\paragraph{Schelling} 
+Within a standard Schelling model (i.e.
+initialised with a uniform density grid), \citet{Gauvinetal2009} have built the phase diagram of segregation patterns depending on the combination of parameter values.
+For high levels of tolerance ($S < 0.25$), there is no segregation.
+For high values of vacancies ($V > 0.65$) and low values of tolerance ($S > 0.5$), these is a diluted segregation state where homogeneous communities are separated from other by large empty buffers.
+Finally, for low values of vacancies ($V < 0.2$) and low values of tolerance ($S > 0.7$), the model is frozen in a state where everyone is unhappy but no-one can express their intolerant behaviour due to the lack of free spaces.
+Between these extreme cases, the model gives rise to segregated states where homogeneous communities adjoin one another.
+The objective of this quantitative experiment is to evaluate to which extent this phase diagram is modified when different density grids are applied.
+\comment[CC]{!!!
+Il faut retrouver les fichiers de simu !!!} \comment[CC]{Juste, tu penses qu'il faudrait faire la meme analyse que toi sur la distance entre phase diagrams?
+Ce serait cool mais si oui, j'aurais besoin d'aide (et des fichiers de simu!)}
 
-We proceed to 4,500,000 simulation runs of the Schelling model (1000 parameter combinations x 45 density grids x 100 replications), using OpenMOLE to distribute the computation, and apply segregation measures to characterise the results.
-We find that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
-
-%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Sugarscape model}
-
-For Sugarscape, 2,500,000 simulations (1000 parameter points x 50 density grids x 50 replications) allow us to show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.
-
-More precisely, we explore a grid of a basic parameter space of the model, which three dimensions are the population of agents $P\in \left[10;510\right]$, the minimal initial agent ressource $s_{-}\in \left[10;100\right]$ and the maximal initial agent ressource $s_{+}\in \left[110;200\right]$.
-Each parameter is binned into 10 values, giving 1000 parameter points.
-We run 50 repetitions for each configuration, what yield reasonable convergence properties.
-The initial spatial configuration varies across 50 different grids, generated by sampling meta-parameters for the generator in a LHS.
-We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model.
-We mesure the distance of all 3-dimensional phase diagrams to the reference phase diagram computed on the default model setup (see Fig.~\ref{fig:sugarscape-distance} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability.
-Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each.
+\paragraph{Sugarscape} We measure the distance of all 3-dimensional phase diagrams to the reference phase diagram computed on the default model setup \comment[CC]{j'ai peut-etre manque une etape, mais c'est quoi le default model setup?
+en termes de grille d'entree: uniforme?
+random?
+ou Netlogo?} (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability.
+\comment[CC]{il manque les equations} Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each.
 Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.
+\comment[CC]{en mode tatillon, on voudrait une double descente chromatique centree sur 1 pour la legende de la distance figure 4, qui distinguerait mieux les situations superieures a 1 et celle inf a 1 (ce qui nous interesse).
+Par exemple, blanc pour 1, et de ++ rose quand c'est >>1 et de ++ vert que c'est <<1.
+En R pour un level plot
+}
 
@@
 
-We obtain a very strong sensitivity to initial conditions, as the distribution of the relative distance to reference across grids ranges from 0.09 to 2.98 with a median of 1.52 and an average of 1.30.
-It means that in average, the model is more sensitive to meta-parameters than to parameters, and the relation variation can reach a factor of 3.
-We plot in Fig.~\ref{fig:sugarscape-distance} their distribution in a morphological space.
-The reduced morphological space is obtained by computing 4 raw indicators of urban form, namely Moran index, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and contextualization), and by reducing the dimension with a principal component analysis for which we keep the first two components (92\% of cumulated variance).
-The first measures a ``level of sprawl'' and of scattering, whereas the second measures aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation.
+We obtain a very strong sensitivity to initial spatial conditions.
+Indeed, the relative distance of phase diagrams of different density grids to the reference case ranges from 0.09 to 2.98 with a median of 1.52 and an average of 1.30.
+The mean distance above 1 means that, on average, the model is more sensitive to meta-parameters of grid generation than to its own parameters (population and sugar).
+Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be 3 times bigger than the variation due to the model's parameters.
+We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in a morphological space.
+Each point represent one of the 50 different density grids used to initialised the distribution of sugar in the model.
+The points are projected with respect to the meta-parameters used to generate the grid, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case.
+Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e.
+with a small number of very high density cells) produce simulation results more different compared to the reference case than within this reference case because of parameter variations.
+This pattern is emphasized when grids are generated with a high $\alpha$ AND a high $\beta$ (i.e.
+with low gradient of density decrease around the kernels of high density).
+These grids have the highest relative distance to the reference case.
+On the contrary, with grids closer to random and uniform patterns (bottom left of the graph), the model's parameters are more important in determining the simulation results.
+
+%%%%%%%%%%%%%
+\begin{figure}
+\centering
+\includegraphics[width=0.5\textwidth]{figures/relativedistance_metaparams}\\
+\includegraphics[width=0.4\textwidth]{figures/alpha_beta_schemas}\\
+\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} (Top) Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).
+\comment[CC]{y'a un truc qui cloche avec cette figure: le nom des axes!
+soit alpha et beta, et on dit ce que c'est en legende, soit strength / diffusion, mais ca ne me parait pas la meilleure solution.
+en tout cas, pas de middle ground sinon c'est confusant, pour parler parfaitement frenglish} (Bottom) Corresponding stylised density grids }
+\label{fig:sugarscape-distance-meta}
+\end{figure}
+%%%%%%%%%%%%%
+
+Another way of quantifying the density grids, instead of looking at the meta-parameters of their generator, is to look at the resulting indicators of urban form, such as Moran index, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and context).
+This 4-dimensional space defined a morphological space.
+For the purpose of interpretability and visualisation, we reduce this space to a bi dimensional space with a principal component analysis.
+The first two components represent 92\% of cumulated variance.
+The first component defines a ``level of sprawl'' and of scattering, whereas the second one represents the level aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation (top left of figure \ref{fig:sugarscape-distance-pca}).
 It is confirmed by the behavior as a function of meta-parameters, as high values of $\alpha$ also yield high distance.
-In terms of model processes, it shows that congestion mechanisms induce rapidly higher levels of inequality.
-To put these results in perspective of our workflow given in Fig.~\ref{fig:method}, we have a sensitivity to spatial parameters in average greater than the sensitivity to model parameters.
+In terms of model processes, it shows that congestion mechanisms in the gathering of the resource induce fast increases of inequality.
+To put these results in perspective of our work-flow given in Fig.~\ref{fig:method}, we have a sensitivity to spatial parameters in average greater than the sensitivity to model parameters.
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figures/relativedistance_metaparams}\\
 \includegraphics[width=0.5\textwidth]{figures/relativedistance_morphspace}
-\caption{\textbf{Relative distances of phase diagrams to the reference across grids.} (Top) Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and diffusion ($\beta$, strength of diffusion process).
-(Bottom) Relative distance as a function of two first principal components of the morphological space (see text).
+\caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text).
 Red point correspond to the reference spatial configuration.
 Green frame and blue frame give respectively the first and second particular phase diagrams shown in Fig.~\ref{fig:sugarscape-phasediagrams}.}
-\label{fig:sugarscape-distance}
+\label{fig:sugarscape-distance-pca}
 \end{figure}
@@
 
-We now check the sensitivity in terms of qualitative behavior of phase diagrams.
+\subsection{Qualitative variations}
+\paragraph{Sugarscape} We now check the sensitivity in terms of qualitative behavior of phase diagrams.
 We show the phase diagrams for two very opposite morphologies in term of sprawling, but controlling for aggregation with the same $PC2$ value.
-These correspond to the green and blue frames in Fig.~\ref{fig:sugarscape-distance}.
+These correspond to the green and blue frames in Fig.~\ref{fig:sugarscape-distance-pca}.
 The behaviors are rather stable for varying $s_+$, what means that the poorest agents have a determinant role in trajectories.
-The two examples have not only a very distant baseline inequality (the ceil of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values.
+The two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values.
 The process is thus completely inverted, what would have significant impacts if one tried to schematize policies from this model.
@@
 \comment{Ajouter un paragraphe sur le lien entre forme de grille et comportement du modèle/résultats.}
+
+\paragraph{Schelling} In this qualitative exploration of the effect of initial spatial conditions on the results of Schelling's model, we use the classification of grids into three morphological types (cf.
+figure \ref{fig:densityTypes}).
+In particular, we want evaluate to which extent the typology summarizes the spatial effects, and so if one type of urban form or the other enhances or mutes the segregation mechanism of the model, or interacts differently with the model parameters.
+This experiment aims at suggesting topical conclusions on urban morphology, beyond the technical conclusions already obtained with respect to simulation sensitivity.
+
+\begin{table*}[htbp]
+\centering
+\begin{tabular}{|l|ll|ll|ll|}
+\hline
+Simulation outcome by segregation index: & \multicolumn{2}{c|}{\textbf{Dissimilarity}} & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
+\textbf{Intercept}                       & -0.007 ***  & -0.03 ***                     & -0.028 ***        & -0.041 ***        & 0.127 ***        & 0.129 ***            \\ \hline
+\textbf{Similarity Wanted (S)}           & -0.007 ***  & -0.007 ***                    & -0.007 ***        & -0.007 ***        & -0.005 ***       & -0.005 ***           \\ 
+\textbf{Vacancy Rate (V)}                & 0.294 ***   & 0.294 ***                     & 0.204 ***         & 0.294 ***         & 0.204 ***        & \textit{0.099}       \\ \hline
+\textbf{Density Grid = Polycentric}      &             & 0.038 ***                     &                   & 0.023 ***         &                  & -0.006 ***           \\ 
+\textbf{Density Grid = Discontinuous}    &             & 0.03 ***                     &                   & 0.017 ***         &                  & -0.001 ***           \\ \hline
+\textbf{R2 (\%)}                         & 15.01       & 15.57  & 7.62              & 7.83              & 2.04             & 2.05                 \\ 
+\textbf{\# of observations (sim. runs)}  & 4,500,000   & 4,500,000                     & 4,500,000         & 4,500,000         & 4,500,000        & 4,500,000            \\ \hline
+\end{tabular}
+\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant.}
+\label{tab:regressionSchelling}
+\end{table*}
+
+In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised determines to a certain extent the level of segregation measured at the end of the simulation run.
+Indeed, compared to compact density patterns, polycentric grids produce more dissimilarity and entropy between the location of green and red agents.
+Discontinuous grids have the same effect, although attenuated.
+The results obtained with Moran's I are opposite, because this index measures spatial autocorrelation at the global level and that compact cities have higher levels of global autocorrelation by construction.
+However, linear models with and without information about the type of density distribution yields the same coefficients for Schelling's parameters $V$ and $S$, the only exception being the vacancy rate $V$ in the Moran's I model with grid types, which becomes non-significant.
+The similarity of the coefficient in both cases means that the effect of the model's parameters (and thus the mechanism by which agents of similar group cluster in space) is the same regardless of the distribution of density.
+The way polycentric and discontinuous density grid exhibit higher segregation is by allowing buffer zones of low density to surround pockets of homogeneity, which is impossible in a compact city, because everyone is at reach of everyone else.
+The buffering process confirms previous results obtained with network structures \citep{Banos2012}.
 
@@
 
+\section{Conclusion.}
+
+With the Schelling experiment, we found that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
+With Sugarscape, we show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.
 %%%%%%%%%%%%%
@@
 \bibliographystyle{SageH}
-\bibliography{spacematters,biblio}
+\bibliography{spacematters}
 

### 2017-08-04 paper.tex
.

@@
-\documentclass[Afour,sageh,times]{sagej}
+\documentclass[Royal,sageh,times]{sagej}
+% Afour
 \input{header.tex}
@@
 % **************  MAIN BODY OF THE PAPER **************
-\newpage
+%\newpage
 
@@
 
+%\cite{ligmann2013spatially}
+
 \comment[clem]{Ce serait cool d'avoir aussi des exemples de simu hors Schelling pour les deux dernieres sous-parties}
@@
 Par ailleurs est ce que la formule est bien homogène du point de vue des dimensions?
-En tous cas je pense qu'on peut se passer de la formule à ce stade, éventuellement à mettre dans l'affichage des résultats si on en a besoin.}
+En tous cas je pense qu'on peut se passer de la formule à ce stade, éventuellement à mettre dans l'affichage des résultats si on en a besoin.}[(JR)]
 
@@
 %\begin{equation}\label{eq:phase-distance}
-%d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}%{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
+%d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
 %\end{equation}
@@
 The objective of this quantitative experiment is to evaluate to which extent this phase diagram is modified when different density grids are applied.
-\comment[CC]{!!!
-Il faut retrouver les fichiers de simu !!!} \comment[CC]{Juste, tu penses qu'il faudrait faire la meme analyse que toi sur la distance entre phase diagrams?
-Ce serait cool mais si oui, j'aurais besoin d'aide (et des fichiers de simu!)}
+
+\comment[CC]{Juste, tu penses qu'il faudrait faire la meme analyse que toi sur la distance entre phase diagrams? Ce serait cool mais si oui, j'aurais besoin d'aide (et des fichiers de simu!)}
 
@@
 random?
-ou Netlogo?} (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability.
+ou Netlogo?}[(JR) c'est Netlogo, une configuration bi-centrique particulière] (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability.
 \comment[CC]{il manque les equations} Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each.
 Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.
+
 \comment[CC]{en mode tatillon, on voudrait une double descente chromatique centree sur 1 pour la legende de la distance figure 4, qui distinguerait mieux les situations superieures a 1 et celle inf a 1 (ce qui nous interesse).
 Par exemple, blanc pour 1, et de ++ rose quand c'est >>1 et de ++ vert que c'est <<1.
-En R pour un level plot
-}
+En R pour un level plot}[(JR) oui tu as carrément raison, j'ai updaté les figures !]
 
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figures/relativedistance_morphspace}
+\includegraphics[width=0.8\textwidth]{figures/relativedistance_morphspace}
 \caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text).
@@
 \centering
-\includegraphics[width=0.24\textwidth]{figures/phasediagram_id27_maxSugar110}
-\includegraphics[width=0.24\textwidth]{figures/phasediagram_id0_maxSugar110}
+\includegraphics[width=0.48\textwidth]{figures/phasediagram_id27_maxSugar110}
+\includegraphics[width=0.48\textwidth]{figures/phasediagram_id0_maxSugar110}
 \caption{\textbf{Examples of phase diagrams.} We show two dimensional phase diagrams on $(P,s_-)$, both at fixed $s_+ = 110$.
@@
 \comment[MLT]{Mise à disposition des grilles générées? D'un échantillon?}
+
+Ref Clem : \cite{alizadeh2016generating} génération de réseau spatiaux / données synthétiques
 

### 2017-08-04 paper.tex
.

@@
 Par ailleurs est ce que la formule est bien homogène du point de vue des dimensions?
-En tous cas je pense qu'on peut se passer de la formule à ce stade, éventuellement à mettre dans l'affichage des résultats si on en a besoin.}[(JR)]
+En tous cas je pense qu'on peut se passer de la formule à ce stade, éventuellement à mettre dans l'affichage des résultats si on en a besoin.}[(JR) pour l'homogénéité oui justement on prend $d^2$ car c'est les variances, on obtient une mesure sans unité ; ok pour omettre la formule ; pour l'interprétation, on regarde la distance relative, pour des classes à une certaine distance, si elles sont concentrées, cette distance sera plus sigificative que si elles ont une grande variabilité et peuvent se recouper ; par exemple si elles ont mêmes variance $\sigma^2$, alors si $d \simeq \sqrt{2} \sigma$ alors les nuages de points vont en gros se recouper sur leur moitié (cas gaussien idéal), tandis que si $d\gg \sigma$, le max d'un diagramme sera très loin du min de l'autre - on compare à la fois la distance relative et la distance de forme.]
 
@@
 
-\comment[JR]{j'ai simplifié au max}\comment[mLT]{Attention, on utilise alpha plus haut.}
-
 The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows.
@@
 Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}).
-In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment \comment[FL]{à quoi?} characterised by its strength of attraction $\alpha$.
+In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
 More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
@@
 Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
-\comment{A detailler en incluant les equations de morphogenese}[(JR) j'ai détaillé le processus, c'est suffisant ?]\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n).
+\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n).
 Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$?

### 2017-08-04 paper.tex
.

@@
 \cite{LeNechet2015}).
-A stochastic exploration of a Latin Hypercube Sampling\comment[FL]{ce n'est pas explicite} of 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n$, $N$} gives a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions.
+We sample the meta-parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a point cloud with high discrepancy.
+We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n$, $N$}.
+It yields a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions.
 These are further clustered into three classes of morphology: compact (e.g.
@@
 \comment[FL]{pourquoi 15?} \comment[FL]{dire explicitement ce qu'on va faire avec ces catégories dans la partie résultats?}
-
-%cite{Gauvinetal2009}
-
-%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Comparing Phase Diagrams}
-\comment[FL]{j'ai proposé de mettre cette partie au début de method et de la simplifier}
-
-%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Model Exploration Workflows as an meta-sensitivity analysis method}
-\comment[FL]{item}
 
@@
 Each parameter is binned into 10 values, giving 1000 parameter points.
+\comment[JR]{apparemment le sampling est sobol ou lhs pour l'espace des paramètres}
 We run 100 repetitions for each configuration, what yield reasonable convergence properties.
@@
 \caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} (Top) Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).
-\comment[CC]{y'a un truc qui cloche avec cette figure: le nom des axes!
-soit alpha et beta, et on dit ce que c'est en legende, soit strength / diffusion, mais ca ne me parait pas la meilleure solution.
-en tout cas, pas de middle ground sinon c'est confusant, pour parler parfaitement frenglish} (Bottom) Corresponding stylised density grids }
+(Bottom) Corresponding stylised density grids }
 \label{fig:sugarscape-distance-meta}
@@
 \begin{table*}[htbp]
+\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant.\label{tab:regressionSchelling}}
+\begin{adjustwidth}{-0.5cm}{-0.5cm}
 \centering
-\begin{tabular}{|l|ll|ll|ll|}
+\begin{tabular}{|m{2cm}|ll|ll|ll|}
 \hline
@@
 \end{tabular}
-\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant.}
-\label{tab:regressionSchelling}
+\end{adjustwidth}
 \end{table*}

### 2017-08-18 paper.tex
Update on Overleaf.

@@
 
-Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}).
+Our spatial generator applies an urban morphogenesis model \cite{Batty2007}, which has been generalised, explored and calibrated \cite{Raimbault2014}.
 In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
@@
 \comment[CC]{en mode tatillon, on voudrait une double descente chromatique centree sur 1 pour la legende de la distance figure 4, qui distinguerait mieux les situations superieures a 1 et celle inf a 1 (ce qui nous interesse).
-Par exemple, blanc pour 1, et de ++ rose quand c'est >>1 et de ++ vert que c'est <<1.
-En R pour un level plot}[(JR) oui tu as carrément raison, j'ai updaté les figures !]
+Par exemple, blanc pour 1, et de ++ rose quand c'est >>1 et de ++ vert que c'est <<1.}[(JR) oui tu as carrément raison, j'ai updaté les figures !]\comment[CC]{magnifique!
+bien joué}
 
@@
 
+\comment[cc]{Resultats Temporaires sur 1/3 des simu}
+\begin{table*}[]
+\centering
+\caption{My caption}
+\begin{adjustwidth}{-0.5cm}{-0.5cm}
+\begin{tabular}{|m{2.5cm}|ll|ll|ll|}
+\hline
+Simulation outcome by segregation index:    & \multicolumn{2}{c|}{\textbf{Dissimilarity}}   & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
+\textbf{Intercept}                          & -0.222 *** & -0.206 ***                       & -0.272 ***        & -0.288 ***        & -0.040 ***           & -0.086 ***               \\ \hline
+\textbf{Similarity Wanted (S)}              & 1.426 ***  & 1.426 ***                        & 1.482 ***         & 1.482 ***         & 0.673 ***            & 0.673 ***                \\ 
+\textbf{quadratic term ($S^2$)}               & -1.158 *** & -1.158 ***                       & -1.179 ***        & -1.179 ***        & -0.545 ***           & -0.545 ***               \\ 
+\textbf{Vacancy Rate (V)}                   & 0.614 ***  & 0.614 ***                        & 0.486 ***         & 0.486 ***         & -0.037 ***           & -0.037 ***               \\ 
+\textbf{Minority Index (\%Maj - \%Min)}     & 0.280 ***  & 0.280 ***                        & 0.094 ***         & 0.094 ***         & -0.090 ***           & -0.090 ***               \\ \hline
+\textbf{Density Grid = Polycentric}         &            & 0.098 ***                        &                   & 0.070 ***         &                      & 0.002 ***                \\ 
+\textbf{Density Grid = Discontinuous}       &            & 0.117 ***                        &                   & 0.082 ***         &                      & -0.001 ***               \\
+\textbf{Attraction meta-parameter $\alpha$} &            & -0.055 ***                       &                   & -0.024 ***        &                      & 0.026 ***                \\ 
+\textbf{Diffusion meta-parameter $\beta$}   &            & 0.270 ***                        &                   & 0.175 ***         &                      & \textit{-0.00}           \\ \hline
+\textbf{R2 (\%)}                            & 34.6       & 38.5                             & 29.1              & 30.8              & 29.8                 & 30.0                     \\ 
+\textbf{\# of observations (sim. runs)}     & 1,403,899  & {\color[HTML]{000000} 1,403,899} & 1,403,899         & 1,403,899         & 1,403,899            & 1,403,899                \\ 
+\textbf{AIC}                                & -5939.02   & -91944.86                        & 246882.4          & 213145.1          & -2829106             & -2832773                 \\ \hline
+\end{tabular}
+\end{adjustwidth}
+N.B. Moran's I applies to the minority population
+\end{table*}
+
 In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised determines to a certain extent the level of segregation measured at the end of the simulation run.
@@
 
-\comment[FL]{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}[(JR) bonne idée, y'a pas mal d'exemples ou il faut pas mal contourner l'usage ``naturel'' de Netlogo pour s'en sortir - sur les représentations raster/vecteur c'est souvent un dilemne et ça peut induire des biais d'implémentation - je vais essayer de trouver un exemple simple (ceux que j'ai là trop usine à gaz, exemple Lutecia)]
+We consider that the method presented holds great potential for strengthening geographical models' validation.
+However, two limits and two areas of opportunities have still not been tackled.
+We discuss them in this section.
 
-\comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}
+\subsection{Limits}
 
-\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.}
-
-\comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
-
-\comment[MLT]{Mise à disposition des grilles générées? D'un échantillon?}
-
-Ref Clem : \cite{alizadeh2016generating} génération de réseau spatiaux / données synthétiques
-
-\paragraph{Comparing phase diagrams}
-
-Comparing phase diagrams is as we saw not straightforward, and further developments of our method imply testing alternative methods for this particular point.
+\paragraph{Comparing phase diagrams.} Comparing phase diagrams is as we saw not straightforward, and further developments of our method imply testing alternative methods for this particular point.
 For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
@@
 Note that with the intuitive index we used, more subtle distances such as the Earth's Mover Distance can be used.
+
+\paragraph{Platform constrains and docking challenges.} An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
+For example, the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements.
+Its toroidal default setting might also have influenced the work of many modellers who did not question explicitly the representation of space.
+This issue is part of the docking challenge \citep{Axtelletal1996} (i.e.
+checking if two models can produce the same results), but more generally, it involves a description of the model and its spatial requirements more detailed than what is currently the rule.
+\comment[FL]{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}[(JR) bonne idée, y'a pas mal d'exemples ou il faut pas mal contourner l'usage ``naturel'' de Netlogo pour s'en sortir - sur les représentations raster/vecteur c'est souvent un dilemne et ça peut induire des biais d'implémentation - je vais essayer de trouver un exemple simple (ceux que j'ai là trop usine à gaz, exemple Lutecia)]
+
+\subsection{Opportunities/extensions}
+
+\paragraph{Reproducibility/Applicability.} \comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
+Although the applications we present here are limited by the simplicity of the models, we think that the method could (and should) be applied to larger models including domain mechanisms and more empirical initialisation data, for example synthetic populations.
+The sensitivity analysis to initial spatial conditions could then be either a replication on the spatial allocation of the synthetic population, or a series of spatial permutations of the empirical spatial inputs.
+We want foster this extension of our work by releasing the density grids also generated, as well as the generating work-flow \comment[CC]{add link to github repo}.
+Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents.
+
+\paragraph{An emancipation opportunity for social sciences.
+\comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}
+\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.} }
 
@@
 
+After reviewing the extensive literature on spatial biases in statistical and simulation models, we presented a method to analyse the sensitivity of a simulation's results to the initial spatial configuration.
+We did so by implementing a spatial generator whose output is used as input for the simulation model.
+We applied this approach to two mythical ABMs: Schelling and Sugarscape.
 With the Schelling experiment, we found that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
 With Sugarscape, we show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.
+We think this method has the potential to increase the arsenal of evaluation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
 %%%%%%%%%%%%%

### 2017-08-23 paper.tex
bib

--- 2017-08-04 paper.tex
@@
 
-Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{Raimbault2014}).
+Our spatial generator applies an urban morphogenesis model (\cite{Batty2007}), which has been generalised, explored and calibrated (\cite{2017arXiv170806743R}).
 In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.

### 2017-08-23 paper.tex

--- 2017-08-18 paper.tex
@@
 
-Our spatial generator applies an urban morphogenesis model \cite{Batty2007}, which has been generalised, explored and calibrated \cite{Raimbault2014}.
+Our spatial generator applies an urban morphogenesis model \cite{Batty2007}, which has been generalised, explored and calibrated \cite{2017arXiv170806743R}.
 In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.

### 2017-09-05 paper.tex
Update on Overleaf.

@@
 
-%\begin{equation}\label{eq:phase-distance}
-%d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
-%\end{equation}
+\begin{equation}\label{eq:phase-distance}
+d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
+\end{equation}
 
@@
 
+\comment[cc]{Nouveaux Resultats}
+\begin{table*}[]
+\centering
+\caption{My caption}
+\begin{adjustwidth}{-0.5cm}{-0.5cm}
+\begin{tabular}{|m{2.5cm}|ll|ll|ll|}
+\hline
+Simulation outcome by segregation index:    & \multicolumn{2}{c|}{\textbf{Dissimilarity}}   & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
+\textbf{Intercept}                          & -0.212 *** & -0.141 ***                       & -0.254 ***        & -0.208 ***        & -0.036 ***           & -0.061 ***               \\ \hline
+\textbf{Similarity Wanted (S)}              & 1.212 ***  & 1.212 ***                        & 1.250 ***         & 1.250 ***         & 0.550 ***            & 0.550 ***                \\ 
+\textbf{quadratic term ($S^2$)}               & -0.942 *** & -0.942 ***                       & -0.963 ***        & -0.963 ***        & -0.428 ***           & -0.438 ***               \\ 
+\textbf{Vacancy Rate (V)}                   & 0.602 ***  & 0.602 ***                        & 0.453 ***         & 0.453 ***         & -0.027 ***           & -0.027 ***               \\ 
+\textbf{Minority Index (\%Maj - \%Min)}     & 0.307å ***  & 0.307 ***                        & 0.130 ***         & 0.130 ***         & -0.067 ***           & -0.067 ***               \\ \hline
+\textbf{Density Grid = Polycentric}         &            & 0.087 ***                        &                   & 0.052 ***         &                      & 0.001 ***                \\ 
+\textbf{Density Grid = Discontinuous}       &            & 0.111 ***                        &                   & 0.068 ***         &                      & \textit{0.00}              \\
+\textbf{Attraction meta-parameter $\alpha$} &            & -0.083 ***                       &                   & -0.053 ***        &                      & 0.014 ***                \\ 
+\textbf{Diffusion meta-parameter $\beta$}   &            & 0.323 ***                        &                   & 0.218 ***         &                      & 0.017 ***           \\ \hline
+\textbf{R2 (\%)}                            & 30.6       & 34.7                             & 24.1              & 25.6              & 23.9                 & 24.0                    \\ 
+\textbf{\# of observations (sim. runs)}     & 2,106,000  & 2,106,000 						 & 2,106,000          & 2,106,000          & 2,106,000             & 2,106,000                \\ 
+\textbf{AIC}                                & -70717.68   & -198748.2  						& 208213.8          & 166048.8          & -4385990             & -4387816                 \\ \hline
+\end{tabular}
+\end{adjustwidth}
+N.B. Moran's I applies to the minority population
+\end{table*}
 In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised determines to a certain extent the level of segregation measured at the end of the simulation run.

### 2017-09-28 paper.tex
.

@@
 
-\comment[CC]{en mode tatillon, on voudrait une double descente chromatique centree sur 1 pour la legende de la distance figure 4, qui distinguerait mieux les situations superieures a 1 et celle inf a 1 (ce qui nous interesse).
-Par exemple, blanc pour 1, et de ++ rose quand c'est >>1 et de ++ vert que c'est <<1.}[(JR) oui tu as carrément raison, j'ai updaté les figures !]\comment[CC]{magnifique!
-bien joué}
-
 % summary stats
@@
 \centering
-\includegraphics[width=0.5\textwidth]{figures/relativedistance_metaparams}\\
-\includegraphics[width=0.4\textwidth]{figures/alpha_beta_schemas}\\
-\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} (Top) Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).
-(Bottom) Corresponding stylised density grids }
+\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}\\
+%\includegraphics[width=0.4\textwidth]{figures/alpha_beta_schemas}(Bottom) Corresponding stylised density grids\\
+\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).
+\comment[JR]{je sais pas si c'est une bonne idee de mettre les grilles stylisees comme ca parce que ca donne une fausse idee des sorties du modele de generation}}
 \label{fig:sugarscape-distance-meta}
@@
 
-\comment[cc]{Resultats Temporaires sur 1/3 des simu}
 \begin{table*}[]
 \centering
-\caption{My caption}
+\caption{\comment[cc]{Resultats Temporaires sur 1/3 des simu}}
 \begin{adjustwidth}{-0.5cm}{-0.5cm}
@@
 
-\comment[cc]{Nouveaux Resultats}
 \begin{table*}[]
 \centering
-\caption{My caption}
+\caption{\comment[CC]{Nouveaux Resultats}}
 \begin{adjustwidth}{-0.5cm}{-0.5cm}
@@
 
-\paragraph{An emancipation opportunity for social sciences.
+\paragraph{An emancipation opportunity for social sciences.}
+
 \comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}
-\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.} }
+\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.}
 

### 2017-09-28 paper.tex
Update on Overleaf.

@@
 In order to test for the influence of spatial initial conditions on model outputs, we use a systematic method to compare phase diagrams.
-We define a phase diagram as a vector of  outputs\comment[FL]{Juste, peut-être des choses qui sont mal dites?
-: par exemple, dans diagramme de phase est-ce qu'on entend pas une "réification" de catégories - des clusters par ex?
-Et du coup est-ce que cela ne manque pas là?}.
+We define a phase diagram as the projection of a vector of model outputs in the parameter space.
+Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values as the median of the final values of an output indicator obtained for the replications of the model initialized with this combination of parameter.\\
 
@@
 We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model.
+Indeed, because the density distribution refers to the distribution of resource rather than to the representation of a city structure, we do not need the typology of urban density in this experiment.
 The full experiment thus equates to 2,500,000 simulations (1000 parameter combinations x 50 density grids x 50 replications).
@@
 
-\begin{table*}[htbp]
-\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant.\label{tab:regressionSchelling}}
-\begin{adjustwidth}{-0.5cm}{-0.5cm}
-\centering
-\begin{tabular}{|m{2cm}|ll|ll|ll|}
-\hline
-Simulation outcome by segregation index: & \multicolumn{2}{c|}{\textbf{Dissimilarity}} & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
-\textbf{Intercept}                       & -0.007 ***  & -0.03 ***                     & -0.028 ***        & -0.041 ***        & 0.127 ***        & 0.129 ***            \\ \hline
-\textbf{Similarity Wanted (S)}           & -0.007 ***  & -0.007 ***                    & -0.007 ***        & -0.007 ***        & -0.005 ***       & -0.005 ***           \\ 
-\textbf{Vacancy Rate (V)}                & 0.294 ***   & 0.294 ***                     & 0.204 ***         & 0.294 ***         & 0.204 ***        & \textit{0.099}       \\ \hline
-\textbf{Density Grid = Polycentric}      &             & 0.038 ***                     &                   & 0.023 ***         &                  & -0.006 ***           \\ 
-\textbf{Density Grid = Discontinuous}    &             & 0.03 ***                     &                   & 0.017 ***         &                  & -0.001 ***           \\ \hline
-\textbf{R2 (\%)}                         & 15.01       & 15.57  & 7.62              & 7.83              & 2.04             & 2.05                 \\ 
-\textbf{\# of observations (sim. runs)}  & 4,500,000   & 4,500,000                     & 4,500,000         & 4,500,000         & 4,500,000        & 4,500,000            \\ \hline
-\end{tabular}
-\end{adjustwidth}
-\end{table*}
-
 \begin{table*}[]
 \centering
-\caption{\comment[cc]{Resultats Temporaires sur 1/3 des simu}}
-\begin{adjustwidth}{-0.5cm}{-0.5cm}
-\begin{tabular}{|m{2.5cm}|ll|ll|ll|}
-\hline
-Simulation outcome by segregation index:    & \multicolumn{2}{c|}{\textbf{Dissimilarity}}   & \multicolumn{2}{c|}{\textbf{Entropy}} & \multicolumn{2}{c|}{\textbf{Moran's I}} \\ \hline
-\textbf{Intercept}                          & -0.222 *** & -0.206 ***                       & -0.272 ***        & -0.288 ***        & -0.040 ***           & -0.086 ***               \\ \hline
-\textbf{Similarity Wanted (S)}              & 1.426 ***  & 1.426 ***                        & 1.482 ***         & 1.482 ***         & 0.673 ***            & 0.673 ***                \\ 
-\textbf{quadratic term ($S^2$)}               & -1.158 *** & -1.158 ***                       & -1.179 ***        & -1.179 ***        & -0.545 ***           & -0.545 ***               \\ 
-\textbf{Vacancy Rate (V)}                   & 0.614 ***  & 0.614 ***                        & 0.486 ***         & 0.486 ***         & -0.037 ***           & -0.037 ***               \\ 
-\textbf{Minority Index (\%Maj - \%Min)}     & 0.280 ***  & 0.280 ***                        & 0.094 ***         & 0.094 ***         & -0.090 ***           & -0.090 ***               \\ \hline
-\textbf{Density Grid = Polycentric}         &            & 0.098 ***                        &                   & 0.070 ***         &                      & 0.002 ***                \\ 
-\textbf{Density Grid = Discontinuous}       &            & 0.117 ***                        &                   & 0.082 ***         &                      & -0.001 ***               \\
-\textbf{Attraction meta-parameter $\alpha$} &            & -0.055 ***                       &                   & -0.024 ***        &                      & 0.026 ***                \\ 
-\textbf{Diffusion meta-parameter $\beta$}   &            & 0.270 ***                        &                   & 0.175 ***         &                      & \textit{-0.00}           \\ \hline
-\textbf{R2 (\%)}                            & 34.6       & 38.5                             & 29.1              & 30.8              & 29.8                 & 30.0                     \\ 
-\textbf{\# of observations (sim. runs)}     & 1,403,899  & {\color[HTML]{000000} 1,403,899} & 1,403,899         & 1,403,899         & 1,403,899            & 1,403,899                \\ 
-\textbf{AIC}                                & -5939.02   & -91944.86                        & 246882.4          & 213145.1          & -2829106             & -2832773                 \\ \hline
-\end{tabular}
-\end{adjustwidth}
-N.B. Moran's I applies to the minority population
-\end{table*}
-
-\begin{table*}[]
-\centering
-\caption{\comment[CC]{Nouveaux Resultats}}
+\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant.}
+\label{tab:regressionSchelling}
 \begin{adjustwidth}{-0.5cm}{-0.5cm}
@@
 \end{table*}
+
 In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised determines to a certain extent the level of segregation measured at the end of the simulation run.

### 2017-10-09 paper.tex
.

@@
 
+We chose different design of experiments, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method.
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-\comment[CC]{Juste, tu penses qu'il faudrait faire la meme analyse que toi sur la distance entre phase diagrams? Ce serait cool mais si oui, j'aurais besoin d'aide (et des fichiers de simu!)}
+%%%%%%%%%%%%%
+\begin{figure}
+\centering
+\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams}
+\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace}
+\caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} \label{fig:schelling-distance-meta}}
+\end{figure}
+%%%%%%%%%%%%%
 
-\paragraph{Sugarscape} We measure the distance of all 3-dimensional phase diagrams to the reference phase diagram computed on the default model setup \comment[CC]{j'ai peut-etre manque une etape, mais c'est quoi le default model setup?
-en termes de grille d'entree: uniforme?
-random?
-ou Netlogo?}[(JR) c'est Netlogo, une configuration bi-centrique particulière] (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability.
+\paragraph{Sugarscape} We measure the distance of all 3-dimensional phase diagrams to the reference phase diagram computed on the default model setup, which is a bi-centric symmetric configuration (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability.
 \comment[CC]{il manque les equations} Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each.
@@
 \includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}\\
-%\includegraphics[width=0.4\textwidth]{figures/alpha_beta_schemas}(Bottom) Corresponding stylised density grids\\
+%\includegraphics[width=0.4\textwidth]{figures/alpha_beta_schemas}(Bottom) Corresponding stylised density grids\\\comment[JR]{je sais pas si c'est une bonne idee de mettre les grilles stylisees comme ca parce que ca donne une fausse idee des sorties du modele de generation}}
 \caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).
-\comment[JR]{je sais pas si c'est une bonne idee de mettre les grilles stylisees comme ca parce que ca donne une fausse idee des sorties du modele de generation}}
 \label{fig:sugarscape-distance-meta}
@@
 \caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text).
-Red point correspond to the reference spatial configuration.
+Black point correspond to the reference spatial configuration.
 Green frame and blue frame give respectively the first and second particular phase diagrams shown in Fig.~\ref{fig:sugarscape-phasediagrams}.}
@@
 The two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values.
-The process is thus completely inverted, what would have significant impacts if one tried to schematize policies from this model.
-This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
+The process is thus completely inverted, what would have significant impacts if one tried to schematise policies from this model.
+To interpret these behavior in terms of grid shape, we observe that the difference between the two grids is mainly on average distance and entropy: in a nutshell, the first grid is much more dispersed and disorganised than the second.
+In sprawled spaces, inequalities are thus fostered by a lack of minimal local ressources, whereas population will drive inequality in concentrated spaces.
 
-\comment{Ajouter un paragraphe sur le lien entre forme de grille et comportement du modèle/résultats.}
+%This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
 
@@
 figure \ref{fig:densityTypes}).
-In particular, we want evaluate to which extent the typology summarizes the spatial effects, and so if one type of urban form or the other enhances or mutes the segregation mechanism of the model, or interacts differently with the model parameters.
+In particular, we want evaluate to which extent the typology summarises the spatial effects, and so if one type of urban form or the other enhances or mutes the segregation mechanism of the model, or interacts differently with the model parameters.
 This experiment aims at suggesting topical conclusions on urban morphology, beyond the technical conclusions already obtained with respect to simulation sensitivity.

### 2017-11-02 paper.tex
Update on Overleaf.

@@
 %\includegraphics[width=0.4\textwidth]{figures/alpha_beta_schemas}(Bottom) Corresponding stylised density grids\\\comment[JR]{je sais pas si c'est une bonne idee de mettre les grilles stylisees comme ca parce que ca donne une fausse idee des sorties du modele de generation}}
-\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).
+\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
 \label{fig:sugarscape-distance-meta}

### 2017-11-03 paper.tex
.

@@
 
-\runninghead{Raimbault \& al.}
+%\runninghead{Raimbault \& al.}
+\runninghead{}
 
@@
 
-\author{Juste Raimbault\affilnum{1,2}, 
-Cl{\' e}mentine Cottineau\affilnum{3},
-Marion Le Texier\affilnum{4}, 
-Florent Le N{\' e}chet\affilnum{2}, 
-Romain Reuillon\affilnum{1,5}
-}
+%\author{Juste Raimbault\affilnum{1,2}, 
+%Cl{\' e}mentine Cottineau\affilnum{3},
+%Marion Le Texier\affilnum{4}, 
+%Florent Le N{\' e}chet\affilnum{2}, 
+%Romain Reuillon\affilnum{1,5}
+%}
+\author{}
 
-\affiliation{\affilnum{1}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
-\affilnum{2}Laboratoire Ville Mobilit{\'e} Transport, Universit{\'e} Paris-Est, France\\
-\affilnum{3}Centre for Advanced Spatial Analysis, University College London, UK\\
-\affilnum{4}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
-\affilnum{5}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}
+%\affiliation{\affilnum{1}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
+%\affilnum{2}Laboratoire Ville Mobilit{\'e} Transport, Universit{\'e} Paris-Est, France\\
+%\affilnum{3}Centre for Advanced Spatial Analysis, University College London, UK\\
+%\affilnum{4}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
+%\affilnum{5}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}
 
-\corrauth{\email{juste.raimbault@polytechnique.edu}}
+%\corrauth{Juste Raimbault, UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
+%13 rue du Four,
+%75006 Paris, France.
+%\email{juste.raimbault@polytechnique.edu}}
+%\corrauth{}
+
+% authors bio
+% Juste Raimbault is a PhD student at Université Paris 7, UMR CNRS 8504 Géographie-cités and UMR-T 9403 IFSTTAR LVMT, under the supervision of Arnaud Banos and Florent Le Néchet.
+He holds an engineer degree from Ecole Polytechnique and Ecole des Ponts et Chaussées, and a Master in Complex Systems Science from Ecole Polytechnique.
+His research focuses on complexity approaches to the relations between networks and territories.
+
+% bios...
+
+% Cover Letter
+% Dear editors,
+%We are pleased with my coauthors to submit an original research article entitled ``Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?'' for consideration for publication in Environment and Planning B.
+This article deals with the central yet poorly explored issue of the sensitivity of geosimulation models to their spatial context, and more precisely their spatial initial conditions.
+It does not rely on previously published work.
+Preliminary results where however presented as an oral parallel session talk at ECTQG 2015, Bari, and at Geocomputation Conference 2017, Leeds.
+We did not have any previous interaction with EPB regarding this manuscript.
+This manuscript has not been published and is not under consideration for publication elsewhere.
+We have no conflicts of interest to disclose.
+We do not have any opposed reviewer.
+% Yours faithfully,
+% Juste Raimbault
 
@@
 Simulation within computer models has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}.
-\comment{+ Seb \& James Doran in the 1970s??} Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of social and ecological systems in space and time.
+%\comment{+ Seb \& James Doran in the 1970s??}
+ Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of social and ecological systems in space and time.
 The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model.
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Previous considerations on the effects of the urban form in simulation models}
+\subsection{Previous considerations on the effects of the spatial configuration in simulation models}
 
@@
 Dynamically, sprawl consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}.
-Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that{\it "deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking"}.
+Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}.
 The discussion of these effects has also been extensive in the field of geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984, FotheringhamWong1991}.
@@
 \begin{itemize}
-\item The accuraty of geo-localised input data;
+\item The accuracy of geo-localised input data;
 \item The spatial system shape, precision and boundaries;
@@
 \citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins.
-\comment[FL]{je dois compléter\\
-\cite{horritt2001effects} identified the effects of grid cell size on the behaviour of a raster flood model
-\cite{vazquez2002effect} ont exploré la meilleure granularité pour la précision d'un modèle d'écoulement hydrologie et le temps de calcul (à reprendre)}
+Similar issues were also dealt with in physical sciences, as for example \cite{horritt2001effects} studies the effects of grid cell size on the behaviour of a raster flood model and shows that increasing resolution does not increase model prediction performance under a certain limit.
+Similar conclusions are obtained by \cite{vazquez2002effect}, unveiling an intermediate optimal spatial resolution regarding model performance and computation time.
+Spatial resolution also plays a role in the Schelling model: \citet{Singhetal2009} show that the segregation patterns for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.
+
+%\comment[FL]{je dois compléter\\
+%\cite{horritt2001effects} identified the effects of grid cell size on the behaviour of a raster flood model
+%\cite{vazquez2002effect} ont exploré la meilleure granularité pour la précision d'un modèle d'écoulement hydrologie et le temps de calcul (à reprendre)}
 
@@
 \citet{HatnaBenenson2012} show that their model replications run on a 50x50 torus with 2\% of empty cells were not sensitive to the initial patterns (random and fully segregated distribution of agents).
+In ecology, \citet{smith2002predicting} study the spread of a disturbance in an heterogeneous landscape using a percolation model, and show that landscape structure has a significant influence on final patterns of contamination outcome.
+In a spatial epidemics model for an infectious disease, parametrised on real data, \citet{smith2002predicting} finds that the physical landscape heterogeneity, in particular the presence of rivers, locally influence the propagation speed.
 
-\paragraph{Combination:} \citet{Singhetal2009} show that the segregation patterns obtained by Schelling for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.
+%\paragraph{Combination:}
 
 %\cite{ligmann2013spatially}
-
-\comment[clem]{Ce serait cool d'avoir aussi des exemples de simu hors Schelling pour les deux dernieres sous-parties}
 
@@
 We wanted to emphasize that spatial effects do not only derive from grid size or shape effects, but indeed from the heterogeneity of relevant geographic entities (people, housing, networks, etc.) through geographical space.
-In models, the initial configuration of space can be either flat (most economic models\comment[FL]{c'est peut être un peu brutal?}), or heterogeneous,  based on external databases and internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
+In models, the initial configuration of space can be either flat, or heterogeneous,  based on external databases and internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
 %%%%%%%%%%%%%%%%%%%%%%
@@
 	\includegraphics{figures/SchemaMeta_1.png}
-} \caption{General workflow of our method} \label{fig:method} \end{center} \end{figure*} %
+} \caption{\textbf{General workflow of our method.}} \label{fig:method} \end{center} \end{figure*} %
 %%%%%%%%%%%%%%%%%%%%%%
@@
 Comparing phase diagrams is useful because it allows us to compare the 
-Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic\comment[FL]{donner un ordre de grandeur du nombre de simulations?}.
+Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic (with around 50 different spatial configurations in our applications).
 A solution is to use systematic quantitative procedures.
 To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature.
-We used a distance between two phase diagram corresponding to the share of between-diagrams variability relative to their internal variability\comment[FL]{je ne comprends pas ce choix : pourquoi faut il considérer que des classes ayant une forte cohérence interne seraient plus éloignées?
+We propose to use a relative distance between two phase diagram corresponding to the share of between-diagrams variability relative to their internal variability, given formally by
+
+%\comment[FL]{je ne comprends pas ce choix : pourquoi faut il considérer que des classes ayant une forte cohérence interne seraient plus éloignées?
 ou alors la question ne se pose pas en ces termes...
@@
 \begin{equation}\label{eq:phase-distance}
-d_r\left(\alpha_1,\alpha_2\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
+d_r\left(\vec{\alpha_1},\vec{\alpha_2}\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
 \end{equation}
 
-%where $f_{\vec{\alpha}}(\vec{x})$ is the phase diagrams on parameters $\vec{x}$ at given meta-parameters $\vec{\alpha}$.
+where $f_{\vec{\alpha}}(\vec{x})$ is the phase diagrams on parameters $\vec{x}$ at given meta-parameters $\vec{\alpha}$.
 $d$ is a functional distance that we take simply as Euclidian distance.

### 2017-11-03 paper.tex
.

@@
 
-In our approach, we have introduced a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions. 
-To allow for the exploration of the relative effects of model parameters and spatial configuration, we clustered initial spatial conditions  to represent types of spaces ex-ante (for example: monocentric or polycentric density grids).
+In our approach, we introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions. 
+To allow for the exploration of the relative effects of model parameters and spatial configuration, we cluster initial spatial conditions  to represent types of spaces ex-ante (for example: monocentric or polycentric density grids).
 The sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types.
@@
 In order to test for the influence of spatial initial conditions on model outputs, we use a systematic method to compare phase diagrams.
-We define a phase diagram as the projection of a vector of model outputs in the parameter space.
-Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values as the median of the final values of an output indicator obtained for the replications of the model initialized with this combination of parameter.\\
+We define a phase diagram as the vector of model outputs considered as a function of model parameters.
+Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values as the median of the final values of an output indicator obtained for the replications of the model initialized with this combination of parameters.\\
 
-Comparing phase diagrams is useful because it allows us to compare the 
+%Comparing phase diagrams is useful because it allows us to compare the 
 Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic (with around 50 different spatial configurations in our applications).
@@
 
-where $f_{\vec{\alpha}}(\vec{x})$ is the phase diagrams on parameters $\vec{x}$ at given meta-parameters $\vec{\alpha}$.
+where $f_{\vec{\alpha_i}}$ are the phase diagrams at given meta-parameters $\vec{\alpha_i}$.
 $d$ is a functional distance that we take simply as Euclidian distance.
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Constructing a spatial generator of density grids}
+\subsection{Spatial generator of density grids}
 
-Our spatial generator applies an urban morphogenesis model \cite{Batty2007}, which has been generalised, explored and calibrated \cite{2017arXiv170806743R}.
+The spatial generator applies an urban morphogenesis model \citep{Batty2007}, which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}.
+An open implementation and a characterisation of urban forms the model can produce are provided, what allows us to integrate it easily in our workflow.
 In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
 More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
-At the end of each time step, this growth process is smoothed $n$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches).
+At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches).
 To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside.
-The procedure stops when a fixed number of steps $k$ is reached.
-The grid then has a population of $k \times N$ (the population lost due to diffusion process to the outside is reallowed through a normalization procedure at the end of the steps).
-Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n$ and $N$, in addition to the random seed.
-To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment of 100,000 units (cf.
-figure \ref{fig:spatialGen}).
-Typical value ranges for the other parameters are $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n\in\left[1,4\right]$.
-\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n).
+The procedure stops when a fixed number of steps $t_f$ is reached.
+The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps).
+Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed.
+To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment.
+We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$.
+Typical value ranges for the  parameters will be taken as, following \citet{2017arXiv170806743R}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$.
+We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configuration that can be generated.
+
+%\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n).
 Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$?

### 2017-11-03 paper.tex
Update on Overleaf.

@@
 %\author{Juste Raimbault\affilnum{1,2}, 
-%Cl{\' e}mentine Cottineau\affilnum{3},
-%Marion Le Texier\affilnum{4}, 
+%Cl{\' e}mentine Cottineau\affilnum{3,4},
+%Marion Le Texier\affilnum{5}, 
 %Florent Le N{\' e}chet\affilnum{2}, 
-%Romain Reuillon\affilnum{1,5}
+%Romain Reuillon\affilnum{1,6}
 %}
@@
 %\affilnum{3}Centre for Advanced Spatial Analysis, University College London, UK\\
-%\affilnum{4}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
-%\affilnum{5}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}
+%\affilnum{4}CNRS, UMR 8097 Centre Maurice Halbwachs, Paris, France\\
+%\affilnum{5}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
+%\affilnum{6}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}
 
@@
 
-% bios...
+% Dr.
+Clémentine Cottineau is a research associate at the Centre for Advanced Spatial Analysis of University College London.
+She studied geography and economics at University Paris 1 Panthéon-Sorbonne and holds a PhD in geography from this university.
+Her thesis aimed at modelling urbanisation in the post-Soviet space with modular agent-based models.
+Since 2014, she has been working at the Centre for Advanced Spatial Analysis on urban economics and parametric scaling in the context of European systems of cities, using tools from complexity science to investigate inequality in cities and economic disparities between cities.
 
 % Cover Letter
+
 % Dear editors,
-%We are pleased with my coauthors to submit an original research article entitled ``Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?'' for consideration for publication in Environment and Planning B.
-This article deals with the central yet poorly explored issue of the sensitivity of geosimulation models to their spatial context, and more precisely their spatial initial conditions.
-It does not rely on previously published work.
-Preliminary results where however presented as an oral parallel session talk at ECTQG 2015, Bari, and at Geocomputation Conference 2017, Leeds.
+
+%My coauthors and I are pleased to submit an original research article entitled ``Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?'' for consideration for publication in Environment and Planning B.
+
+%This article deals with the central yet poorly explored issue of the sensitivity of geosimulation models to their spatial context (i.e.
+the spatial configuration of agents and their environment), and more precisely their spatial initial conditions.
+We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of EPB, a journal dealing with urban modelling and city science.
+
+%Despite some preliminary results being presented as at ECTQG 2015, Bari, and at Geocomputation Conference 2017, Leeds, we confirm that neither the manuscript nor any parts of its content are currently under consideration or published in another journal.
 We did not have any previous interaction with EPB regarding this manuscript.
-This manuscript has not been published and is not under consideration for publication elsewhere.
 We have no conflicts of interest to disclose.
 We do not have any opposed reviewer.
+
 % Yours faithfully,
-% Juste Raimbault
+% Juste Raimbault and co-authors.
 
@@
 In this paper, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create the density grids with which spatial simulation models are initialised.
-We show, with the example of two very classical agent-based models (Schelling's models of segregation and Sugarscape's model of resource harvesting) and a straightforward open-source work-flow using high performance computing, that the effect of space is significant in the two types of simulation.
+We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of resource harvesting) and a straightforward open-source work-flow using high performance computing, that the effect of space is significant in the two types of simulation.
 Furthermore, this effect is sometimes larger than the effect of parameters' value change.
@@
 
-\keywords{Space, Initial conditions, Sensitivity, ABM}
+\keywords{Space, Initial conditions, Sensitivity, ABM, Validation}
 
@@
 Indeed, they usually seek to understand and model the role of space in social interactions and environmental processes, in a city for example.
-We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial description and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled.
+We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled.
 Unfortunately, the first condition is not always met, and the second is seldom even mentioned.

### 2017-11-03 paper.tex
.

--- 2017-11-03 paper.tex
@@
 To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature.
-We propose to use a relative distance between two phase diagram corresponding to the share of between-diagrams variability relative to their internal variability, given formally by
+We propose to use a relative distance between two phase diagram corresponding to the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by
 
@@
 The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$.
+For a multi-dimensional phase diagram, we sum these relative distances over the components.
 
@@
 Vienna), polycentric (Liege) and discontinuous (Augsburg) in order to evaluate the non-trivial effects of urban form on simulation results.
-We select 15 grids of each type to work with in the computation of sensitivity analysis.
-\comment[FL]{pourquoi 15?} \comment[FL]{dire explicitement ce qu'on va faire avec ces catégories dans la partie résultats?}
+We select 15 grids of each type to work with in the computation of sensitivity analysis, obtaining a total of 45 spatial configurations to compare (as we will see, the significant number of model runs needed to establish each phase diagram imposes a computational limit on this number of configurations).
+
+%\comment[FL]{pourquoi 15?} \comment[FL]{dire explicitement ce qu'on va faire avec ces catégories dans la partie résultats?}
 
@@
 In addition, we are interested in testing the impact of the spatial distribution of housing capacity in this project, using the generated grids.
-\comment{We extend the implementation of...
-The outcome of the model is measured as a phase diagram of segregation level (Dissimilarity, Moran's I, Entropy, Exposure, all four of them?).
-.}\\
+The outcome of the model is measured as a phase diagram of segregation level, described by the combination of several indicators: Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.
 
@@
 For Schelling's model, we explore a grid of a basic parameter space of the model, which is composed of two dimensions: the minimum proportion of similar agents required in the neighbourhood for the agent to be satisfied (or intolerance level) $S\in \left[0;1\right]$ and the vacancy rate of the city $V\in \left[0;1\right]$, which then defines the number of green and red agents, present in equal proportion for this experiment.
-Each parameter is binned into 10 values, giving 1000 parameter points.
-\comment[JR]{apparemment le sampling est sobol ou lhs pour l'espace des paramètres}
-We run 100 repetitions for each configuration, what yield reasonable convergence properties.
+We sample 1000 parameter values using a Sobol sampling and run 100 repetitions for each configuration.
 The initial spatial configuration varies across 45 different grids, chosen among the ones which are most representative of the three types of urban morphology: 15 compact grids (like Vienna, cf figure \ref{fig:densityTypes}), 15 polycentric grids (like Liege) and 15 discontinuous grids (like Augsburg).
@@
 
-We chose different design of experiments, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method.
+We choose different design of experiments, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices.
+In principle, our workflow applies with any manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.
 
@@
 
-\comment[FL]{il faudrait peut être davantage renforcer l'analyse a propos de figure 1 de si c'est plutôt les mu parameters ou les spatial parameters qui ont une grosse influence sur les résultats.}[(JR) assez explicite pour sugarscape, j'ai rajouté un peu tout de même.]
+%\comment[FL]{il faudrait peut être davantage renforcer l'analyse a propos de figure 1 de si c'est plutôt les mu parameters ou les spatial parameters qui ont une grosse influence sur les résultats.}[(JR) assez explicite pour sugarscape, j'ai rajouté un peu tout de même.]
 
 \subsection{Quantitative variation across grids}
-\paragraph{Schelling} 
-Within a standard Schelling model (i.e.
-initialised with a uniform density grid), \citet{Gauvinetal2009} have built the phase diagram of segregation patterns depending on the combination of parameter values.
-For high levels of tolerance ($S < 0.25$), there is no segregation.
-For high values of vacancies ($V > 0.65$) and low values of tolerance ($S > 0.5$), these is a diluted segregation state where homogeneous communities are separated from other by large empty buffers.
-Finally, for low values of vacancies ($V < 0.2$) and low values of tolerance ($S > 0.7$), the model is frozen in a state where everyone is unhappy but no-one can express their intolerant behaviour due to the lack of free spaces.
-Between these extreme cases, the model gives rise to segregated states where homogeneous communities adjoin one another.
-The objective of this quantitative experiment is to evaluate to which extent this phase diagram is modified when different density grids are applied.
 
-%%%%%%%%%%%%%
-\begin{figure}
-\centering
-\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams}
-\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace}
-\caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} \label{fig:schelling-distance-meta}}
-\end{figure}
-%%%%%%%%%%%%%
-
-\paragraph{Sugarscape} We measure the distance of all 3-dimensional phase diagrams to the reference phase diagram computed on the default model setup, which is a bi-centric symmetric configuration (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using equation~\ref{eq:phase-distance} with the L2 distance to ensure direct interpretability.
-\comment[CC]{il manque les equations} Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each.
+\paragraph{Sugarscape} We measure the distance of all phase diagrams to the reference phase diagram computed on the default model setup, which is a bi-centric symmetric configuration (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using the relative measure defined in equation~\ref{eq:phase-distance}.
+Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each.
 Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.
@@
 
+\paragraph{Schelling} 
+Within a standard Schelling model (i.e.
+initialised with a uniform density grid), \citet{Gauvinetal2009} have built the phase diagram of segregation patterns depending on the combination of parameter values.
+For high levels of tolerance ($S < 0.25$), there is no segregation.
+For high values of vacancies ($V > 0.65$) and low values of tolerance ($S > 0.5$), these is a diluted segregation state where homogeneous communities are separated from other by large empty buffers.
+Finally, for low values of vacancies ($V < 0.2$) and low values of tolerance ($S > 0.7$), the model is frozen in a state where everyone is unhappy but no-one can express their intolerant behaviour due to the lack of free spaces.
+Between these extreme cases, the model gives rise to segregated states where homogeneous communities adjoin one another.
+The objective of this quantitative experiment is to evaluate to which extent this phase diagram is modified when different density grids are applied.
+We show in Fig.\ref{fig:schelling-distance-meta} the values of the relative distance as a function of meta-parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape before.
+Variations are less considerable than for Sugarscape across phase diagrams, but values close to 1 show that several configurations are as much sensitive to space than to their parameters.
+We will focus in the following in a qualitative characterisation of variations.
+
+%%%%%%%%%%%%%
+\begin{figure}
+\centering
+\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams}
+\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace}
+\caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams.
+We show them in the meta-parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
+\end{figure}
+%%%%%%%%%%%%%
+
+\subsection{Qualitative variations}
+\paragraph{Sugarscape} We now check the sensitivity in terms of qualitative behavior of phase diagrams.
+We show the phase diagrams for two very opposite morphologies in term of sprawling, but controlling for aggregation with the same $PC2$ value.
+These correspond to the green and blue frames in Fig.~\ref{fig:sugarscape-distance-pca}.
+The behaviors are rather stable for varying $s_+$, what means that the poorest agents have a determinant role in trajectories.
+The two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values.
+The process is thus completely inverted, what would have significant impacts if one tried to schematise policies from this model.
+To interpret these behavior in terms of grid shape, we observe that the difference between the two grids is mainly on average distance and entropy: in a nutshell, the first grid is much more dispersed and disorganised than the second.
+In sprawled spaces, inequalities are thus fostered by a lack of minimal local ressources, whereas population will drive inequality in concentrated spaces.
+
+%This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
+
 % phase diagrams -> ok well different qualitatively
@@
 
-\subsection{Qualitative variations}
-\paragraph{Sugarscape} We now check the sensitivity in terms of qualitative behavior of phase diagrams.
-We show the phase diagrams for two very opposite morphologies in term of sprawling, but controlling for aggregation with the same $PC2$ value.
-These correspond to the green and blue frames in Fig.~\ref{fig:sugarscape-distance-pca}.
-The behaviors are rather stable for varying $s_+$, what means that the poorest agents have a determinant role in trajectories.
-The two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values.
-The process is thus completely inverted, what would have significant impacts if one tried to schematise policies from this model.
-To interpret these behavior in terms of grid shape, we observe that the difference between the two grids is mainly on average distance and entropy: in a nutshell, the first grid is much more dispersed and disorganised than the second.
-In sprawled spaces, inequalities are thus fostered by a lack of minimal local ressources, whereas population will drive inequality in concentrated spaces.
-
-%This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
-
 \paragraph{Schelling} In this qualitative exploration of the effect of initial spatial conditions on the results of Schelling's model, we use the classification of grids into three morphological types (cf.
@@
 \centering
-\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant.}
+\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant at the 0.01 level.}
 \label{tab:regressionSchelling}
-\begin{adjustwidth}{-0.5cm}{-0.5cm}
+\begin{adjustwidth}{-1cm}{-1cm}
 \begin{tabular}{|m{2.5cm}|ll|ll|ll|}
@@
 The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
-Map comparison methods, popular in environmental sciences \comment[MLT]{Et pas en géo?
-Qu'est-ce que ces méthodes ont de particulier?}, provide numeral tools to compare two dimensional fields~\cite{visser2006map}.
+Map comparison methods, popular in environmental sciences %\comment[MLT]{Et pas en géo?
+Qu'est-ce que ces méthodes ont de particulier?}
+, provide numeral tools to compare two dimensional fields~\cite{visser2006map}.
 To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}.
-Note that with the intuitive index we used, more subtle distances such as the Earth's Mover Distance can be used.
+The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.
 
@@
 checking if two models can produce the same results), but more generally, it involves a description of the model and its spatial requirements more detailed than what is currently the rule.
-\comment[FL]{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}[(JR) bonne idée, y'a pas mal d'exemples ou il faut pas mal contourner l'usage ``naturel'' de Netlogo pour s'en sortir - sur les représentations raster/vecteur c'est souvent un dilemne et ça peut induire des biais d'implémentation - je vais essayer de trouver un exemple simple (ceux que j'ai là trop usine à gaz, exemple Lutecia)]
 
-\subsection{Opportunities/extensions}
+% \comment[FL]{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}[(JR) bonne idée, y'a pas mal d'exemples ou il faut pas mal contourner l'usage ``naturel'' de Netlogo pour s'en sortir - sur les représentations raster/vecteur c'est souvent un dilemne et ça peut induire des biais d'implémentation - je vais essayer de trouver un exemple simple (ceux que j'ai là trop usine à gaz, exemple Lutecia)]
 
-\paragraph{Reproducibility/Applicability.} \comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
+\subsection{Opportunities and extensions}
+
+\paragraph{Reproducibility and Applicability.} \comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
 Although the applications we present here are limited by the simplicity of the models, we think that the method could (and should) be applied to larger models including domain mechanisms and more empirical initialisation data, for example synthetic populations.

### 2017-11-03 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 An urban system can be defined as a large number of diverse social agents interacting with one another, within a limited portion of space.
-Social agents thus constitute the microscopic level of urban systems, which are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000}, and, in the  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex
Update on Overleaf.

@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000}, and, in the  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framewo emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex

--- 2017-11-03 paper.tex
@@
 An urban system can be defined as a large number of diverse social agents interacting with one another, within a limited portion of space.
-Social agents thus constitute the microscopic level of urban systems, which are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framewo emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000}, and, in the  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems,  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framewo emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex

--- 2017-11-03 paper.tex
@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framewo emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems,  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems,  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit macroscale  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex

--- 2017-11-03 paper.tex
@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems,  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit macroscale  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit macroscale characteristics  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit macroscale  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit characteristics  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex
Update on Overleaf.

@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit characteristics  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit characteristics that emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit characteristics  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit qcharacteristics that emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
 Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.

### 2017-11-03 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 
+\subsection{Definition of the problem}
+
 An urban system can be defined as a large number of diverse social agents interacting with one another, within a limited portion of space.
 Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the framework of complex systems, cities exhibit characteristics  emergent properties at qualitatively differentiated scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002, Batty2007}.
-Complexity is thus partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}.
+Complexity is partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.
 Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures.
 \\
+
 Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity~\citep{pumain2012urban}, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average.
-On the contrary, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time provide the use of ergodic models.
 Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
+Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models...
+which are the main models to be found in geosimulation!
 \\
 
-Although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) in cities is far from isotropic, even in the most sprawled cities.
+Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) in cities is far from isotropic, even in the most sprawled cities.
 On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities.
@@
 The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
-Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which has been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies, which tend to focus on single aspects of the debate.
+Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies.
 In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.
@@
 
-Different spatial configurations produce different distributions of income, carbon emissions, educational outcomes, etc.
-For example, \citet{wheeler2006urban} shows that, in the US, sprawling cities tend to be more unequal than their compact counterparts with respects to income.
+In 'real-life', different spatial configurations of people in a city tend to be associated with different distributions of income, carbon emissions, educational outcomes, etc.
+For example, \citet{wheeler2006urban} shows that, in the US, sprawling cities are more unequal than their compact counterparts with respect to income.
 Dynamically, sprawl consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}.
 Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}.
-The discussion of these effects has also been extensive in the field of geostatistics, meanly since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984, FotheringhamWong1991}.
-Recently, \citet{Kwan2012} has argued for a careful examination of what she coins the uncertain geographic context problem (UGCoP), i.e.
+The discussion of these effects has also been associated with the field of geostatistics since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984, FotheringhamWong1991}.
+For example, \citet{Kwan2012} has argued for a careful examination of what she coins the 'uncertain geographic context problem' (UGCoP), i.e.
 of the spatial configuration of geographical units even when the size and delineation of the area are the same.\\
@@
 \item The accuracy of geo-localised input data;
-\item The spatial system shape, precision and boundaries;
-\item Spatial heterogeneity.
+\item The shape, precision and boundaries of the modelled spatial system;
+\item The degree of spatial heterogeneity modelled.
 \end{itemize}
@@
 \citet{moreno2009integrating} who explore the role of the topology of possible interactions in a Schelling model).
-Generally, sensitivity analyses focus either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely both at the same time.
+Generally, sensitivity analyses focus either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
 
@@
 
-In this paper, we build on the contributions of these previous studies and go further by systematically measuring the impact of the city lattice on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures.
-We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model of residential segregation and Sugarscape.
+In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures.
+We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.
 
@@
 
-In this section, we detail the method developed to analyse the sensitivity of simulation models to initial spatial conditions.
-The general method workflow is illustrated in Figure \ref{fig:method}.
-In addition to the usual protocol (upper branch in figure \ref{fig:method}, which consists of running a model $\mu$ with various values of its parameters, we introduced the lower branch in figure \ref{fig:method}.
-We wanted to emphasize that spatial effects do not only derive from grid size or shape effects, but indeed from the heterogeneity of relevant geographic entities (people, housing, networks, etc.) through geographical space.
-In models, the initial configuration of space can be either flat, or heterogeneous,  based on external databases and internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
+%In this section, we detail the method developed to analyse the sensitivity of simulation models to initial spatial conditions. 
+The general method workflow of our method is illustrated in Figure \ref{fig:method}.
+In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with various values of its parameters, we introduced a spatial generator which depends on its own meta-parameters and feeds the model $\mu$ at initialisation (lower branch).
+The resulting configurations can be clustered into qualitative types of spatial patterns, so that the sensitivity analysis relates to the model's parameters, how the density spatial distribution was generated and the pattern of density in order to produce qualitative conclusion about the model and its behaviour.
+In particular, we want to emphasize that spatial effects only derive not only from grid size or shape effects, but also from the heterogeneity of relevant geographic entities (people, housing, networks, etc.) in geographical space.
+In the models we produced, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-In our approach, we introduce a spatial generator (lower branch in figure \ref{fig:method}), which itself is determined by input parameters and produces sets of spatial initial conditions. 
-To allow for the exploration of the relative effects of model parameters and spatial configuration, we cluster initial spatial conditions  to represent types of spaces ex-ante (for example: monocentric or polycentric density grids).
-The sensitivity analysis of the model is now run against $\mu$ parameters as well as spatial parameters or spatial types.
-It allows the sensitivity analysis to produce qualitative conclusions regarding the influence of spatial distribution on the outputs of simulation models, alongside the classic variation of parameter values.
-
-In order to test for the influence of spatial initial conditions on model outputs, we use a systematic method to compare phase diagrams.
+In order to test for the influence of initial spatial conditions on model outputs, we use a systematic method to compare phase diagrams.
 We define a phase diagram as the vector of model outputs considered as a function of model parameters.
-Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values as the median of the final values of an output indicator obtained for the replications of the model initialized with this combination of parameters.\\
+Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\overrightarrow{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\overrightarrow{P}$.\\
 
@@
 
-\paragraph{Reproducibility and Applicability.} \comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
+\paragraph{Reproducibility and Applicability.} 
+%\comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
+
 Although the applications we present here are limited by the simplicity of the models, we think that the method could (and should) be applied to larger models including domain mechanisms and more empirical initialisation data, for example synthetic populations.
 The sensitivity analysis to initial spatial conditions could then be either a replication on the spatial allocation of the synthetic population, or a series of spatial permutations of the empirical spatial inputs.
-We want foster this extension of our work by releasing the density grids also generated, as well as the generating work-flow \comment[CC]{add link to github repo}.
+We want foster this extension of our work by releasing the density grids also generated, as well as the generating work-flow and the model implementation.
+They are available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
 Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents.
@@
 
-\comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}
-\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.}
+%\comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}
+%\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.}
+
+As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions, that become senseless taken out of context.
+Geography and social science in general have been in the last decades strongly influenced by insights from physics, that beside their enriching
 
@@
 
-\begin{acks}
-The authors acknowledge the funding of their institutions and the EPSRC project number EP/M023583/1.
+% acks commentés pour la soumission (anonymous)
+%\begin{acks}
+%The authors acknowledge the funding of their institutions and the EPSRC project number EP/M023583/1.
 Results obtained in this paper were computed on the vo.complex-system.eu virtual organization of the European Grid Infrastructure ( http://www.egi.eu ).
 We thank the European Grid Infrastructure and its supporting National Grid Initiatives (France-Grilles in particular) for providing the technical support and infrastructure.
-\end{acks}
+%\end{acks}
 

### 2017-11-03 paper.tex
.

@@
 As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions, that become senseless taken out of context.
-Geography and social science in general have been in the last decades strongly influenced by insights from physics, that beside their enriching
+Geography and social science in general have been in the last decades strongly influenced by insights from physics, that beside their highly enriching impact~\citep{o2015physicists}, may have softly impose strong assumptions such as homogeneity and isotropy of space.
+We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity for social science to emancipate and continue to build their own stream of methodologies in the modelling domain.
 

### 2017-11-03 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 
-In order to test for the influence of initial spatial conditions on model outputs, we use a systematic method to compare phase diagrams.
+In order to test for the influence of initial spatial conditions on model outputs, we use a systematic method to compare {\bf phase diagrams}.
 We define a phase diagram as the vector of model outputs considered as a function of model parameters.
-Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\overrightarrow{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\overrightarrow{P}$.\\
+Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
 
 %Comparing phase diagrams is useful because it allows us to compare the 
-Indeed, we have as many phase diagrams than we have spatial grids, what makes a qualitative visual comparison not realistic (with around 50 different spatial configurations in our applications).
-A solution is to use systematic quantitative procedures.
-To our knowledge there exist no well established method to compare phase diagrams in agent-based modeling and geosimulation literature.
-We propose to use a relative distance between two phase diagram corresponding to the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by
+We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment).
+A solution is to use systematic quantitative procedures to compare them.
+To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.
+We propose to define the relative distance $d_r$ between two phase diagrams $\vec{\alpha_1}$ and $\vec{\alpha_2}$ as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by
 
@@
 
-The last methodological point which we need to emphasis is the relation between the workflow we introduce and model exploration workflows.
+The last methodological point which we need to emphasize is the relation between the workflow we introduce and model exploration workflows.
 The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities.

### 2017-11-09 paper.tex
Update on Overleaf.

--- 2017-11-03 paper.tex
@@
 
-\title{Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?}
+\title{Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models
+%Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?
+}
 
@@
 %\affilnum{2}Laboratoire Ville Mobilit{\'e} Transport, Universit{\'e} Paris-Est, France\\
+%NB : mon affiliation "officielle" c'est : 
+%Le Néchet, F., Université Paris-Est, Laboratoire Ville Mobilité Transport, UPEMLV : 5 boulevard Copernic, Cité Descartes F 77454 Marne-la-Vallée cedex 2 France
 %\affilnum{3}Centre for Advanced Spatial Analysis, University College London, UK\\
@@
 
+% Dr.
+Marion Le Texier is associate professor at University of Rouen Normandie.
+She holds a PhD in geography from both Paris Diderot University and Luxembourg University.
+Her thesis aimed at modelling the circulation of euro coins through European borders with statistical and agent-based models.
+Previous to joining Rouen Normandie University in 2016, she has been a Jean Monet fellow at the European University Institute in Florence and a post-doc at Luxembourg University.
+Her work deals with data and modelling issues in different thematic contexts: access to nature in cities, international online media coverage of earthquakes, to name but two.
+
+% Dr.
+Florent Le Néchet is associate professor at  Paris-Est Marne-la-Vallée University (UPEM).
+He holds a PhD in planning from Paris-Est University (UPE).
+His thesis aimed at modelling the relationships between urban form and mobility patterns at metropolitan level, comparing several European cities, especially in France and Germany.
+His work deals with charactering and measuring mobility in cities, in relation to the built environment, and involves statistical techniques as well as multi-agent modelling.
+
+% Dr.
+Romain Reuillon is a research fellow at UMR CNRS 8504 Géographie-cités and at the Complex System Institute Paris Ile-de-France.
+He holds a PhD in Computer Science from Université Clermont II.
+His work deals with intensive computation model exploration, focusing on new techniques and methods to extract knowledge from models of simulation.
+He is the creator of the OpenMole platform for model exploration.
+
 % Cover Letter
@@
 
-%My coauthors and I are pleased to submit an original research article entitled ``Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?'' for consideration for publication in Environment and Planning B.
+%My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Environment and Planning B.
 
@@
 This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions.
-To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model, both defined at the intra-urban level.\\
+To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\
 
@@
 \\
+% détail de chez détail : j'enlèverais le ! les ... parlent d'eux-mêmes il me semble.
 
@@
 
-%The example of transportation networks is a good illustration, as their spatial shape and hierarchy is strongly influenced by past investment decisions, technical choices, or political decisions sometimes not rational~(\cite{zembri2010new}).
-Some aggregated indicators will not take into account positions and trajectories of each agent (such as segregation in the Schelling model) but others, as in the case of spatial patterns of accessibility in a system of cities, fully capture the path-dependency and may therefore be highly dependent of the initial spatial configuration.
-It is not clear for example what shifted the economical and political capital of France from Lyon to Paris in the early Middle Age, some assumptions being the reconfiguration of trade patterns from South to North of Europe and thus an increased centrality for Paris due to its spatial position: the bifurcation induced by socio-economic and political factors took a deep significance with worldwide repercussions until today when magnified by the spatial configuration.
-\comment[JR]{quelqu'un sait si d'une part c'est vrai et d'autre part si des references backupent ca ?
-- je sors ca completement du chapeau pour l'instant mais ça me semble intuitivement raisonnable et une bonne illustration.}[(MLT)L'idée de donner un exemple me semble bonne mais je ne suis pas sûre qu'on l'amène de la bonne manière?
---> en fait je crois que l'on a pas besoin de ce paragraphe illustratif, ça complique le papier.
-Qu'en pensez-vous?]  [(FL) : illustrer par les transports c'est une bonne idée mais pas besoin de parler du moyen âge!
-la faible croissance de Dijon depuis 30 ans peut être une illustration tarte à la crême.
-mais pas sûr effectivement qu'on ait absolument besoin d'illustration, en tous cas il n'y en a pas dans le § d'avant (bientôt après ;o)]
-
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-%In this section, we detail the method developed to analyse the sensitivity of simulation models to initial spatial conditions. 
 The general method workflow of our method is illustrated in Figure \ref{fig:method}.
@@
 
-In order to test for the influence of initial spatial conditions on model outputs, we use a systematic method to compare {\bf phase diagrams}.
+In order to test for the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}.
 We define a phase diagram as the vector of model outputs considered as a function of model parameters.
@@
 To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.
-We propose to define the relative distance $d_r$ between two phase diagrams $\vec{\alpha_1}$ and $\vec{\alpha_2}$ as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by
+We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$.
+Phase diagrams are denoted by a common function $\mu$ indexed by what we call \emph{meta-parameters} $\vec{\alpha}$, that capture the spatial configuration (in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself).
+The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by
 
@@
 \begin{equation}\label{eq:phase-distance}
-d_r\left(\vec{\alpha_1},\vec{\alpha_2}\right) = 2 \cdot \frac{d(f_{\vec{\alpha_1}},f_{\vec{\alpha_2}})^2}{Var\left[f_{\vec{\alpha_1}}\right] + Var\left[f_{\vec{\alpha_2}}\right]}
+d_r\left(\mu_{\vec{\alpha}_1},\mu_{\vec{\alpha}_2}\right) = 2 \cdot \frac{d(\mu_{\vec{\alpha}_1},\mu_{\vec{\alpha}_2})^2}{Var\left[\mu_{\vec{\alpha}_1}\right] + Var\left[\mu_{\vec{\alpha}_2}\right]}
 \end{equation}
 
-where $f_{\vec{\alpha_i}}$ are the phase diagrams at given meta-parameters $\vec{\alpha_i}$.
+where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given meta-parameters $\vec{\alpha}_i$.
 $d$ is a functional distance that we take simply as Euclidian distance.
@@
 
-The last methodological point which we need to emphasize is the relation between the workflow we introduce and model exploration workflows.
-The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the apparition of new methods and tools together with an explosion of computation capabilities.
-The model exploration platform OpenMole~\citep{reuillon2013openmole} allows to embed any model as a blackbox, write modulable exploration workflow using advanced methodologies such as genetic algorithms and distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
-In our case, the workflow tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and allow to couple any generator with any model in a straightforward way as soon as the model can take its spatial configuration as input or from an input file.
-In this paper, we use the OpenMole platform for spatial environment and model coupling, placing ourselves in the renewed framework of multi-modelling claimed by \citet{cottineau2015modular}.
+The last methodological point which we need to emphasize is the relation between the present workflow and model exploration workflows in general.
+The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities.
+The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write modulable exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
+In our case, this tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file.
+In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the renewed framework of multi-modelling \citep{cottineau2015modular}.
 %%%%%%%%%%%%%%%%%%%%%%
@@
 The spatial generator applies an urban morphogenesis model \citep{Batty2007}, which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}.
-An open implementation and a characterisation of urban forms the model can produce are provided, what allows us to integrate it easily in our workflow.
+An open implementation and a characterisation of the urban forms which the model can produce are provided, which allows us to integrate it easily in our workflow.
 In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
 More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
-At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its extended neighborhood (8 surrounding patches).
+At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (8 surrounding patches).
 To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside.
@@
 Typical value ranges for the  parameters will be taken as, following \citet{2017arXiv170806743R}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$.
-We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configuration that can be generated.
-
-%\comment[MLT]{Il me semble que l'on pourrait mieux expliciter la relation N/n (et peut-être du coup inverser N et n).
-Plus il me semble que l'on devrait dire plus clairement à quoi se réfère alpha : à une cellule $i$?
-au pas de temps $t$?
-Pourquoi ne pas indexer alpha?
-Enfin, je crois que l'on gagnerait encore plus en clarté si on expliquait formellement comment tous les paramètres sont reliés entre eux, notamment pour expliquer l'arrêt de la procédure d'attribution d'une nouvelle pop à la grille.} \comment[FL]{je crois que N c'est l'incrément de pop et n le nombre de fois où c'est smoothé mais c'est quoi la pop totale?}
+We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configurations that can be generated.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
+%\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}
+
+In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure: their concentration index, hierarchy index, centrality index and continuity index (cf.
+\cite{LeNechet2015}).
+We sample the meta-parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a point cloud with high discrepancy.
+We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}.
+It yields a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions.
+These are further clustered into three classes of morphology (figure \ref{fig:densityTypes}): compact (e.g.
+Vienna), polycentric (Liege) and discontinuous (Augsburg).
+This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results.
+We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model, obtaining a total of 45 spatial configurations to compare (as we will see, the significant number of model runs needed to establish each phase diagram imposes a computational limit on this number of configurations).
+
+%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure*}[htbp] \begin{center} 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}
-
-In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure : their concentration index, hierarchy index, centrality index and continuity index (cf.
-\cite{LeNechet2015}).
-We sample the meta-parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a point cloud with high discrepancy.
-We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n$, $N$}.
-It yields a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions.
-These are further clustered into three classes of morphology: compact (e.g.
-Vienna), polycentric (Liege) and discontinuous (Augsburg) in order to evaluate the non-trivial effects of urban form on simulation results.
-We select 15 grids of each type to work with in the computation of sensitivity analysis, obtaining a total of 45 spatial configurations to compare (as we will see, the significant number of model runs needed to establish each phase diagram imposes a computational limit on this number of configurations).
-
-%\comment[FL]{pourquoi 15?} \comment[FL]{dire explicitement ce qu'on va faire avec ces catégories dans la partie résultats?}
-
-In the following section, we briefly recall the main components of the two ``classical'' agent-based simulation models used to test how spatial density variations may impact simulation models behaviour and results, and how general the method proposed is.
+In the following section, we briefly recall the main components of the two ``classical'' agent-based simulation models used to test how spatial density variations may impact simulation models behaviour and results, and how general the method is.
 
@@
 
-Schelling's model consists in an abstract urban housing market where agents of different nature sense their environment, evaluate their satisfaction in terms of neighbourhood composition, and relocate if unsatisfied.
+Schelling's model consists in an abstract urban housing market where agents of different attributes (for example: red or green) sense their environment, evaluate their satisfaction in terms of neighbourhood composition (how many reds and greens?), and relocate if unsatisfied.
 It has been shown by \cite{Schelling1969} that even tolerant agents tended to produce segregated patterns because of the complexity of their local interactions.
-The main parameters of this model are the tolerance level (\% of agents similar to {\it ego}), the scope of sensing and the percentage of vacant spaces in the housing market.
+The main parameters of this model are the tolerance level (\% of agents similar to {\it ego}), the scope of sensing, the minority split and the percentage of vacant spaces in the housing market.
 In addition, we are interested in testing the impact of the spatial distribution of housing capacity in this project, using the generated grids.
@@
 Agents of different vision scopes and different metabolisms harvest a self-regenerating resource available heterogeneously in the initial landscape, they settle and collect this resource, which leads some of them to survive and others to perish.
-The main parameters of this model are the number of agents, their minimal and maximal resource.
-In addition, we are interested in testing the impact of the spatial distribution of the resource in this project, using the generated grids.
+The main parameters of this model are the number of agents, their minimal and maximal resource levels.
+In addition, we are interested in testing the impact of the spatial distribution of the resource, using the generated grids.
 We extend the implementation with agents wealth distribution of~\citet{li2009netlogo}.
@@
 \subsection{Experiment design}
-For Schelling's model, we explore a grid of a basic parameter space of the model, which is composed of two dimensions: the minimum proportion of similar agents required in the neighbourhood for the agent to be satisfied (or intolerance level) $S\in \left[0;1\right]$ and the vacancy rate of the city $V\in \left[0;1\right]$, which then defines the number of green and red agents, present in equal proportion for this experiment.
-We sample 1000 parameter values using a Sobol sampling and run 100 repetitions for each configuration.
-The initial spatial configuration varies across 45 different grids, chosen among the ones which are most representative of the three types of urban morphology: 15 compact grids (like Vienna, cf figure \ref{fig:densityTypes}), 15 polycentric grids (like Liege) and 15 discontinuous grids (like Augsburg).
-The full experiment thus equates to 4,500,000 simulations (1000 parameter combinations x 45 density grids x 100 replications).
-We use OpenMOLE to distribute the computation, and apply segregation measures to characterise the results.
-
-For Sugarscape, we explore a grid of a basic parameter space of the model, which is composed of three dimensions: the population of agents $P\in \left[10;510\right]$, the minimal initial agent resource $s_{-}\in \left[10;100\right]$ and the maximal initial agent resource $s_{+}\in \left[110;200\right]$.
+For Sugarscape, we explore three dimensions of the parameter space: the total population of agents $P\in \left[10;510\right]$, the minimal initial agent resource $s_{-}\in \left[10;100\right]$ and the maximal initial agent resource $s_{+}\in \left[110;200\right]$.
 Each parameter is binned into 10 values, giving 1000 parameter points.
-We run 50 repetitions for each configuration, what yield reasonable convergence properties.
+We run 50 repetitions for each configuration, which yields reasonable convergence properties.
 The initial spatial configuration varies across 50 different grids, generated by sampling meta-parameters for the generator in a LHS.
@@
 
-We choose different design of experiments, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices.
+For Schelling's model, we also explore three dimensions of the parameter space of the model: the minimum proportion of similar agents required in the neighbourhood for the agent to be satisfied (or intolerance level) $S\in \left[0;1\right]$, the initial split of population, derived from the proportion of green population, $G\in \left[0;1\right]$ and the vacancy rate of the city $V\in \left[0;1\right]$.
+We sample 1000 parameter values using a Sobol sampling and run 100 repetitions for each configuration.
+We first try the same experiment design (50 density grids generated on the fly), then look at clustered grids representing urban densities.
+We choose 45 different grids among the ones which are most representative of the three types of urban morphology: 15 compact grids (like Vienna, cf figure \ref{fig:densityTypes}), 15 polycentric grids (like Liege) and 15 discontinuous grids (like Augsburg).
+The last experiment thus equates to 4,500,000 simulations (1000 parameter combinations x 45 density grids x 100 replications).
+We use OpenMOLE to distribute the computation, and apply segregation measures to characterise the results.
+
+We choose different experiment designs, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices.
 In principle, our workflow applies with any manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.
@@
 
-\paragraph{Sugarscape} We measure the distance of all phase diagrams to the reference phase diagram computed on the default model setup, which is a bi-centric symmetric configuration (see Fig.~\ref{fig:sugarscape-distance-meta} for its morphological positioning regarded generated grids), using the relative measure defined in equation~\ref{eq:phase-distance}.
-Indeed, it gives in that case the average squared distance between corresponding points of the phase diagrams, relative to the average of the variance of each.
+We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default model setup (a bi-centric symmetric configuration) using the relative measure defined in equation~\ref{eq:phase-distance}.
+For each density grid, we then obtain the average squared distance between corresponding points of the phase diagrams, i.e.
+the median value of the final output measure, such as segregation or inequality, for a given value of parameters.
+This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry).
 Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.
+
+\paragraph{Sugarscape}
 
@@
 
-We obtain a very strong sensitivity to initial spatial conditions.
-Indeed, the relative distance of phase diagrams of different density grids to the reference case ranges from 0.09 to 2.98 with a median of 1.52 and an average of 1.30.
-The mean distance above 1 means that, on average, the model is more sensitive to meta-parameters of grid generation than to its own parameters (population and sugar).
-Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be 3 times bigger than the variation due to the model's parameters.
+We obtain a very strong sensitivity to initial spatial conditions for the Sugarscape model.
+Indeed, the relative distance between the phase diagrams of different density grids and the phase diagram of the reference case ranges from 0.09 to 2.98 with a median of 1.52 and an average value of 1.30.
+The mean distance above 1 means that, on average, the model is more sensitive to meta-parameters of grid generation than to its own parameters (population and sugar endowment) in the reference model.
+Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be up to three times bigger than the variation due to the model parameters.
 We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in a morphological space.
-Each point represent one of the 50 different density grids used to initialised the distribution of sugar in the model.
+Each point represent one of the 50 different density grids used to initialise the distribution of sugar in the model.
 The points are projected with respect to the meta-parameters used to generate the grid, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case.
@@
 with a small number of very high density cells) produce simulation results more different compared to the reference case than within this reference case because of parameter variations.
-This pattern is emphasized when grids are generated with a high $\alpha$ AND a high $\beta$ (i.e.
+This pattern is emphasized when grids are generated with a high $\alpha$ and a high $\beta$ (i.e.
 with low gradient of density decrease around the kernels of high density).
 These grids have the highest relative distance to the reference case.
-On the contrary, with grids closer to random and uniform patterns (bottom left of the graph), the model's parameters are more important in determining the simulation results.
+On the contrary, with grids closer to the uniform pattern of the reference case (bottom left of the graph), the model parameters are more important in determining the simulation results.
 
@@
 This 4-dimensional space defined a morphological space.
-For the purpose of interpretability and visualisation, we reduce this space to a bi dimensional space with a principal component analysis.
+For the purpose of interpretability and visualisation, we reduce this space to a bi-dimensional space with a principal component analysis.
 The first two components represent 92\% of cumulated variance.
@@
 It is confirmed by the behavior as a function of meta-parameters, as high values of $\alpha$ also yield high distance.
-In terms of model processes, it shows that congestion mechanisms in the gathering of the resource induce fast increases of inequality.
+In terms of model processes, it shows that congestion mechanisms in the gathering of the resource induces fast increases of inequality.
 To put these results in perspective of our work-flow given in Fig.~\ref{fig:method}, we have a sensitivity to spatial parameters in average greater than the sensitivity to model parameters.
@@
 The objective of this quantitative experiment is to evaluate to which extent this phase diagram is modified when different density grids are applied.
-We show in Fig.\ref{fig:schelling-distance-meta} the values of the relative distance as a function of meta-parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape before.
+We show in Fig.\ref{fig:schelling-distance-meta} (Supplementary Material) the values of the relative distance as a function of meta-parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape before.
 Variations are less considerable than for Sugarscape across phase diagrams, but values close to 1 show that several configurations are as much sensitive to space than to their parameters.
@@
 
-%%%%%%%%%%%%%
-\begin{figure}
-\centering
-\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams}
-\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace}
-\caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams.
-We show them in the meta-parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
-\end{figure}
-%%%%%%%%%%%%%
-
 \subsection{Qualitative variations}
-\paragraph{Sugarscape} We now check the sensitivity in terms of qualitative behavior of phase diagrams.
-We show the phase diagrams for two very opposite morphologies in term of sprawling, but controlling for aggregation with the same $PC2$ value.
-These correspond to the green and blue frames in Fig.~\ref{fig:sugarscape-distance-pca}.
-The behaviors are rather stable for varying $s_+$, what means that the poorest agents have a determinant role in trajectories.
-The two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values.
-The process is thus completely inverted, what would have significant impacts if one tried to schematise policies from this model.
-To interpret these behavior in terms of grid shape, we observe that the difference between the two grids is mainly on average distance and entropy: in a nutshell, the first grid is much more dispersed and disorganised than the second.
-In sprawled spaces, inequalities are thus fostered by a lack of minimal local ressources, whereas population will drive inequality in concentrated spaces.
-
-%This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
-
-% phase diagrams -> ok well different qualitatively
-%          spAlpha spDiffsteps spDiffusion spGrowth spPopulation
-% id=27 : 0.7913103    2.376837   0.1440293 157.4147 4852.746
-% id=0 : 2.562398    3.753032   0.1316788 128.4632 4753.983
-% maxSugar = 110
-
-%%%%%%%%%%%%%
-\begin{figure}
-\centering
-\includegraphics[width=0.48\textwidth]{figures/phasediagram_id27_maxSugar110}
-\includegraphics[width=0.48\textwidth]{figures/phasediagram_id0_maxSugar110}
-\caption{\textbf{Examples of phase diagrams.} We show two dimensional phase diagrams on $(P,s_-)$, both at fixed $s_+ = 110$.
-(Left) Green frame, obtained with $\alpha = 0.79$, $n=2$, $\beta = 0.14$, $N=157$; (Right) Blue frame, obtained with $\alpha = 2.56$, $n=3$, $\beta = 0.13$, $N=128$.}
-\label{fig:sugarscape-phasediagrams}
-\end{figure}
-%%%%%%%%%%%%%
 
@@
 
+\paragraph{Sugarscape} We now check the sensitivity in terms of qualitative behavior of phase diagrams.
+We show the phase diagrams for two very opposite morphologies in terms of sprawling, but controlling for aggregation with the same $PC2$ value.
+These correspond to the green and blue frames in Fig.~\ref{fig:sugarscape-distance-pca}.
+In terms of grid shape, we observe that the difference between the two grids is mainly on average distance and entropy: in a nutshell, the first grid is much more dispersed and disorganised than the second.
+Although the behaviours are rather stable for varying $s_+$, the initial maximum endowment in sugar, which means that the poorest agents have a determinant role in trajectories, the two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values.
+In sprawled spaces, inequalities are thus fostered by a lack of minimal local resources, whereas population will drive inequality in concentrated spaces.
+The process is thus completely reversed depending on the grid chosen to run the model, which would have significant impacts if one tried to deduce policies from this model.
+%This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
+
+% phase diagrams -> ok well different qualitatively
+%          spAlpha spDiffsteps spDiffusion spGrowth spPopulation
+% id=27 : 0.7913103    2.376837   0.1440293 157.4147 4852.746
+% id=0 : 2.562398    3.753032   0.1316788 128.4632 4753.983
+% maxSugar = 110
+
+%%%%%%%%%%%%%
+\begin{figure}
+\centering
+\includegraphics[width=0.48\textwidth]{figures/phasediagram_id27_maxSugar110}
+\includegraphics[width=0.48\textwidth]{figures/phasediagram_id0_maxSugar110}
+\caption{\textbf{Examples of phase diagrams.} We show two dimensional phase diagrams on $(P,s_-)$, both at fixed $s_+ = 110$.
+(Left) Green frame, obtained with $\alpha = 0.79$, $n=2$, $\beta = 0.14$, $N=157$; (Right) Blue frame, obtained with $\alpha = 2.56$, $n=3$, $\beta = 0.13$, $N=128$.}
+\label{fig:sugarscape-phasediagrams}
+\end{figure}
+%%%%%%%%%%%%%
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-We consider that the method presented holds great potential for strengthening geographical models' validation.
-However, two limits and two areas of opportunities have still not been tackled.
-We discuss them in this section.
+We consider that the method presented holds great potential for strengthening geographical models' validation. However, two limits and two areas of opportunities have still not been tackled.
 
@@
 
-\paragraph{Comparing phase diagrams.} Comparing phase diagrams is as we saw not straightforward, and further developments of our method imply testing alternative methods for this particular point.
+\paragraph{Comparing phase diagrams.} Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point.
 For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
-The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\cite{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
-Map comparison methods, popular in environmental sciences %\comment[MLT]{Et pas en géo?
-Qu'est-ce que ces méthodes ont de particulier?}
-, provide numeral tools to compare two dimensional fields~\cite{visser2006map}.
-To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\cite{10.1371/journal.pone.0178165}.
+The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\citep{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
+Map comparison methods, popular in environmental sciences, provide numeral tools to compare two dimensional fields~\citep{visser2006map}.
+To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\citep{10.1371/journal.pone.0178165}.
 The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.
 
-\paragraph{Platform constrains and docking challenges.} An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
+\paragraph{Platform constraints and docking challenges.} An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
 For example, the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements.
@@
 The sensitivity analysis to initial spatial conditions could then be either a replication on the spatial allocation of the synthetic population, or a series of spatial permutations of the empirical spatial inputs.
-We want foster this extension of our work by releasing the density grids also generated, as well as the generating work-flow and the model implementation.
+We want to foster this extension of our work by releasing the density grids also generated, as well as the generating work-flow and the model implementation.
 They are available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
@@
 As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions, that become senseless taken out of context.
-Geography and social science in general have been in the last decades strongly influenced by insights from physics, that beside their enriching
+Geography and the social sciences in general have been in the last decades strongly influenced by insights from physics, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space.
+We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity the for social sciences to build their own stream of methodologies in the modelling domain.
 
@@
 With the Schelling experiment, we found that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
-With Sugarscape, we show that the model is more sensitive to space than to its other parameters, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behavior of phase diagram is qualitatively different in different regions of the morphological space.
-We think this method has the potential to increase the arsenal of evaluation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
+With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space.
+We think that this method has the potential to increase the arsenal of evaluation/validation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
 %%%%%%%%%%%%%
@@
 
+\newpage
+
+\begin{sm}
+
+%%%%%%%%%%%%%
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams_red}
+\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace_red}
+\caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams.
+We show them in the meta-parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
+\end{figure}
+%%%%%%%%%%%%%
+
+\end{sm}
+
 \end{document}

### 2018-04-10 paper.tex
Update on Overleaf.

@@
-\documentclass[Royal,sageh,times]{sagej}
-% Afour
+%% The '3p' and 'times' class options of elsarticle are used for Elsevier CRC
+%% The 'procedia' option causes ecrc to approximate to the Word template
+\documentclass[3p,times,procedia]{elsarticle}
+\flushbottom
+
+%% The `ecrc' package must be called to make the CRC functionality available
+\usepackage{ecrc}
+
 \input{header.tex}
+
+%\runauth{Raimbault et al.}
+\runauth{}
 
@@
 
-%\runninghead{Raimbault \& al.}
-\runninghead{}
+\begin{frontmatter}
+
+\dochead{}
 
@@
 
-%\author{Juste Raimbault\affilnum{1,2}, 
-%Cl{\' e}mentine Cottineau\affilnum{3,4},
-%Marion Le Texier\affilnum{5}, 
-%Florent Le N{\' e}chet\affilnum{2}, 
-%Romain Reuillon\affilnum{1,6}
-%}
-\author{}
+\author[a,b]{Juste Raimbault\corref{cor1}}
+\author[c,d]{Cl{\' e}mentine Cottineau}
+\author[e]{Marion Le Texier}
+\author[f]{Florent Le N{\' e}chet}
+\author[a]{Romain Reuillon}
 
-%\affiliation{\affilnum{1}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
-%\affilnum{2}Laboratoire Ville Mobilit{\'e} Transport, Universit{\'e} Paris-Est, France\\
+\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
+\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
+\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
+\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
+\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
+\address[f]{Université Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
 %NB : mon affiliation "officielle" c'est : 
 %Le Néchet, F., Université Paris-Est, Laboratoire Ville Mobilité Transport, UPEMLV : 5 boulevard Copernic, Cité Descartes F 77454 Marne-la-Vallée cedex 2 France
-%\affilnum{3}Centre for Advanced Spatial Analysis, University College London, UK\\
-%\affilnum{4}CNRS, UMR 8097 Centre Maurice Halbwachs, Paris, France\\
-%\affilnum{5}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
-%\affilnum{6}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}
+
+\cortext[cor1]{Corresponding author.}
 
@@
 %75006 Paris, France.
-%\email{juste.raimbault@polytechnique.edu}}
+%}
 %\corrauth{}
@@
 Her thesis aimed at modelling urbanisation in the post-Soviet space with modular agent-based models.
-Since 2014, she has been working at the Centre for Advanced Spatial Analysis on urban economics and parametric scaling in the context of European systems of cities, using tools from complexity science to investigate inequality in cities and economic disparities between cities.
+Since 2014, she has been working at the Centre for Advanced Spatial Analysis on urban economics and parametric scaling in the context of European systems of cities, using tools from complexity science to investigate inequality in cities and economic disparities between cities.Conc
 
@@
 
+% ********************************************
+% Review EPB :
+
+%Referee: 1
+
+%Comments to the Author
+
+%One sentence in the abstract of the paper, which is a part of the review invitation, was indeed confusing “However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment.” What can be that special in initial conditions?
+There are standard notions of equilibrium and, more general “attractor” of the dynamic system and there are “domains” of “basins” of attraction for each attractor that defines to which of the attractors the solution will converge in time (see http://www.scholarpedia.org/article/Basin_of_attraction).
+May “spatial” in “spatial conditions” make a difference?
+%I believe that after reading the paper I understand the way of author(s) thinking.
+Regrettably, their view of the dynamic system as well as their use of the basic notions and examples, are wrong.
+In what follows I will try to point to the major steps where the authors deviate, further and further away, from the proper view of the complex dynamic system theory.
+%The introduction reads very well while raises some general questions.
+E.g., cities’ “non-ergodicity” is a kind of a philosophical statements, some urban processes, as buildup area dynamics, can be indeed considered as path-dependent, while some, as urban traffic, are most probably, not.
+Whatever, introduction does not cause problems.
+%Then the methodical section comes, posing the first red flag.
+I mean the choice of the Schelling and Sugarscape modes as the examples.
+On the one hand, none of these models deals directly with the physical structure of space; even more, studies of the e.g.
+Schelling model on graphs show that its dynamics do depend on the physical structure of space.
+On the other, each of these models has many versions and, as dynamic spatial systems, these versions are very different.
+The dynamics of some versions of the Schelling model are essentially history-dependent, while the dynamics of the others are not.
+The objectives of the paper demand very careful recognition between the system with path-dependent and path-independent dynamics.
+%From this point on, the paper took a “believe me” path.
+The impression of the reader is that the authors are absolutely sure in their views and, thus, ignore numerous critical points that may cause reader’s doubts.
+Even definitions became problematic.
+“We define a phase diagram as the vector of model outputs considered as a function of model parameters.” Sorry, phase diagram is a standard notion and it is something else, see https://en.wikipedia.org/wiki/Phase_diagram.
+Further on, “Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values P as the median of the final values of an output indicator O obtained for the replications of the model initialized with P”.
+Well, the reader can guess now that the intention of the author is to investigate the dynamics of the aggregate characteristics of the output as dependent on the parameter(s) P of the system.
+Putting aside wrong use of “phase diagram,” this is manageable, while quite standard.
+However, the part of this statement raises deeper concerns: The dependence of the “median O” on P may work well in case when the stochastic system converges to a globally stable stochastic equilibrium.
+But what if it doesn’t?
+Let’s consider the simplest case of a stochastic system with two attractors, each being locally stable stochastic equilibrium.
+In this case, initial conditions that are close to the border between equilibrium’s basins would evidently result in very different values of O.
+To avoid a vicious circle, one cannot use median or any other aggregate characteristic before the attractors are properly recognized.
+That is, one cannot use median before system’s properties are known well enough.
+%Ok, let’s skip “median” and follow the paper based on the single solutions.
+Let us essentially reduce the class of systems to keep in mind too, assuming that (1) for every P, every attractor is just an equilibrium maybe, stochastic one; (2) variation in P can result in variation in the number of equilibriums and their location in a phase space.
+In this case, conceptually, given K output characteristics, we can indeed recognize the basins of attraction based on the distance between solutions as far as they stabilize in time.
+Formula (1) is ok in this respect, while I would consider multidimensional O and standard measures, e.g.,   https://math.stackexchange.com/questions/1358633/malalanobis-distance-between-two-multivariate-gaussian-distributions.
+%From this point, let us continue with the Schelling model only.
+Sugarscape is more complex and less studied.
+In respect to the papers goals, the background property of the Schelling model was formulated in Vinkovic, Kirman, 2006, A physical analogue of the Schelling model, Proceedings of the National Academy of Sciences, 103(51), 19261–19265.
+Namely, they distinguish between models, which rules enable relocation to a better location only and the rules that enable relocation to a cell of the same utility.
+Models of the first subclass always stall, with the non-zero (and, often, essential) fraction of discontent agents who are unable to find a better location.
+Models of the second subclass converge to simple stochastic equilibrium patterns that can be segregated or integrated depending on parameters.
+There may be other classifications of the Schelling-like models into those, which solutions stall and those, which solutions are simpler.
+However, the basic problem of the reviewed paper is that the Galvin’s et al (2009) version of the Schelling model belongs to the first class – its solutions are strongly path-dependent and, indeed, can stall in a great variety of unstable steady states.
+That is, in mathematical terms, the number of attractors of the version of Schelling model that is investigated in the paper is very large and the attractors’ patterns are complicated.
+One can follow here Chris Langton’s ideas of 1990s or recall Game of Life or elementary CA that served to Steven Wolfram for his “New Kind of Science” https://www.wolframscience.com/.
+Whatever, we cannot just average outputs of the dynamic systems for arbitrarily/randomly chosen set of parameters.
+Just as we cannot study the dynamics of the elementary CA until we recognize the correspondence between the rules that govern them and the patterns they produce.
+Compare the 50, 75, 86, etc.
+rules and completely different dynamic systems they generate http://mathworld.wolfram.com/ElementaryCellularAutomaton.html.
+At this point, the major idea of the paper is destroyed by: (1) path-dependent version of the Schelling (and Sugarscape) model; (2) extremely wide spectrum of initial conditions and (3) aggregate view of solutions.
+The authors overlook the consequences of the (1) – (3) and this takes them too far away from the correct view of the complex systems.
+The result is totally wrong result section.
+The observations of this section all describe averages over extremely heterogeneous sets of solutions that, just as the average characteristics of all elementary CA, they do not represent any meaningful phenomena.
+%Minor comment: I have mentioned above some specific paper’s flaws. There are more.
+
+%Referee: 2
+
+%Comments to the Author
+
+%I totally agree with the authors that “space matters” in models and this is one issue that makes geographical models complicated compared to say economic models that don’t consider spatial relationships in them.
+To me this is one of the first papers that to show systemically how initial spatial conditions (and variations in such) impact model results.
+By taking two well know agent-based models (Schelling’s Segregation model and Sugarscape) the authors are able to show quantitatively and qualitatively how different spatial conditions impact the results.
+
+%However, I find it hard to follow your argument especially based on what you write: “We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled.” Is unclear, in the sense, for example for your first condition, if I have a geographically explicit model of a real world place using GIS, say within an agent-based model, is space not directly accounted for and the behavior of the agents are also explicitly spelled out?
+There are many examples of this, as you note / cite in the paper.
+With respect to the second condition “provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled” you write this is seldom met.
+In what sense?
+Maybe this might be the case for abstract models like the ones you have used here but how can you alter space in geographically explicit models using real data?
+Most geographically explicit models are focused on a specific area therefore how could one alter the spatial environments without applying it to another area which is often a difficult task and seldom done.
+
+%Moreover, it is unclear in the current version of the paper if you turned off the “toroidal default setting” in NetLogo when using your different spaces in the Schelling model.
+I write this because if you are using realistic spaces does it make sense to have agents say at the south of the city also communicating with agents (i.e.
+checking their neighbors) in the north of the city?
+Which would be the case the Schelling NetLogo model, which would not be the case in reality.
+
+% One of your key words you use “model validation” but in what sense?
+You don’t really come back to this in the paper, also you write about “the validity of generative models is uncertain until their results are proven robust and representative of ’real-world’ conditions.” What do you consider real world conditions to be?
+You seem to allude that you need to have the spatial generator to generate realistic (styles) city shapes for testing on the Schelling model but I am not convinced.
+
+%This seems to be superfluous on your main purpose of paper.
+Your paper could be much shorter if you just show how different spatial configurations impact the results rather than spending time talking about spatial patterns.
+Don’t get me wrong, I like how your spatial generator produces patterns seen within European cities and captures compact, polycentric, etc.
+urban forms under different parameter configurations but this appears to be moving away from the focus of the paper (at least as its portrayed in the introduction) specifically how initial spatial conditions impact the results.
+
+%I guess, I am wondering if the paper could not be shorter and simpler if this generator was removed?
+Or I would suggest you refocus the paper/ or better articulate in your introduction the objective, how different urban forms (initial conditions impact model results).
+You have clear objectives: “1/ to test the robustness of simulation results to small variations of meta-parameters and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric vs.
+polycentric for example) on the results of a given model.” But your introduction does not lead clearly to this.
+
+%Basically, I think the paper needs a more clearer focus, a better introduction which focus specifically on its objectives and clearer structure from background to results.
+
+%Response to reviewers (Juste) :
+%rev 1 :
+%This review is quite arrogant ad totally misses the point of the paper, trying to fit a inappropriate framework on our work.
+%- we do not study at all the attractors of the dynamic, but aggregated indicators to have a synthetic view of model behavior.
+The first remark is difficult to understand, and yes the spatial makes a difference (mostly in practice).
+%- of course we deviate from dynamic system theory because we are not AT ALL in this theoretical framework.
+It is strange to try to fit a given framework to something that is positioned in a totally different one.
+%- choice of the model : precisely it has not been shown systematically how model behavior depends on space, thus the choice.
+And these models fundamentally deal with space as it is in the core of agents dynamics.
+%- the term phase diagram is used following the Gauvin paper, and we assume this extension of the concept.
+%- the theoretical example of the saddle hypersurface makes no sense here in practice, as indicators we took precisely converge to narrow unimodal distributions (easily distinguishable when parameters vary)
+%- we know that the attractor patterns are complicated, that's exactly why we use simulation and study aggregated indicators.
+%- references on CA are not useful here (none of these models are CA and it would be difficult to fit within a CA fwk, by forcing the neighborhood to be all space).
+(looks like an argument from authority)
+%- yes we can average macro outputs, that the standard way to do in simulation and empirically it shows no problem here -> maybe add a paragraph to show the intra-run variablity compared to variation between parameters, and convergence rate etc.
+%- the conclusion are not "destroyed" as : (1) this is true for the model state, but we study aggregated outputs that are NOT path-dependent - did the reviewer entirely read the paper ?
+; (2) idem, we average on initial distributions ; (3) idem.
+There is no "correct view of complex systems", there are different approaches, and yes we are far from dynamical systems because we take an other approach.
+%- "do not represent meaningful phenomena" : seriously this reviewer never saw any work in simulation ; finding appropriate macro indicators for a model, that have a thematic sense (segregation in our case) is at the core of the study of simulation models.
+%- please give us the other flaws that's the aim of a review (and 6 month to do a 1 page out-of-scope review, we could have expected more).
+
+%rev 2 :
+%- on the alteration of a real space, that is out of the scope of the current paper, but yes we aim at producing synthetic geographical configurations that capture a typical spatial configuration -> maybe explicit that in the discussion
+%- on model validation -> yes we should precise the concept
+% - we need to be more precise on the schelling implementation/specification
+% - ok, we can be shorter on the generator (or put more explicitly that we provide it and the grids for other modelers to do similar analysis
+
+%Response to reviewers (Marion) :
+
+%rev1
+
+%May“spatial” in “spatial conditions” make a difference?
+%=> mouai, je ne suis pas non plus convaincue de la pertinence de cette remarque, à laquelle on répond partiellement dans la literature review.
+Idem que Juste, je pense qu’il s’agit d’un problème de cadrage théorique.
+En lien avec la dernière remarque du reviewer 2, je pense que l’on peut ajouter un court paragraphe sur le positionnement du papier en intro.
+
+% cities’ “non-ergodicity” isa kind of a philosophical statements, some urban processes, as buildup area dynamics, can be indeed considered as path-dependent, while some, as urban traffic, are most probably, not.
+% =>  On peut reprendre cette remarque pour illustrer la question de la non-ergodicité des villes ?
+
+%The objectives of the paper demand very careful recognition between the system with path-dependent and path-independent dynamics.
+% => Bon, ici comme on dit clairement que l’on est pas là pour discuter des comportements particuliers de ces deux modèles, je ne reprendrais pas les remarques ci-dessus.
+On peut les garder en mémoire si on nous repose la question lors de la 2 nd soumission du papier.
+
+%  Sorry, phase diagram is a standard notion and it is something else, see https://en.wikipedia.org/wiki/Phase_diagram.
+% => je préfère aussi citer Gauvin à Wikipedia ;)
+
+% the reader can guess now that the intention of the author is to investigate the dynamics of the aggregate characteristics of the output as dependent on the parameter(s) P of the system.
+Putting aside wrong use of “phase diagram,” this is manageable, while quite standard.
+% => « the reader can guess now that » : on doit le dire plus clairement et en amont
+
+% That is, one cannot use median before system’s properties are known well enough.
+% => Sa remarque n’est pas non plus sans justifications… je pense que l’on doit afficher plus clairement notre cadre d’analyse en introduction, pour que l’on ne nous reproche pas par la suite de ne pas avoir exploré entièrement le comportement du modèle avant de tirer des conclusions de l’analyse d’indicateurs agrégés.
+Mais on montre justement que certains espaces de paramètres sont plus ou moins sensibles aux variations de conditions spatiales initiales… je ne pense pas que la résolution analytique du modèle viendrait contredire une telle conclusion ?
+
+% CAs and number of attractors 
+% => Ok, on peut garder ce type d’éléments pour la partie discussion, en disant bien que le papier est un premier proof of concept et que notre but n’est pas d’étudier le comportement de ces deux modèles… je ne pense pas que l’on sur-interprétait nos résultats, mais seulement que l’on pointait les variations de sensibilité aux conditions initiales.
+Il me semble que le fait que cela soit dû à différents facteurs de variabilité du comportement du modèle ne pose pas trop de problèmes vu le niveau de généralité de l’analyse que l’on propose ?
+
+% The observations of this section all describe averages over extremely heterogeneous sets of solutions that, just as the average characteristics of all elementary CA, they do not represent any meaningful phenomena.
+% => idem que plus haut, avec en plus l’idée que on dit bien que la path-denpendency est justement une des raisons pour lesquelles la prise en compte de l’ensemble des conditions initiales dont l’espace est primordiale…
+
+% rev 2 :
+
+%Most geographically explicit models are focused on a specific area therefore how could one alter the spatial environments without applying it to another area which is often a difficult task and seldom done.
+% => Justement, un papier comme celui de Isabelle Thomas et al.
+montre que même avec des modèles spatially explicit comme il dit, on doit prendre en compte l’effet de la configuration spatiale choisie (résolution de l’information géographique, effets de bord, etc.), bref c’est pas parce que tu as des données spatialisées que tu évites le MAUP, loin de là.
+En plus, ces modèles dressent des conclusions qui sortent souvent du cadre de leur espace d’études, sur la relation entre deux phénomènes par ex.
+
+% Which would be the case the Schelling NetLogo model, which would not be the case in reality.
+% => Euh, on utilise un torus nous ???
+
+% model validation 
+% C’est marrant, je dis souvent à mes étudiants que lorsqu’on utilise des « » ça met l’accent sur une notion non définie, on s’est fait piéger à notre tour !
+La question est posée de savoir si l’on parle d’exploration du modèle ou de validation dans le papier ?
+La première solution nous éviterait de tomber dans un débat épistémologique sur les conditions nécessaires à la validation d’un modèle, et l’existence même d’un tel état de validation…
+
+% spatial generator and focus
+% Je suis d’accord avec cette partie, après il me semble que le générateur de formes urbaines est un vrai plus du papier et donc je le garderais à moins que d’autres reviewers nous fassent le même commentaire.
+
 % **************  ABSTRACT  **************
@@
 
-\keywords{Space, Initial conditions, Sensitivity, ABM, Validation}
+\begin{keyword}
+Space \sep Initial conditions \sep Sensitivity \sep ABM \sep Validation
+\end{keyword}
 
-\maketitle
+\end{frontmatter}
 
-% **************  MAIN BODY OF THE PAPER **************
-%\newpage
+\email{juste.raimbault@polytechnique.edu}
+
+%\maketitle
+
+%% main text
+
+\enlargethispage{-15mm}
 
@@
 
-\bibliographystyle{SageH}
+\bibliographystyle{elsarticle-harv}
 \bibliography{spacematters}
@@
 
-\begin{sm}
+\section*{Supplementary material}
 
@@
 
-\end{sm}
-
 \end{document}

### 2018-05-31 paper.tex
Update on Overleaf.

@@
 Her thesis aimed at modelling urbanisation in the post-Soviet space with modular agent-based models.
-Since 2014, she has been working at the Centre for Advanced Spatial Analysis on urban economics and parametric scaling in the context of European systems of cities, using tools from complexity science to investigate inequality in cities and economic disparities between cities.Conc
+Since 2014, she has been working at the Centre for Advanced Spatial Analysis on urban economics and parametric scaling in the context of European systems of cities, using tools from complexity science to investigate inequality in cities and economic disparities between cities.
 
@@
 
-%My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Environment and Planning B.
+%My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Computer, Environment and Urban Systems.
 
@@
 the spatial configuration of agents and their environment), and more precisely their spatial initial conditions.
-We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of EPB, a journal dealing with urban modelling and city science.
+We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of CEUS, a journal dealing with computerised models and urban systems.
 
 %Despite some preliminary results being presented as at ECTQG 2015, Bari, and at Geocomputation Conference 2017, Leeds, we confirm that neither the manuscript nor any parts of its content are currently under consideration or published in another journal.
-We did not have any previous interaction with EPB regarding this manuscript.
+We did not have any previous interaction with CEUS regarding this manuscript.
 We have no conflicts of interest to disclose.
@@
 %\setlength{\parindent}{0pt}
-Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust and representative of 'real-world' conditions.
+Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain \replaced{unless}{until} their results are proven robust and representative of 'real-world' conditions.
 Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes.
 However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment.
-In this paper, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create the density grids with which spatial simulation models are initialised.
-We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of resource harvesting) and a straightforward open-source work-flow using high performance computing, that the effect of space is significant in the two types of simulation.
+In this paper, we present a method to assess the effect of \added{some} initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create \deleted{the} density grids with which spatial simulation models are initialised.
+We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of \replaced{unequal societies}{resource harvesting}) and a straightforward open-source work-flow using high performance computing, that the effect of \replaced{initial spatial arrangements}{space} is significant in the two \replaced{models}{types of simulation}.
 Furthermore, this effect is sometimes larger than the effect of parameters' value change.
@@
 
+\todo{Response to reviewers
+\begin{itemize}
+\item For reviewer 1, explicitly mention the simulation framework at the beginning (and position it regarding others such as dynamical systems) with literature
+\item Similarly, clarify the aggregation step, how the indicators are macro, have a thematic meaning, and distribution converge clearly (possibly : add Sharpe ratios statistics and type of distrib for better fits)
+\item Rev 1 : (complementary to above) : positioning regarding some analytical works / cite some ?
+Precise that we do not aim at understanding fully model behavior, but finding some sensitivity to space
+\item Rev 2 : precise which Schelling we use.
+\item On real configs that can not be modified : remark by Marion on MAUP etc ; and add in discussion a layus on synthetic data resembling real configs etc.
+\item Remove or adapt the concept of validation to avoid the epistemological trap
+\item To be debated : do we keep the detailed description of the spatial generator ? (knowing that it is otherwise fully described and explored here : \url{https://arxiv.org/abs/1708.06743})
+\end{itemize}
+}
+
+\textbf{Example track changes : }\added{this is added text}, \deleted{removed} text or \replaced{new}{old} replaced text
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-Simulation within computer models has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}.
+\added{Computer} simulation \deleted{within computer models} has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}.
 %\comment{+ Seb \& James Doran in the 1970s??}
- Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of social and ecological systems in space and time.
-The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model.
-Indeed, they usually seek to understand and model the role of space in social interactions and environmental processes, in a city for example.
-We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled.
+ Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of \added{path-dependent} social and ecological systems\deleted{ in space and time}.
+The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model \replaced{in}{.
+Indeed, they usually seek to} understand\replaced{ing}{ and model} the role of space \replaced{on}{in} social interactions and environmental processes, in a city for example.
+We think simulation can become a very good tool to achieve this, provided that models \deleted{do} include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is \replaced{represented}{modelled}.
 Unfortunately, the first condition is not always met, and the second is seldom even mentioned.
@@
 
-An urban system can be defined as a large number of diverse social agents interacting with one another, within a limited portion of space.
-Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
-Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}.
+\replaced{Geographical systems}{An urban system} can be \replaced{crudely described as}{defined as a large number of diverse} social agents interacting with one another, within a limited portion of space.
+Social agents thus constitute the microscopic level of \replaced{the}{urban} system\deleted{s}, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects\added{, also known as path-dependent \citep{arthur1994increasing}.
+Therefore, observing a system at different points in time does not equate to observing different system at a single point in time.
+This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}}.
+Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
+Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models...
+which are the main models to be found in geosimulation.
+
+Self-organization has been shown to be a central feature of \replaced{geographical}{urban} systems \added{in general and cities in particular} ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
+In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}.
 Complexity is partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.
 Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures.
+\added{In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as other initial conditions, but of more interest to the geographer}.
 \\
 
-Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity~\citep{pumain2012urban}, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average.
-Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
-Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models...
-which are the main models to be found in geosimulation!
-\\
-% détail de chez détail : j'enlèverais le ! les ... parlent d'eux-mêmes il me semble.
+\deleted{Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average.}
 
-Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) in cities is far from isotropic, even in the most sprawled cities.
+Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) \deleted{in cities} is far from isotropic, even in the most sprawled cities.
 On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities.
 In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions.
-However, most geographical models, especially cellular automata, represent cities as uniform grids of isotropic densities.
+However, most geographical models, especially cellular automata, \added{still} represent cities as uniform grids of isotropic densities.
 Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}.
@@
 For example, households can have different preferences with respect to the built-environment they want to live in \citep{SpielmanHarrison2014}, or agents can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}.
-The way modellers abstract space is therefore a central element of geographical simulation models.
-A meaningful way of abstracting space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures.
+The way modellers \replaced{represent}{abstract} space is therefore a central element of geographical simulation models.
+\added{However, this step is rarely explicit and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one}.
+A meaningful way of \replaced{representing}{abstracting} space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures\added{ so as to estimate the variability of the model behaviour in plausible spatial arrangements}.
 
@@
 
-In this paper, we suggest a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation.
+In this paper, \replaced{we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape).
+In no way do we want to provide a full exploration of these two particular models, their attractors and/or potential policy implications.
+Instead, we present}{we suggest} a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation.
 The generator being controlled by meta-parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations.
@@
 
-Considerations of such issues in the geographical modelling literature are rather scarce, although there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions.
-Mainly, they have focused on the impact of:
+Considerations of such issues in the geographical modelling literature are rather scarce.\replaced{Despite the opportunity of simulation in the representation of space and its potential observation biases,}{ although} there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions\replaced{, which}{.
+Mainly, they} have focused on the impact of:
 \begin{itemize}
@@
 \end{itemize}
-We detail these approaches in the paragraphs below.
+
+\deleted{We detail these approaches in the paragraphs below.
 Other approaches, not included in the scope of this article, include the effects of the topology of space on the processes (cf.
-\citet{moreno2009integrating} who explore the role of the topology of possible interactions in a Schelling model).
-Generally, sensitivity analyses focus either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
+moreno who explore the role of the topology of possible interactions in a Schelling model).
+}
 
-\paragraph{Geo-localised input data accuracy:} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and scale of analysis.
+\paragraph{Geo-localised input data accuracy} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and scale of analysis.
 They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes.
-In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.
+In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and \added{the} computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.
 
-\paragraph{Spatial system shape, precision, and boundaries:} \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e.
+\paragraph{Spatial system shape, precision, and boundaries} \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e.
 made equivalent to) to Axelrod Culture Model.
 \citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are higher in a rectangular grid and lower in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents.
-\citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours.
+\citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favours segregationist behaviours.
 \citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins.
@@
 
-\paragraph{Spatial heterogeneity:} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
+\paragraph{Spatial heterogeneity} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
 They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns.
@@
 %\paragraph{Combination:} 
-
+ Generally, sensitivity analyses \replaced{have focused}{focus} either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
 %\cite{ligmann2013spatially}
 
-In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures.
+In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution \replaced{aggregate model behaviours (e.g.
+the final segregation level for Schelling)}{on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures}.
 We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.
@@
 The general method workflow of our method is illustrated in Figure \ref{fig:method}.
-In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with various values of its parameters, we introduced a spatial generator which depends on its own meta-parameters and feeds the model $\mu$ at initialisation (lower branch).
-The resulting configurations can be clustered into qualitative types of spatial patterns, so that the sensitivity analysis relates to the model's parameters, how the density spatial distribution was generated and the pattern of density in order to produce qualitative conclusion about the model and its behaviour.
-In particular, we want to emphasize that spatial effects only derive not only from grid size or shape effects, but also from the heterogeneity of relevant geographic entities (people, housing, networks, etc.) in geographical space.
-In the models we produced, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
+In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with \replaced{different}{various} values of its parameters, we introduce a spatial generator which depends \deleted{on its} own meta-parameters and feeds the model $\mu$ at initialisation (lower branch).
+The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.\deleted{so that} The sensitivity analysis relates to the model's parameters \added{to} how the density spatial distribution was generated and the pattern of density \replaced{generated in order to produce qualitative knowledge from}{in order to produce qualitative conclusion about} the model and its behaviour.
+In particular, we want to emphasize that spatial effects \deleted{only} derive not only from grid size or shape effects, but also from the heterogeneity of \replaced{the distribution of}{relevant} geographical entities (people, housing, networks, etc.) in \deleted{geographical} space.
+In the models we \replaced{used as examples}{produced}, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
 %%%%%%%%%%%%%%%%%%%%%%
-\begin{figure*}[htbp] \begin{center} 
+\begin{figure*}[htbp] 
+\begin{center} 
 \resizebox{0.9\textwidth}{!}{ 
 	\includegraphics{figures/SchemaMeta_1.png}
-} \caption{\textbf{General workflow of our method.}} \label{fig:method} \end{center} \end{figure*} %
+} \caption{\textbf{General workflow.}} \label{fig:method}
+\end{center}
+\end{figure*} %
 %%%%%%%%%%%%%%%%%%%%%%
 
-In order to test for the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}.
-We define a phase diagram as the vector of model outputs considered as a function of model parameters.
+In order to test the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}.
+\added{Following} \citet{Gauvinetal2009}, we define a phase diagram as the vector of \added{final aggregated} model outputs considered as a function of model parameters.
+We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment).
+A solution is to use systematic quantitative procedures to compare them.
 Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
 
-%Comparing phase diagrams is useful because it allows us to compare the 
-We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment).
-A solution is to use systematic quantitative procedures to compare them.
 To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.
 We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$.
-Phase diagrams are denoted by a common function $\mu$ indexed by what we call \emph{meta-parameters} $\vec{\alpha}$, that capture the spatial configuration (in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself).
+Phase diagrams are denoted by a common function $\mu$ indexed by what we call \emph{meta-parameters} $\vec{\alpha}$, that capture the spatial configuration \footnote{in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself}.
 The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by
@@
 $d$ is a functional distance that we take simply as Euclidian distance.
-The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$.
+The \replaced{internal variabilities}{variance} are estimated \added{as the variance} within each phase diagram $f_{\vec{\alpha_i}}$.
 For a multi-dimensional phase diagram, we sum these relative distances over the components.
@@
 In our case, this tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file.
-In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the renewed framework of multi-modelling \citep{cottineau2015modular}.
+In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the  framework of multi-modelling \citep{cottineau2015modular}.
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-The spatial generator applies an urban morphogenesis model \citep{Batty2007}, which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}.
-An open implementation and a characterisation of the urban forms which the model can produce are provided, which allows us to integrate it easily in our workflow.
-In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
-More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
+The spatial generator applies an urban morphogenesis model \citep{Batty2007} which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}.
+An open implementation and a characterisation of the urban forms which the model can produce \deleted{are provided, which} allow us to integrate it easily in our workflow.
+\deleted{In a nutshell, g}Grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
+More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to \added{a} patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
 At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (8 surrounding patches).
@@
 This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results.
-We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model, obtaining a total of 45 spatial configurations to compare (as we will see, the significant number of model runs needed to establish each phase diagram imposes a computational limit on this number of configurations).
+We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model \deleted{, obtaining a total of 45 spatial configurations to compare (as we will see, the significant number of model runs needed to establish each phase diagram imposes a computational limit on this number of configurations)}(cf.
+section \ref{sec:qualResults}).
 
@@
 Schelling's model consists in an abstract urban housing market where agents of different attributes (for example: red or green) sense their environment, evaluate their satisfaction in terms of neighbourhood composition (how many reds and greens?), and relocate if unsatisfied.
-It has been shown by \cite{Schelling1969} that even tolerant agents tended to produce segregated patterns because of the complexity of their local interactions.
-The main parameters of this model are the tolerance level (\% of agents similar to {\it ego}), the scope of sensing, the minority split and the percentage of vacant spaces in the housing market.
-In addition, we are interested in testing the impact of the spatial distribution of housing capacity in this project, using the generated grids.
-The outcome of the model is measured as a phase diagram of segregation level, described by the combination of several indicators: Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.
+It has been shown by \cite{Schelling1969} that even tolerant agents tend to produce segregated patterns because of the complexity of their local interactions \added{and the snowball effect of individual moves on the global distribution of agents in the city}.
+The main parameters of this model are the tolerance level (\added{maximum} \% of agents different to {\it ego} \added{accepted in the neighbourhood}), the scope of sensing, the minority split and the percentage of vacant spaces in the housing market.
+In addition, we are interested in testing the impact of the \added{initial} spatial distribution of housing capacity in this project, using the generated grids.
+The outcome of the model is measured \replaced{by a combination of three segregation indices:}{as a phase diagram of segregation level, described by the combination of several indicators:} Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.
+We translated the Netlogo implementation of Schelling into Scala for this project.
 
 Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \citep{EpsteinAxtell1996}.
-Although it {\it "is designed to study the interaction of many plausible social mechanisms"} \citep[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where {\it "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its toroidal array of cells"}.
+Although it {\it "is designed to study the interaction of many plausible social mechanisms"} \citep[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where {\it "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its \replaced{[...]}{toroidal} array of cells"}.
 Agents of different vision scopes and different metabolisms harvest a self-regenerating resource available heterogeneously in the initial landscape, they settle and collect this resource, which leads some of them to survive and others to perish.
@@
 We extend the implementation with agents wealth distribution of~\citet{li2009netlogo}.
-The outcome of the model is measured as a phase diagram of an index of inequality for resource distribution (Gini index).
+The outcome of the model is measured \replaced{by a Gini}{ as a phase diagram of an} index of inequality for resource distribution.\deleted{(Gini index)}.
 
@@
 We first try the same experiment design (50 density grids generated on the fly), then look at clustered grids representing urban densities.
-We choose 45 different grids among the ones which are most representative of the three types of urban morphology: 15 compact grids (like Vienna, cf figure \ref{fig:densityTypes}), 15 polycentric grids (like Liege) and 15 discontinuous grids (like Augsburg).
+We choose 45 different grids among the ones which are most representative of the three types of urban morphology: 15 compact grids \deleted{(like Vienna, cf figure 3)}, 15 polycentric grids \deleted{(like Liege)} and 15 discontinuous grids \deleted{(like Augsburg)}.
 The last experiment thus equates to 4,500,000 simulations (1000 parameter combinations x 45 density grids x 100 replications).
@@
 We choose different experiment designs, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices.
-In principle, our workflow applies with any manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.
+In principle, our workflow applies with \replaced{to}{any} manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.
 
@@
 
-We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default model setup (a bi-centric symmetric configuration) using the relative measure defined in equation~\ref{eq:phase-distance}.
-For each density grid, we then obtain the average squared distance between corresponding points of the phase diagrams, i.e.
-the median value of the final output measure, such as segregation or inequality, for a given value of parameters.
+We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default \replaced{initial spatial condition}{model} setup (a bi-centric symmetric\added{al non toroidal} configuration) using the \deleted{relative} measure defined in equation~\ref{eq:phase-distance}.
+For each density grid, we \deleted{then} obtain the average squared distance between corresponding points of the phase diagrams, i.e.
+the median value of the final output measure, such as segregation or inequality, for a given value of parameters \added{in the two setups (isotropic and generated)}.
 This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry).
@@
 Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be up to three times bigger than the variation due to the model parameters.
-We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in a morphological space.
+We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in \replaced{the meta-parameter}{a morphological} space.
 Each point represent one of the 50 different density grids used to initialise the distribution of sugar in the model.
@@
 Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e.
-with a small number of very high density cells) produce simulation results more different compared to the reference case than within this reference case because of parameter variations.
+with a small number of very high density cells) produce simulation results \replaced{that vary more between the reference case and the generated grid with the sa,e values of parameters than within the reference case because}{more different compared to the reference case than within this reference case because} of parameter variations.
 This pattern is emphasized when grids are generated with a high $\alpha$ and a high $\beta$ (i.e.
@@
 These grids have the highest relative distance to the reference case.
-On the contrary, with grids closer to the uniform pattern of the reference case (bottom left of the graph), the model parameters are more important in determining the simulation results.
+On the contrary, with grids closer to the uniform pattern of the reference case (bottom left of the graph), the model parameters are more important in determining the \replaced{final inequality levels than the initial spatial distribution of sugar}{simulation results}.
 
@@
 The first component defines a ``level of sprawl'' and of scattering, whereas the second one represents the level aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation (top left of figure \ref{fig:sugarscape-distance-pca}).
-It is confirmed by the behavior as a function of meta-parameters, as high values of $\alpha$ also yield high distance.
+It is confirmed by the behaviour as a function of meta-parameters, as high values of $\alpha$ also yield high distance.
 In terms of model processes, it shows that congestion mechanisms in the gathering of the resource induces fast increases of inequality.
-To put these results in perspective of our work-flow given in Fig.~\ref{fig:method}, we have a sensitivity to spatial parameters in average greater than the sensitivity to model parameters.
+To put these results in perspective of our workflow given in Fig.~\ref{fig:method}, we have a sensitivity to spatial parameters on average greater than the sensitivity to model parameters.
 
@@
 The objective of this quantitative experiment is to evaluate to which extent this phase diagram is modified when different density grids are applied.
-We show in Fig.\ref{fig:schelling-distance-meta} (Supplementary Material) the values of the relative distance as a function of meta-parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape before.
-Variations are less considerable than for Sugarscape across phase diagrams, but values close to 1 show that several configurations are as much sensitive to space than to their parameters.
-We will focus in the following in a qualitative characterisation of variations.
+We show in Fig.\ref{fig:schelling-distance-meta} (Supplementary Material) the values of the relative distance as a function of meta-parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape\deleted{ before}.
+Variations are less considerable than for Sugarscape across phase diagrams, but values close to 1 show that several configurations are as \deleted{much} sensitive to \replaced{initial spatial conditions}{space} than to their parameters.
+We \deleted{will} focus in the following \replaced{on}{in} a qualitative characterisation of \added{these} variations.
 
 \subsection{Qualitative variations}
+\label{sec:qualResults}
 
@@
 figure \ref{fig:densityTypes}).
-In particular, we want evaluate to which extent the typology summarises the spatial effects, and so if one type of urban form or the other enhances or mutes the segregation mechanism of the model, or interacts differently with the model parameters.
-This experiment aims at suggesting topical conclusions on urban morphology, beyond the technical conclusions already obtained with respect to simulation sensitivity.
+In particular, we want \added{to} evaluate to which extent the typology summarises the spatial effects, and \deleted{so} if one type of urban form or \replaced{another}{the other} enhances\deleted{ or mutes} the segregation mechanism of the model, or interacts differently with the model parameters.
+This experiment \replaced{attempts at drawing}{aims at suggesting topical} conclusions on urban morphology, beyond the technical conclusions already obtained with respect to simulation sensitivity.
 
@@
 \centering
-\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant at the 0.01 level.}
+\begin{threeparttable}
+\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.}}
 \label{tab:regressionSchelling}
-\begin{adjustwidth}{-1cm}{-1cm}
+%\begin{adjustwidth}{-1cm}{-1cm}
 \begin{tabular}{|m{2.5cm}|ll|ll|ll|}
@@
 \end{tabular}
-\end{adjustwidth}
-N.B. Moran's I applies to the minority population
+%\end{adjustwidth}
+\begin{tablenotes}
+ \item Moran's I applies to the minority population
+ \item *** means that the estimate is significant at the 0.01 level.
+\end{tablenotes}
+  \end{threeparttable}
 \end{table*}
 
-In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised determines to a certain extent the level of segregation measured at the end of the simulation run.
+In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised \replaced{correlates}{determines} to a certain extent \added{with} the level of segregation measured at the end of the simulation run.
 Indeed, compared to compact density patterns, polycentric grids produce more dissimilarity and entropy between the location of green and red agents.
@@
 The results obtained with Moran's I are opposite, because this index measures spatial autocorrelation at the global level and that compact cities have higher levels of global autocorrelation by construction.
-However, linear models with and without information about the type of density distribution yields the same coefficients for Schelling's parameters $V$ and $S$, the only exception being the vacancy rate $V$ in the Moran's I model with grid types, which becomes non-significant.
+However, linear models with and without \deleted{information about }the type of density distribution yield the same coefficients for Schelling's parameters $V$ and $S$, the only exception being the vacancy rate $V$ in the Moran's I model with grid types, which becomes non-significant.
 The similarity of the coefficient in both cases means that the effect of the model's parameters (and thus the mechanism by which agents of similar group cluster in space) is the same regardless of the distribution of density.
 The way polycentric and discontinuous density grid exhibit higher segregation is by allowing buffer zones of low density to surround pockets of homogeneity, which is impossible in a compact city, because everyone is at reach of everyone else.
-The buffering process confirms previous results obtained with network structures \citep{Banos2012}.
+The buffering process confirms previous results obtained with network structures \citep{Banos2012} \added{and supports the conclusion that space acts here on top of mechanisms rather than in interaction with them}.
 
@@
 In terms of grid shape, we observe that the difference between the two grids is mainly on average distance and entropy: in a nutshell, the first grid is much more dispersed and disorganised than the second.
-Although the behaviours are rather stable for varying $s_+$, the initial maximum endowment in sugar, which means that the poorest agents have a determinant role in trajectories, the two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values.
+Although the behaviours are rather stable for varying $s_+$, the initial maximum endowment in sugar, which means that the poorest agents have a determinant role in trajectories, the two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values (fig.
+\ref{fig:sugarscape-phasediagrams}.
 In sprawled spaces, inequalities are thus fostered by a lack of minimal local resources, whereas population will drive inequality in concentrated spaces.
@@
 
-We consider that the method presented holds great potential for strengthening geographical models' validation. However, two limits and two areas of opportunities have still not been tackled.
+We consider that the method presented \added{in this paper} holds great potential for strengthening geographical models' exploration.
+However, two limits and two areas of opportunities have still not been tackled.
 
@@
 
-\paragraph{Comparing phase diagrams.} Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point.
+\paragraph{Comparing phase diagrams} Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point.
 For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
@@
 
-\paragraph{Platform constraints and docking challenges.} An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
+\paragraph{Platform constraints and docking challenges} An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
 For example, the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements.
@@
 
-\paragraph{Reproducibility and Applicability.} 
+\paragraph{Reproducibility and Applicability} 
 %\comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
@@
 We applied this approach to two mythical ABMs: Schelling and Sugarscape.
-With the Schelling experiment, we found that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
+With the Schelling experiment, we found that the different urban morphologies impact the \deleted{parameter} interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
 With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space.
-We think that this method has the potential to increase the arsenal of evaluation/validation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
+We think that this method has the potential to increase the arsenal of evaluation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
 %%%%%%%%%%%%%

### 2018-06-13 paper.tex
.

@@
 
-% ********************************************
-% Review EPB :
-
-%Referee: 1
-
-%Comments to the Author
-
-%One sentence in the abstract of the paper, which is a part of the review invitation, was indeed confusing “However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment.” What can be that special in initial conditions?
-There are standard notions of equilibrium and, more general “attractor” of the dynamic system and there are “domains” of “basins” of attraction for each attractor that defines to which of the attractors the solution will converge in time (see http://www.scholarpedia.org/article/Basin_of_attraction).
-May “spatial” in “spatial conditions” make a difference?
-%I believe that after reading the paper I understand the way of author(s) thinking.
-Regrettably, their view of the dynamic system as well as their use of the basic notions and examples, are wrong.
-In what follows I will try to point to the major steps where the authors deviate, further and further away, from the proper view of the complex dynamic system theory.
-%The introduction reads very well while raises some general questions.
-E.g., cities’ “non-ergodicity” is a kind of a philosophical statements, some urban processes, as buildup area dynamics, can be indeed considered as path-dependent, while some, as urban traffic, are most probably, not.
-Whatever, introduction does not cause problems.
-%Then the methodical section comes, posing the first red flag.
-I mean the choice of the Schelling and Sugarscape modes as the examples.
-On the one hand, none of these models deals directly with the physical structure of space; even more, studies of the e.g.
-Schelling model on graphs show that its dynamics do depend on the physical structure of space.
-On the other, each of these models has many versions and, as dynamic spatial systems, these versions are very different.
-The dynamics of some versions of the Schelling model are essentially history-dependent, while the dynamics of the others are not.
-The objectives of the paper demand very careful recognition between the system with path-dependent and path-independent dynamics.
-%From this point on, the paper took a “believe me” path.
-The impression of the reader is that the authors are absolutely sure in their views and, thus, ignore numerous critical points that may cause reader’s doubts.
-Even definitions became problematic.
-“We define a phase diagram as the vector of model outputs considered as a function of model parameters.” Sorry, phase diagram is a standard notion and it is something else, see https://en.wikipedia.org/wiki/Phase_diagram.
-Further on, “Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values P as the median of the final values of an output indicator O obtained for the replications of the model initialized with P”.
-Well, the reader can guess now that the intention of the author is to investigate the dynamics of the aggregate characteristics of the output as dependent on the parameter(s) P of the system.
-Putting aside wrong use of “phase diagram,” this is manageable, while quite standard.
-However, the part of this statement raises deeper concerns: The dependence of the “median O” on P may work well in case when the stochastic system converges to a globally stable stochastic equilibrium.
-But what if it doesn’t?
-Let’s consider the simplest case of a stochastic system with two attractors, each being locally stable stochastic equilibrium.
-In this case, initial conditions that are close to the border between equilibrium’s basins would evidently result in very different values of O.
-To avoid a vicious circle, one cannot use median or any other aggregate characteristic before the attractors are properly recognized.
-That is, one cannot use median before system’s properties are known well enough.
-%Ok, let’s skip “median” and follow the paper based on the single solutions.
-Let us essentially reduce the class of systems to keep in mind too, assuming that (1) for every P, every attractor is just an equilibrium maybe, stochastic one; (2) variation in P can result in variation in the number of equilibriums and their location in a phase space.
-In this case, conceptually, given K output characteristics, we can indeed recognize the basins of attraction based on the distance between solutions as far as they stabilize in time.
-Formula (1) is ok in this respect, while I would consider multidimensional O and standard measures, e.g.,   https://math.stackexchange.com/questions/1358633/malalanobis-distance-between-two-multivariate-gaussian-distributions.
-%From this point, let us continue with the Schelling model only.
-Sugarscape is more complex and less studied.
-In respect to the papers goals, the background property of the Schelling model was formulated in Vinkovic, Kirman, 2006, A physical analogue of the Schelling model, Proceedings of the National Academy of Sciences, 103(51), 19261–19265.
-Namely, they distinguish between models, which rules enable relocation to a better location only and the rules that enable relocation to a cell of the same utility.
-Models of the first subclass always stall, with the non-zero (and, often, essential) fraction of discontent agents who are unable to find a better location.
-Models of the second subclass converge to simple stochastic equilibrium patterns that can be segregated or integrated depending on parameters.
-There may be other classifications of the Schelling-like models into those, which solutions stall and those, which solutions are simpler.
-However, the basic problem of the reviewed paper is that the Galvin’s et al (2009) version of the Schelling model belongs to the first class – its solutions are strongly path-dependent and, indeed, can stall in a great variety of unstable steady states.
-That is, in mathematical terms, the number of attractors of the version of Schelling model that is investigated in the paper is very large and the attractors’ patterns are complicated.
-One can follow here Chris Langton’s ideas of 1990s or recall Game of Life or elementary CA that served to Steven Wolfram for his “New Kind of Science” https://www.wolframscience.com/.
-Whatever, we cannot just average outputs of the dynamic systems for arbitrarily/randomly chosen set of parameters.
-Just as we cannot study the dynamics of the elementary CA until we recognize the correspondence between the rules that govern them and the patterns they produce.
-Compare the 50, 75, 86, etc.
-rules and completely different dynamic systems they generate http://mathworld.wolfram.com/ElementaryCellularAutomaton.html.
-At this point, the major idea of the paper is destroyed by: (1) path-dependent version of the Schelling (and Sugarscape) model; (2) extremely wide spectrum of initial conditions and (3) aggregate view of solutions.
-The authors overlook the consequences of the (1) – (3) and this takes them too far away from the correct view of the complex systems.
-The result is totally wrong result section.
-The observations of this section all describe averages over extremely heterogeneous sets of solutions that, just as the average characteristics of all elementary CA, they do not represent any meaningful phenomena.
-%Minor comment: I have mentioned above some specific paper’s flaws. There are more.
-
-%Referee: 2
-
-%Comments to the Author
-
-%I totally agree with the authors that “space matters” in models and this is one issue that makes geographical models complicated compared to say economic models that don’t consider spatial relationships in them.
-To me this is one of the first papers that to show systemically how initial spatial conditions (and variations in such) impact model results.
-By taking two well know agent-based models (Schelling’s Segregation model and Sugarscape) the authors are able to show quantitatively and qualitatively how different spatial conditions impact the results.
-
-%However, I find it hard to follow your argument especially based on what you write: “We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled.” Is unclear, in the sense, for example for your first condition, if I have a geographically explicit model of a real world place using GIS, say within an agent-based model, is space not directly accounted for and the behavior of the agents are also explicitly spelled out?
-There are many examples of this, as you note / cite in the paper.
-With respect to the second condition “provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled” you write this is seldom met.
-In what sense?
-Maybe this might be the case for abstract models like the ones you have used here but how can you alter space in geographically explicit models using real data?
-Most geographically explicit models are focused on a specific area therefore how could one alter the spatial environments without applying it to another area which is often a difficult task and seldom done.
-
-%Moreover, it is unclear in the current version of the paper if you turned off the “toroidal default setting” in NetLogo when using your different spaces in the Schelling model.
-I write this because if you are using realistic spaces does it make sense to have agents say at the south of the city also communicating with agents (i.e.
-checking their neighbors) in the north of the city?
-Which would be the case the Schelling NetLogo model, which would not be the case in reality.
-
-% One of your key words you use “model validation” but in what sense?
-You don’t really come back to this in the paper, also you write about “the validity of generative models is uncertain until their results are proven robust and representative of ’real-world’ conditions.” What do you consider real world conditions to be?
-You seem to allude that you need to have the spatial generator to generate realistic (styles) city shapes for testing on the Schelling model but I am not convinced.
-
-%This seems to be superfluous on your main purpose of paper.
-Your paper could be much shorter if you just show how different spatial configurations impact the results rather than spending time talking about spatial patterns.
-Don’t get me wrong, I like how your spatial generator produces patterns seen within European cities and captures compact, polycentric, etc.
-urban forms under different parameter configurations but this appears to be moving away from the focus of the paper (at least as its portrayed in the introduction) specifically how initial spatial conditions impact the results.
-
-%I guess, I am wondering if the paper could not be shorter and simpler if this generator was removed?
-Or I would suggest you refocus the paper/ or better articulate in your introduction the objective, how different urban forms (initial conditions impact model results).
-You have clear objectives: “1/ to test the robustness of simulation results to small variations of meta-parameters and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric vs.
-polycentric for example) on the results of a given model.” But your introduction does not lead clearly to this.
-
-%Basically, I think the paper needs a more clearer focus, a better introduction which focus specifically on its objectives and clearer structure from background to results.
-
-%Response to reviewers (Juste) :
-%rev 1 :
-%This review is quite arrogant ad totally misses the point of the paper, trying to fit a inappropriate framework on our work.
-%- we do not study at all the attractors of the dynamic, but aggregated indicators to have a synthetic view of model behavior.
-The first remark is difficult to understand, and yes the spatial makes a difference (mostly in practice).
-%- of course we deviate from dynamic system theory because we are not AT ALL in this theoretical framework.
-It is strange to try to fit a given framework to something that is positioned in a totally different one.
-%- choice of the model : precisely it has not been shown systematically how model behavior depends on space, thus the choice.
-And these models fundamentally deal with space as it is in the core of agents dynamics.
-%- the term phase diagram is used following the Gauvin paper, and we assume this extension of the concept.
-%- the theoretical example of the saddle hypersurface makes no sense here in practice, as indicators we took precisely converge to narrow unimodal distributions (easily distinguishable when parameters vary)
-%- we know that the attractor patterns are complicated, that's exactly why we use simulation and study aggregated indicators.
-%- references on CA are not useful here (none of these models are CA and it would be difficult to fit within a CA fwk, by forcing the neighborhood to be all space).
-(looks like an argument from authority)
-%- yes we can average macro outputs, that the standard way to do in simulation and empirically it shows no problem here -> maybe add a paragraph to show the intra-run variablity compared to variation between parameters, and convergence rate etc.
-%- the conclusion are not "destroyed" as : (1) this is true for the model state, but we study aggregated outputs that are NOT path-dependent - did the reviewer entirely read the paper ?
-; (2) idem, we average on initial distributions ; (3) idem.
-There is no "correct view of complex systems", there are different approaches, and yes we are far from dynamical systems because we take an other approach.
-%- "do not represent meaningful phenomena" : seriously this reviewer never saw any work in simulation ; finding appropriate macro indicators for a model, that have a thematic sense (segregation in our case) is at the core of the study of simulation models.
-%- please give us the other flaws that's the aim of a review (and 6 month to do a 1 page out-of-scope review, we could have expected more).
-
-%rev 2 :
-%- on the alteration of a real space, that is out of the scope of the current paper, but yes we aim at producing synthetic geographical configurations that capture a typical spatial configuration -> maybe explicit that in the discussion
-%- on model validation -> yes we should precise the concept
-% - we need to be more precise on the schelling implementation/specification
-% - ok, we can be shorter on the generator (or put more explicitly that we provide it and the grids for other modelers to do similar analysis
-
-%Response to reviewers (Marion) :
-
-%rev1
-
-%May“spatial” in “spatial conditions” make a difference?
-%=> mouai, je ne suis pas non plus convaincue de la pertinence de cette remarque, à laquelle on répond partiellement dans la literature review.
-Idem que Juste, je pense qu’il s’agit d’un problème de cadrage théorique.
-En lien avec la dernière remarque du reviewer 2, je pense que l’on peut ajouter un court paragraphe sur le positionnement du papier en intro.
-
-% cities’ “non-ergodicity” isa kind of a philosophical statements, some urban processes, as buildup area dynamics, can be indeed considered as path-dependent, while some, as urban traffic, are most probably, not.
-% =>  On peut reprendre cette remarque pour illustrer la question de la non-ergodicité des villes ?
-
-%The objectives of the paper demand very careful recognition between the system with path-dependent and path-independent dynamics.
-% => Bon, ici comme on dit clairement que l’on est pas là pour discuter des comportements particuliers de ces deux modèles, je ne reprendrais pas les remarques ci-dessus.
-On peut les garder en mémoire si on nous repose la question lors de la 2 nd soumission du papier.
-
-%  Sorry, phase diagram is a standard notion and it is something else, see https://en.wikipedia.org/wiki/Phase_diagram.
-% => je préfère aussi citer Gauvin à Wikipedia ;)
-
-% the reader can guess now that the intention of the author is to investigate the dynamics of the aggregate characteristics of the output as dependent on the parameter(s) P of the system.
-Putting aside wrong use of “phase diagram,” this is manageable, while quite standard.
-% => « the reader can guess now that » : on doit le dire plus clairement et en amont
-
-% That is, one cannot use median before system’s properties are known well enough.
-% => Sa remarque n’est pas non plus sans justifications… je pense que l’on doit afficher plus clairement notre cadre d’analyse en introduction, pour que l’on ne nous reproche pas par la suite de ne pas avoir exploré entièrement le comportement du modèle avant de tirer des conclusions de l’analyse d’indicateurs agrégés.
-Mais on montre justement que certains espaces de paramètres sont plus ou moins sensibles aux variations de conditions spatiales initiales… je ne pense pas que la résolution analytique du modèle viendrait contredire une telle conclusion ?
-
-% CAs and number of attractors 
-% => Ok, on peut garder ce type d’éléments pour la partie discussion, en disant bien que le papier est un premier proof of concept et que notre but n’est pas d’étudier le comportement de ces deux modèles… je ne pense pas que l’on sur-interprétait nos résultats, mais seulement que l’on pointait les variations de sensibilité aux conditions initiales.
-Il me semble que le fait que cela soit dû à différents facteurs de variabilité du comportement du modèle ne pose pas trop de problèmes vu le niveau de généralité de l’analyse que l’on propose ?
-
-% The observations of this section all describe averages over extremely heterogeneous sets of solutions that, just as the average characteristics of all elementary CA, they do not represent any meaningful phenomena.
-% => idem que plus haut, avec en plus l’idée que on dit bien que la path-denpendency est justement une des raisons pour lesquelles la prise en compte de l’ensemble des conditions initiales dont l’espace est primordiale…
-
-% rev 2 :
-
-%Most geographically explicit models are focused on a specific area therefore how could one alter the spatial environments without applying it to another area which is often a difficult task and seldom done.
-% => Justement, un papier comme celui de Isabelle Thomas et al.
-montre que même avec des modèles spatially explicit comme il dit, on doit prendre en compte l’effet de la configuration spatiale choisie (résolution de l’information géographique, effets de bord, etc.), bref c’est pas parce que tu as des données spatialisées que tu évites le MAUP, loin de là.
-En plus, ces modèles dressent des conclusions qui sortent souvent du cadre de leur espace d’études, sur la relation entre deux phénomènes par ex.
-
-% Which would be the case the Schelling NetLogo model, which would not be the case in reality.
-% => Euh, on utilise un torus nous ???
-
-% model validation 
-% C’est marrant, je dis souvent à mes étudiants que lorsqu’on utilise des « » ça met l’accent sur une notion non définie, on s’est fait piéger à notre tour !
-La question est posée de savoir si l’on parle d’exploration du modèle ou de validation dans le papier ?
-La première solution nous éviterait de tomber dans un débat épistémologique sur les conditions nécessaires à la validation d’un modèle, et l’existence même d’un tel état de validation…
-
-% spatial generator and focus
-% Je suis d’accord avec cette partie, après il me semble que le générateur de formes urbaines est un vrai plus du papier et donc je le garderais à moins que d’autres reviewers nous fassent le même commentaire.
-
-% **************  ABSTRACT  **************
-
 \begin{abstract}
@@
 %\setlength{\parindent}{0pt}
-Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain \replaced{unless}{until} their results are proven robust and representative of 'real-world' conditions.
+Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions.
 Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes.
 However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment.
-In this paper, we present a method to assess the effect of \added{some} initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create \deleted{the} density grids with which spatial simulation models are initialised.
-We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of \replaced{unequal societies}{resource harvesting}) and a straightforward open-source work-flow using high performance computing, that the effect of \replaced{initial spatial arrangements}{space} is significant in the two \replaced{models}{types of simulation}.
+In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create density grids with which spatial simulation models are initialised.
+We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant in the two models.
 Furthermore, this effect is sometimes larger than the effect of parameters' value change.
@@
 
-\todo{Response to reviewers
-\begin{itemize}
-\item For reviewer 1, explicitly mention the simulation framework at the beginning (and position it regarding others such as dynamical systems) with literature
-\item Similarly, clarify the aggregation step, how the indicators are macro, have a thematic meaning, and distribution converge clearly (possibly : add Sharpe ratios statistics and type of distrib for better fits)
-\item Rev 1 : (complementary to above) : positioning regarding some analytical works / cite some ?
-Precise that we do not aim at understanding fully model behavior, but finding some sensitivity to space
-\item Rev 2 : precise which Schelling we use.
-\item On real configs that can not be modified : remark by Marion on MAUP etc ; and add in discussion a layus on synthetic data resembling real configs etc.
-\item Remove or adapt the concept of validation to avoid the epistemological trap
-\item To be debated : do we keep the detailed description of the spatial generator ? (knowing that it is otherwise fully described and explored here : \url{https://arxiv.org/abs/1708.06743})
-\end{itemize}
-}
-
-\textbf{Example track changes : }\added{this is added text}, \deleted{removed} text or \replaced{new}{old} replaced text
+%\textbf{Example track changes : }\added{this is added text}, \deleted{removed} text or \replaced{new}{old} replaced text
 
@@
 
-\added{Computer} simulation \deleted{within computer models} has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}.
-%\comment{+ Seb \& James Doran in the 1970s??}
- Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of \added{path-dependent} social and ecological systems\deleted{ in space and time}.
+Computer simulation has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}.
+Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
 The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model \replaced{in}{.
 Indeed, they usually seek to} understand\replaced{ing}{ and model} the role of space \replaced{on}{in} social interactions and environmental processes, in a city for example.
-We think simulation can become a very good tool to achieve this, provided that models \deleted{do} include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is \replaced{represented}{modelled}.
+We think simulation can become a very good tool to achieve this, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is \replaced{represented}{modelled}.
 Unfortunately, the first condition is not always met, and the second is seldom even mentioned.
@@
 \replaced{Geographical systems}{An urban system} can be \replaced{crudely described as}{defined as a large number of diverse} social agents interacting with one another, within a limited portion of space.
-Social agents thus constitute the microscopic level of \replaced{the}{urban} system\deleted{s}, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects\added{, also known as path-dependent \citep{arthur1994increasing}.
+Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects\added{, also known as path-dependent \citep{arthur1994increasing}.
 Therefore, observing a system at different points in time does not equate to observing different system at a single point in time.
@@
 In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions.
-However, most geographical models, especially cellular automata, \added{still} represent cities as uniform grids of isotropic densities.
+However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic densities.
 Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}.
@@
 The way modellers \replaced{represent}{abstract} space is therefore a central element of geographical simulation models.
-\added{However, this step is rarely explicit and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one}.
-A meaningful way of \replaced{representing}{abstracting} space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures\added{ so as to estimate the variability of the model behaviour in plausible spatial arrangements}.
+However, this step is rarely explicit and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one.
+A meaningful way of \replaced{representing}{abstracting} space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour in plausible spatial arrangements.
 
@@
 
-In this paper, \replaced{we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape).
+In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape).
 In no way do we want to provide a full exploration of these two particular models, their attractors and/or potential policy implications.
-Instead, we present}{we suggest} a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation.
+Instead, we present a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation.
 The generator being controlled by meta-parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations.

### 2018-06-13 paper.tex
Update on Overleaf.

@@
 
-\author[a,b]{Juste Raimbault\corref{cor1}}
-\author[c,d]{Cl{\' e}mentine Cottineau}
-\author[e]{Marion Le Texier}
-\author[f]{Florent Le N{\' e}chet}
-\author[a]{Romain Reuillon}
+%\author[a,b]{Juste Raimbault\corref{cor1}}
+%\author[c,d]{Cl{\' e}mentine Cottineau}
+%\author[e]{Marion Le Texier}
+%\author[f]{Florent Le N{\' e}chet}
+%\author[a]{Romain Reuillon}
+\author{}
 
-\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
-\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
-\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
-\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
-\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
-\address[f]{Université Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
+%\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
+%\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
+%\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
+%\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
+%\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
+%\address[f]{Université Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
 %NB : mon affiliation "officielle" c'est : 
@@
 
-\cortext[cor1]{Corresponding author.}
+%\cortext[cor1]{Corresponding author.}
 
@@
 
-\email{juste.raimbault@polytechnique.edu}
+%\email{juste.raimbault@polytechnique.edu}
 

### 2018-06-13 highlights.tex
submission

\documentclass[11pt,a4paper]{article}        % possible options include font size ('10pt', '11pt' and '12pt'), paper size ('a4paper', 'letterpaper', 'a5paper', 'legalpaper', 'executivepaper' and 'landscape') and font family ('sans' and 'roman')

\usepackage[document]{ragged2e}
\usepackage[utf8]{inputenc}              
\begin{document}

\section*{Highlights}

\begin{itemize}
	\item We investigate the sensitivity of geosimulation models to initial spatial conditions.
	\item We generate stylized population density grids.
	\item We assess the sensitivity of two classical agent-based models, Schelling's model and Sugarscape's model.
	\item Strong quantitative and qualitative effects of spatial initial conditions are shown for these models.
\end{itemize}

\end{document}

%% end of file `template.tex'.

### 2018-08-08 paper.tex
submission
.

--- 2018-06-13 paper.tex
@@
-%% The '3p' and 'times' class options of elsarticle are used for Elsevier CRC
-%% The 'procedia' option causes ecrc to approximate to the Word template
-\documentclass[3p,times,procedia]{elsarticle}
-\flushbottom
-
-%% The `ecrc' package must be called to make the CRC functionality available
-\usepackage{ecrc}
-
+\documentclass[Royal,sageh,times]{sagej}
+% Afour
 \input{header.tex}
-
-%\runauth{Raimbault et al.}
-\runauth{}
 
@@
 
-\begin{frontmatter}
-
-\dochead{}
+%\runninghead{Raimbault \& al.}
+\runninghead{}
 
@@
 
-\author[a,b]{Juste Raimbault\corref{cor1}}
-\author[c,d]{Cl{\' e}mentine Cottineau}
-\author[e]{Marion Le Texier}
-\author[f]{Florent Le N{\' e}chet}
-\author[a]{Romain Reuillon}
+%\author{Juste Raimbault\affilnum{1,2}, 
+%Cl{\' e}mentine Cottineau\affilnum{3,4},
+%Marion Le Texier\affilnum{5}, 
+%Florent Le N{\' e}chet\affilnum{2}, 
+%Romain Reuillon\affilnum{1,6}
+%}
+\author{}
 
-\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
-\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
-\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
-\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
-\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
-\address[f]{Université Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
+%\affiliation{\affilnum{1}UMR 8504 G{\'e}ographie-cit{\'e}s, Paris, France\\
+%\affilnum{2}Universit{\'e} Paris-Est, Laboratoire Ville Mobilit{\'e} Transport, France\\
 %NB : mon affiliation "officielle" c'est : 
 %Le Néchet, F., Université Paris-Est, Laboratoire Ville Mobilité Transport, UPEMLV : 5 boulevard Copernic, Cité Descartes F 77454 Marne-la-Vallée cedex 2 France
-
-\cortext[cor1]{Corresponding author.}
+%\affilnum{3}Centre for Advanced Spatial Analysis, University College London, UK\\
+%\affilnum{4}CNRS, UMR 8097 Centre Maurice Halbwachs, Paris, France\\
+%\affilnum{5}UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France\\
+%\affilnum{6}Institut des Syst{\`e}mes Complexes Paris Ile-de-France, France}
 
@@
 %75006 Paris, France.
-%}
-%\corrauth{}
+%\email{juste.raimbault@polytechnique.edu}}
+\corrauth{}
 
@@
 
-%My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Computer, Environment and Urban Systems.
+%My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Environment and Planning B.
 
@@
 the spatial configuration of agents and their environment), and more precisely their spatial initial conditions.
-We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of CEUS, a journal dealing with computerised models and urban systems.
+We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of EPB, a journal dealing with urban modelling and city science.
 
 %Despite some preliminary results being presented as at ECTQG 2015, Bari, and at Geocomputation Conference 2017, Leeds, we confirm that neither the manuscript nor any parts of its content are currently under consideration or published in another journal.
-We did not have any previous interaction with CEUS regarding this manuscript.
+We did not have any previous interaction with EPB regarding this manuscript.
 We have no conflicts of interest to disclose.
@@
 
+% **************  ABSTRACT  **************
+
 \begin{abstract}
@@
 %\setlength{\parindent}{0pt}
-Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions.
+Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain until their results are proven robust and representative of 'real-world' conditions.
 Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes.
 However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment.
-In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create density grids with which spatial simulation models are initialised.
-We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant in the two models.
+In this paper, we present a method to assess the effect of initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create the density grids with which spatial simulation models are initialised.
+We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of resource harvesting) and a straightforward open-source work-flow using high performance computing, that the effect of space is significant in the two types of simulation.
 Furthermore, this effect is sometimes larger than the effect of parameters' value change.
@@
 
-\begin{keyword}
-Space \sep Initial conditions \sep Sensitivity \sep ABM \sep Validation
-\end{keyword}
+\keywords{Space, Initial conditions, Sensitivity, ABM, Validation}
 
-\end{frontmatter}
+\maketitle
 
-\email{juste.raimbault@polytechnique.edu}
-
-%\maketitle
-
-%% main text
-
-\enlargethispage{-15mm}
-
-%\textbf{Example track changes : }\added{this is added text}, \deleted{removed} text or \replaced{new}{old} replaced text
+% **************  MAIN BODY OF THE PAPER **************
+%\newpage
 
@@
 
-Computer simulation has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}.
-Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
-The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model \replaced{in}{.
-Indeed, they usually seek to} understand\replaced{ing}{ and model} the role of space \replaced{on}{in} social interactions and environmental processes, in a city for example.
-We think simulation can become a very good tool to achieve this, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is \replaced{represented}{modelled}.
+Simulation within computer models has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}.
+%\comment{+ Seb \& James Doran in the 1970s??}
+ Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of social and ecological systems in space and time.
+The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model.
+Indeed, they usually seek to understand and model the role of space in social interactions and environmental processes, in a city for example.
+We think simulation can become a very good tool to achieve this, provided that models do include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is modelled.
 Unfortunately, the first condition is not always met, and the second is seldom even mentioned.
@@
 
-\replaced{Geographical systems}{An urban system} can be \replaced{crudely described as}{defined as a large number of diverse} social agents interacting with one another, within a limited portion of space.
-Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects\added{, also known as path-dependent \citep{arthur1994increasing}.
-Therefore, observing a system at different points in time does not equate to observing different system at a single point in time.
-This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}}.
+An urban system can be defined as a large number of diverse social agents interacting with one another, within a limited portion of space.
+Social agents thus constitute the microscopic level of urban systems, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects.
+Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,saint1989villes, Portugali2000} and, in the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}.
+Complexity is partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.
+Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures.
+\\
+
+Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity~\citep{pumain2012urban}, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average.
 Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
 Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models...
-which are the main models to be found in geosimulation.
+which are the main models to be found in geosimulation!
+\\
+% détail de chez détail : j'enlèverais le ! les ... parlent d'eux-mêmes il me semble.
 
-Self-organization has been shown to be a central feature of \replaced{geographical}{urban} systems \added{in general and cities in particular} ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
-In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}.
-Complexity is partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.
-Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures.
-\added{In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as other initial conditions, but of more interest to the geographer}.
-\\
-
-\deleted{Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average.}
-
-Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) \deleted{in cities} is far from isotropic, even in the most sprawled cities.
+Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) in cities is far from isotropic, even in the most sprawled cities.
 On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities.
 In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions.
-However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic densities.
+However, most geographical models, especially cellular automata, represent cities as uniform grids of isotropic densities.
 Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}.
@@
 For example, households can have different preferences with respect to the built-environment they want to live in \citep{SpielmanHarrison2014}, or agents can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}.
-The way modellers \replaced{represent}{abstract} space is therefore a central element of geographical simulation models.
-However, this step is rarely explicit and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one.
-A meaningful way of \replaced{representing}{abstracting} space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour in plausible spatial arrangements.
+The way modellers abstract space is therefore a central element of geographical simulation models.
+A meaningful way of abstracting space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures.
 
@@
 
-In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape).
-In no way do we want to provide a full exploration of these two particular models, their attractors and/or potential policy implications.
-Instead, we present a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation.
+In this paper, we suggest a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation.
 The generator being controlled by meta-parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations.
@@
 
-Considerations of such issues in the geographical modelling literature are rather scarce.\replaced{Despite the opportunity of simulation in the representation of space and its potential observation biases,}{ although} there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions\replaced{, which}{.
-Mainly, they} have focused on the impact of:
+Considerations of such issues in the geographical modelling literature are rather scarce, although there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions.
+Mainly, they have focused on the impact of:
 \begin{itemize}
@@
 \end{itemize}
+We detail these approaches in the paragraphs below.
+Other approaches, not included in the scope of this article, include the effects of the topology of space on the processes (cf.
+\citet{moreno2009integrating} who explore the role of the topology of possible interactions in a Schelling model).
+Generally, sensitivity analyses focus either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
 
-\deleted{We detail these approaches in the paragraphs below.
-Other approaches, not included in the scope of this article, include the effects of the topology of space on the processes (cf.
-moreno who explore the role of the topology of possible interactions in a Schelling model).
-}
+\paragraph{Geo-localised input data accuracy:} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and scale of analysis.
+They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes.
+In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.
 
-\paragraph{Geo-localised input data accuracy} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and scale of analysis.
-They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes.
-In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and \added{the} computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.
-
-\paragraph{Spatial system shape, precision, and boundaries} \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e.
+\paragraph{Spatial system shape, precision, and boundaries:} \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e.
 made equivalent to) to Axelrod Culture Model.
 \citet{FlacheHegselmann2001} show that chances for random emergence of a stable cluster of similar agents in a Schelling-like model are higher in a rectangular grid and lower in a hexagonal grid and that an irregular (Voronoi-diagram) city lattice structure favours migration stabilisation around decentralised clusters of similar agents.
-\citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favours segregationist behaviours.
+\citet{Banos2012} compares the behaviour of Schelling segregation model on city lattices formalized as either grid, random, scale-free and Sierpinski networks and concludes that the presence of cliques in graph-based urban structures favor segregationist behaviours.
 \citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins.
@@
 
-\paragraph{Spatial heterogeneity} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
+\paragraph{Spatial heterogeneity:} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
 They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns.
@@
 %\paragraph{Combination:}
- Generally, sensitivity analyses \replaced{have focused}{focus} either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
+
 %\cite{ligmann2013spatially}
 
-In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution \replaced{aggregate model behaviours (e.g.
-the final segregation level for Schelling)}{on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures}.
+In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures.
 We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.
@@
 The general method workflow of our method is illustrated in Figure \ref{fig:method}.
-In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with \replaced{different}{various} values of its parameters, we introduce a spatial generator which depends \deleted{on its} own meta-parameters and feeds the model $\mu$ at initialisation (lower branch).
-The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.\deleted{so that} The sensitivity analysis relates to the model's parameters \added{to} how the density spatial distribution was generated and the pattern of density \replaced{generated in order to produce qualitative knowledge from}{in order to produce qualitative conclusion about} the model and its behaviour.
-In particular, we want to emphasize that spatial effects \deleted{only} derive not only from grid size or shape effects, but also from the heterogeneity of \replaced{the distribution of}{relevant} geographical entities (people, housing, networks, etc.) in \deleted{geographical} space.
-In the models we \replaced{used as examples}{produced}, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
+In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with various values of its parameters, we introduced a spatial generator which depends on its own meta-parameters and feeds the model $\mu$ at initialisation (lower branch).
+The resulting configurations can be clustered into qualitative types of spatial patterns, so that the sensitivity analysis relates to the model's parameters, how the density spatial distribution was generated and the pattern of density in order to produce qualitative conclusion about the model and its behaviour.
+In particular, we want to emphasize that spatial effects only derive not only from grid size or shape effects, but also from the heterogeneity of relevant geographic entities (people, housing, networks, etc.) in geographical space.
+In the models we produced, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
 %%%%%%%%%%%%%%%%%%%%%%
-\begin{figure*}[htbp] 
-\begin{center} 
+\begin{figure*}[htbp] \begin{center} 
 \resizebox{0.9\textwidth}{!}{ 
 	\includegraphics{figures/SchemaMeta_1.png}
-} \caption{\textbf{General workflow.}} \label{fig:method}
-\end{center}
-\end{figure*} %
+} \caption{\textbf{General workflow of our method.}} \label{fig:method} \end{center} \end{figure*} %
 %%%%%%%%%%%%%%%%%%%%%%
 
-In order to test the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}.
-\added{Following} \citet{Gauvinetal2009}, we define a phase diagram as the vector of \added{final aggregated} model outputs considered as a function of model parameters.
+In order to test for the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}.
+We define a phase diagram as the vector of model outputs considered as a function of model parameters.
+Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
+
+%Comparing phase diagrams is useful because it allows us to compare the 
 We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment).
 A solution is to use systematic quantitative procedures to compare them.
-Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
-
 To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.
 We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$.
-Phase diagrams are denoted by a common function $\mu$ indexed by what we call \emph{meta-parameters} $\vec{\alpha}$, that capture the spatial configuration \footnote{in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself}.
+Phase diagrams are denoted by a common function $\mu$ indexed by what we call \emph{meta-parameters} $\vec{\alpha}$, that capture the spatial configuration (in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself).
 The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by
@@
 $d$ is a functional distance that we take simply as Euclidian distance.
-The \replaced{internal variabilities}{variance} are estimated \added{as the variance} within each phase diagram $f_{\vec{\alpha_i}}$.
+The variances are estimated within each phase diagram $f_{\vec{\alpha_i}}$.
 For a multi-dimensional phase diagram, we sum these relative distances over the components.
@@
 In our case, this tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file.
-In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the  framework of multi-modelling \citep{cottineau2015modular}.
-
+In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the renewed framework of multi-modelling \citep{cottineau2015modular}.
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-The spatial generator applies an urban morphogenesis model \citep{Batty2007} which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}.
-An open implementation and a characterisation of the urban forms which the model can produce \deleted{are provided, which} allow us to integrate it easily in our workflow.
-\deleted{In a nutshell, g}Grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
-More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to \added{a} patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
+The spatial generator applies an urban morphogenesis model \citep{Batty2007}, which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}.
+An open implementation and a characterisation of the urban forms which the model can produce are provided, which allows us to integrate it easily in our workflow.
+In a nutshell, grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
+More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
 At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (8 surrounding patches).
@@
 This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results.
-We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model \deleted{, obtaining a total of 45 spatial configurations to compare (as we will see, the significant number of model runs needed to establish each phase diagram imposes a computational limit on this number of configurations)}(cf.
-section \ref{sec:qualResults}).
+We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model, obtaining a total of 45 spatial configurations to compare (as we will see, the significant number of model runs needed to establish each phase diagram imposes a computational limit on this number of configurations).
 
@@
 Schelling's model consists in an abstract urban housing market where agents of different attributes (for example: red or green) sense their environment, evaluate their satisfaction in terms of neighbourhood composition (how many reds and greens?), and relocate if unsatisfied.
-It has been shown by \cite{Schelling1969} that even tolerant agents tend to produce segregated patterns because of the complexity of their local interactions \added{and the snowball effect of individual moves on the global distribution of agents in the city}.
-The main parameters of this model are the tolerance level (\added{maximum} \% of agents different to {\it ego} \added{accepted in the neighbourhood}), the scope of sensing, the minority split and the percentage of vacant spaces in the housing market.
-In addition, we are interested in testing the impact of the \added{initial} spatial distribution of housing capacity in this project, using the generated grids.
-The outcome of the model is measured \replaced{by a combination of three segregation indices:}{as a phase diagram of segregation level, described by the combination of several indicators:} Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.
-We translated the Netlogo implementation of Schelling into Scala for this project.
+It has been shown by \cite{Schelling1969} that even tolerant agents tended to produce segregated patterns because of the complexity of their local interactions.
+The main parameters of this model are the tolerance level (\% of agents similar to {\it ego}), the scope of sensing, the minority split and the percentage of vacant spaces in the housing market.
+In addition, we are interested in testing the impact of the spatial distribution of housing capacity in this project, using the generated grids.
+The outcome of the model is measured as a phase diagram of segregation level, described by the combination of several indicators: Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.
 
 Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \citep{EpsteinAxtell1996}.
-Although it {\it "is designed to study the interaction of many plausible social mechanisms"} \citep[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where {\it "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its \replaced{[...]}{toroidal} array of cells"}.
+Although it {\it "is designed to study the interaction of many plausible social mechanisms"} \citep[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where {\it "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its toroidal array of cells"}.
 Agents of different vision scopes and different metabolisms harvest a self-regenerating resource available heterogeneously in the initial landscape, they settle and collect this resource, which leads some of them to survive and others to perish.
@@
 We extend the implementation with agents wealth distribution of~\citet{li2009netlogo}.
-The outcome of the model is measured \replaced{by a Gini}{ as a phase diagram of an} index of inequality for resource distribution.\deleted{(Gini index)}.
+The outcome of the model is measured as a phase diagram of an index of inequality for resource distribution (Gini index).
 
@@
 We first try the same experiment design (50 density grids generated on the fly), then look at clustered grids representing urban densities.
-We choose 45 different grids among the ones which are most representative of the three types of urban morphology: 15 compact grids \deleted{(like Vienna, cf figure 3)}, 15 polycentric grids \deleted{(like Liege)} and 15 discontinuous grids \deleted{(like Augsburg)}.
+We choose 45 different grids among the ones which are most representative of the three types of urban morphology: 15 compact grids (like Vienna, cf figure \ref{fig:densityTypes}), 15 polycentric grids (like Liege) and 15 discontinuous grids (like Augsburg).
 The last experiment thus equates to 4,500,000 simulations (1000 parameter combinations x 45 density grids x 100 replications).
@@
 We choose different experiment designs, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices.
-In principle, our workflow applies with \replaced{to}{any} manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.
+In principle, our workflow applies with any manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.
 
@@
 
-We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default \replaced{initial spatial condition}{model} setup (a bi-centric symmetric\added{al non toroidal} configuration) using the \deleted{relative} measure defined in equation~\ref{eq:phase-distance}.
-For each density grid, we \deleted{then} obtain the average squared distance between corresponding points of the phase diagrams, i.e.
-the median value of the final output measure, such as segregation or inequality, for a given value of parameters \added{in the two setups (isotropic and generated)}.
+We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default model setup (a bi-centric symmetric configuration) using the relative measure defined in equation~\ref{eq:phase-distance}.
+For each density grid, we then obtain the average squared distance between corresponding points of the phase diagrams, i.e.
+the median value of the final output measure, such as segregation or inequality, for a given value of parameters.
 This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry).
@@
 Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be up to three times bigger than the variation due to the model parameters.
-We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in \replaced{the meta-parameter}{a morphological} space.
+We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in a morphological space.
 Each point represent one of the 50 different density grids used to initialise the distribution of sugar in the model.
@@
 Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e.
-with a small number of very high density cells) produce simulation results \replaced{that vary more between the reference case and the generated grid with the sa,e values of parameters than within the reference case because}{more different compared to the reference case than within this reference case because} of parameter variations.
+with a small number of very high density cells) produce simulation results more different compared to the reference case than within this reference case because of parameter variations.
 This pattern is emphasized when grids are generated with a high $\alpha$ and a high $\beta$ (i.e.
@@
 These grids have the highest relative distance to the reference case.
-On the contrary, with grids closer to the uniform pattern of the reference case (bottom left of the graph), the model parameters are more important in determining the \replaced{final inequality levels than the initial spatial distribution of sugar}{simulation results}.
+On the contrary, with grids closer to the uniform pattern of the reference case (bottom left of the graph), the model parameters are more important in determining the simulation results.
 
@@
 The first component defines a ``level of sprawl'' and of scattering, whereas the second one represents the level aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation (top left of figure \ref{fig:sugarscape-distance-pca}).
-It is confirmed by the behaviour as a function of meta-parameters, as high values of $\alpha$ also yield high distance.
+It is confirmed by the behavior as a function of meta-parameters, as high values of $\alpha$ also yield high distance.
 In terms of model processes, it shows that congestion mechanisms in the gathering of the resource induces fast increases of inequality.
-To put these results in perspective of our workflow given in Fig.~\ref{fig:method}, we have a sensitivity to spatial parameters on average greater than the sensitivity to model parameters.
+To put these results in perspective of our work-flow given in Fig.~\ref{fig:method}, we have a sensitivity to spatial parameters in average greater than the sensitivity to model parameters.
 
@@
 The objective of this quantitative experiment is to evaluate to which extent this phase diagram is modified when different density grids are applied.
-We show in Fig.\ref{fig:schelling-distance-meta} (Supplementary Material) the values of the relative distance as a function of meta-parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape\deleted{ before}.
-Variations are less considerable than for Sugarscape across phase diagrams, but values close to 1 show that several configurations are as \deleted{much} sensitive to \replaced{initial spatial conditions}{space} than to their parameters.
-We \deleted{will} focus in the following \replaced{on}{in} a qualitative characterisation of \added{these} variations.
+We show in Fig.\ref{fig:schelling-distance-meta} (Supplementary Material) the values of the relative distance as a function of meta-parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape before.
+Variations are less considerable than for Sugarscape across phase diagrams, but values close to 1 show that several configurations are as much sensitive to space than to their parameters.
+We will focus in the following in a qualitative characterisation of variations.
 
 \subsection{Qualitative variations}
-\label{sec:qualResults}
 
@@
 figure \ref{fig:densityTypes}).
-In particular, we want \added{to} evaluate to which extent the typology summarises the spatial effects, and \deleted{so} if one type of urban form or \replaced{another}{the other} enhances\deleted{ or mutes} the segregation mechanism of the model, or interacts differently with the model parameters.
-This experiment \replaced{attempts at drawing}{aims at suggesting topical} conclusions on urban morphology, beyond the technical conclusions already obtained with respect to simulation sensitivity.
+In particular, we want evaluate to which extent the typology summarises the spatial effects, and so if one type of urban form or the other enhances or mutes the segregation mechanism of the model, or interacts differently with the model parameters.
+This experiment aims at suggesting topical conclusions on urban morphology, beyond the technical conclusions already obtained with respect to simulation sensitivity.
 
@@
 \centering
-\begin{threeparttable}
-\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.}}
+\caption{\textbf{Regression of the segregation level of Schelling simulation with order parameters and type of city grid.} *** means that the estimate is significant at the 0.01 level.}
 \label{tab:regressionSchelling}
-%\begin{adjustwidth}{-1cm}{-1cm}
+\begin{adjustwidth}{-1cm}{-1cm}
 \begin{tabular}{|m{2.5cm}|ll|ll|ll|}
@@
 \end{tabular}
-%\end{adjustwidth}
-\begin{tablenotes}
- \item Moran's I applies to the minority population
- \item *** means that the estimate is significant at the 0.01 level.
-\end{tablenotes}
-  \end{threeparttable}
+\end{adjustwidth}
+N.B. Moran's I applies to the minority population
 \end{table*}
 
-In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised \replaced{correlates}{determines} to a certain extent \added{with} the level of segregation measured at the end of the simulation run.
+In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised determines to a certain extent the level of segregation measured at the end of the simulation run.
 Indeed, compared to compact density patterns, polycentric grids produce more dissimilarity and entropy between the location of green and red agents.
@@
 The results obtained with Moran's I are opposite, because this index measures spatial autocorrelation at the global level and that compact cities have higher levels of global autocorrelation by construction.
-However, linear models with and without \deleted{information about }the type of density distribution yield the same coefficients for Schelling's parameters $V$ and $S$, the only exception being the vacancy rate $V$ in the Moran's I model with grid types, which becomes non-significant.
+However, linear models with and without information about the type of density distribution yields the same coefficients for Schelling's parameters $V$ and $S$, the only exception being the vacancy rate $V$ in the Moran's I model with grid types, which becomes non-significant.
 The similarity of the coefficient in both cases means that the effect of the model's parameters (and thus the mechanism by which agents of similar group cluster in space) is the same regardless of the distribution of density.
 The way polycentric and discontinuous density grid exhibit higher segregation is by allowing buffer zones of low density to surround pockets of homogeneity, which is impossible in a compact city, because everyone is at reach of everyone else.
-The buffering process confirms previous results obtained with network structures \citep{Banos2012} \added{and supports the conclusion that space acts here on top of mechanisms rather than in interaction with them}.
+The buffering process confirms previous results obtained with network structures \citep{Banos2012}.
 
@@
 In terms of grid shape, we observe that the difference between the two grids is mainly on average distance and entropy: in a nutshell, the first grid is much more dispersed and disorganised than the second.
-Although the behaviours are rather stable for varying $s_+$, the initial maximum endowment in sugar, which means that the poorest agents have a determinant role in trajectories, the two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values (fig.
-\ref{fig:sugarscape-phasediagrams}.
+Although the behaviours are rather stable for varying $s_+$, the initial maximum endowment in sugar, which means that the poorest agents have a determinant role in trajectories, the two examples have not only a very distant baseline inequality (the ceiling of the first 0.35 is roughly the floor of the second 0.3), but their qualitative behavior is also radically opposite: the sprawled configuration gives inequalities decreasing as population decreases and decreasing as minimal wealth increases, whereas the concentrated one gives inequalities strongly increasing as population decreases and also decreasing with minimal weights but significantly only for large population values.
 In sprawled spaces, inequalities are thus fostered by a lack of minimal local resources, whereas population will drive inequality in concentrated spaces.
@@
 
-We consider that the method presented \added{in this paper} holds great potential for strengthening geographical models' exploration.
-However, two limits and two areas of opportunities have still not been tackled.
+We consider that the method presented holds great potential for strengthening geographical models' validation. However, two limits and two areas of opportunities have still not been tackled.
 
@@
 
-\paragraph{Comparing phase diagrams} Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point.
+\paragraph{Comparing phase diagrams.} Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point.
 For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
@@
 
-\paragraph{Platform constraints and docking challenges} An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
+\paragraph{Platform constraints and docking challenges.} An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
 For example, the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements.
@@
 
-\paragraph{Reproducibility and Applicability} 
+\paragraph{Reproducibility and Applicability.} 
 %\comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
@@
 We applied this approach to two mythical ABMs: Schelling and Sugarscape.
-With the Schelling experiment, we found that the different urban morphologies impact the \deleted{parameter} interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
+With the Schelling experiment, we found that the different urban morphologies impact the parameter interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
 With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space.
-We think that this method has the potential to increase the arsenal of evaluation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
+We think that this method has the potential to increase the arsenal of evaluation/validation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
 %%%%%%%%%%%%%
@@
 
-\bibliographystyle{elsarticle-harv}
+\bibliographystyle{SageH}
 \bibliography{spacematters}
@@
 
-\section*{Supplementary material}
+\begin{sm}
 
@@
 
+\end{sm}
+
 \end{document}

### 2018-08-08 coverletter.tex

\documentclass[11pt,a4paper,sans]{moderncv}        % possible options include font size ('10pt', '11pt' and '12pt'), paper size ('a4paper', 'letterpaper', 'a5paper', 'legalpaper', 'executivepaper' and 'landscape') and font family ('sans' and 'roman')

\usepackage[document]{ragged2e}
% pour justifier

% moderncv themes
\moderncvstyle{banking}                            % style options are 'casual' (default), 'classic', 'oldstyle' and 'banking'
\moderncvcolor{red}                                % color options 'blue' (default), 'orange', 'green', 'red', 'purple', 'grey' and 'black'
\renewcommand{\familydefault}{\rmdefault}         % to set the default font; use '\sfdefault' for the default sans serif font, '\rmdefault' for the default roman one, or any tex font name
%\nopagenumbers{}                                  % uncomment to suppress automatic page numbering for CVs longer than one page

% character encoding
\usepackage[utf8]{inputenc}                       % if you are not using xelatex ou lualatex, replace by the encoding you are using

%----------------------------------------------------------------------------------
%            content
%----------------------------------------------------------------------------------
\firstname{}
\lastname{}
\begin{document}
%-----       letter       ---------------------------------------------------------

% recipient data
\recipient{Editors CEUS}{}
\date{June 13, 2018}
\opening{Dear Editors,}
\closing{Yours faithfully,\\
Juste Raimbault and co-authors\\
UPS CNRS 3611 ISC-PIF
}
         % use an optional argument to use a string other than "Enclosure", or redefine \enclname
\makelettertitle

\justify
My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Computer, Environment and Urban Systems.

This article deals with the central yet poorly explored issue of the sensitivity of geosimulation models to their spatial context (i.e. the spatial configuration of agents and their environment), and more precisely their spatial initial conditions. We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of CEUS, a journal dealing with computerised models and urban systems.

Despite some preliminary results being presented as at ECTQG 2015, Bari, and at Geocomputation Conference 2017, Leeds, we confirm that neither the manuscript nor any parts of its content are currently under consideration or published in another journal. We did not have any previous interaction with CEUS regarding this manuscript. We have no conflicts of interest to disclose. We do not have any opposed reviewer.
\justify

%

\makeletterclosing

\end{document}

%% end of file `template.tex'.

### 2018-08-20 Response_to_Reviewers.tex
added_response_letter

%% start of file `template.tex'.
%% Copyright 2006-2013 Xavier Danaux (xdanaux@gmail.com).
%
% This work may be distributed and/or modified under the
% conditions of the LaTeX Project Public License version 1.3c,
% available at http://www.latex-project.org/lppl/.

\documentclass[11pt,a4paper,sans]{moderncv}        % possible options include font size ('10pt', '11pt' and '12pt'), paper size ('a4paper', 'letterpaper', 'a5paper', 'legalpaper', 'executivepaper' and 'landscape') and font family ('sans' and 'roman')

\usepackage[document]{ragged2e}
% pour justifier

% moderncv themes
\moderncvstyle{banking}                            % style options are 'casual' (default), 'classic', 'oldstyle' and 'banking'
\moderncvcolor{red}                                % color options 'blue' (default), 'orange', 'green', 'red', 'purple', 'grey' and 'black'
\renewcommand{\familydefault}{\rmdefault}         % to set the default font; use '\sfdefault' for the default sans serif font, '\rmdefault' for the default roman one, or any tex font name
%\nopagenumbers{}                                  % uncomment to suppress automatic page numbering for CVs longer than one page

% character encoding
\usepackage[utf8]{inputenc}                       % if you are not using xelatex ou lualatex, replace by the encoding you are using
%\usepackage{CJKutf8}                              % if you need to use CJK to typeset your resume in Chinese, Japanese or Korean

% adjust the page margins
%\usepackage[scale=0.9]{geometry}
%\setlength{\hintscolumnwidth}{3cm}                % if you want to change the width of the column with the dates
%\setlength{\makecvtitlenamewidth}{10cm}           % for the 'classic' style, if you want to force the width allocated to your name and avoid line breaks. be careful though, the length is normally calculated to avoid any overlap with your personal info; use this at your own typographical risks...

\usepackage{xparse}
\DeclareDocumentCommand{\comment}{m o o o o}
{%
    \textcolor{red}{#1}
    \IfValueT{#2}{\textcolor{blue}{#2}}
    \IfValueT{#3}{\textcolor{ForestGreen}{#3}}
    \IfValueT{#4}{\textcolor{red!50!blue}{#4}}
    \IfValueT{#5}{\textcolor{Aquamarine}{#5}}
}

% to show numerical labels in the bibliography (default is to show no labels); only useful if you make citations in your resume
%\makeatletter
%\renewcommand*{\bibliographyitemlabel}{\@biblabel{\arabic{enumiv}}}
%\makeatother
%\renewcommand*{\bibliographyitemlabel}{[\arabic{enumiv}]}% CONSIDER REPLACING THE ABOVE BY THIS

% bibliography with mutiple entries
%\usepackage{multibib}
%\newcites{book,misc}{{Books},{Others}}
%----------------------------------------------------------------------------------
%            content
%----------------------------------------------------------------------------------
%-----       letter       ---------------------------------------------------------

\firstname{}
\lastname{}
\begin{document}

% recipient data
\recipient{Editor CEUS}{}
\date{\today}
\opening{Dear Editor,}
\closing{Yours faithfully,\\
Juste Raimbault and co-authors%\\
%Université Paris 7 - UMR CNRS 8504 Géographie-cités
}
         % use an optional argument to use a string other than "Enclosure", or redefine \enclname

%\makelettertitle

%Dear Editor,

\justify

Thank you for the feedback on the draft paper ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models''. The suggestions and comments will undoubtedly be of great value to the paper. It was updated accordingly.

\medskip

% Thank you for submitting your manuscript to Computers, Environment and Urban Systems. I have completed the review of your manuscript and a summary is appended below. The reviewers recommend reconsideration of your paper following major revision. I invite you to resubmit your manuscript after addressing all reviewer comments.

% When revising your paper, pay special attention to the explicit presentation of every model that is investigated in the paper. Present each model as an algorithm/unambiguous pseudocode and supply the Netlogo code for downloading (all this can be in appendix). The Schelling model, for example, has several Netlogo implementation, and it is critically important to know which of them is investigated. The same is true for the Sugarscape model. See comments of the reviewer #1 in this respect.

% When resubmitting your manuscript, please carefully consider all issues mentioned in the reviewers' comments, outline every change made point by point, and provide suitable rebuttals for any comments not addressed.

First, the following general changes were done:

\begin{itemize}
	\item
\end{itemize}

\medskip

We provide now you point-by-point responses to referees comments.

\medskip

Concerning specifically the comments made by the first referee:

\begin{itemize}
	% 1. I think it would be better if the author would focus on a single model. This way they could describe the model and the analyses in more detail. As it stands, the models are treated as black boxes. I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
	\item

	% 2. It is very important to describe the models. This is especially relevant for the Schelling model, which has many variants; each has different dependencies on the initial pattern. In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
	\item

	% 3. I am not sure why the Sugarscape model has been chosen given that the paper seems to be focused on cities.
	\item

	% 4. Consider replacing “meta parameters” with something like “generator parameters” or “initial pattern parameters.”
	\item

	% 5. Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index
	\item

	% 6. Line 232: comparing phase diagrams is related to the topic of comparing patterns. It would make sense to mention the methods that are usually used to compare patterns
	\item

	% 7. Line 238: Please explain why this particular index was chosen and discuss its properties. Connect the discussion to other indices like L1, and Linf norms, etc. Would we get qualitatively different results from different indices? 
	\item

	% 8. Regarding the spatial generator: the authors should explain why this generator is relevant to these two models? 
	\item

	% 9. The authors should illustrate the contribution of each parameter of the generator to the generated initial pattern using a figure. Details of the generators should be included in an appendix
	\item

	% 10. Explain how the density grids are used in the Schelling model, do you allow more than one agent to be in the same cell? 
	\item

	% 11. Line 297: I am not sure why empirical density destitutions of cities are important in the context of this study. 
	\item

	% 12. Line 641: NetLogo supports grids, continuous space, and GIS elements. It does not restrict the user in any way.
	\item

	% 13. Line 301: I am not sure what the authors mean by “interesting grids”
	\item

	% 14. Line 671: The same comment regarding “mythical ABMs” 
	\item
\end{itemize}

-Reviewer 2

  -
This is an interesting article that I read with true interest. The paper fits the profile of this journal and can reach a wider audience.

I also have some concerns:

In the abstract the authors claim that initial spatial conditions are often taken for granted. I do not agree to this statement. Initial spatial condition can have a large impact on the results of a model, however I do not think that the initial spatial conditions are often taken for granted.

Page 2 line 88: the authors state that “Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density”, I do not think this is true. Many models generate a heterogeneous synthetic population that is based on local data (replicating the actual population as closely as possible) and in far more detail then what is done in this paper. I am thinking of work in the line of disease diffusion, crime etc.

Page 2 line 98: “However, this step is rarely explicit and authors tend to generalize the conclusion of the model from a single, spatial case study”. I do not understand the first part of the this statement. The steps taken are often very explicit. Due to the amount of time it takes to correctly represent all spatial data, including population densities and the variation of other population variables over space, many authors indeed restrict themselves to a single case study due to the time intensity of this task. Many agent-based models are also not valid for other areas.

The authors do not differentiate between data driven empirical models versus theory-driven models. I think a clear differentiation between the two groups and the positioning of the two case studies in the second group would make the paper more balanced.

In section 1.3 (Previous considerations on the effects of the spatial configuration in simulation models) authors provide a lot of evidence on previous work on the impact of initial spatial conditions on the output of models. This is contrary to the statement in the abstract that initial conditions are often taken for granted.

The most interesting part of the research to me, is not the way the density layers are generated but the use of the phase diagrams, defined as a vector of final aggregated model outputs. As these are vectors, they are not so different from time series and many techniques exist to compare vectors in this domain. I wonder if the Euclidean distance is the best measure to determine the difference between the phase vectors of different spatial inputs.

Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.

In section 3.1, page 8, line 404-405 the authors mention that the distance of the phase diagrams was measured with respect to the reference phase diagram computed on the default initial spatial conditions. Actually this came as a surprise to me. I had the impression that all phase diagrams would be evaluated against all other diagrams. Perhaps this should be explained earlier in the paper.

Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.

I appreciate the fact that the authors in section 4 clearly indicate the limitations of their work (and I share the limitations listed under “comparing phase diagrams”). Personally I do not think that releasing the density grids is an important opportunity. As mentioned before I think that the core of the work is in the comparison of the phase vectors. I am not so sure the second opportunity listed in 4.2 is very valid. I do not think that the CEUS journal targets social sciences exactly.

-Reviewer 3

  -
I found the research question posed by the author(s) interesting and I agree with the author(s) that many modelers don’t consider how spatial arrangements impact the outcomes of the model and tend to focus only on the parameters. The paper makes good use of the complexity and spatial literature in defining the problem and demonstrates how urban form impacts the results. However, while the paper is innovative, I found it really difficult to interpret and follow the results. I would suggest to the author to rewrite the results and methodology so that they flow better and also provide more context with respect to how to interpret the results.

On a side note, one of your key words you use “validation” but in what sense?  You don’t really come back to this in the paper, also you write about “the validity of generative models is uncertain until their results are proven robust and representative of ’real-world’ conditions.” In the abstract but this is it.  What do you consider real world conditions to be? You seem to allude that you need to have the spatial generator to generate realistic (styles) city shapes for testing on the Schelling model but I am not convinced.  After the abstract all notion of validation is missing.

Geosimulation is introduced but not defined or referenced.  This should be clarified as the term has several meanings etc.

In places the English along with the overall argument could be better articulated. For example, on page 1 you write: “We think simulation can become a very good tool to achieve this,”  Could this be rewritten or better argued. You are not the only one to believe this but you don’t show this in your paper.

Page 4: you write “Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time” This makes little sense. What it should be is “Generally, sensitivity analyses have focused either on the spatial context or the heterogeneity of space” or something like that. Basically you need the “or”.

You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields?

In the Schelling and sugarscape models how are the initial population of agents distributed? Is it random? I ask as this is unclear in the current version of the paper but it is vital if your purpose is to explore how the initial spatial configuration impacts the results of the model. Should the agents not be all located at the same location for each run to rule out the impact of initial placement of agents on the resulting patterns? This is something that is not addressed in the paper. I know for example, the Schelling model in NetLogo normally places the agents at random locations but in your reimplementation how is this done? The same goes for the sugarscape model in NetLogo. I believe a clearer description of the model initialization would be good and how its not the initial random placement of agents but the actual environment that impacts the outcomes. Maybe this was explained in the paper but I missed it.

In Section 1.2 talk about “(monocentric vs. polycentric for example)” but in the results you don’t discuss the monocentric versions explicitly using these terms.

Also I appreciate you are trying to show rather complex results but reading the results several times I found the presentation of them (both written and visual) really hard to follow and understand. I was wondering if maybe you could find a better way to show / discuss them. Maybe a  worked example of how each set of parameters leads to a point in say Figure 4. Also it is confusing why the results from both models are not presented using the same styles of figures. For example, for the sugar scape model you different styles of figures than for the segregation model.

-Reviewer 4

  -
Review of "Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models"

In sensitivity analysis of spatial models, stochastic effects and small parameter changes are normally applied. What about initial conditions? The authors test the effects of different arrangements (e.g. monocentric vs. polycentric density patterns) on agent-based Schelling segregation models and Sugarscape's unequal societies model.

Overall, I find this to be a very good paper. Very well written in general. Good review of the problem, clearly outlining (with references) the need for focus on initial conditions in geosimulation. The chief factors (data uncertainty, nature of spatial units (e.g. shape) and spatial heterogeneity) on initial spatial conditions are covered:

-how many of these do you test? I see heterogeneity addressed, but what about shape and uncertainty? You touch upon shape when talking about NetLogo's use of raster as opposed to vector in the Discussion - could you expand on this?

Rigorous experimental method described. In the matching of European metropolitan areas to your generated outputs, I wonder what you are missing by not using cities from other continents. i.e. by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).

Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?

You say your models are open source - do you give details of that source?

Small edits:

line 68: "...a different system"

figure 1: modify font size so that labels do not go to the edge of their respective objects

figure 2: can you add the parameters that led to each of the grids

figure 3: check spelling of "Augburg" (should be Augsburg)

line 350: "...consists of..."

line 455: "sa,e" = "same"?

figure 4 and figure 5 captions: make it explicit that we're seeing the Sugarscape results here.

line 515: "analyse" = "analysis"

line 593: "at reach" = "within reach"

line 671: would "classic" or "hypothetical" be better words (choice dependent on your meaning) than "mythical", which implies these are models of centuries ago and somehow out of reach..

\justify

\makeletterclosing

\end{document}

%% end of file `template.tex'.

### 2018-08-20 Response_to_Reviewers.tex
reviews

@@
 
--Reviewer 2
+Concerning the comments made by the second referee:
 
-  -
-This is an interesting article that I read with true interest. The paper fits the profile of this journal and can reach a wider audience.
+%This is an interesting article that I read with true interest. The paper fits the profile of this journal and can reach a wider audience.
 
-I also have some concerns:
+%I also have some concerns:
 
-In the abstract the authors claim that initial spatial conditions are often taken for granted.
+\begin{itemize}
+	% In the abstract the authors claim that initial spatial conditions are often taken for granted.
 I do not agree to this statement.
 Initial spatial condition can have a large impact on the results of a model, however I do not think that the initial spatial conditions are often taken for granted.
+	\item
 
-Page 2 line 88: the authors state that “Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density”, I do not think this is true.
+	% Page 2 line 88: the authors state that “Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density”, I do not think this is true.
 Many models generate a heterogeneous synthetic population that is based on local data (replicating the actual population as closely as possible) and in far more detail then what is done in this paper.
 I am thinking of work in the line of disease diffusion, crime etc.
+	\item
 
-Page 2 line 98: “However, this step is rarely explicit and authors tend to generalize the conclusion of the model from a single, spatial case study”.
+	% Page 2 line 98: “However, this step is rarely explicit and authors tend to generalize the conclusion of the model from a single, spatial case study”.
 I do not understand the first part of the this statement.
@@
 Many agent-based models are also not valid for other areas.
+	\item
 
-The authors do not differentiate between data driven empirical models versus theory-driven models.
+	% The authors do not differentiate between data driven empirical models versus theory-driven models.
 I think a clear differentiation between the two groups and the positioning of the two case studies in the second group would make the paper more balanced.
+	\item
 
-In section 1.3 (Previous considerations on the effects of the spatial configuration in simulation models) authors provide a lot of evidence on previous work on the impact of initial spatial conditions on the output of models.
+	% In section 1.3 (Previous considerations on the effects of the spatial configuration in simulation models) authors provide a lot of evidence on previous work on the impact of initial spatial conditions on the output of models.
 This is contrary to the statement in the abstract that initial conditions are often taken for granted.
+	\item
 
-The most interesting part of the research to me, is not the way the density layers are generated but the use of the phase diagrams, defined as a vector of final aggregated model outputs.
+	% The most interesting part of the research to me, is not the way the density layers are generated but the use of the phase diagrams, defined as a vector of final aggregated model outputs.
 As these are vectors, they are not so different from time series and many techniques exist to compare vectors in this domain.
 I wonder if the Euclidean distance is the best measure to determine the difference between the phase vectors of different spatial inputs.
+	% -> 
+	\item
 
-Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.
+	% Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.
+	\item
 
-In section 3.1, page 8, line 404-405 the authors mention that the distance of the phase diagrams was measured with respect to the reference phase diagram computed on the default initial spatial conditions.
+	% In section 3.1, page 8, line 404-405 the authors mention that the distance of the phase diagrams was measured with respect to the reference phase diagram computed on the default initial spatial conditions.
 Actually this came as a surprise to me.
@@
 Perhaps this should be explained earlier in the paper.
+	\item
 
-Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.
+	% Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.
+	\item
 
-I appreciate the fact that the authors in section 4 clearly indicate the limitations of their work (and I share the limitations listed under “comparing phase diagrams”).
+	% I appreciate the fact that the authors in section 4 clearly indicate the limitations of their work (and I share the limitations listed under “comparing phase diagrams”).
 Personally I do not think that releasing the density grids is an important opportunity.
@@
 
--Reviewer 3
+\end{itemize}
 
-  -
-I found the research question posed by the author(s) interesting and I agree with the author(s) that many modelers don’t consider how spatial arrangements impact the outcomes of the model and tend to focus only on the parameters.
+Concerning the comments made by the third referee:
+
+\begin{itemize}
+	% I found the research question posed by the author(s) interesting and I agree with the author(s) that many modelers don’t consider how spatial arrangements impact the outcomes of the model and tend to focus only on the parameters.
 The paper makes good use of the complexity and spatial literature in defining the problem and demonstrates how urban form impacts the results.
@@
 I would suggest to the author to rewrite the results and methodology so that they flow better and also provide more context with respect to how to interpret the results.
+	\item
 
-On a side note, one of your key words you use “validation” but in what sense?
+	% On a side note, one of your key words you use “validation” but in what sense?
 You don’t really come back to this in the paper, also you write about “the validity of generative models is uncertain until their results are proven robust and representative of ’real-world’ conditions.” In the abstract but this is it.
@@
 After the abstract all notion of validation is missing.
+	\item
 
-Geosimulation is introduced but not defined or referenced.  This should be clarified as the term has several meanings etc.
+	% Geosimulation is introduced but not defined or referenced.  This should be clarified as the term has several meanings etc.
+	\item
 
-In places the English along with the overall argument could be better articulated.
+	% In places the English along with the overall argument could be better articulated.
 For example, on page 1 you write: “We think simulation can become a very good tool to achieve this,”  Could this be rewritten or better argued.
 You are not the only one to believe this but you don’t show this in your paper.
+	\item
 
-Page 4: you write “Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time” This makes little sense.
+	% Page 4: you write “Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time” This makes little sense.
 What it should be is “Generally, sensitivity analyses have focused either on the spatial context or the heterogeneity of space” or something like that.
 Basically you need the “or”.
+	\item
 
-You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields?
+	% You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields?
+	\item
 
-In the Schelling and sugarscape models how are the initial population of agents distributed?
+	% In the Schelling and sugarscape models how are the initial population of agents distributed?
 Is it random?
@@
 Maybe this was explained in the paper but I missed it.
+	\item
 
-In Section 1.2 talk about “(monocentric vs. polycentric for example)” but in the results you don’t discuss the monocentric versions explicitly using these terms.
+	% In Section 1.2 talk about “(monocentric vs. polycentric for example)” but in the results you don’t discuss the monocentric versions explicitly using these terms.
+	\item
 
-Also I appreciate you are trying to show rather complex results but reading the results several times I found the presentation of them (both written and visual) really hard to follow and understand.
+	% Also I appreciate you are trying to show rather complex results but reading the results several times I found the presentation of them (both written and visual) really hard to follow and understand.
 I was wondering if maybe you could find a better way to show / discuss them.
@@
 For example, for the sugar scape model you different styles of figures than for the segregation model.
+	\item 
+\end{itemize}
 
--Reviewer 4
+Concerning the comments made by the fourth referee:
 
-  -
-Review of "Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models"
-
-In sensitivity analysis of spatial models, stochastic effects and small parameter changes are normally applied.
+\begin{itemize}
+	% In sensitivity analysis of spatial models, stochastic effects and small parameter changes are normally applied.
 What about initial conditions?
@@
 polycentric density patterns) on agent-based Schelling segregation models and Sugarscape's unequal societies model.
+	\item
 
-Overall, I find this to be a very good paper.
+	% Overall, I find this to be a very good paper.
 Very well written in general.
@@
 The chief factors (data uncertainty, nature of spatial units (e.g.
-shape) and spatial heterogeneity) on initial spatial conditions are covered:
-
--how many of these do you test?
+shape) and spatial heterogeneity) on initial spatial conditions are covered: -how many of these do you test?
 I see heterogeneity addressed, but what about shape and uncertainty?
 You touch upon shape when talking about NetLogo's use of raster as opposed to vector in the Discussion - could you expand on this?
+	\item
 
-Rigorous experimental method described.
+	% Rigorous experimental method described.
 In the matching of European metropolitan areas to your generated outputs, I wonder what you are missing by not using cities from other continents.
@@
 by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).
+	\item
 
-Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?
+	% Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?
+	\item
 
-You say your models are open source - do you give details of that source?
+	% You say your models are open source - do you give details of that source?  
+	\item
 
-Small edits:
-
-line 68: "...a different system"
-
-figure 1: modify font size so that labels do not go to the edge of their respective objects
-
-figure 2: can you add the parameters that led to each of the grids
-
-figure 3: check spelling of "Augburg" (should be Augsburg)
-
-line 350: "...consists of..."
-
-line 455: "sa,e" = "same"?
-
-figure 4 and figure 5 captions: make it explicit that we're seeing the Sugarscape results here.
-
-line 515: "analyse" = "analysis"
-
-line 593: "at reach" = "within reach"
-
-line 671: would "classic" or "hypothetical" be better words (choice dependent on your meaning) than "mythical", which implies these are models of centuries ago and somehow out of reach..
+	% Small edits:
+	% line 68: "...a different system"
+	% figure 1: modify font size so that labels do not go to the edge of their respective objects
+	% figure 2: can you add the parameters that led to each of the grids
+	% figure 3: check spelling of "Augburg" (should be Augsburg)
+	%line 350: "...consists of..."
+	% line 455: "sa,e" = "same"?
+	%figure 4 and figure 5 captions: make it explicit that we're seeing the Sugarscape results here.
+	%line 515: "analyse" = "analysis"
+	%line 593: "at reach" = "within reach"
+	%line 671: would "classic" or "hypothetical" be better words (choice dependent on your meaning) than "mythical", which implies these are models of centuries ago and somehow out of reach..
+	\item
+\end{itemize}
 

### 2018-08-20 paper.tex
Update on Overleaf.

--- 2018-06-13 paper.tex
@@
 \author[a]{Romain Reuillon}
+%\author{}
 
@@
 
+%%%%%%%
+%% Highlights
+%
+% - We investigate the sensitivity of geosimulation models to initial spatial conditions.
+% - We generate stylized population density grids.
+% - We assess the sensitivity of two classical agent-based models, Schelling's model and Sugarscape's model.
+% - Strong quantitative and qualitative effects of spatial initial conditions are shown for these models.
+
 %\maketitle
@@
 
-\replaced{Geographical systems}{An urban system} can be \replaced{crudely described as}{defined as a large number of diverse} social agents interacting with one another, within a limited portion of space.
-Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects\added{, also known as path-dependent \citep{arthur1994increasing}.
+Geographical systems can be crudely described as social agents interacting with one another, within a limited portion of space.
+Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependent \citep{arthur1994increasing}.
 Therefore, observing a system at different points in time does not equate to observing different system at a single point in time.
-This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}}.
+This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}.
 Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
@@
 
-Self-organization has been shown to be a central feature of \replaced{geographical}{urban} systems \added{in general and cities in particular} ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
+Self-organization has been shown to be a central feature of \replaced{geographical}{urban} systems in general and cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
 In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}.

### 2018-08-20 paper.tex

@@
 Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
-The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model \replaced{in}{.
-Indeed, they usually seek to} understand\replaced{ing}{ and model} the role of space \replaced{on}{in} social interactions and environmental processes, in a city for example.
-We think simulation can become a very good tool to achieve this, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is \replaced{represented}{modelled}.
+The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model in understanding the role of space on social interactions and environmental processes, in a city for example.
+We think simulation can become a very good tool to achieve this, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is represented.
 Unfortunately, the first condition is not always met, and the second is seldom even mentioned.
@@
 
-Self-organization has been shown to be a central feature of \replaced{geographical}{urban} systems in general and cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
+Self-organization has been shown to be a central feature of geographical systems in general and cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
 In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}.
@@
 Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures.
-\added{In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as other initial conditions, but of more interest to the geographer}.
+In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as other initial conditions, but of more interest to the geographer.
 \\
 
-\deleted{Another crucial aspect of spatio-temporal complex systems in general, and cities in particular, is their non-ergodicity, which means that cross-sectional samples in space (the different cities of a country at time t) are not equivalent to samples in time (the same city at time t, t+1, t+2...) to compute statistics such as average.}
-
-Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) \deleted{in cities} is far from isotropic, even in the most sprawled cities.
+Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in the most sprawled cities.
 On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities.
@@
 For example, households can have different preferences with respect to the built-environment they want to live in \citep{SpielmanHarrison2014}, or agents can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}.
-The way modellers \replaced{represent}{abstract} space is therefore a central element of geographical simulation models.
+The way modellers represent space is therefore a central element of geographical simulation models.
 However, this step is rarely explicit and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one.
-A meaningful way of \replaced{representing}{abstracting} space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour in plausible spatial arrangements.
+A meaningful way of representing space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour in plausible spatial arrangements.
 
@@
 
-%\comment[FL]{Je pense que cette partie est vraiment utile, et devrait être étoffée, en procédant peut être à une typologie plus systématique des "initial condition" qui nous intéressent.
-Il y a déjà une différence entre l'hétérogénéité de l'espace (densités réparties mono ou poly, etc) et les différentes façons de représenter l'espace (grilles carrées, hexagones).
-Par ailleurs aussi des discussions sur l'incertitude de l'information (est ce que par exemple les emplois downscalés au carreau INSEE par Pivano 2016 c'est vraiment fiable ou très ?) et la précision de la grille}
-
 In 'real-life', different spatial configurations of people in a city tend to be associated with different distributions of income, carbon emissions, educational outcomes, etc.
@@
 
-Considerations of such issues in the geographical modelling literature are rather scarce.\replaced{Despite the opportunity of simulation in the representation of space and its potential observation biases,}{ although} there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions\replaced{, which}{.
-Mainly, they} have focused on the impact of:
+Considerations of such issues in the geographical modelling literature are rather scarce.Despite the opportunity of simulation in the representation of space and its potential observation biases, there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions, which have focused on the impact of:
 \begin{itemize}
@@
 
-\deleted{We detail these approaches in the paragraphs below.
-Other approaches, not included in the scope of this article, include the effects of the topology of space on the processes (cf.
-moreno who explore the role of the topology of possible interactions in a Schelling model).
-}
-
 \paragraph{Geo-localised input data accuracy} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and scale of analysis.
 They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes.
-In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and \added{the} computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.
+In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and the computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.
 
@@
 
-%\comment[FL]{je dois compléter\\
-%\cite{horritt2001effects} identified the effects of grid cell size on the behaviour of a raster flood model
-%\cite{vazquez2002effect} ont exploré la meilleure granularité pour la précision d'un modèle d'écoulement hydrologie et le temps de calcul (à reprendre)}
-
 \paragraph{Spatial heterogeneity} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
@@
 
-%\paragraph{Combination:} 
- Generally, sensitivity analyses \replaced{have focused}{focus} either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
+ Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
 %\cite{ligmann2013spatially}
 
-In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution \replaced{aggregate model behaviours (e.g.
-the final segregation level for Schelling)}{on model behaviour, which has so far only been assessed through a small set of illustrative spatial structures}.
+In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution aggregate model behaviours (e.g.
+the final segregation level for Schelling).
 We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.
@@
 The general method workflow of our method is illustrated in Figure \ref{fig:method}.
-In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with \replaced{different}{various} values of its parameters, we introduce a spatial generator which depends \deleted{on its} own meta-parameters and feeds the model $\mu$ at initialisation (lower branch).
-The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.\deleted{so that} The sensitivity analysis relates to the model's parameters \added{to} how the density spatial distribution was generated and the pattern of density \replaced{generated in order to produce qualitative knowledge from}{in order to produce qualitative conclusion about} the model and its behaviour.
-In particular, we want to emphasize that spatial effects \deleted{only} derive not only from grid size or shape effects, but also from the heterogeneity of \replaced{the distribution of}{relevant} geographical entities (people, housing, networks, etc.) in \deleted{geographical} space.
-In the models we \replaced{used as examples}{produced}, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
+In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends own meta-parameters and feeds the model $\mu$ at initialisation (lower branch).
+The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.
+The sensitivity analysis relates to the model's parameters to how the density spatial distribution was generated and the pattern of density generated in order to produce qualitative knowledge from the model and its behaviour.
+In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space.
+In the models we used as examples, the initial spatial configurations  can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
 %%%%%%%%%%%%%%%%%%%%%%
@@
 In order to test the influence of initial spatial conditions on model outputs, we use a systematic method to compare \emph{phase diagrams}.
-\added{Following} \citet{Gauvinetal2009}, we define a phase diagram as the vector of \added{final aggregated} model outputs considered as a function of model parameters.
+Following \citet{Gauvinetal2009}, we define a phase diagram as the vector of final aggregated model outputs considered as a function of model parameters.
 We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment).
@@
 
-%\comment[FL]{je ne comprends pas ce choix : pourquoi faut il considérer que des classes ayant une forte cohérence interne seraient plus éloignées?
-ou alors la question ne se pose pas en ces termes...
-Par ailleurs est ce que la formule est bien homogène du point de vue des dimensions?
-En tous cas je pense qu'on peut se passer de la formule à ce stade, éventuellement à mettre dans l'affichage des résultats si on en a besoin.}[(JR) pour l'homogénéité oui justement on prend $d^2$ car c'est les variances, on obtient une mesure sans unité ; ok pour omettre la formule ; pour l'interprétation, on regarde la distance relative, pour des classes à une certaine distance, si elles sont concentrées, cette distance sera plus sigificative que si elles ont une grande variabilité et peuvent se recouper ; par exemple si elles ont mêmes variance $\sigma^2$, alors si $d \simeq \sqrt{2} \sigma$ alors les nuages de points vont en gros se recouper sur leur moitié (cas gaussien idéal), tandis que si $d\gg \sigma$, le max d'un diagramme sera très loin du min de l'autre - on compare à la fois la distance relative et la distance de forme.]
-
-%Several potential methods from other fields such as environmental science could be used, but we keep it simple and such methodological considerations are furthermore auxiliary to the main purpose of this paper.
-We propose therefore an intuitive measure corresponding to the share of between-diagrams variability relative to their internal variability.
-More formally, the distance is given by
-
 \begin{equation}\label{eq:phase-distance}
@@
 $d$ is a functional distance that we take simply as Euclidian distance.
-The \replaced{internal variabilities}{variance} are estimated \added{as the variance} within each phase diagram $f_{\vec{\alpha_i}}$.
+The internal variabilities are estimated as the variance within each phase diagram $f_{\vec{\alpha_i}}$.
 For a multi-dimensional phase diagram, we sum these relative distances over the components.
@@
 The spatial generator applies an urban morphogenesis model \citep{Batty2007} which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}.
-An open implementation and a characterisation of the urban forms which the model can produce \deleted{are provided, which} allow us to integrate it easily in our workflow.
-\deleted{In a nutshell, g}Grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
-More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to \added{a} patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
+An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily in our workflow.
+Grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
+More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
 At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (8 surrounding patches).
@@
 
-%%%%%%%%%%%%%%%%%%%%%%
-%\subsection{Including a (typical) variety of initial conditions in the sensitivity analysis}
-
 In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure: their concentration index, hierarchy index, centrality index and continuity index (cf.
@@
 This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results.
-We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model \deleted{, obtaining a total of 45 spatial configurations to compare (as we will see, the significant number of model runs needed to establish each phase diagram imposes a computational limit on this number of configurations)}(cf.
+We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model (cf.
 section \ref{sec:qualResults}).
@@
 Schelling's model consists in an abstract urban housing market where agents of different attributes (for example: red or green) sense their environment, evaluate their satisfaction in terms of neighbourhood composition (how many reds and greens?), and relocate if unsatisfied.
-It has been shown by \cite{Schelling1969} that even tolerant agents tend to produce segregated patterns because of the complexity of their local interactions \added{and the snowball effect of individual moves on the global distribution of agents in the city}.
-The main parameters of this model are the tolerance level (\added{maximum} \% of agents different to {\it ego} \added{accepted in the neighbourhood}), the scope of sensing, the minority split and the percentage of vacant spaces in the housing market.
-In addition, we are interested in testing the impact of the \added{initial} spatial distribution of housing capacity in this project, using the generated grids.
-The outcome of the model is measured \replaced{by a combination of three segregation indices:}{as a phase diagram of segregation level, described by the combination of several indicators:} Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.
+It has been shown by \cite{Schelling1969} that even tolerant agents tend to produce segregated patterns because of the complexity of their local interactions and the snowball effect of individual moves on the global distribution of agents in the city.
+The main parameters of this model are the tolerance level (maximum \% of agents different to {\it ego} accepted in the neighbourhood), the scope of sensing, the minority split and the percentage of vacant spaces in the housing market.
+In addition, we are interested in testing the impact of the initial spatial distribution of housing capacity in this project, using the generated grids.
+The outcome of the model is measured by a combination of three segregation indices: Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.
 We translated the Netlogo implementation of Schelling into Scala for this project.
@@
 Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \citep{EpsteinAxtell1996}.
-Although it {\it "is designed to study the interaction of many plausible social mechanisms"} \citep[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where {\it "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its \replaced{[...]}{toroidal} array of cells"}.
+Although it {\it "is designed to study the interaction of many plausible social mechanisms"} \citep[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where {\it "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its [...] array of cells"}.
 Agents of different vision scopes and different metabolisms harvest a self-regenerating resource available heterogeneously in the initial landscape, they settle and collect this resource, which leads some of them to survive and others to perish.
@@
 We extend the implementation with agents wealth distribution of~\citet{li2009netlogo}.
-The outcome of the model is measured \replaced{by a Gini}{ as a phase diagram of an} index of inequality for resource distribution.\deleted{(Gini index)}.
+The outcome of the model is measured by a Gini index of inequality for resource distribution.
 
@@
 We first try the same experiment design (50 density grids generated on the fly), then look at clustered grids representing urban densities.
-We choose 45 different grids among the ones which are most representative of the three types of urban morphology: 15 compact grids \deleted{(like Vienna, cf figure 3)}, 15 polycentric grids \deleted{(like Liege)} and 15 discontinuous grids \deleted{(like Augsburg)}.
+We choose 45 different grids among the ones which are most representative of the three types of urban morphology: 15 compact grids, 15 polycentric grids and 15 discontinuous grids.
 The last experiment thus equates to 4,500,000 simulations (1000 parameter combinations x 45 density grids x 100 replications).
@@
 We choose different experiment designs, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices.
-In principle, our workflow applies with \replaced{to}{any} manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.
+In principle, our workflow applies with to manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.
 
@@
 
-%\comment[FL]{il faudrait peut être davantage renforcer l'analyse a propos de figure 1 de si c'est plutôt les mu parameters ou les spatial parameters qui ont une grosse influence sur les résultats.}[(JR) assez explicite pour sugarscape, j'ai rajouté un peu tout de même.]
-
 \subsection{Quantitative variation across grids}
 
-We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default \replaced{initial spatial condition}{model} setup (a bi-centric symmetric\added{al non toroidal} configuration) using the \deleted{relative} measure defined in equation~\ref{eq:phase-distance}.
-For each density grid, we \deleted{then} obtain the average squared distance between corresponding points of the phase diagrams, i.e.
-the median value of the final output measure, such as segregation or inequality, for a given value of parameters \added{in the two setups (isotropic and generated)}.
+We measure the distance of the phase diagrams for all density grids with respect to the reference phase diagram computed on the default initial spatial condition setup (a bi-centric symmetrical non toroidal configuration) using the measure defined in equation~\ref{eq:phase-distance}.
+For each density grid, we obtain the average squared distance between corresponding points of the phase diagrams, i.e.
+the median value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated).
 This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry).
@@
 Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be up to three times bigger than the variation due to the model parameters.
-We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in \replaced{the meta-parameter}{a morphological} space.
+We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in the meta-parameter space.
 Each point represent one of the 50 different density grids used to initialise the distribution of sugar in the model.
@@
 Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e.
-with a small number of very high density cells) produce simulation results \replaced{that vary more between the reference case and the generated grid with the sa,e values of parameters than within the reference case because}{more different compared to the reference case than within this reference case because} of parameter variations.
+with a small number of very high density cells) produce simulation results that vary more between the reference case and the generated grid with the same values of parameters than within the reference case because of parameter variations.
 This pattern is emphasized when grids are generated with a high $\alpha$ and a high $\beta$ (i.e.
@@
 These grids have the highest relative distance to the reference case.
-On the contrary, with grids closer to the uniform pattern of the reference case (bottom left of the graph), the model parameters are more important in determining the \replaced{final inequality levels than the initial spatial distribution of sugar}{simulation results}.
+On the contrary, with grids closer to the uniform pattern of the reference case (bottom left of the graph), the model parameters are more important in determining the final inequality levels than the initial spatial distribution of sugar.
 
@@
 The objective of this quantitative experiment is to evaluate to which extent this phase diagram is modified when different density grids are applied.
-We show in Fig.\ref{fig:schelling-distance-meta} (Supplementary Material) the values of the relative distance as a function of meta-parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape\deleted{ before}.
-Variations are less considerable than for Sugarscape across phase diagrams, but values close to 1 show that several configurations are as \deleted{much} sensitive to \replaced{initial spatial conditions}{space} than to their parameters.
-We \deleted{will} focus in the following \replaced{on}{in} a qualitative characterisation of \added{these} variations.
+We show in Fig.\ref{fig:schelling-distance-meta} (Supplementary Material) the values of the relative distance as a function of meta-parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape.
+Variations are less considerable than for Sugarscape across phase diagrams, but values close to 1 show that several configurations are as sensitive to initial spatial conditions than to their parameters.
+We focus in the following on a qualitative characterisation of these variations.
 
@@
 
-\paragraph{Schelling} In this qualitative exploration of the effect of initial spatial conditions on the results of Schelling's model, we use the classification of grids into three morphological types (cf.
+\paragraph{Schelling}
+
+In this qualitative exploration of the effect of initial spatial conditions on the results of Schelling's model, we use the classification of grids into three morphological types (cf.
 figure \ref{fig:densityTypes}).
-In particular, we want \added{to} evaluate to which extent the typology summarises the spatial effects, and \deleted{so} if one type of urban form or \replaced{another}{the other} enhances\deleted{ or mutes} the segregation mechanism of the model, or interacts differently with the model parameters.
-This experiment \replaced{attempts at drawing}{aims at suggesting topical} conclusions on urban morphology, beyond the technical conclusions already obtained with respect to simulation sensitivity.
+In particular, we want to evaluate to which extent the typology summarises the spatial effects, and if one type of urban form or another enhances the segregation mechanism of the model, or interacts differently with the model parameters.
+This experiment attempts at drawing conclusions on urban morphology, beyond the technical conclusions already obtained with respect to simulation sensitivity.
 
@@
 
-In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised \replaced{correlates}{determines} to a certain extent \added{with} the level of segregation measured at the end of the simulation run.
+In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised correlates to a certain extent with the level of segregation measured at the end of the simulation run.
 Indeed, compared to compact density patterns, polycentric grids produce more dissimilarity and entropy between the location of green and red agents.
@@
 The results obtained with Moran's I are opposite, because this index measures spatial autocorrelation at the global level and that compact cities have higher levels of global autocorrelation by construction.
-However, linear models with and without \deleted{information about }the type of density distribution yield the same coefficients for Schelling's parameters $V$ and $S$, the only exception being the vacancy rate $V$ in the Moran's I model with grid types, which becomes non-significant.
+However, linear models with and without the type of density distribution yield the same coefficients for Schelling's parameters $V$ and $S$, the only exception being the vacancy rate $V$ in the Moran's I model with grid types, which becomes non-significant.
 The similarity of the coefficient in both cases means that the effect of the model's parameters (and thus the mechanism by which agents of similar group cluster in space) is the same regardless of the distribution of density.
 The way polycentric and discontinuous density grid exhibit higher segregation is by allowing buffer zones of low density to surround pockets of homogeneity, which is impossible in a compact city, because everyone is at reach of everyone else.
-The buffering process confirms previous results obtained with network structures \citep{Banos2012} \added{and supports the conclusion that space acts here on top of mechanisms rather than in interaction with them}.
+The buffering process confirms previous results obtained with network structures \citep{Banos2012} and supports the conclusion that space acts here on top of mechanisms rather than in interaction with them.
 
-\paragraph{Sugarscape} We now check the sensitivity in terms of qualitative behavior of phase diagrams.
+\paragraph{Sugarscape}
+
+We now check the sensitivity in terms of qualitative behavior of phase diagrams.
 We show the phase diagrams for two very opposite morphologies in terms of sprawling, but controlling for aggregation with the same $PC2$ value.
@@
 
-We consider that the method presented \added{in this paper} holds great potential for strengthening geographical models' exploration.
+We consider that the method presented in this paper holds great potential for strengthening geographical models' exploration.
 However, two limits and two areas of opportunities have still not been tackled.
@@
 We applied this approach to two mythical ABMs: Schelling and Sugarscape.
-With the Schelling experiment, we found that the different urban morphologies impact the \deleted{parameter} interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
+With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
 With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space.

### 2018-08-20 paper.tex
Update on Overleaf.

@@
 Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependent \citep{arthur1994increasing}.
-Therefore, observing a system at different points in time does not equate to observing different system at a single point in time.
+Therefore, observing a system at different points in time does not equate to observing \added{a} different system at a single point in time.
 This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}.
@@
 
+\todo{add code and data availability statement here}
+
 \subsection{Quantitative variation across grids}
@@
 
-\paragraph{Platform constraints and docking challenges} An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
+\added{\cite{kuhnert2005comparing} map comparison techniques}
+
+\added{\cite{kruskal1983overview} parallel with time-series comparison techniques (see reviewer comment - however more constrained as two dimensional indexing)}
+
+\paragraph{Platform constraints and docking challenges}
+
+An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
 For example, the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements.
@@
 
-% \comment[FL]{NetLogo qui influence la formalisation de l'espace dans les modèles alors que formalisation de ce dernier peut être plus poussée dans la description du modèle original.}[(JR) bonne idée, y'a pas mal d'exemples ou il faut pas mal contourner l'usage ``naturel'' de Netlogo pour s'en sortir - sur les représentations raster/vecteur c'est souvent un dilemne et ça peut induire des biais d'implémentation - je vais essayer de trouver un exemple simple (ceux que j'ai là trop usine à gaz, exemple Lutecia)]
-
 \subsection{Opportunities and extensions}
@@
 \paragraph{Reproducibility and Applicability}
-%\comment[CC]{Lancer des pistes sur comment travailler l'effet de la grille pour d'autres types de modélisation type modèles appliqués.}
 
@@
 \paragraph{An emancipation opportunity for social sciences.}
-
-%\comment[MLT]{Au-delà de l'outil NetLogo, parler de la formalisation de l'espace par les physiciens qui a influencé les modèles en géographie (cf: commentaire de Florent que je propose de bouger ici)?}
-%\comment[FLN]{opportunity for social science to emancipate from the very strong hypotheses of physicists that have become standards (homogeneity and isotropy of space), even though they are never valid in social systems.}
 
@@
 
+\subsection*{\added{Additional figures for the Schelling model}}
+
+\added{The Fig.~\ref{fig:schelling-distance-meta} gives the phase diagrams distances as a function of generator parameters and morphological components, similarly to the Sugarscape model in main text.
+In absolute, this version of the Schelling model seems less sensitive to density grids than the sugarscape model, as we do not obtain a high range of values here.}
+
 %%%%%%%%%%%%%
@@
 
+\subsection*{\added{Pseudo-code for models}}
+
+\added{We give below the pseudo-code for the implementations we used of both models.}
+
 \end{document}

### 2018-08-21 paper.tex
Update on Overleaf.

@@
 
-% **************  TITLE AND AUTHOR INFORMATION **************
-
 \begin{frontmatter}
@@
 
-\title{Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models
-%Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?
-}
+\title{Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models}
 
@@
 \address[f]{Université Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
-%NB : mon affiliation "officielle" c'est : 
-%Le Néchet, F., Université Paris-Est, Laboratoire Ville Mobilité Transport, UPEMLV : 5 boulevard Copernic, Cité Descartes F 77454 Marne-la-Vallée cedex 2 France
 
@@
 In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the  framework of multi-modelling \citep{cottineau2015modular}.
+\added{We use therefore OpenMOLE's functionalities for model embedding through workflow, design of experiments (parameter sampling) and high performance environment access.}
 
@@
 \begin{figure*}[htbp] \begin{center} 
-\resizebox{0.9\textwidth}{!}{ 
-	\includegraphics{figures/spatialGen.png}
-} \caption{Four examples of grids produced by the spatial generator.
-The lighter the red, the denser the area.
-Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$.} \label{fig:spatialGen} \end{center} \end{figure*} %
+	\includegraphics[width=\linewidth]{figures/spatialGen.png}
+ \caption{\textbf{Four examples of grids produced by the spatial generator.} The lighter the red, the denser the area.
+Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$.
+\added{Parameter values used here are for the first grid $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 971)$, the second $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 176)$, the third $(\alpha = 1.4,\beta=0.045, N = 102, n_d = 2, t_f = 618)$, and the fourth $(\alpha = 1.8,\beta=0.114, N = 108, n_d = 1, t_f = 227)$.}}
+\label{fig:spatialGen}
+\end{center}
+\end{figure*} %
 %%%%%%%%%%%%%%%%%%%%%%
+
+% Generator parameters values:
+%  1) ex_sp-diffusion=0.0060_sp-growth-rate=25_sp-diffusion-steps=1_sp-alpha-localization=0.4_ticks=971_sp-population=24275.000000000102
+%  2) ex_sp-diffusion=0.0060_sp-growth-rate=25_sp-diffusion-steps=1_sp-alpha-localization=0.4_ticks=176_sp-population=4400.000000000003
+%  3) ex_sp-max-pop=80260_sp-diffusion=0.045_sp-growth-rate=102_sp-diffusion-steps=2_sp-alpha-localization=1.4_ticks=618
+%  4) ex_sp-max-pop=382170_sp-diffusion=0.114_sp-growth-rate=108_sp-diffusion-steps=1_sp-alpha-localization=1.8_ticks=227
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}\\
-%\includegraphics[width=0.4\textwidth]{figures/alpha_beta_schemas}(Bottom) Corresponding stylised density grids\\\comment[JR]{je sais pas si c'est une bonne idee de mettre les grilles stylisees comme ca parce que ca donne une fausse idee des sorties du modele de generation}}
+\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}
 \caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
@@
 We did so by implementing a spatial generator whose output is used as input for the simulation model.
-We applied this approach to two mythical ABMs: Schelling and Sugarscape.
+We applied this approach to two \replaced{well-known}{mythical} ABMs: Schelling and Sugarscape.
 With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
@@
 
-% acks commentés pour la soumission (anonymous)
 %\begin{acks}
@@
 \section*{Supplementary material}
+
+\subsection*{\added{Behavior of the density grid generator}}
+
+\added{We give in Fig.~\ref{fig:stylized-generator} plots roughly summarizing the behavior of the density grid generator, according to parameters $(\alpha,\beta)$.}
+
+%%%%%%%%%%%%%%
+\begin{figure}
+	\includegraphics[width=0.49\linewidth]{figures/alpha_beta_examples.png}
+    \includegraphics[width=0.49\linewidth]{figures/alpha_beta_schemas.png}
+    \caption{\added{\textbf{Stylized behavior of the density grid generator.}}\label{fig:stylized-generator}}
+\end{figure}
+%%%%%%%%%%%%%%
 
@@
 
+% Minimal netlogo code for the sugarscape model
+\begin{lstlisting}
+	to main-loop
+    	regrow-patches
+        update-agetns
+    end
+
+    to regrow-patches
+    	ask patches [set psugar min (list max-psugar (psugar + 1))]
+    end
+
+    to update-agents
+    	ask turtles [
+    		let move-candidates (patch-set patch-here (patches at-points vision-points) with [not any? turtles-here])
+  let possible-winners move-candidates with-max [psugar]
+  if any? possible-winners [
+    ;; if there are any such patches move to one of the patches that is closest
+    move-to min-one-of possible-winners [distance myself]
+  ]
+
+            ; 
+            set sugar (sugar - metabolism + psugar)
+  			set psugar 0
+
+    		set age (age + 1)
+    		if sugar <= 0 or age > max-age [
+      			hatch 1 [ new-turtle ]
+      			die
+    		]
+  		]
+    end
+
+\end{lstlisting}
+
 \end{document}

### 2018-09-03 Response_to_Reviewers.tex
.
Update on Overleaf.

@@
 As it stands, the models are treated as black boxes.
-I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
+	\item This is indeed a qualified comment.
+However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies than with a single model.
+We acknowledge the fact that this choice led us to shorten their presentations.
+Nevertheless, we have been taking heed to your remark to improve the description of the models by providing hyperlinks towards the NetLogo pages and Scala code lines (to be done).
+We believe this is sufficient as the models have not been adjusted from these textbook cases implementations.
+
+   % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
    \item
@@
 In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
-	\item
+	\item Thank you for this comment.
+The reason why we have decided to use the NetLogo implementations was again to help the readers to distance themselves from the case studies and focus on the genericity of the approach.
+This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms (in the NetLogo scenarios) and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs.
+This is again an argument in favour of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.
 
 	% 3. I am not sure why the Sugarscape model has been chosen given that the paper seems to be focused on cities.
-	\item
+	\item See above.
+Besides, we clarified our choice by adding the following sentence to the model description: "In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants."
 
 	% 4. Consider replacing “meta parameters” with something like “generator parameters” or “initial pattern parameters.”
-	\item
+	\item Thank you for the suggestion.
+In order to clarify the concept, we now propose a short definition after the first use of the term (section Methods): "In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
+We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process".
 
@@
 	% Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.
-	\item
+	\item The functionalities of OpenMOLE we used were detailed.
 
@@
 	% Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?
-	\item
+	\item \comment[JR]{no justification $\rightarrow$ new runs ?}
 
@@
 	% Small edits:
-	% line 68: "...a different system"
+	% X line 68: "...a different system"
 	% figure 1: modify font size so that labels do not go to the edge of their respective objects
+	% TODO Clem
 	% figure 2: can you add the parameters that led to each of the grids
+	% TODO Juste
 	% figure 3: check spelling of "Augburg" (should be Augsburg)
-	%line 350: "...consists of..."
-	% line 455: "sa,e" = "same"?
+	% TODO Clem
+	% line 350: "...consists of..."
+	% X line 455: "sa,e" = "same"? OK
 	%figure 4 and figure 5 captions: make it explicit that we're seeing the Sugarscape results here.
@@
 	%line 671: would "classic" or "hypothetical" be better words (choice dependent on your meaning) than "mythical", which implies these are models of centuries ago and somehow out of reach..
-	\item
+	\item Spelling errors and approximations in the vocabulary chosen were corrected.
 \end{itemize}

### 2018-09-03 paper.tex
Update on Overleaf.

@@
 Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependent \citep{arthur1994increasing}.
-Therefore, observing a system at different points in time does not equate to observing \added{a} different system at a single point in time.
+Therefore, observing a system at different points in time does not equate to observing \added{a} different systems at a single point in time.
 This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}.
@@
 The general method workflow of our method is illustrated in Figure \ref{fig:method}.
-In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends own meta-parameters and feeds the model $\mu$ at initialisation (lower branch).
+In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on \replaced{meta-parameters and feeds the model $\mu$ at initialisation (lower branch)}{its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
+We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process}.
 The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.
@@
 
-Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \citep{EpsteinAxtell1996}.
-Although it {\it "is designed to study the interaction of many plausible social mechanisms"} \citep[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where {\it "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its [...] array of cells"}.
+Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \cite{EpsteinAxtell1996}.
+Although it "is designed to study the interaction of many plausible social mechanisms" \cite[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its $[...]$ array of cells".
 Agents of different vision scopes and different metabolisms harvest a self-regenerating resource available heterogeneously in the initial landscape, they settle and collect this resource, which leads some of them to survive and others to perish.
 The main parameters of this model are the number of agents, their minimal and maximal resource levels.
-In addition, we are interested in testing the impact of the spatial distribution of the resource, using the generated grids.
-We extend the implementation with agents wealth distribution of~\citet{li2009netlogo}.
+\added{In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants.} We extend the implementation with agents wealth distribution of \cite{li2009netlogo}.
 The outcome of the model is measured by a Gini index of inequality for resource distribution.
+\replaced{We are interested in testing the impact of the spatial distribution of the resource, using the generated grids.}
 

### 2018-09-03 paper.tex
.
new_temoplate

@@
-%% The '3p' and 'times' class options of elsarticle are used for Elsevier CRC
-%% The 'procedia' option causes ecrc to approximate to the Word template
-\documentclass[3p,times,procedia]{elsarticle}
-\flushbottom
+\documentclass[preprint,5p,times,twocolumn,authoryear]{elsarticle}
 
-%% The `ecrc' package must be called to make the CRC functionality available
-\usepackage{ecrc}
+\usepackage{amssymb,amsmath,bbm}
+\usepackage{url}
 
-\input{header.tex}
+%%%%%
+% variable to include comments or not in the compilation ; set to 1 to include
+\def \draft {1}
 
-%\runauth{Raimbault et al.}
-\runauth{}
+\usepackage{xparse}
+\usepackage{ifthen}
+\DeclareDocumentCommand{\comment}{m o o o o}
+{\ifthenelse{\draft=1}{
+    \textcolor{red}{\textbf{C : }#1}
+    \IfValueT{#2}{\textcolor{blue}{\textbf{A1 : }#2}}
+    \IfValueT{#3}{\textcolor{ForestGreen}{\textbf{A2 : }#3}}
+    \IfValueT{#4}{\textcolor{red!50!blue}{\textbf{A3 : }#4}}
+    \IfValueT{#5}{\textcolor{Aquamarine}{\textbf{A4 : }#5}}
+ }{}
+}
+\newcommand{\todo}[1]{
+\ifthenelse{\draft=1}{\textcolor{red!50!blue}{\textbf{TODO : \textit{#1}}}}{}
+}
+
+\usepackage[flushleft]{threeparttable}
+\usepackage{multirow}
+
+\usepackage{listings}
+
+\usepackage{soul}
+\soulregister\cite7
+\soulregister\citep7
+\soulregister\ref7
+
+%\usepackage[final]{changes}
+\usepackage{changes}
+
+\setaddedmarkup{\textcolor{black}{\hl{#1}}}
+\setdeletedmarkup{\textcolor{red}{\sout{#1}}}
+
+\journal{Computers, Environment and Urban Systems}
 
@@
 \begin{frontmatter}
-
-\dochead{}
 
@@
 \address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
-\address[f]{Université Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
+\address[f]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
 
-\cortext[cor1]{Corresponding author.}
+\cortext[cor1]{Corresponding author: juste.raimbault@polytechnique.edu}
 
@@
 \end{frontmatter}
-
-\email{juste.raimbault@polytechnique.edu}
 
@@
 The general method workflow of our method is illustrated in Figure \ref{fig:method}.
-In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on \replaced{meta-parameters and feeds the model $\mu$ at initialisation (lower branch)}{its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
-We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process}.
+In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on \replaced{its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
+We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process}{meta-parameters and feeds the model $\mu$ at initialisation (lower branch)}.
 The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.
@@
 The outcome of the model is measured by a Gini index of inequality for resource distribution.
-\replaced{We are interested in testing the impact of the spatial distribution of the resource, using the generated grids.}
+\added{We are interested in testing the impact of the spatial distribution of the resource, using the generated grids.}
 
@@
 
-% \hfill \break
-% \itshape{This is a sub, subheading}\normalfont
-
-% \hfill\break
-
-%
-%
-%\begin{table}[htp]
-%
-%\begin{center}
-%\begin{tabular}{c c c c}
-%\arrayrulecolor{black}
-%\hline 
-%This & Is & A & Table\\
-%\arrayrulecolor{lightgray}
-%\hline 
-%\arrayrulecolor{black}
-%Label & 0.1 & 0.2 & 0.3\\
-%Label & 1.0 & 2.0 & 3.0\\
-%\hline
-%\end{tabular}
-%\end{center}
-%\label{first_table}
-%\caption{This is a table caption}
-%\end{table}%
-%
-%
-%
-%\begin{equation}
-%a^2 + b^2 = c^2
-%\tag*{Equation 1}
-%\end{equation}
-
 \section{Conclusion.}

### 2018-09-03 Response_to_Reviewers.tex
Update on Overleaf.

@@
 In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
-	\item Thank you for this comment.
+	\item \comment[MLT]{Thank you for this comment.
 The reason why we have decided to use the NetLogo implementations was again to help the readers to distance themselves from the case studies and focus on the genericity of the approach.
 This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms (in the NetLogo scenarios) and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs.
-This is again an argument in favour of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.
+This is again an argument in favour of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.}
 

### 2018-09-03 paper.tex
compiles
.

@@
 
-%\corrauth{Juste Raimbault, UMR CNRS 8504 G{\'e}ographie-cit{\'e}s
-%13 rue du Four,
-%75006 Paris, France.
-%}
-%\corrauth{}
-
 % authors bio
@@
 \begin{abstract}
-%\noindent
-%\setlength{\parindent}{0pt}
 Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions.
@@
 % - Strong quantitative and qualitative effects of spatial initial conditions are shown for these models.
-
-%\maketitle
-
-%% main text
-
-\enlargethispage{-15mm}
 
@@
 %%%%%%%%%%%%%
-\begin{figure}
+\begin{figure*}
 \centering
@@
 \label{fig:sugarscape-distance-meta}
-\end{figure}
+\end{figure*}
 %%%%%%%%%%%%%
@@
 \label{fig:sugarscape-distance-pca}
-\end{figure}
+\end{figure*}
 %%%%%%%%%%%%%
@@
 %\begin{adjustwidth}{-1cm}{-1cm}
-\begin{tabular}{|m{2.5cm}|ll|ll|ll|}
+\begin{tabular}{|p{2.5cm}|ll|ll|ll|}
 \hline
@@
 \label{fig:sugarscape-phasediagrams}
-\end{figure}
+\end{figure*}
 %%%%%%%%%%%%%

### 2018-09-03 Response_to_Reviewers.tex
Update on Overleaf.

--- 2018-08-20 Response_to_Reviewers.tex
@@
 \moderncvcolor{red}                                % color options 'blue' (default), 'orange', 'green', 'red', 'purple', 'grey' and 'black'
-\renewcommand{\familydefault}{\rmdefault}         % to set the default font; use '\sfdefault' for the default sans serif font, '\rmdefault' for the default roman one, or any tex font name
-%\nopagenumbers{}                                  % uncomment to suppress automatic page numbering for CVs longer than one page
+\renewcommand{\familydefault}{\rmdefault} 
+%\nopagenumbers{} 
+\usepackage[utf8]{inputenc} 
+\usepackage[scale=0.9]{geometry}
 
-% character encoding
-\usepackage[utf8]{inputenc}                       % if you are not using xelatex ou lualatex, replace by the encoding you are using
-%\usepackage{CJKutf8}                              % if you need to use CJK to typeset your resume in Chinese, Japanese or Korean
-
-% adjust the page margins
-%\usepackage[scale=0.9]{geometry}
-%\setlength{\hintscolumnwidth}{3cm}                % if you want to change the width of the column with the dates
-%\setlength{\makecvtitlenamewidth}{10cm}           % for the 'classic' style, if you want to force the width allocated to your name and avoid line breaks.
-be careful though, the length is normally calculated to avoid any overlap with your personal info; use this at your own typographical risks...
-
+\def \draft {1}
 \usepackage{xparse}
-\DeclareDocumentCommand{\comment}{m o o o o}
-{%
-    \textcolor{red}{#1}
-    \IfValueT{#2}{\textcolor{blue}{#2}}
-    \IfValueT{#3}{\textcolor{ForestGreen}{#3}}
-    \IfValueT{#4}{\textcolor{red!50!blue}{#4}}
-    \IfValueT{#5}{\textcolor{Aquamarine}{#5}}
+\usepackage{ifthen}
+\DeclareDocumentCommand{\comment}{o m o o o o}
+{\ifthenelse{\draft=1}{
+  \IfValueT{#1}{
+      \textcolor{red}{\textbf{C (#1) : }#2}
+      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
+      \IfValueT{#4}{\textcolor{ForestGreen}{\textbf{A2 : }#4}}
+      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
+      \IfValueT{#6}{\textcolor{Aquamarine}{\textbf{A4 : }#6}}
     }
-
-% to show numerical labels in the bibliography (default is to show no labels); only useful if you make citations in your resume
-%\makeatletter
-%\renewcommand*{\bibliographyitemlabel}{\@biblabel{\arabic{enumiv}}}
-%\makeatother
-%\renewcommand*{\bibliographyitemlabel}{[\arabic{enumiv}]}% CONSIDER REPLACING THE ABOVE BY THIS
-
-% bibliography with mutiple entries
-%\usepackage{multibib}
-%\newcites{book,misc}{{Books},{Others}}
-%----------------------------------------------------------------------------------
-%            content
-%----------------------------------------------------------------------------------
-%-----       letter       ---------------------------------------------------------
+    \IfNoValueT{#1}{
+      \textcolor{red}{\textbf{C : }#2}
+      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
+      \IfValueT{#4}{\textcolor{ForestGreen}{\textbf{A2 : }#4}}
+      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
+      \IfValueT{#6}{\textcolor{Aquamarine}{\textbf{A4 : }#6}}
+    }
+ }{}
+}
 
@@
 \begin{itemize}
-	\item
+	\item Several comments, including the ones of the first and third referee, were linked to clarity and explicit description of models: we clarified the descriptions and better justified our choices, and also provided pseudo-codes in Appendix and made clearer that source code are openly available.
+    \item The feedback on the third referee that the most interesting was the phase diagram comparison was not directly compatible with other referees comments; we however added some developments and comments on this point to conciliate as most as possible these different comments.
 \end{itemize}
@@
 As it stands, the models are treated as black boxes.
-I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
-	\item
+	\item The suggestion to focus on a single model is indeed a qualified comment.
+However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies than with a single model.
+In that perspective, the ``black-box'' is seen as an advantage as it suggests this genericity.
+We acknowledge the fact that this choice led us to shorten their presentations.
+Nevertheless, we have been taking heed to your remark to improve the description of the models by adding pseudo-codes in Appendix and recalling more clearly the link to model source we used.
+    %We believe this is sufficient as the models have not been adjusted from these textbook cases implementations.
+
+   % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
+   \item \comment[JR]{Possibility to find configuration with an ``eyeball'' difference ? i'm not sure of that..}
 
@@
 In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
-	\item
+	%\comment[MLT]{Thank you for this comment.
+The reason why we have decided to use the NetLogo implementations was again to help the readers to distance themselves from the case studies and focus on the genericity of the approach.
+This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms (in the NetLogo scenarios) and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs.
+This is again an argument in favour of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.}
+    \item There was a misunderstanding for the Schelling implementation actually used, as we used our own implementation which pseudo-code has been added in Appendix of the paper and source code is openly available on the git repository of the project.
+This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.
+It does not either reach a segregated state for high values of intolerance, but it does not cause a problem in our case as indicators can be computed on any model state (and actually reflect the fact that it is not segregated).
+    \item More generally concerning the choice of these two models, we aim at helping the readers to distance themselves from the case studies and focus on the genericity of the approach.
+This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs.
+This is again an argument in favour of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.
 
 	% 3. I am not sure why the Sugarscape model has been chosen given that the paper seems to be focused on cities.
-	\item
+	\item Furthermore concerning the choice of the Sugarscape model, we think it is a good example of a stylized model of a spatialized socio-economic system, and according to (Batty, 2005) a good metaphor of an urban system.
+Our choice was accordingly clarified in the paper.
+%: "In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants." % pas besoin de mettre les modifs ici
 
 	% 4. Consider replacing “meta parameters” with something like “generator parameters” or “initial pattern parameters.”
-	\item
+	%In order to clarify the concept of ``meta-parameter'' that we prefer to keep, we now propose a short definition after the first use of the term (section Methods): "In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
+We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process".
+    \item \comment[JR]{TB la def Marion ; là on peut choisir de virer en effet le terme, ou dire que les 3 sont interchangeable mais que "meta-" est plus pratique, qu'en pensez vous ?}
 
 	% 5. Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index
-	\item
+	\item \comment[JR]{je pensais qu'on prenait la moyenne ? je verifie...}
 
 	% 6. Line 232: comparing phase diagrams is related to the topic of comparing patterns. It would make sense to mention the methods that are usually used to compare patterns
-	\item
+	\item \comment[JR]{biblio a rajouter}
 
@@
 Would we get qualitatively different results from different indices?
-	\item
+	\item \comment[JR]{on avait teste avec le Monge ca donnait a peut pres pareil ; donc les normes alpha surement encore plus. je regarde.}
 
 	% 8. Regarding the spatial generator: the authors should explain why this generator is relevant to these two models? 
-	\item
+	\item This spatial generator is relevant to both model as it produces density grids at a ``metropolitan scale'' (see detailed discussion added in Appendix and the original paper from which we used the generator), which is the scale at which both model were initially intended to be.
+We made this aspect more explicit.
 
 	% 9. The authors should illustrate the contribution of each parameter of the generator to the generated initial pattern using a figure. Details of the generators should be included in an appendix
-	\item
+	\item The behavior of the spatial generator was described with more details in Appendix.
 
 	% 10. Explain how the density grids are used in the Schelling model, do you allow more than one agent to be in the same cell? 
-	\item
+	\item \comment[JR]{il a mal lu ou on est pas assez explicites, a detailler}
 
@@
 	% 12. Line 641: NetLogo supports grids, continuous space, and GIS elements. It does not restrict the user in any way.
-	\item
+	\item \comment[JR]{plutot d'accord avec lui mais parce que proficient en NetLogo, peut etre que des debutants sont plus contraints ? quelquun aurait une ref là dessus ?}
 
 	% 13. Line 301: I am not sure what the authors mean by “interesting grids”
-	\item
+	\item The term ``interesting grids'' was a typo and corrected, ``grids'' only was meant.
 
 	% 14. Line 671: The same comment regarding “mythical ABMs” 
-	\item
+	\item The term ``mythical'' line 671 was a language mistake and corrected, ``well-known'' ABM was meant.
 \end{itemize}
@@
 Initial spatial condition can have a large impact on the results of a model, however I do not think that the initial spatial conditions are often taken for granted.
-	\item
+	\item \comment[JR]{il s'agirait de nuancer ici}
 
@@
 I am thinking of work in the line of disease diffusion, crime etc.
-	\item
+	\item \comment[JR]{il a un point car ce sont des classes de modeles qu'on a pas trop cité ?
+par contre la sensibilite a ces distribs precises n'est pas testée, ou pas systeématiquement : il faut qu'on clarifie ou reformule}
 
@@
 Many agent-based models are also not valid for other areas.
-	\item
+	\item \comment[JR]{idem la on parle pas exactement du meme type de modeles, mais son exemple prouve notre propos : le modele tourne sur une situation unique, et on teste pas l'effet de l'espace.}
 
@@
 I think a clear differentiation between the two groups and the positioning of the two case studies in the second group would make the paper more balanced.
-	\item
+	\item \comment[JR]{oui du coup en disant ça ça repond à ses deux points précédents}
 
@@
 	% Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.
-	\item
+	\item The functionalities of OpenMOLE we used were detailed; we did not use genetic algorithms but sampling methods and high performance computing environments.
 
@@
 	% Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.
-	\item
+	\item \comment[JR]{Clem tu es d'accord là dessus ?}
 
@@
 	% You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields?
-	\item
+	\item \comment[JR]{cf remarques other reviewers sur ce point}
 
@@
 Maybe this was explained in the paper but I missed it.
-	\item
+	\item \comment[JR]{on regarde des indicateurs aggreges qui ne dependent pas de la trajectoire de chaque agent, si le placement initial est random il na aucune influence sur ceux-ci (cf les distribs) -> ajouter des graphes de comportement stat des modeles en Appendix ?}
 

### 2018-09-03 paper.tex
Update on Overleaf.

@@
 \usepackage{ifthen}
-\DeclareDocumentCommand{\comment}{m o o o o}
+\DeclareDocumentCommand{\comment}{o m o o o o}
 {\ifthenelse{\draft=1}{
-    \textcolor{red}{\textbf{C : }#1}
-    \IfValueT{#2}{\textcolor{blue}{\textbf{A1 : }#2}}
-    \IfValueT{#3}{\textcolor{ForestGreen}{\textbf{A2 : }#3}}
-    \IfValueT{#4}{\textcolor{red!50!blue}{\textbf{A3 : }#4}}
-    \IfValueT{#5}{\textcolor{Aquamarine}{\textbf{A4 : }#5}}
+  \IfValueT{#1}{
+      \textcolor{red}{\textbf{C (#1) : }#2}
+      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
+      \IfValueT{#4}{\textcolor{ForestGreen}{\textbf{A2 : }#4}}
+      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
+      \IfValueT{#6}{\textcolor{Aquamarine}{\textbf{A4 : }#6}}
+    }
+    \IfNoValueT{#1}{
+      \textcolor{red}{\textbf{C : }#2}
+      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
+      \IfValueT{#4}{\textcolor{ForestGreen}{\textbf{A2 : }#4}}
+      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
+      \IfValueT{#6}{\textcolor{Aquamarine}{\textbf{A4 : }#6}}
+    }
  }{}
 }
+
 \newcommand{\todo}[1]{
@@
 \usepackage{listings}
+\usepackage{mdframed}
 
@@
 \cortext[cor1]{Corresponding author: juste.raimbault@polytechnique.edu}
-
-% authors bio
-% Juste Raimbault is a PhD student at Université Paris 7, UMR CNRS 8504 Géographie-cités and UMR-T 9403 IFSTTAR LVMT, under the supervision of Arnaud Banos and Florent Le Néchet.
-He holds an engineer degree from Ecole Polytechnique and Ecole des Ponts et Chaussées, and a Master in Complex Systems Science from Ecole Polytechnique.
-His research focuses on complexity approaches to the relations between networks and territories.
-
-% Dr.
-Clémentine Cottineau is a research associate at the Centre for Advanced Spatial Analysis of University College London.
-She studied geography and economics at University Paris 1 Panthéon-Sorbonne and holds a PhD in geography from this university.
-Her thesis aimed at modelling urbanisation in the post-Soviet space with modular agent-based models.
-Since 2014, she has been working at the Centre for Advanced Spatial Analysis on urban economics and parametric scaling in the context of European systems of cities, using tools from complexity science to investigate inequality in cities and economic disparities between cities.
-
-% Dr.
-Marion Le Texier is associate professor at University of Rouen Normandie.
-She holds a PhD in geography from both Paris Diderot University and Luxembourg University.
-Her thesis aimed at modelling the circulation of euro coins through European borders with statistical and agent-based models.
-Previous to joining Rouen Normandie University in 2016, she has been a Jean Monet fellow at the European University Institute in Florence and a post-doc at Luxembourg University.
-Her work deals with data and modelling issues in different thematic contexts: access to nature in cities, international online media coverage of earthquakes, to name but two.
-
-% Dr.
-Florent Le Néchet is associate professor at  Paris-Est Marne-la-Vallée University (UPEM).
-He holds a PhD in planning from Paris-Est University (UPE).
-His thesis aimed at modelling the relationships between urban form and mobility patterns at metropolitan level, comparing several European cities, especially in France and Germany.
-His work deals with charactering and measuring mobility in cities, in relation to the built environment, and involves statistical techniques as well as multi-agent modelling.
-
-% Dr.
-Romain Reuillon is a research fellow at UMR CNRS 8504 Géographie-cités and at the Complex System Institute Paris Ile-de-France.
-He holds a PhD in Computer Science from Université Clermont II.
-His work deals with intensive computation model exploration, focusing on new techniques and methods to extract knowledge from models of simulation.
-He is the creator of the OpenMole platform for model exploration.
-
-% Cover Letter
-
-% Dear editors,
-
-%My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Computer, Environment and Urban Systems.
-
-%This article deals with the central yet poorly explored issue of the sensitivity of geosimulation models to their spatial context (i.e.
-the spatial configuration of agents and their environment), and more precisely their spatial initial conditions.
-We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of CEUS, a journal dealing with computerised models and urban systems.
-
-%Despite some preliminary results being presented as at ECTQG 2015, Bari, and at Geocomputation Conference 2017, Leeds, we confirm that neither the manuscript nor any parts of its content are currently under consideration or published in another journal.
-We did not have any previous interaction with CEUS regarding this manuscript.
-We have no conflicts of interest to disclose.
-We do not have any opposed reviewer.
-
-% Yours faithfully,
-% Juste Raimbault and co-authors.
 
@@
 The generator being controlled by meta-parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations.
-The purpose is two-fold: 1/ to test the robustness of simulation results to small variations of meta-parameters and 2/ to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
+The purpose is two-fold: (i) to test the robustness of simulation results to small variations of meta-parameters and (ii) to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
 Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies.
@@
 We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}.
-It yields a subset of 170 interesting grids matching empirical densities, which constituted our set of different initial spatial conditions.
+It yields a subset of 170 \deleted{interesting} grids matching empirical densities, which constituted our set of different initial spatial conditions.
 These are further clustered into three classes of morphology (figure \ref{fig:densityTypes}): compact (e.g.
@@
 section \ref{sec:qualResults}).
+
+\todo{explain why this generator is relevant for both models (discussion on scale ?)}
 
@@
 The outcome of the model is measured by a combination of three segregation indices: Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.
-We translated the Netlogo implementation of Schelling into Scala for this project.
+\replaced{We use a ad-hoc implementation of the Schelling model, both in Scala for performance reasons and in NetLogo to ensure visualization of model dynamics.
+The pseudo-code of the implemented model is available in Appendix~\ref{app:code} and source code are available on the repository of the project.}{We translated the Netlogo implementation of Schelling into Scala for this project.}
 
@@
 The main parameters of this model are the number of agents, their minimal and maximal resource levels.
-\added{In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants.} We extend the implementation with agents wealth distribution of \cite{li2009netlogo}.
+\added{In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants.
+Following \cite{batty2005agents}, it is a good metaphor of an urban system.} We extend the implementation with agents wealth distribution of \cite{li2009netlogo}.
 The outcome of the model is measured by a Gini index of inequality for resource distribution.
@@
 
-\newpage
+\appendix
 
-\section*{Supplementary material}
-
-\subsection*{\added{Behavior of the density grid generator}}
+\section{\added{Behavior of the density grid generator}}
 
 \added{We give in Fig.~\ref{fig:stylized-generator} plots roughly summarizing the behavior of the density grid generator, according to parameters $(\alpha,\beta)$.}
+
+\todo{bla bla on generator scale / real configs in the initial paper}
 
@@
 
-\subsection*{\added{Additional figures for the Schelling model}}
+\section{\added{Additional figures for the Schelling model}}
 
@@
 %%%%%%%%%%%%%
-\begin{figure}[h!]
+\begin{figure*}[h!]
 \centering
@@
 We show them in the meta-parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
-\end{figure}
+\end{figure*}
 %%%%%%%%%%%%%
 
-\subsection*{\added{Pseudo-code for models}}
+\section{\added{Pseudo-code for models}}
+\label{app:code}
 
@@
 
+\begin{figure*}
+%\begin{frame}
 % Minimal netlogo code for the sugarscape model
@@
 \end{lstlisting}
+%\end{frame}
+\end{figure*}
+
+%%%%%%%%%%%%%%%
+\begin{figure*}
+	\begin{lstlisting}
+
+    \end{lstlisting}
+\end{figure*}
+%%%%%%%%%%%%%%%
 

### 2018-09-03 Response_to_Reviewers.tex
Update on Overleaf.

@@
 In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
-	%\comment[MLT]{Thank you for this comment.
-The reason why we have decided to use the NetLogo implementations was again to help the readers to distance themselves from the case studies and focus on the genericity of the approach.
-This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms (in the NetLogo scenarios) and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs.
-This is again an argument in favour of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.}
 	\item There was a misunderstanding for the Schelling implementation actually used, as we used our own implementation which pseudo-code has been added in Appendix of the paper and source code is openly available on the git repository of the project.
@@
 	% 5. Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index
-	\item \comment[JR]{je pensais qu'on prenait la moyenne ? je verifie...}
+	\item \comment[JR]{je pensais qu'on prenait la moyenne ? je verifie... ligne 175 on écrit la médiane en effet}
 
@@
 	% 12. Line 641: NetLogo supports grids, continuous space, and GIS elements. It does not restrict the user in any way.
-	\item \comment[JR]{plutot d'accord avec lui mais parce que proficient en NetLogo, peut etre que des debutants sont plus contraints ? quelquun aurait une ref là dessus ?}
+    % \comment[JR]{plutot d'accord avec lui mais parce que proficient en NetLogo, peut etre que des debutants sont plus contraints ? quelquun aurait une ref là dessus ?}
+	\item We agree that NetLogo does not restrict the use on what implementation to use.
+However, some features are much more easier as raster or the contrary, what could influence the choice of beginners.
+We clarified that statement in the paper.
 
@@
 	\item The term ``mythical'' line 671 was a language mistake and corrected, ``well-known'' ABM was meant.
+    \comment[MLT]{Use "textbook ABM instead?}
 \end{itemize}
@@
 After the abstract all notion of validation is missing.
-	\item
+	\item \comment[JR]{enlever le concept de validation}
 
@@
 For example, for the sugar scape model you different styles of figures than for the segregation model.
-	\item 
+	\item Additional figures for the Schelling model presenting the results in a similar way were provided.
+Figures showing grid estimates of phase diagrams could not be realized for the Schelling model because of the different sampling of the parameter space used.
 \end{itemize}

### 2018-09-03 paper.tex
Update on Overleaf.

@@
 
-Considerations of such issues in the geographical modelling literature are rather scarce.Despite the opportunity of simulation in the representation of space and its potential observation biases, there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions, which have focused on the impact of:
+Considerations of such issues in the geographical modelling literature are rather scarce.
+Despite the opportunity of simulation in the representation of space and its potential observation biases, there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions, which have focused on the impact of:
 \begin{itemize}
@@
 An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
-For example, the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements.
+For example, \replaced{spatial structure may be easier to implement as a raster rather a vector in NetLogo models, what could influence the implementation choices of some non-experienced modelers}{the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements}.
 Its toroidal default setting might also have influenced the work of many modellers who did not question explicitly the representation of space.
@@
 %%%%%%%%%%%%%%
-\begin{figure}
+\begin{figure*}
 	\includegraphics[width=0.49\linewidth]{figures/alpha_beta_examples.png}
@@
     \caption{\added{\textbf{Stylized behavior of the density grid generator.}}\label{fig:stylized-generator}}
-\end{figure}
+\end{figure*}
 %%%%%%%%%%%%%%

### 2018-09-03 Response_to_Reviewers.tex
Update on Overleaf.

@@
     % \comment[JR]{plutot d'accord avec lui mais parce que proficient en NetLogo, peut etre que des debutants sont plus contraints ? quelquun aurait une ref là dessus ?}
+    %\comment[MLT]{En fait, c'est surtout le cas des pre-built simulations proposées dans la librairie de NetLogo.
+J'ai un peu cherché, je trouve pas de références qui montrent explicitement la surabondance des grilles dans les modèles hors librairie...}[(JR) ok yes en effet !
+cela devrait suffire comme argument ?]
 	\item We agree that NetLogo does not restrict the use on what implementation to use.
 However, some features are much more easier as raster or the contrary, what could influence the choice of beginners.
+Furthermore, models in the built-in NetLogo model library mostly rely on rasters.
 We clarified that statement in the paper.
@@
 	% 14. Line 671: The same comment regarding “mythical ABMs” 
-	\item The term ``mythical'' line 671 was a language mistake and corrected, ``well-known'' ABM was meant.
-    \comment[MLT]{Use "textbook ABM instead?}
+	\item The term ``mythical'' line 671 was a language mistake and corrected, ``textbook'' ABM was meant.
+    \comment[MLT]{Use "textbook ABM instead?}[(JR) yes, better]
 \end{itemize}
@@
 	% Geosimulation is introduced but not defined or referenced.  This should be clarified as the term has several meanings etc.
-	\item
+	\item For the definition of geosimulation, we meant the one coined by (Benenson and Torrens, 2004), this was made explicit and the reference was added.
 

### 2018-09-03 paper.tex
Update on Overleaf.

@@
 
-Computer simulation has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}.
-Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
+Computer simulation has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}\added{, what has been synthesized as the emerging field of geosimulation by \cite{benenson2004geosimulation}.} Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
 The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model in understanding the role of space on social interactions and environmental processes, in a city for example.
@@
 We did so by implementing a spatial generator whose output is used as input for the simulation model.
-We applied this approach to two \replaced{well-known}{mythical} ABMs: Schelling and Sugarscape.
+We applied this approach to two \replaced{textbook}{mythical} ABMs: Schelling and Sugarscape.
 With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.

### 2018-09-11 Response_to_Reviewers.tex
Update on Overleaf.

@@
 
-First, the following general changes were done:
+As a response to several reviewers, we made the following general changes:
 
 \begin{itemize}
-	\item Several comments, including the ones of the first and third referee, were linked to clarity and explicit description of models: we clarified the descriptions and better justified our choices, and also provided pseudo-codes in Appendix and made clearer that source code are openly available.
-    \item The feedback on the third referee that the most interesting was the phase diagram comparison was not directly compatible with other referees comments; we however added some developments and comments on this point to conciliate as most as possible these different comments.
+	\item Several comments, from the first and third referees especially, suggested that we clarity and explicit the description of our two models: we clarified the descriptions and justified our choices more thoroughly, and also provided pseudo-codes for each model in the Appendix so as to help the reader navigate the implementation chosen compared to other articles.
+Finally, we mentioned more clearly in the text the fact that source codes are openly available in a github repository.
+    \item There was a conflict between the comments of the third referee and the other referees about the comparison of phase diagram.
+In order to conciliate as much as possible the different parties, we added some developments and comments about how we defined, built, justified and compared phase diagrams in the paper.
     \end{itemize}
@@
 
-We provide now you point-by-point responses to referees comments.
+We now turn to point-by-point responses to referees comments.
 
@@
 As it stands, the models are treated as black boxes.
-	\item The suggestion to focus on a single model is indeed a qualified comment.
-However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies than with a single model.
-In that perspective, the ``black-box'' is seen as an advantage as it suggests this genericity.
-We acknowledge the fact that this choice led us to shorten their presentations.
-Nevertheless, we have been taking heed to your remark to improve the description of the models by adding pseudo-codes in Appendix and recalling more clearly the link to model source we used.
+	\item The suggestion to focus on a single model is a qualified comment.
+However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies rather than with a single model.
+In that perspective, the ``black-box'' is seen as an advantage as it reinforces genericity.
+We acknowledge the fact that this choice led us to shorten the model presentations.
+Nevertheless, we have taken notice of your remark by improving the description of the models and their results through the addition of pseudo-codes in Appendix, result descriptions in the text and by recalling more clearly the link to the model source we used.
+Furthermore, we have focused on some of the most classical models of the ABM literature so as to lower the reader's cost of entry into the methodology.
     %We believe this is sufficient as the models have not been adjusted from these textbook cases implementations.
@@
    % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
-   \item \comment[JR]{Possibility to find configuration with an ``eyeball'' difference ? i'm not sure of that..}
+   \item \comment[JR]{Possibility to find configuration with an ``eyeball'' difference ? i'm not sure of that..} \comment[CC]{peut-être en prenant des cas extrêmes sur Sugarscape?}
 
@@
 	\item There was a misunderstanding for the Schelling implementation actually used, as we used our own implementation which pseudo-code has been added in Appendix of the paper and source code is openly available on the git repository of the project.
-This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.
-It does not either reach a segregated state for high values of intolerance, but it does not cause a problem in our case as indicators can be computed on any model state (and actually reflect the fact that it is not segregated).
+This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.% It does not reach a segregated state for high values of intolerance, but it does not cause a problem in our case as indicators can be computed on any model state (and actually reflect the fact that it is not segregated).
+	 \comment[CC]{le modèle ne produit pas de ségrégation quand les individus sont très intolérants? Comment ça? Bien sûr qu'il le fait (sous contrainte du taux de vacance).
+	  Cela dit y'a bien un point sur lequel l'implem Netlogo n'est pas bonne: c'est que la relocalisation des mécontents est random alors que dans le bouquin de Schelling, apparemment, c'était dans un voisinage proche.}
+
     \item More generally concerning the choice of these two models, we aim at helping the readers to distance themselves from the case studies and focus on the genericity of the approach.
@@
     \item \comment[JR]{TB la def Marion ; là on peut choisir de virer en effet le terme, ou dire que les 3 sont interchangeable mais que "meta-" est plus pratique, qu'en pensez vous ?}
+    \comment[CC]{d'acc.}
 
@@
 	\item \comment[JR]{je pensais qu'on prenait la moyenne ? je verifie... ligne 175 on écrit la médiane en effet}
+	\comment[CC]{ça doit être moi parce que je pensais qu'on prenait effectivement la médiane, qui est plus robuste aux extrêmes, mais si c'est la moyenne et que ça reste proche entre différentes réplications why not.}
 
@@
 
+	In the case of Schelling, this is the scale at which most empirical segregation indices are computed.
+In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty....), or to a generic landscape where a resource is grown otherwise.
+In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density, and that not acknowledging this diversity leads to potentially false results.
+Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most harmful one (e.g.
+the least realistic, although this is the case of most Schelling models).
+	\comment[CC]{argumentaire à ajouter dans le papier}
+
 	% 9. The authors should illustrate the contribution of each parameter of the generator to the generated initial pattern using a figure. Details of the generators should be included in an appendix
@@
 	% 10. Explain how the density grids are used in the Schelling model, do you allow more than one agent to be in the same cell? 
-	\item \comment[JR]{il a mal lu ou on est pas assez explicites, a detailler}
+	\item Yes, we allow more agents to be in the same cell.
+The potential density is defined by the density grid generated, and is population with the same number of agents in all grids.
+This distribution is the one at initialisation.
+If the potential density is not reached, more agents can move in the cell, otherwise it is deemed full and unavailable for further moves.
+The satisfaction and segregation indices are computed with regard to the people in the cell and the people present in neighbouring cells.
+	\comment[JR]{il a mal lu ou on est pas assez explicites, a detailler}
+	\comment[CC]{ça va comme réponse? }
+	\comment[CC]{argumentaire à ajouter dans le papier}
 
 	% 11. Line 297: I am not sure why empirical density destitutions of cities are important in the context of this study. 
-	\item
+	\item cf infra. 
+	\item density distributions of empirical cities are important because we want to test models with realistic ranges of initial patterns of density distribution.
+Therefore we cannot limit ourselves to a single 2-peak mountain in the classical Sugarscape model and an isotropic square city in Schelling.
+We chose to use actual distribution of European cities to constraint our density generation.
+\comment[CC]{argumentaire à ajouter dans le papier}
 
@@
 	\item We agree that NetLogo does not restrict the use on what implementation to use.
-However, some features are much more easier as raster or the contrary, what could influence the choice of beginners.
+However, some features are much more easier as raster or the contrary, which could influence the choice of beginners.
 Furthermore, models in the built-in NetLogo model library mostly rely on rasters.
@@
 Initial spatial condition can have a large impact on the results of a model, however I do not think that the initial spatial conditions are often taken for granted.
-	\item \comment[JR]{il s'agirait de nuancer ici}
+	\item We forcefully agree with the referee in that initial spatial conditions can have a large impact on the results of the model.
+Maybe not all papers take them for granted, but what we meant is that compared to other parameters, initial spatial conditions are much less often explored using sensitivity analysis.
+Actually, we know only a few papers which have explored the impact of initial spatial conditions as a systematic criterion in a sensitivity analysis of model results.
+We clarified this point in the introduction.
+    \comment[JR]{il s'agirait de nuancer ici}
+\comment[CC]{argumentaire à ajouter dans le papier}
 
@@
 This is contrary to the statement in the abstract that initial conditions are often taken for granted.
-	\item
+	\item Indeed, we have tried to review articles where some kind of sensitivity to initial spatial conditions are tackled.
+But first, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.
+Second, the references cited mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but none provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper.
+We have tried to argue this point at the end of the section.
+    \comment[CC]{argumentaire à ajouter dans le papier}
 

### 2018-09-11 Response_to_Reviewers.tex
Update on Overleaf.

@@
 	\item The functionalities of OpenMOLE we used were detailed; we did not use genetic algorithms but sampling methods and high performance computing environments.
+    \comment[CC]{argumentaire à ajouter dans le papier}
 
@@
 	% Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.
+
 	\item \comment[JR]{Clem tu es d'accord là dessus ?}
+    \comment[CC]{Non mais il chipote complètement à ce niveau!
+Vu la figure pour moi c'est totalement équivalent dans un sens et dans l'autre, le tout c'était qu'il y'a une double descente chromatique et y'a même pas de densité = rouge ou quoi puisque c'est un rapport.
+Si tu veux lui faire plaisir on change mais je vois pas trop la nécessité.}
 
@@
 I do not think that the CEUS journal targets social sciences exactly.
+	Thanks for the remark. We have tried to shorten the considerations which you have found not too useful and will let the readers decide if the grids could be of any use to them.
+	\comment[CC]{argumentaire à ajouter dans le papier}
 

### 2018-09-11 Response_to_Reviewers.tex
.

--- 2018-09-11 Response_to_Reviewers.tex
@@
    % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
-   \item \comment[JR]{Possibility to find configuration with an ``eyeball'' difference ? i'm not sure of that..} \comment[CC]{peut-être en prenant des cas extrêmes sur Sugarscape?}
+   \item \comment[JR]{Possibility to find configuration with an ``eyeball'' difference ? i'm not sure of that..} \comment[CC]{peut-être en prenant des cas extrêmes sur Sugarscape?}[(JR) je regarde...]
 
@@
 	 \comment[CC]{le modèle ne produit pas de ségrégation quand les individus sont très intolérants? Comment ça? Bien sûr qu'il le fait (sous contrainte du taux de vacance).
-	  Cela dit y'a bien un point sur lequel l'implem Netlogo n'est pas bonne: c'est que la relocalisation des mécontents est random alors que dans le bouquin de Schelling, apparemment, c'était dans un voisinage proche.}
+	  Cela dit y'a bien un point sur lequel l'implem Netlogo n'est pas bonne: c'est que la relocalisation des mécontents est random alors que dans le bouquin de Schelling, apparemment, c'était dans un voisinage proche.}[(JR) ah oui cest selon la vacance ; mais il y a des regimes ou ca converge jamais sous forme de cluster et ca reste un etat dynamique sans structure ; oui par contre on a du reprendre la random reloc, c'est ce que font les physiciens aussi]
 
@@
 	\item \comment[JR]{je pensais qu'on prenait la moyenne ? je verifie... ligne 175 on écrit la médiane en effet}
-	\comment[CC]{ça doit être moi parce que je pensais qu'on prenait effectivement la médiane, qui est plus robuste aux extrêmes, mais si c'est la moyenne et que ça reste proche entre différentes réplications why not.}
+	\comment[CC]{ça doit être moi parce que je pensais qu'on prenait effectivement la médiane, qui est plus robuste aux extrêmes, mais si c'est la moyenne et que ça reste proche entre différentes réplications why not.}[(JR) il me semble que les distribs sont assez centrées et que ca revient quasiment au même ; par contre Clem si tu as pris median pour les stats et moi mean pour les plots c'est pas fou ; tu as encore le code des analyses stats ?
+(c'est sur le git ?)]
 
@@
 I am thinking of work in the line of disease diffusion, crime etc.
-	\item \comment[JR]{il a un point car ce sont des classes de modeles qu'on a pas trop cité ?
+	\item \comment[JR]{il a un point car ce sont des classes de modèles qu'on a pas trop cité ?
 par contre la sensibilite a ces distribs precises n'est pas testée, ou pas systeématiquement : il faut qu'on clarifie ou reformule}

### 2018-09-11 Response_to_Reviewers.tex
Update on Overleaf.

@@
 I would suggest to the author to rewrite the results and methodology so that they flow better and also provide more context with respect to how to interpret the results.
-	\item
+	\item We have tried to make the methodology and result section flow better in the corresponding sections.
+    \comment[CC]{à faire dans le papier}
 
@@
 After the abstract all notion of validation is missing.
-	\item \comment[JR]{enlever le concept de validation}
+	\item We have agreed with the reviewer and removed the concept of validation altogether, replacing it with model evaluation or sensitivity analysis.
+    \comment[JR]{enlever le concept de validation}
 
@@
 	\item For the definition of geosimulation, we meant the one coined by (Benenson and Torrens, 2004), this was made explicit and the reference was added. 
+    \comment[CC]{à ajouter dans le papier}
 

### 2018-09-11 Response_to_Reviewers.tex
.

--- 2018-09-11 Response_to_Reviewers.tex
@@
 by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).
-	\item
+	\item \comment[JR]{}
 
 	% Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?
-	\item \comment[JR]{no justification $\rightarrow$ new runs ?}
+	\item \comment[JR]{no justification for the difference in number of runs (just experimental setup contingence\ldots ) $\rightarrow$ either new runs or redo computations with 50 for Schelling}
 
 	% You say your models are open source - do you give details of that source?  
-	\item
+	\item Concerning the source code of the models, an open public repository was put into place and the link provided in the paper (to an anonymized version to ensure blind review).
 

### 2018-09-11 Response_to_Reviewers.tex
.

@@
 	\comment[JR]{il a mal lu ou on est pas assez explicites, a detailler}
-	\comment[CC]{ça va comme réponse? }
+	\comment[CC]{ça va comme réponse? }[(JR) parfait]
 	\comment[CC]{argumentaire à ajouter dans le papier}
@@
 I think a clear differentiation between the two groups and the positioning of the two case studies in the second group would make the paper more balanced.
-	\item \comment[JR]{oui du coup en disant ça ça repond à ses deux points précédents}
+	\item Concerning the remarks on synthetic population generation and the different types of models, we clearly stated our positioning regarding models built from theory, for which the proposed sensitivity analysis workflow is more suited and that our two case studies illustrate.\comment[JR]{oui du coup en disant ça ça repond à ses deux points précédents}\comment[JR]{positionement a plus souligner dans le papier}
 
@@
 	% -> 
-	\item
+	\item The comment on other possible techniques to compare phase diagrams echoes one other reviewer comment, and we indeed expanded the discussion on this point, showing first that other distance do not change much qualitative results and with more literature for possible extensions.
+\comment[JR]{analyse à ajouter}
 
@@
 Perhaps this should be explained earlier in the paper.
-	\item
+	\item \comment[JR]{ce qu'il suggère c'est un indice intégré sur l'ensemble des phases diagramme, donc un ``meta-meta-indice'' en quelque sorte, nous on reste à un niveau de meta seulement car on regarde juste pour chaque s'il bouge par rapport à un point de reference.
+En termes thematique c'est plus coherent pour nous, lui doit avoir en tete des indices de sensibilite globaux type Saltelli.
+$\rightarrow$ a voir comment justifier ca bien}
 
@@
 
-	\item \comment[JR]{Clem tu es d'accord là dessus ?}
-    \comment[CC]{Non mais il chipote complètement à ce niveau!
+	\item Color scale for distance was reversed.
+    \comment[JR]{Clem tu es d'accord là dessus ?}[(CC) Non mais il chipote complètement à ce niveau!
 Vu la figure pour moi c'est totalement équivalent dans un sens et dans l'autre, le tout c'était qu'il y'a une double descente chromatique et y'a même pas de densité = rouge ou quoi puisque c'est un rapport.
-Si tu veux lui faire plaisir on change mais je vois pas trop la nécessité.}
+Si tu veux lui faire plaisir on change mais je vois pas trop la nécessité.][(JR) bon je vais refaire la figure pour caresser dans le sens du poil - TODO]
 
@@
 	\item For the definition of geosimulation, we meant the one coined by (Benenson and Torrens, 2004), this was made explicit and the reference was added. 
-    \comment[CC]{à ajouter dans le papier}
+    \comment[CC]{à ajouter dans le papier}[(JR) fait au début]
 
@@
 by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).
-	\item \comment[JR]{}
+	\item \comment[JR]{to the best of my knowledge there is no full-world morphological classification of cities (and surely because global 1km density dataset is not robust enough for that - already with the European grid it is not perfect)}
 

### 2018-09-18 Response_to_Reviewers.tex
Update on Overleaf.

@@
 \begin{itemize}
-	\item Several comments, from the first and third referees especially, suggested that we clarity and explicit the description of our two models: we clarified the descriptions and justified our choices more thoroughly, and also provided pseudo-codes for each model in the Appendix so as to help the reader navigate the implementation chosen compared to other articles.
+	\item Several comments, from the first and third referees especially, suggested that we clarify and explicit the description of our two models: we clarified the descriptions and justified our choices more thoroughly, and also provided pseudo-codes for each model in the Appendix so as to help the reader navigate the implementation chosen compared to other articles.
 Finally, we mentioned more clearly in the text the fact that source codes are openly available in a github repository.
@@
     \comment[CC]{d'acc.}
+    \comment[MLT]{Du coup, dans nos échanges par mails on disait qu'on était ok pour changer, mais finalement cette solution vous convient?
+Je comprends pourquoi il/elle a tiqué sur le terme, mais c'est pas non plus hors de propos vu l'application qu'on en fait...
+bref, je suis partagée!}
 
@@
 
-	In the case of Schelling, this is the scale at which most empirical segregation indices are computed.
+	\item In the case of Schelling, this is the scale at which most empirical segregation indices are computed.
 In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty....), or to a generic landscape where a resource is grown otherwise.
@@
 the least realistic, although this is the case of most Schelling models).
-	\comment[CC]{argumentaire à ajouter dans le papier}
+	\comment[CC]{Ref du papier de batty à ajouter (je ne l'ai pas trouvée}
 
@@
 	\item Yes, we allow more agents to be in the same cell.
-The potential density is defined by the density grid generated, and is population with the same number of agents in all grids.
+The potential density is defined by the density grid generated, and its population with the same number of agents in all grids.
 This distribution is the one at initialisation.
@@
 	\comment[CC]{ça va comme réponse? }[(JR) parfait]
-	\comment[CC]{argumentaire à ajouter dans le papier}
+	\comment[CC]{argumentaire à ajouter dans le papier}[(cc)done]
 
@@
 We chose to use actual distribution of European cities to constraint our density generation.
-\comment[CC]{argumentaire à ajouter dans le papier}
+\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
 
@@
     \comment[JR]{il s'agirait de nuancer ici}
-\comment[CC]{argumentaire à ajouter dans le papier}
+\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
 
@@
 par contre la sensibilite a ces distribs precises n'est pas testée, ou pas systeématiquement : il faut qu'on clarifie ou reformule}
+    \comment[MLT]{En fait, c'est là aussi une question de reformulation.
+Je crois qu'il faudrait répondre quelque chose comme : "We acknowledge the fact that many ABM applications use "real" data and spatial features for their simulations.
+However, as explained by (Thomas et al.
+2017) in the case of LUTI models for instance, most simulations do not evaluate the impact of city delineations nor MAUP on the output of the models, which constitutes a potential limit in the interpretation of results.}
 
@@
 	\item \comment[JR]{idem la on parle pas exactement du meme type de modeles, mais son exemple prouve notre propos : le modele tourne sur une situation unique, et on teste pas l'effet de l'espace.}
+    \comment[MLT]{Bon, c'est clairement quelqu'un qui fait de l'appliqué et qui veut pas entendre notre argument ;) On peut aller dans son sens et modifier notre propos en écrivant : "However, this step is rarely explicit and authors are therefore not able to generalize the conclusion of the model to other spatial entities."}
 
@@
 We have tried to argue this point at the end of the section.
-    \comment[CC]{argumentaire à ajouter dans le papier}
+    \comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
 
@@
 	\item The functionalities of OpenMOLE we used were detailed; we did not use genetic algorithms but sampling methods and high performance computing environments.
-    \comment[CC]{argumentaire à ajouter dans le papier}
 
@@
 	Thanks for the remark. We have tried to shorten the considerations which you have found not too useful and will let the readers decide if the grids could be of any use to them.
-	\comment[CC]{argumentaire à ajouter dans le papier}
 
@@
 	\item We have tried to make the methodology and result section flow better in the corresponding sections.
-    \comment[CC]{à faire dans le papier}
+    \comment[CC]{à faire en derniere relecture}
 
@@
 	\item For the definition of geosimulation, we meant the one coined by (Benenson and Torrens, 2004), this was made explicit and the reference was added.
-    \comment[CC]{à ajouter dans le papier}[(JR) fait au début]
 
@@
 You are not the only one to believe this but you don’t show this in your paper.
-	\item
+	\item We have worked towards improving the English along with the overall argument. 
+      \comment[CC]{ok pour cette phrase + derniere relecture a faire}
 
@@
 Basically you need the “or”.
-	\item
+	\item Thank you for the suggestion. We have modified the paragraph accordingly.
 
 	% You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields?
-	\item \comment[JR]{cf remarques other reviewers sur ce point}
+	\item
+    \comment[JR]{cf remarques other reviewers sur ce point}[(MLT) Sa remarque fait sens, mais j'avoue ne pas être trop au point sur cette partie de la littérature... vous avez des références?]
 
@@
 Maybe this was explained in the paper but I missed it.
-	\item \comment[JR]{on regarde des indicateurs aggreges qui ne dependent pas de la trajectoire de chaque agent, si le placement initial est random il na aucune influence sur ceux-ci (cf les distribs) -> ajouter des graphes de comportement stat des modeles en Appendix ?}
+	\item Yes, classically, the initial distribution of agents is random in both models and, most importantly for our point, it follows a uniform distribution of density.
+In our implementation, the initial location of agents is also random in Schelling, but it follows a distribution of density given by the generated grid.
+In Sugarscape, the agents are initialised at random but the resources follow a distribution of density given by the generated grid.
+We have clarified this point in section 2.3
+    \item
+    \comment[JR]{on regarde des indicateurs aggreges qui ne dependent pas de la trajectoire de chaque agent, si le placement initial est random il na aucune influence sur ceux-ci (cf les distribs) -> ajouter des graphes de comportement stat des modeles en Appendix ?}
+	       \comment[CC]{à modifier dans le papier section 2.3}[(cc) done]
 
 	% In Section 1.2 talk about “(monocentric vs. polycentric for example)” but in the results you don’t discuss the monocentric versions explicitly using these terms.
-	\item
+	\item We believe that the monocentric versions are explicitly analysed in table 1: the term "monocentric" itself does not appear in the regression results but it is only because it is taken as the reference case under the term "compact' in the regression.
+We have clarified this in section 3.2.
+           \comment[CC]{à modifier dans le papier section 3.2}[(cc) done]
 
@@
 For example, for the sugar scape model you different styles of figures than for the segregation model.
-	\item Additional figures for the Schelling model presenting the results in a similar way were provided.
-Figures showing grid estimates of phase diagrams could not be realized for the Schelling model because of the different sampling of the parameter space used.
+	\item Additional figures for the Schelling model presenting the results in a similar way were provided and commented in a way that tries to make it easier to grasp for the reader.
+Unfortunately, the figures showing grid estimates of phase diagrams could not be realized for the Schelling model because of the different sampling of the parameter space used.
+
 \end{itemize}
@@
 polycentric density patterns) on agent-based Schelling segregation models and Sugarscape's unequal societies model.
-	\item
-
-	% Overall, I find this to be a very good paper.
+Overall, I find this to be a very good paper.
 Very well written in general.
@@
 You touch upon shape when talking about NetLogo's use of raster as opposed to vector in the Discussion - could you expand on this?
-	\item
+	\item We have not addressed the shape of cells and the uncertainty of data in this paper, only the diversity of shapes of urban morphologies (e.g. heterogeneity).
 
@@
 by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).
-	\item \comment[JR]{to the best of my knowledge there is no full-world morphological classification of cities (and surely because global 1km density dataset is not robust enough for that - already with the European grid it is not perfect)}
+	\item \comment[JR]{to the best of my knowledge there is no full-world morphological classification of cities (and surely because global 1km density dataset is not robust enough for that - already with the European grid it is not perfect)}[(MLT) Peut-être qu'on peut intégrer sa remarque dans la partie discussion?
+Je ne connais pas non plus de classification morphologique des villes à l'échelle du monde.
+Les cas européens sont privilégiés du fait de la disponibilité de données comme Corine Land Cover et Urban Atlas (pour l'usage du sol) ou Geostat (grilles de densité de pop de résolution 1km), mais ça doit aussi exister pour les US au moins?
+Il y a aussi le projet gridded population of the world de l'université de Columbia mais la résolution n'est pas assez fine (2.5 arcminute) pour une analyse de la morpho urbaine.]
 

### 2018-09-18 paper.tex
Update on Overleaf.

@@
       \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
-      \IfValueT{#4}{\textcolor{ForestGreen}{\textbf{A2 : }#4}}
+      \IfValueT{#4}{\textcolor{green}{\textbf{A2 : }#4}}
       \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
-      \IfValueT{#6}{\textcolor{Aquamarine}{\textbf{A4 : }#6}}
+      \IfValueT{#6}{\textcolor{blue}{\textbf{A4 : }#6}}
     }
@@
 The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model in understanding the role of space on social interactions and environmental processes, in a city for example.
-We think simulation can become a very good tool to achieve this, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is represented.
+We think \replaced{that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions}{simulation can become a very good tool to achieve this}, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is represented.
 Unfortunately, the first condition is not always met, and the second is seldom even mentioned.
@@
 The way modellers represent space is therefore a central element of geographical simulation models.
-However, this step is rarely explicit and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one.
+However, this step is rarely explicit.
+For example
+
+and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one.
 A meaningful way of representing space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour in plausible spatial arrangements.
@@
 
- Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
+ Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.) \replaced{or}{either} on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time, \added{offering a systematic inquiry into the impact of systematically defined spatial inputs on the model outputs.
+More so, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.}
+
 %\cite{ligmann2013spatially}
@@
 
-\todo{explain why this generator is relevant for both models (discussion on scale ?)}
+\added{The spatial generator and its resulting grids are relevant to the case study models we have picked (Schelling and Sugarscape) because it produces density grids at a ``metropolitan scale'', which is the scale at which both model were initially intended to be.
+In the case of Schelling's segregation model for example, this scale is the one at which most empirical segregation indices are computed and compared to the model outputs.
+In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources [REF BATTY], or to a generic landscape where a resource is grown otherwise.
+In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density and that acknowledging this diversity might leads to a variation in the model outputs.
+Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most unrealistic one, although unfortunately the most common one in Schelling implementations.}
+\todo{add ref to Batty's paper}
 
@@
 The pseudo-code of the implemented model is available in Appendix~\ref{app:code} and source code are available on the repository of the project.}{We translated the Netlogo implementation of Schelling into Scala for this project.}
+
+\added{In general, the implementations of Schelling models allow only one agent per cell, and their initial distribution is random, therefore following a uniform distribution across the modelled city.
+In this experiment, we allow more than one agent to be in a given same cell.
+The potential density of a cell is defined by the density grid generated.
+If the potential density of a cell is not reached at initialisation, more agents can move into the cell during the course of the simulation, otherwise it is deemed full and unavailable for movers.
+The satisfaction and segregation indices are computed with regard to the people in the cell and the people present in neighbouring cells.
+Empirical distributions of density in cities are important in our framework because we want to test models with realistic ranges of initial patterns of density distribution.
+Therefore we cannot limit ourselves to a single 2-peak mountain in the classical Sugarscape model and an isotropic square city in Schelling.
+We chose to use actual distribution of European cities to constraint our density generation.}
 
@@
 In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised correlates to a certain extent with the level of segregation measured at the end of the simulation run.
-Indeed, compared to compact density patterns, polycentric grids produce more dissimilarity and entropy between the location of green and red agents.
+Indeed, compared to \added{the reference case of} compact \added{(monocentric)} density patterns, polycentric grids produce more dissimilarity and entropy between the location of green and red agents.
 Discontinuous grids have the same effect, although attenuated.
@@
 
-As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions, that become senseless taken out of context.
-Geography and the social sciences in general have been in the last decades strongly influenced by insights from physics, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space.
+As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions.
+Geography and the social sciences in general have been strongly influenced by physics in the last decades, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space.
 We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity the for social sciences to build their own stream of methodologies in the modelling domain.

### 2018-09-19 Response_to_Reviewers.tex
.

@@
 
-%\makelettertitle
+\makelettertitle
 
@@
    % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
-   \item \comment[JR]{Possibility to find configuration with an ``eyeball'' difference ? i'm not sure of that..} \comment[CC]{peut-être en prenant des cas extrêmes sur Sugarscape?}[(JR) je regarde...]
+   \item Concerning basic illustration of the influence of initial patterns, we think that it is not possible with single runs to show a significant effect other than the direct effect of space on the localization of agents for example.
+It is indeed the point of our method that such tests must be done in a systematic way rather than ``eyeball validations''.
+   %\comment[JR]{Possibility to find configuration with an ``eyeball'' difference ?
+i'm not sure of that..}[(CC)peut-être en prenant des cas extrêmes sur Sugarscape?][(JR) je regarde...
+$\rightarrow$ bof, on voit rien a part que les agents se repartissent differement mais qui est une consequence directe d'ajouter l'espace, en en comparant les valeurs finales des ginis, ce qu'on fait plus bas plus rigoureusement.
+je suggère poliment que sa suggestion est idiote.]
 
@@
 This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.% It does not reach a segregated state for high values of intolerance, but it does not cause a problem in our case as indicators can be computed on any model state (and actually reflect the fact that it is not segregated).
-	 \comment[CC]{le modèle ne produit pas de ségrégation quand les individus sont très intolérants? Comment ça? Bien sûr qu'il le fait (sous contrainte du taux de vacance).
+	 %\comment[CC]{le modèle ne produit pas de ségrégation quand les individus sont très intolérants?
+Comment ça?
+Bien sûr qu'il le fait (sous contrainte du taux de vacance).
 Cela dit y'a bien un point sur lequel l'implem Netlogo n'est pas bonne: c'est que la relocalisation des mécontents est random alors que dans le bouquin de Schelling, apparemment, c'était dans un voisinage proche.}[(JR) ah oui cest selon la vacance ; mais il y a des regimes ou ca converge jamais sous forme de cluster et ca reste un etat dynamique sans structure ; oui par contre on a du reprendre la random reloc, c'est ce que font les physiciens aussi]
@@
 We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process".
-    \item \comment[JR]{TB la def Marion ; là on peut choisir de virer en effet le terme, ou dire que les 3 sont interchangeable mais que "meta-" est plus pratique, qu'en pensez vous ?}
+	\item We agree that the term of meta-parameters may have been confusing, and following the suggestion we replaced it with ``generator parameters''.
+     \comment[JR]{TB la def Marion ; là on peut choisir de virer en effet le terme, ou dire que les 3 sont interchangeable mais que "meta-" est plus pratique, qu'en pensez vous ?}
     \comment[CC]{d'acc.}
@@
 Je comprends pourquoi il/elle a tiqué sur le terme, mais c'est pas non plus hors de propos vu l'application qu'on en fait...
-bref, je suis partagée!}
+bref, je suis partagée!}[(JR) moi ca me va de virer le meta]
 
@@
 (c'est sur le git ?)]
+	\comment[JR]{c'est la merde, on prend certaines fois la mediane et certaines fois la moyenne...}
 
@@
 the least realistic, although this is the case of most Schelling models).
-	\comment[CC]{Ref du papier de batty à ajouter (je ne l'ai pas trouvée}
+	\comment[CC]{Ref du papier de batty à ajouter (je ne l'ai pas trouvée}[(c'est le Batty, 2005)]
 
@@
 The satisfaction and segregation indices are computed with regard to the people in the cell and the people present in neighbouring cells.
-	\comment[JR]{il a mal lu ou on est pas assez explicites, a detailler}
-	\comment[CC]{ça va comme réponse? }[(JR) parfait]
-	\comment[CC]{argumentaire à ajouter dans le papier}[(cc)done]
+	%\comment[JR]{il a mal lu ou on est pas assez explicites, a detailler}
+	%\comment[CC]{ça va comme réponse? }[(JR) parfait]
+	%\comment[CC]{argumentaire à ajouter dans le papier}[(cc)done]
 
@@
 We chose to use actual distribution of European cities to constraint our density generation.
-\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
+%\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
 
@@
 	\item The term ``mythical'' line 671 was a language mistake and corrected, ``textbook'' ABM was meant.
-    \comment[MLT]{Use "textbook ABM instead?}[(JR) yes, better]
+    %\comment[MLT]{Use "textbook ABM instead?}[(JR) yes, better]
 \end{itemize}
@@
 We clarified this point in the introduction.
-    \comment[JR]{il s'agirait de nuancer ici}
-\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
+    %\comment[JR]{il s'agirait de nuancer ici}
+	%\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
 
@@
 We have tried to argue this point at the end of the section.
-    \comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
+    %\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
 
@@
 I do not think that the CEUS journal targets social sciences exactly.
-	Thanks for the remark. We have tried to shorten the considerations which you have found not too useful and will let the readers decide if the grids could be of any use to them.
-
+	\item Thanks for the remarks on discussion, in particular on releasing density grids.
+We have tried to shorten the considerations which you have found not too useful and will let the readers decide if the grids could be of any use to them.
 \end{itemize}
@@
 	\item We have agreed with the reviewer and removed the concept of validation altogether, replacing it with model evaluation or sensitivity analysis.
-    \comment[JR]{enlever le concept de validation}
+    %\comment[JR]{enlever le concept de validation}
 
@@
 Basically you need the “or”.
-	\item Thank you for the suggestion. We have modified the paragraph accordingly.
+	\item Thank you for the suggestion on page 4. We have modified the paragraph accordingly.
 
@@
 We have clarified this point in section 2.3
-    \item
-    \comment[JR]{on regarde des indicateurs aggreges qui ne dependent pas de la trajectoire de chaque agent, si le placement initial est random il na aucune influence sur ceux-ci (cf les distribs) -> ajouter des graphes de comportement stat des modeles en Appendix ?}
-	       \comment[CC]{à modifier dans le papier section 2.3}[(cc) done]
+    %\comment[JR]{on regarde des indicateurs aggreges qui ne dependent pas de la trajectoire de chaque agent, si le placement initial est random il na aucune influence sur ceux-ci (cf les distribs) -> ajouter des graphes de comportement stat des modeles en Appendix ?}
+	%\comment[CC]{à modifier dans le papier section 2.3}[(cc) done]
 
@@
 We have clarified this in section 3.2.
-           \comment[CC]{à modifier dans le papier section 3.2}[(cc) done]
+    %\comment[CC]{à modifier dans le papier section 3.2}[(cc) done]
 
@@
 	% Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?
-	\item \comment[JR]{no justification for the difference in number of runs (just experimental setup contingence\ldots ) $\rightarrow$ either new runs or redo computations with 50 for Schelling}
+	\item \comment[JR]{no justification for the difference in number of runs (just experimental setup contingence\ldots ) $\rightarrow$ either new runs or redo computations with 50 for Schelling}\comment[JR]{ca va etre chaud patate pour les new computations ; je vais regarder les ratio de sharpe, avec un peu de chance il y a besoin de plus de repets pour schelling...}
 
@@
 	\item Concerning the source code of the models, an open public repository was put into place and the link provided in the paper (to an anonymized version to ensure blind review).
+Similarly, the large simulation datasets used for results presented in the paper are openly available on a dataverse repository (provisorily anonymized also).
+	\comment[JR]{TODO dataverse. / statement open code and data}
 

### 2018-09-19 paper.tex
.

@@
 However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment.
-In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic generator controlled by meta-parameters in order to create density grids with which spatial simulation models are initialised.
+In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic \added{spatial configuration} generator \deleted{controlled by meta-parameters} in order to create density grids with which spatial simulation models are initialised.
 We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant in the two models.
@@
 \begin{keyword}
-Space \sep Initial conditions \sep Sensitivity \sep ABM \sep Validation
+Space \sep Initial conditions \sep Sensitivity \sep ABM
 \end{keyword}
@@
 Instead, we present a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation.
-The generator being controlled by meta-parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations.
-The purpose is two-fold: (i) to test the robustness of simulation results to small variations of meta-parameters and (ii) to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
+The generator being controlled by \replaced{it own}{meta-} parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations.
+The purpose is two-fold: (i) to test the robustness of simulation results to small variations of \replaced{generator}{meta-} parameters and (ii) to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
 Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies.
@@
 In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on \replaced{its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
-We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process}{meta-parameters and feeds the model $\mu$ at initialisation (lower branch)}.
+We call these parameters ``generator parameters'' to distinguish from the standard parameters of the models and highlight their higher position in the simulation process}{meta-parameters and feeds the model $\mu$ at initialisation (lower branch)}.
 The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.
@@
 We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$.
-Phase diagrams are denoted by a common function $\mu$ indexed by what we call \emph{meta-parameters} $\vec{\alpha}$, that capture the spatial configuration \footnote{in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself}.
+Phase diagrams are denoted by \replaced{the same}{a common} function $\mu$ indexed by \replaced{the generator parameters}{what we call \emph{meta-parameters}} $\vec{\alpha}$, \replaced{which}{that} capture the spatial configuration \added{(in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself)}.
 The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by
@@
 
-where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given meta-parameters $\vec{\alpha}_i$.
+where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given \replaced{generator}{meta-} parameters $\vec{\alpha}_i$.
 $d$ is a functional distance that we take simply as Euclidian distance.
@@
 The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write modulable exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
-In our case, this tool is a powerful way to embed both the sensitivity analysis and the meta-sensitivity analysis, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file.
+In our case, this tool is a powerful way to embed both the sensitivity analysis and the \deleted{meta-}sensitivity analysis \added{to initial spatial conditions}, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file.
 In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the  framework of multi-modelling \citep{cottineau2015modular}.
@@
 
-The spatial generator applies an urban morphogenesis model \citep{Batty2007} which has been generalised, explored and calibrated by~\citet{2017arXiv170806743R}.
+The spatial generator applies an urban morphogenesis model \citep{Batty2007} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}.
 An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily in our workflow.
@@
 The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps).
-Grids are thus generated from the combination of the values of these four meta-parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed.
+Grids are thus generated from the combination of the values of these four \replaced{generator}{meta-} parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed.
 To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment.
 We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$.
-Typical value ranges for the  parameters will be taken as, following \citet{2017arXiv170806743R}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$.
+Typical value ranges for the  parameters will be taken as, following \citet{raimbault2018calibration}, $\alpha\in\left[0.5,4.0\right]$, $\beta \in\left[0,0.3\right] $, $N\in \left[100,10000\right]$, $n_d\in\left[1,4\right]$.
 We illustrate in Fig.\ref{fig:spatialGen} the variety of spatial configurations that can be generated.
@@
 \cite{LeNechet2015}).
-We sample the meta-parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a point cloud with high discrepancy.
+We sample the \replaced{generator}{meta-} parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a point cloud with high discrepancy.
 We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}.
@@
 In the case of Schelling's segregation model for example, this scale is the one at which most empirical segregation indices are computed and compared to the model outputs.
-In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources [REF BATTY], or to a generic landscape where a resource is grown otherwise.
+In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources \citep{batty2005agents}, or to a generic landscape where a resource is grown otherwise.
 In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density and that acknowledging this diversity might leads to a variation in the model outputs.
 Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most unrealistic one, although unfortunately the most common one in Schelling implementations.}
-\todo{add ref to Batty's paper}
 
@@
 We run 50 repetitions for each configuration, which yields reasonable convergence properties.
-The initial spatial configuration varies across 50 different grids, generated by sampling meta-parameters for the generator in a LHS.
+The initial spatial configuration varies across 50 different grids, generated by sampling \replaced{generator parameters}{meta-parameters for the generator} in a LHS.
 We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model.
@@
 
-We choose different experiment designs, both for meta-parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices.
+We choose different experiment designs, both for \replaced{generator parameters}{meta-parameters} and for the phase diagram, to demonstrate the robustness of the method to technical choices.
 In principle, our workflow applies with to manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.
@@
 Indeed, the relative distance between the phase diagrams of different density grids and the phase diagram of the reference case ranges from 0.09 to 2.98 with a median of 1.52 and an average value of 1.30.
-The mean distance above 1 means that, on average, the model is more sensitive to meta-parameters of grid generation than to its own parameters (population and sugar endowment) in the reference model.
+The mean distance above 1 means that, on average, the model is more sensitive to \replaced{the generator parameters}{meta-parameters of grid generation} than to its own parameters (population and sugar endowment) in the reference model.
 Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be up to three times bigger than the variation due to the model parameters.
-We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in the meta-parameter space.
+We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in the \replaced{generator parameter}{meta-parameter} space.
 Each point represent one of the 50 different density grids used to initialise the distribution of sugar in the model.
-The points are projected with respect to the meta-parameters used to generate the grid, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case.
+The points are projected with respect to the \replaced{generator parameters}{meta-parameters used to generate the grid}, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case.
 Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e.
@@
 \includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}
-\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by the meta-parameters of their generator.} Relative distance as a function of meta-parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
+\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by \replaced{their generator parameters}{the meta-parameters of their generator}.} Relative distance as a function of \replaced{generator parameters}{meta-parameters} $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
 \label{fig:sugarscape-distance-meta}
@@
 
-Another way of quantifying the density grids, instead of looking at the meta-parameters of their generator, is to look at the resulting indicators of urban form, such as Moran index, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and context).
+Another way of quantifying the density grids, instead of looking at the \replaced{generator parameters}{meta-parameters of their generator}, is to look at the resulting indicators of urban form, such as Moran index, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and context).
 This 4-dimensional space defined a morphological space.
@@
 The first component defines a ``level of sprawl'' and of scattering, whereas the second one represents the level aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation (top left of figure \ref{fig:sugarscape-distance-pca}).
-It is confirmed by the behaviour as a function of meta-parameters, as high values of $\alpha$ also yield high distance.
+It is confirmed by the behaviour as a function of \replaced{generator parameters}{meta-parameters}, as high values of $\alpha$ also yield high distance.
 In terms of model processes, it shows that congestion mechanisms in the gathering of the resource induces fast increases of inequality.
@@
 \textbf{Density Grid = Discontinuous}       &            & 0.111 ***                        &                   & 0.068 ***         &                      & \textit{0.00}              \\
-\textbf{Attraction meta-parameter $\alpha$} &            & -0.083 ***                       &                   & -0.053 ***        &                      & 0.014 ***                \\ 
-\textbf{Diffusion meta-parameter $\beta$}   &            & 0.323 ***                        &                   & 0.218 ***         &                      & 0.017 ***           \\ \hline
+\textbf{Attraction \replaced{generator }{meta-}parameter $\alpha$} &            & -0.083 ***                       &                   & -0.053 ***        &                      & 0.014 ***                \\
+\textbf{Diffusion \replaced{generator }{meta-}parameter $\beta$}   &            & 0.323 ***                        &                   & 0.218 ***         &                      & 0.017 ***           \\ \hline
 \textbf{R2 (\%)}                            & 30.6       & 34.7                             & 24.1              & 25.6              & 23.9                 & 24.0                    \\ 
@@
 \paragraph{Comparing phase diagrams} Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point.
-For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \emph{meta phase transitions} (phase transitions in the space of meta parameters).
+For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \deleted{\emph{meta phase transitions}} phase transitions in the space of \replaced{generator}{meta} parameters.
 The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\citep{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
@@
 \caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams.
-We show them in the meta-parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
+We show them in the \replaced{generator }{meta-}parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
 \end{figure*}

### 2018-09-19 Response_to_Reviewers.tex
.

@@
 the least realistic, although this is the case of most Schelling models).
-	\comment[CC]{Ref du papier de batty à ajouter (je ne l'ai pas trouvée}[(c'est le Batty, 2005)]
+	%\comment[CC]{Ref du papier de batty à ajouter (je ne l'ai pas trouvée}[(c'est le Batty, 2005)]
 
@@
 	% 11. Line 297: I am not sure why empirical density destitutions of cities are important in the context of this study. 
-	\item cf infra. 
-	\item density distributions of empirical cities are important because we want to test models with realistic ranges of initial patterns of density distribution.
+	%\item cf infra. 
+	\item Density distributions of empirical cities are important because we want to test models with realistic ranges of initial patterns of density distribution.
 Therefore we cannot limit ourselves to a single 2-peak mountain in the classical Sugarscape model and an isotropic square city in Schelling.
@@
 However, as explained by (Thomas et al.
-2017) in the case of LUTI models for instance, most simulations do not evaluate the impact of city delineations nor MAUP on the output of the models, which constitutes a potential limit in the interpretation of results.}
+2017) in the case of LUTI models for instance, most simulations do not evaluate the impact of city delineations nor MAUP on the output of the models, which constitutes a potential limit in the interpretation of results.}{(JR) ca me va comme reformulation}
 
@@
 	\item The comment on other possible techniques to compare phase diagrams echoes one other reviewer comment, and we indeed expanded the discussion on this point, showing first that other distance do not change much qualitative results and with more literature for possible extensions.
-\comment[JR]{analyse à ajouter}
+	%\comment[JR]{analyse à ajouter}
 
@@
 En termes thematique c'est plus coherent pour nous, lui doit avoir en tete des indices de sensibilite globaux type Saltelli.
-$\rightarrow$ a voir comment justifier ca bien}
+$\rightarrow$ a voir comment justifier ca bien}{(JR) inclure en mise en contexte ce lien avec les Saltelli et indexes globaux}
 
@@
 After the abstract all notion of validation is missing.
-	\item We have agreed with the reviewer and removed the concept of validation altogether, replacing it with model evaluation or sensitivity analysis.
+	\item We have agreed with the reviewer on ``validation'' and removed the concept of validation altogether, replacing it with model evaluation or sensitivity analysis.
     %\comment[JR]{enlever le concept de validation}
@@
 	\item 
-    \comment[JR]{cf remarques other reviewers sur ce point}[(MLT) Sa remarque fait sens, mais j'avoue ne pas être trop au point sur cette partie de la littérature... vous avez des références?]
+    \comment[JR]{cf remarques other reviewers sur ce point}[(MLT) Sa remarque fait sens, mais j'avoue ne pas être trop au point sur cette partie de la littérature...
+vous avez des références?][(JR)concernant la spatial sensitivity : justement c'est pas trop fait non plus dans les autres champs !
+je parlais avec un ecologue, leur espace est jamais varié autre que donnees reelles ; j'ai l'impressions que geosciences, hydrologie etc idem.
+Pour comparer phase diagrams j'ai pas en tete de domaine ou ca se ferait (faute d'avoir cherché, justement quand on cherchait une methode) - le plus proche c'est les patterns comparison etc.
+comme l'on suggere les reviewers.]
 
@@
 by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).
-	\item \comment[JR]{to the best of my knowledge there is no full-world morphological classification of cities (and surely because global 1km density dataset is not robust enough for that - already with the European grid it is not perfect)}[(MLT) Peut-être qu'on peut intégrer sa remarque dans la partie discussion?
+	\item To the best of our knowledge, there does not exist a study of urban morphologies similar to the one we used but at a global scale (this should be linked to the absence of a raster of population with a resolution fine enough).
+Working on other regional systems of cities such as the US could be possible but remains out of the scope of this paper.
+We added this interesting suggestion to the discussion.
+	%\comment[JR]{to the best of my knowledge there is no full-world morphological classification of cities (and surely because global 1km density dataset is not robust enough for that - already with the European grid it is not perfect)}[(MLT) Peut-être qu'on peut intégrer sa remarque dans la partie discussion?
 Je ne connais pas non plus de classification morphologique des villes à l'échelle du monde.
@@
 Similarly, the large simulation datasets used for results presented in the paper are openly available on a dataverse repository (provisorily anonymized also).
-	\comment[JR]{TODO dataverse. / statement open code and data}
+	%\comment[JR]{TODO dataverse. / statement open code and data}
 

### 2018-09-19 paper.tex
.

@@
 \usepackage{listings}
+\lstset{escapeinside={(*@}{@*)}}
+
 \usepackage{mdframed}
@@
 
-\todo{add code and data availability statement here}
+\added{The implementations of the models were done in NetLogo.
+We modified the sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model.
+Both pseudo-codes are available in Appendix~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
+Density grids are also available at this address.
+Simulation data are available for reproducibility on the dataverse repository at \url{https://doi.org/10.7910/DVN/A0N5GV}.}
 
@@
 The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\citep{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
-Map comparison methods, popular in environmental sciences, provide numeral tools to compare two dimensional fields~\citep{visser2006map}.
+Map comparison methods, popular in environmental sciences, provide numeral tools to compare two dimensional fields~\citep{visser2006map,kuhnert2005comparing}.
 To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\citep{10.1371/journal.pone.0178165}.
-The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.
-
-\added{\cite{kuhnert2005comparing} map comparison techniques}
-
-\added{\cite{kruskal1983overview} parallel with time-series comparison techniques (see reviewer comment - however more constrained as two dimensional indexing)}
+\added{The transfer of methods used to compare sequences \citep{kruskal1983overview} or time-series \citep{liao2005clustering} is a possible way to develop measures between phase diagrams.
+The higher dimension of the phase diagram we study must however be considered with caution when transferring method, in a way analog to the application of global sensitivity indices to spatial data \citep{lilburne2009sensitivity}.} The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.
 
@@
 They are available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
-Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents.
+\added{Future work could be done to compare these or generate grids with a larger morphological span, covering other typical urban forms that can be found in the world.}
+
+Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents, \added{transportation networks generators \citep{raimbault2018multi}, or coupled road network and population raster generators \citep{raimbault2018urban}.}
 
@@
 
+\newpage
+
 \section{\added{Behavior of the density grid generator}}
 
-\added{We give in Fig.~\ref{fig:stylized-generator} plots roughly summarizing the behavior of the density grid generator, according to parameters $(\alpha,\beta)$.}
-
-\todo{bla bla on generator scale / real configs in the initial paper}
+\added{We give in Fig.~\ref{fig:stylized-generator} plots roughly summarizing the behavior of the density grid generator, according to parameters $(\alpha,\beta)$.
+To put it in a simple way, high values of $\alpha$ give highly hierarchical configurations, and diminishing $\beta$ increase the number of centers.
+Low values of $\alpha$ give diffuse patterns, with however clear centers for a high diffusion.
+We do not discuss it the role of other parameters, but according to \cite{raimbault2018calibration}, diffusion steps give smoother forms, and the rate between total population and the population increment at each step (which is equivalent to the total number of steps) is crucial to select non-stationary distributions that are closer to real configurations.
+\cite{raimbault2018calibration} also shows the existence of non-linear behaviors in some regions of the parameter space, so the description we gave here shall not be interpreted as a linear link between generator parameters and the morphological properties of the generated grids.}
 
@@
 	\includegraphics[width=0.49\linewidth]{figures/alpha_beta_examples.png}
-    \includegraphics[width=0.49\linewidth]{figures/alpha_beta_schemas.png}
-    \caption{\added{\textbf{Stylized behavior of the density grid generator.}}\label{fig:stylized-generator}}
+    \includegraphics[width=0.47\linewidth]{figures/alpha_beta_schemas.png}
+    \caption{\added{\textbf{Stylized behavior of the density grid generator.} (Left) Examples of grids with their position in the $(\alpha,\beta)$ parameter space; (Right) Stylized interpretation of the forms obtained as a function of $\alpha$ and $\beta$.}\label{fig:stylized-generator}}
 \end{figure*}
@@
 \added{The Fig.~\ref{fig:schelling-distance-meta} gives the phase diagrams distances as a function of generator parameters and morphological components, similarly to the Sugarscape model in main text.
-In absolute, this version of the Schelling model seems less sensitive to density grids than the sugarscape model, as we do not obtain a high range of values here.}
+In absolute, this version of the Schelling model seems less sensitive to density grids than the sugarscape model, as we do not obtain a high range of values here.
+We however obtain measures ranging from 0.3 to 1, what is however characteristic of a significant sensitivity to space.}
 
 %%%%%%%%%%%%%
-\begin{figure*}[h!]
+\begin{figure*}
 \centering
@@
 
+\section{\added{Comparison of phase diagram with other distances}}
+
+\added{We describe here the tests done with other distances to compare phase diagrams.}
+
 \section{\added{Pseudo-code for models}}
@@
 
-\added{We give below the pseudo-code for the implementations we used of both models.}
+\added{We give below the pseudo-code for the implementations we used of both sugarscape and schelling models.
+We recall that source code is openly available at} \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
 
+%%%%%%%%%%%%%%%
 \begin{figure*}
-%\begin{frame}
+	\caption{\added{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.}}}
+	\begin{mdframed}
+	\begin{lstlisting}
+	while (*@\textit{not stopping-condition}@*) [
+    	ask turtles [
+    		set unsatisfied? false
+    		let neighboragents turtles in-radius neighborhood-radius
+    		let wantedcolor color
+    		let stranger-rate count neighboragents with [color = wantedcolor] / count neighboragents
+    		if stranger-rate < similar-wanted [
+    set unsatisfied? true random-move
+    ]
+  ]
+    \end{lstlisting}
+    \end{mdframed}
+\end{figure*}
+%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%
+\begin{figure*}
 % Minimal netlogo code for the sugarscape model
+\caption{\added{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}}
+\begin{mdframed}
 \begin{lstlisting}
-	to main-loop
-    	regrow-patches
-        update-agetns
-    end
+while [time < final-time] [
+	; regrow the patch resources
+    ask patches [set psugar min (list max-psugar (psugar + 1))]
 
-    to regrow-patches
-    	ask patches [set psugar min (list max-psugar (psugar + 1))]
-    end
+    ; make agents exploit resources
+    ask turtles [
+    	; candidate patches to which the turtle can move
+    	let move-candidates (*@\textit{patches in vision range with no turtles}@*)
 
-    to update-agents
-    	ask turtles [
-    		let move-candidates (patch-set patch-here (patches at-points vision-points) with [not any? turtles-here])
+    	; select the patch with maximal resource
  		let possible-winners move-candidates with-max [psugar]
   		if any? possible-winners [
-    ;; if there are any such patches move to one of the patches that is closest
-    move-to min-one-of possible-winners [distance myself]
+  			move-to one-of (*@\textit{possible-winners at minimal distance}@*)
   		]
 
-            ; 
+        ; eat the sugar on the patch
         set sugar (sugar - metabolism + psugar)
@@
 
+        ; age and die if necessary
     	set age (age + 1)
     	if sugar <= 0 or age > max-age [
+    		; hatch an other turtle before dying
       		hatch 1 [ new-turtle ]
@@
   	]
-    end
-
+]
 \end{lstlisting}
-%\end{frame}
-\end{figure*}
-
-%%%%%%%%%%%%%%%
-\begin{figure*}
-	\begin{lstlisting}
-
-    \end{lstlisting}
+\end{mdframed}
 \end{figure*}

### 2018-09-19 paper.tex
.

@@
 \replaced{We use a ad-hoc implementation of the Schelling model, both in Scala for performance reasons and in NetLogo to ensure visualization of model dynamics.
-The pseudo-code of the implemented model is available in Appendix~\ref{app:code} and source code are available on the repository of the project.}{We translated the Netlogo implementation of Schelling into Scala for this project.}
+The pseudo-code of the implemented model is available in~\ref{app:code} and source code are available on the repository of the project.}{We translated the Netlogo implementation of Schelling into Scala for this project.}
 
@@
 We modified the sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model.
-Both pseudo-codes are available in Appendix~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
+Both pseudo-codes are available in~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
 Density grids are also available at this address.

### 2018-09-27 Response_to_Reviewers.tex
Update on Overleaf.

@@
 Finally, we mentioned more clearly in the text the fact that source codes are openly available in a github repository.
-    \item There was a conflict between the comments of the third referee and the other referees about the comparison of phase diagram.
-In order to conciliate as much as possible the different parties, we added some developments and comments about how we defined, built, justified and compared phase diagrams in the paper.
+    \item In order to conciliate as much as possible the different points of view about the comparison of phase diagram, we added some developments and comments about how we defined, built, justified and compared phase diagrams in the paper.
+    %\comment[FL]{je ne sais pas si dire "conflict" sera productif?, de même que de parler de "parties"? ci-dessous en commentaires la réponse précédente, j'ai juste reformulé}[(JR) ok nickel]
+    % \item There was a conflict between the comments of the third referee and the other referees about the comparison of phase diagram.
+In order to conciliate as much as possible the different parties, we added some developments and comments about how we defined, built, justified and compared phase diagrams in the paper
     \end{itemize}
@@
 It is indeed the point of our method that such tests must be done in a systematic way rather than ``eyeball validations''.
+   %\comment[FL]{je ne suis pas sûr d'être d'accord avec cette réponse : si on est pas en mesure de fournir des illustrations frappantes du truc d'entrée de jeu, c'est dommage.
+Pas sûr que se cacher derrière les grilles complètes soit le bon argument.}[(JR) du coup on est ok que ca compliquerait trop l'autre Schelling encore et qu'on le met pas ?]
    %\comment[JR]{Possibility to find configuration with an ``eyeball'' difference ?
@@
 	\item We agree that the term of meta-parameters may have been confusing, and following the suggestion we replaced it with ``generator parameters''.
-     \comment[JR]{TB la def Marion ; là on peut choisir de virer en effet le terme, ou dire que les 3 sont interchangeable mais que "meta-" est plus pratique, qu'en pensez vous ?}
-    \comment[CC]{d'acc.}
-    \comment[MLT]{Du coup, dans nos échanges par mails on disait qu'on était ok pour changer, mais finalement cette solution vous convient?
+     %\comment[JR]{TB la def Marion ; là on peut choisir de virer en effet le terme, ou dire que les 3 sont interchangeable mais que "meta-" est plus pratique, qu'en pensez vous ?}
+    %\comment[CC]{d'acc.}
+    %\comment[MLT]{Du coup, dans nos échanges par mails on disait qu'on était ok pour changer, mais finalement cette solution vous convient?
 Je comprends pourquoi il/elle a tiqué sur le terme, mais c'est pas non plus hors de propos vu l'application qu'on en fait...
-bref, je suis partagée!}[(JR) moi ca me va de virer le meta]
+bref, je suis partagée!}[(JR) moi ca me va de virer le meta]\comment[FL]{moi aussi ca me va de virer méta}
 
@@
 	\comment[JR]{c'est la merde, on prend certaines fois la mediane et certaines fois la moyenne...}
+    \comment[FL]{certes c'est pas fou mais ca peut se défendre en disant que la médiane sert pour communiquer de façon accessible, sur une dimension pour laquelle c'était pertinent, mais que de façon générique on a préféré travailler sur la loi de Gauss car tout le monde à en tête ces indicateurs moyenne / ecartype c'est plus "passe partout"?
+}
 
@@
 Would we get qualitatively different results from different indices?
-	\item \comment[JR]{on avait teste avec le Monge ca donnait a peut pres pareil ; donc les normes alpha surement encore plus. je regarde.}
+	\item \comment[JR]{on avait teste avec le Monge ca donnait a peut pres pareil ; donc les normes alpha surement encore plus.
+je regarde.}\comment[FL]{je pense qu'on peut assumer le côté exploratoire de l'approche : on a deux modèles déjà et on présente un framework : on peut ouvrir en disant qu'ensuite en fonction des indicateurs on pourrait avoir des agrégateurs mathématiques ad hoc, parfois L2, L1, Linf, mais que là pour faire "passe partout" on présente L2 que tout le monde connait?
+}
 
@@
 We made this aspect more explicit.
+\comment[FL]{c'est léger comme justification : je dirais que ton générateur est crédible il produit des indicateurs dont l'espace de variation est crédible quand on le confronte aux observations empiriques : il faut que tu parles je pense de la ref où tu as les deux nuages de points superposés.
+Sinon cela fait un peu "cela marche car cela marche"}
 
 	\item In the case of Schelling, this is the scale at which most empirical segregation indices are computed.
-In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty....), or to a generic landscape where a resource is grown otherwise.
+In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty, 2005), or to a generic landscape where a resource is grown otherwise.
 In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density, and that not acknowledging this diversity leads to potentially false results.
-Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most harmful one (e.g.
-the least realistic, although this is the case of most Schelling models).
+Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most common simplification of the heterogeneity of space, with consequences on the model outputs often not evaluated.
+    %(e.g.
+the least realistic, although this is the case of most Schelling models).\comment[FL]{reformuler en enlevant harmful : simplement "is potentially the most common simplification of the heterogeneity of space, with consequences on the model outputs often not evaluated}
 	%\comment[CC]{Ref du papier de batty à ajouter (je ne l'ai pas trouvée}[(c'est le Batty, 2005)]
@@
 We chose to use actual distribution of European cities to constraint our density generation.
-%\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
+%\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]\comment[FL]{top}
 
@@
 	\item \comment[JR]{idem la on parle pas exactement du meme type de modeles, mais son exemple prouve notre propos : le modele tourne sur une situation unique, et on teste pas l'effet de l'espace.}
-    \comment[MLT]{Bon, c'est clairement quelqu'un qui fait de l'appliqué et qui veut pas entendre notre argument ;) On peut aller dans son sens et modifier notre propos en écrivant : "However, this step is rarely explicit and authors are therefore not able to generalize the conclusion of the model to other spatial entities."}
+    \comment[MLT]{Bon, c'est clairement quelqu'un qui fait de l'appliqué et qui veut pas entendre notre argument ;) On peut aller dans son sens et modifier notre propos en écrivant : "However, this step is rarely explicit and authors are therefore not able to generalize the conclusion of the model to other spatial entities."}\comment[FL]{cette formulation me va, peut être "spatial context" plutôt que "spatial entities", qui serait encore plus général?}
 
@@
 	\item Indeed, we have tried to review articles where some kind of sensitivity to initial spatial conditions are tackled.
-But first, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.
+\comment[FL]{ci dessous en commentaire l'ancienne version, j'ai essayé d'arrondir la formulation ca fait agacé genre "bon j'ai raison : d'abord..., ensuite..."}.
+However we would like to emphasize that the references cited in this section are few compared to the mass of geosimulation articles published with hardly any mention of dependence of output on initial spatial condition.
+Moreover, even if such sensitivity is tester, it mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but to the best of our knowledge no article provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper.
+We have tried to argue this point at the end of the section.
+    %\item Indeed, we have tried to review articles where some kind of sensitivity to initial spatial conditions are tackled.
+But first\comment[FL]{arrondir la formulation ca fait agacé genre "bon j'ai raison : d'abord..., ensuite..."}, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.
 Second, the references cited mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but none provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper.
@@
 	% Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.
-	\item The functionalities of OpenMOLE we used were detailed; we did not use genetic algorithms but sampling methods and high performance computing environments.
+	\item The functionalities of OpenMOLE we used were detailed in the article; we did not use genetic algorithms but sampling methods and high performance computing environments.
 
@@
 En termes thematique c'est plus coherent pour nous, lui doit avoir en tete des indices de sensibilite globaux type Saltelli.
-$\rightarrow$ a voir comment justifier ca bien}{(JR) inclure en mise en contexte ce lien avec les Saltelli et indexes globaux}
+$\rightarrow$ a voir comment justifier ca bien}{(JR) inclure en mise en contexte ce lien avec les Saltelli et indexes globaux}\comment[FL]{oui je trouve qu'on a le droit de rester limiter à cela c'est déjà pas mal!}
 

### 2018-09-27 paper.tex
Update on Overleaf.

@@
 
-\author[a,b]{Juste Raimbault\corref{cor1}}
-\author[c,d]{Cl{\' e}mentine Cottineau}
-\author[e]{Marion Le Texier}
-\author[f]{Florent Le N{\' e}chet}
-\author[a]{Romain Reuillon}
-%\author{}
+%\author[a,b]{Juste Raimbault\corref{cor1}}
+%\author[c,d]{Cl{\' e}mentine Cottineau}
+%\author[e]{Marion Le Texier}
+%\author[f]{Florent Le N{\' e}chet}
+%\author[a]{Romain Reuillon}
+\author{}
 
-\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
-\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
-\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
-\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
-\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
-\address[f]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
+%\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
+%\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
+%\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
+%\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
+%\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
+%\address[f]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
 
-\cortext[cor1]{Corresponding author: juste.raimbault@polytechnique.edu}
+%\cortext[cor1]{Corresponding author: juste.raimbault@polytechnique.edu}
 
@@
 Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes.
-However, initial spatial conditions are usually taken for granted in geographical models, thus leaving completely unexplored the effect of spatial arrangements on the interactions of agents and of that of agents with their environment.
+However, initial spatial conditions are usually \replaced{not modified systematically}{taken for granted} in geographical models, thus leaving \deleted{completely} unexplored the effect of \added{initial} spatial arrangements on the interactions of agents \replaced{with one another as well as}{and of that of agents} with their environment.
 In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic \added{spatial configuration} generator \deleted{controlled by meta-parameters} in order to create density grids with which spatial simulation models are initialised.
-We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant in the two models.
+We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant \replaced{on}{in} the two models.
 Furthermore, this effect is sometimes larger than the effect of parameters' value change.
@@
 % - Strong quantitative and qualitative effects of spatial initial conditions are shown for these models.
-
-%\textbf{Example track changes : }\added{this is added text}, \deleted{removed} text or \replaced{new}{old} replaced text
 
@@
 We modified the sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model.
-Both pseudo-codes are available in~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
-Density grids are also available at this address.
-Simulation data are available for reproducibility on the dataverse repository at \url{https://doi.org/10.7910/DVN/A0N5GV}.}
+Both pseudo-codes are available in~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at} \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
+\added{Density grids are also available at this address.
+Simulation data are available for reproducibility on the dataverse repository at }\url{https://doi.org/10.7910/DVN/A0N5GV}.
 
@@
 \textbf{Vacancy Rate (V)}                   & 0.602 ***  & 0.602 ***                        & 0.453 ***         & 0.453 ***         & -0.027 ***           & -0.027 ***               \\ 
-\textbf{Minority Index (\%Maj - \%Min)}     & 0.307å ***  & 0.307 ***                        & 0.130 ***         & 0.130 ***         & -0.067 ***           & -0.067 ***               \\ \hline
+\textbf{Minority Index (\%Maj - \%Min)}     & 0.307 ***  & 0.307 ***                        & 0.130 ***         & 0.130 ***         & -0.067 ***           & -0.067 ***               \\ \hline
 \textbf{Density Grid = Polycentric}         &            & 0.087 ***                        &                   & 0.052 ***         &                      & 0.001 ***                \\ 
@@
 
+%%%%%%%%%%%%%%%%%
+\section{\added{Additional statistical analysis}}
+
+%%%
+%% Sugarscape
+
+% Sharpes
+%summary(sres$giniSharpe)
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%  2.343  14.413  20.993  20.098  26.208  49.808
+
+% # ks
+% # table(sres$distrib)
+% #gamma  geom lnorm  norm  unif 
+% #1090  9794 16212 14263 13201 
+% AIC
+% table(sres$distrib)
+% exp gamma lnorm  norm 
+%  1451  9707 22825 20577
+
+%%%
+%% Schelling
+
+%%%%%%%%%%%%%%%%%
 \section{\added{Additional figures for the Schelling model}}
@@
 In absolute, this version of the Schelling model seems less sensitive to density grids than the sugarscape model, as we do not obtain a high range of values here.
-We however obtain measures ranging from 0.3 to 1, what is however characteristic of a significant sensitivity to space.}
+We however obtain measures ranging from 0 to 0.85 with the euclidian distance, what is however characteristic of a significant sensitivity to space.}
+
+%summary(dists)
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+% 0.0000  0.2164  0.3325  0.3700  0.4846  0.8473
 
@@
 \centering
-\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams_red}
-\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace_red}
+\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams_mean.png}
+\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace_mean.png}
 \caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams.

### 2018-09-27 paper.tex
.

@@
 
-%\author[a,b]{Juste Raimbault\corref{cor1}}
-%\author[c,d]{Cl{\' e}mentine Cottineau}
-%\author[e]{Marion Le Texier}
-%\author[f]{Florent Le N{\' e}chet}
-%\author[a]{Romain Reuillon}
-\author{}
+\author[a,b]{Juste Raimbault\corref{cor1}}
+\author[c,d]{Cl{\' e}mentine Cottineau}
+\author[e]{Marion Le Texier}
+\author[f]{Florent Le N{\' e}chet}
+\author[a]{Romain Reuillon}
+%\author{}
 
-%\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
-%\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
-%\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
-%\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
-%\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
-%\address[f]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
+\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
+\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
+\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
+\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
+\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
+\address[f]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
 
-%\cortext[cor1]{Corresponding author: juste.raimbault@polytechnique.edu}
+\cortext[cor1]{Corresponding author: juste.raimbault@polytechnique.edu}
 

### 2018-09-27 paper.tex
Update on Overleaf.

@@
 
-Computer simulation has been recognised and increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}\added{, what has been synthesized as the emerging field of geosimulation by \cite{benenson2004geosimulation}.} Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
-The specificity of geographical models compared to other social sciences' models is that space and spatial interactions are given a prime role, as geographers are driven by an explicit interest in studying the way space influences the outcomes of the model in understanding the role of space on social interactions and environmental processes, in a city for example.
-We think \replaced{that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions}{simulation can become a very good tool to achieve this}, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of output variations to the way space is represented.
+Computer simulation has been recognised and \added{is }increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}\added{.
+It has been identified as an emerging field and coined under the term geosimulation by \cite{benenson2004geosimulation}.} Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
+The specificity of geographical models compared to other social science models is that space and spatial interactions are given a prime role, \deleted{as }geographers \replaced{being}{are} driven by an explicit interest in studying the way space influences the outcomes of \replaced{social processes modelled.}{the model in understanding the role of space on social interactions and environmental processes, in a city for example}.
+We think \replaced{that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions}{simulation can become a very good tool to achieve this}, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of \replaced{the model outputs}{ variations} to the way space is represented.
 Unfortunately, the first condition is not always met, and the second is seldom even mentioned.
@@
 Geographical systems can be crudely described as social agents interacting with one another, within a limited portion of space.
-Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependent \citep{arthur1994increasing}.
-Therefore, observing a system at different points in time does not equate to observing \added{a} different systems at a single point in time.
+Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependency \citep{arthur1994increasing}.
+Therefore, observing a system at different points in time does not equate to observing different systems at a single point in time.
 This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}.
-Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex system generally, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
-Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models...
-which are the main models to be found in geosimulation.
+Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex systems generally, a given configuration contains clues \replaced{about}{on} past bifurcations, that can have had dramatic effects on the state of the system.
+Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models.
+\replaced{Ironically, these very models tend to be the models most frequently used}{..
+which are the main models to be found} in geosimulation.
 
-Self-organization has been shown to be a central feature of geographical systems in general and cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
+Self-organization has been shown to be a central feature of geographical systems in general and of cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
 In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}.
@@
 Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures.
-In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as other initial conditions, but of more interest to the geographer.
+In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as \added{any }other initial conditions, but of more interest to the geographer.
 \\
 
-Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in the most sprawled cities.
+Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in \deleted{the most }sprawled cities.
 On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities.
 In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions.
-However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic densities.
-Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}.
-The isotropic simplification is potentially harmful to the representation of the urban processes under inquiry because density and accessibility have environmental, economic and social consequences.
+However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic density.
+Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents \replaced{tend to be}{is still} approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}.
+The isotropic simplification is potentially harmful to the representation of urban processes \deleted{under inquiry }because density and accessibility have environmental, economic and social consequences.
 Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment.
-For example, households can have different preferences with respect to the built-environment they want to live in \citep{SpielmanHarrison2014}, or agents can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}.
-The way modellers represent space is therefore a central element of geographical simulation models.
+For example, households can have different preferences with respect to the built-environment they \added{might} want to live in \citep{SpielmanHarrison2014}, or agents \replaced{moving around will sense a different set of objects within the same fixed radius depending on}{can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on} the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}.
+The way modellers represent \added{initial} space is therefore a central element of geographical simulation models.
 However, this step is rarely explicit.
-For example
+\deleted{For example
 
 and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one.
-A meaningful way of representing space might thus be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour in plausible spatial arrangements.
+}A meaningful way of representing \added{initial} space might \deleted{thus }be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour \replaced{to different}{in} plausible spatial arrangements.
 
@@
 In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape).
-In no way do we want to provide a full exploration of these two particular models, their attractors and/or potential policy implications.
-Instead, we present a way of performing a sensitivity analysis to the spatial conditions of models by using a spatial generator to produce a variety of density grids, which are then fed into simulation models at initialisation.
-The generator being controlled by \replaced{it own}{meta-} parameters, we can then relate the parameters used to generate initial spatial conditions to the variation in outcome of the simulations.
+In no way do we \replaced{pretend}{want} to provide a full exploration of these two particular models, their attractors and/or potential policy implications.
+Instead, we present a way of performing a sensitivity analysis to \replaced{initial}{the} spatial conditions of models by using a spatial generator to produce a variety of density grids, which are \replaced{taken as input by}{then fed into} simulation models at initialisation.
+The generator being controlled by \replaced{its own}{meta-} parameters, we can then relate the parameters used to generate initial spatial conditions to the variation \replaced{simulation outcomes}{in outcome of the simulations}.
 The purpose is two-fold: (i) to test the robustness of simulation results to small variations of \replaced{generator}{meta-} parameters and (ii) to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
-Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies.
+Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies \added{to our knowledge}.
 In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.
@@
 The internal variabilities are estimated as the variance within each phase diagram $f_{\vec{\alpha_i}}$.
-For a multi-dimensional phase diagram, we sum these relative distances over the components.
+For a multi-dimensional phase diagram, we \replaced{average}{sum} these relative distances over the components.
 
@@
 
-\section{Conclusion.}
+\section{Conclusion}
 
@@
 %% Schelling
+
+For both models we estimate the Sharpe ratios for each indicator by
 

### 2018-09-27 paper.tex
Update on Overleaf.

@@
 For example, \citet{wheeler2006urban} shows that, in the US, sprawling cities are more unequal than their compact counterparts with respect to income.
-Dynamically, sprawl consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}.
+Dynamically, sprawl \added{in American cities} consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}.
 Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}.
@@
 
-Considerations of such issues in the geographical modelling literature are rather scarce.
-Despite the opportunity of simulation in the representation of space and its potential observation biases, there have been some noticeable attempts to evaluate the effects of the same set of mechanisms run from different initial spatial conditions, which have focused on the impact of:
+Considerations of such issues in the \replaced{geosimulation}{geographical modelling} literature are rather scarce.
+\replaced{However,}{Despite the opportunity of simulation in the representation of space and its potential observation biases,} there have been some noticeable attempts \replace{at analysing the impact of three types of initial spatial characteristics on model outcomes}{to evaluate the effects of the same set of mechanisms run from different initial spatial conditions, which have focused on the impact of}:
 \begin{itemize}
@@
 
-\paragraph{Geo-localised input data accuracy} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and scale of analysis.
-They provide a few examples of how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes.
+\paragraph{Geo-localised input data accuracy} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and \added{the} scale of analysis.
+They provide a few examples on how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes.
 In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and the computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.
@@
 \citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins.
-Similar issues were also dealt with in physical sciences, as for example \cite{horritt2001effects} studies the effects of grid cell size on the behaviour of a raster flood model and shows that increasing resolution does not increase model prediction performance under a certain limit.
+Similar issues were also dealt with in physical sciences: for example, \cite{horritt2001effects} studies the effects of grid cell size on the behaviour of a raster flood model and shows that increasing resolution does not increase model prediction performance \replaced{below a certain level}{under a certain limit}.
 Similar conclusions are obtained by \cite{vazquez2002effect}, unveiling an intermediate optimal spatial resolution regarding model performance and computation time.
-Spatial resolution also plays a role in the Schelling model: \citet{Singhetal2009} show that the segregation patterns for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinaisons of tolerance threshold and vacancy density values.
+Spatial resolution also plays a role in the Schelling model: \citet{Singhetal2009} show that the segregation patterns for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinations of tolerance threshold and vacancy density values.
 
@@
 In ecology, \citet{smith2002predicting} study the spread of a disturbance in an heterogeneous landscape using a percolation model, and show that landscape structure has a significant influence on final patterns of contamination outcome.
-In a spatial epidemics model for an infectious disease, parametrised on real data, \citet{smith2002predicting} finds that the physical landscape heterogeneity, in particular the presence of rivers, locally influence the propagation speed.
+In a spatial epidemics model for an infectious disease, parameterised on real data, \citet{smith2002predicting} finds that the physical landscape heterogeneity, in particular the presence of rivers, locally influence the propagation speed.
 
- Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.) \replaced{or}{either} on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time, \added{offering a systematic inquiry into the impact of systematically defined spatial inputs on the model outputs.
-More so, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.}
+Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.) \replaced{or}{either} on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
+\added{More so, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.}
 
@@
 In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on \replaced{its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
-We call these parameters ``generator parameters'' to distinguish from the standard parameters of the models and highlight their higher position in the simulation process}{meta-parameters and feeds the model $\mu$ at initialisation (lower branch)}.
+We call these parameters ``generator parameters'' to distinguish them from the standard parameters of the models and to highlight their higher position in the simulation process}{meta-parameters and feeds the model $\mu$ at initialisation (lower branch)}.
 The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.
-The sensitivity analysis relates to the model's parameters to how the density spatial distribution was generated and the pattern of density generated in order to produce qualitative knowledge from the model and its behaviour.
+The sensitivity analysis relates the \added{variations in the} model's \replaced{outcomes}{parameters} to how the density spatial distribution was generated and \added{to} the patterns of density generated.
+\deleted{in order to produce qualitative knowledge from the model and its behaviour.
 In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space.
@@
 
+%%%%%%%%%%%%%%%
 \section{\added{Comparison of phase diagram with other distances}}
 
-\added{We describe here the tests done with other distances to compare phase diagrams.}
+\added{We describe here the tests done with other distances to compare phase diagrams.
+We tested normalized Minkovski distances, defined by $d(x,y) = \left(\frac{1}{N}\cdot \sum_i \left|x_i - y_i\right|^{\alpha}\right)^{\frac{1}{\alpha}}$, for varying values of $\alpha$ from $\alpha = 1$ (Manhattan distance) to $\alpha = 1000$ (close to the infinite norm), including $\alpha = 2$ (Euclidian distance) which is}
 
+% Schelling
+% [1] "alpha = 1"
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%0.00000 0.02838 0.06263 0.10608 0.16818 0.37018 
+%[1] "alpha = 2"
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+% 0.0000  0.1520  0.2517  0.3107  0.4359  0.8155 
+%[1] "alpha = 3"
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+% 0.0000  0.3709  0.5133  0.5860  0.7435  1.2930 
+%[1] "alpha = 10"
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%  0.000   2.083   2.431   2.380   2.713   3.664 
+%[1] "alpha = 1000"
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%  0.000   3.481   3.868   3.960   4.262   6.781
+
+%%%%%%%%%%%%%%%
 \section{\added{Pseudo-code for models}}

### 2018-09-28 Response_to_Reviewers.tex
Update on Overleaf.

--- 2018-09-19 Response_to_Reviewers.tex
@@
 Furthermore, we have focused on some of the most classical models of the ABM literature so as to lower the reader's cost of entry into the methodology.
-    %We believe this is sufficient as the models have not been adjusted from these textbook cases implementations.
 
    % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
-   \item Concerning basic illustration of the influence of initial patterns, we think that it is not possible with single runs to show a significant effect other than the direct effect of space on the localization of agents for example.
-It is indeed the point of our method that such tests must be done in a systematic way rather than ``eyeball validations''.
-   %\comment[JR]{Possibility to find configuration with an ``eyeball'' difference ?
-i'm not sure of that..}[(CC)peut-être en prenant des cas extrêmes sur Sugarscape?][(JR) je regarde...
-$\rightarrow$ bof, on voit rien a part que les agents se repartissent differement mais qui est une consequence directe d'ajouter l'espace, en en comparant les valeurs finales des ginis, ce qu'on fait plus bas plus rigoureusement.
-je suggère poliment que sa suggestion est idiote.]
+   \item We agree that illustrative examples would provide a relevant starting point to  the analysis but we were afraid that their addition to the paper would make the introduction less straightforward and have the opposite effect to the one we were seeking.
+As we were nevertheless very interested by your remark, we provide a short answer to your proposition.
+
+In order to get a first idea of the potential impact of spatial initial conditions on to the results of the simulation, we implemented on Segregation model a geographical constraint: some cells cannot be inhabited (yellow cells in Figure \ref{Fig1}).
+The way those “forbidden” cells are distributed throughout space will indeed influence the steps of the model.
+We introduced various spatial initial conditions in the model that we controlled by one clustering parameter on the forbidden cells.
+A value of 0 would indicate a spatial distribution of yellow cells so as to minimize spatial autocorrelation.
+A value of 1 indicates on the contrary a concentration process that tends towards a high value of autocorrelation (see below Figure \ref{Fig1}).
+
+We then compute the number of communities arising when the model converge, hence the number of different related components of similar neighbors.
+Using 270 replications of 11 values of the clustering parameter (from 0 to 1 by step 0.1), we obtained a significant growth in terms of number of communities when the constraints on residential locations are concentrated in space, which might be interpreted in such example as a bigger segregation when density of housing is heterogeneously distributed through space.
+%In the rest of the article, we wish to further investigate such spatial effects in a much more systematic way.
+
+\begin{figure*}[htbp] 
+\begin{center} 
+\resizebox{0.9\textwidth}{!}{ 
+	\includegraphics{figures/SpaceMatters2.bmp}
+} \caption{\textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (implementation on Netlogo).}} \label{fig:Fig1}
+\end{center}
+\end{figure*}
 
@@
 	\item There was a misunderstanding for the Schelling implementation actually used, as we used our own implementation which pseudo-code has been added in Appendix of the paper and source code is openly available on the git repository of the project.
-This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.% It does not reach a segregated state for high values of intolerance, but it does not cause a problem in our case as indicators can be computed on any model state (and actually reflect the fact that it is not segregated).
+This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.
+% It does not reach a segregated state for high values of intolerance, but it does not cause a problem in our case as indicators can be computed on any model state (and actually reflect the fact that it is not segregated).
 	 %\comment[CC]{le modèle ne produit pas de ségrégation quand les individus sont très intolérants?
@@
 This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs.
-This is again an argument in favour of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.
+This is again an argument in favor of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.
 
@@
 Our choice was accordingly clarified in the paper.
-%: "In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants." % pas besoin de mettre les modifs ici
 
 	% 4. Consider replacing “meta parameters” with something like “generator parameters” or “initial pattern parameters.”
-	%In order to clarify the concept of ``meta-parameter'' that we prefer to keep, we now propose a short definition after the first use of the term (section Methods): "In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
-We call these parameters meta-parameters to distinguish from the standard parameters of the models and highlight their higher position in the simulation process".
 	\item We agree that the term of meta-parameters may have been confusing, and following the suggestion we replaced it with ``generator parameters''.
-     \comment[JR]{TB la def Marion ; là on peut choisir de virer en effet le terme, ou dire que les 3 sont interchangeable mais que "meta-" est plus pratique, qu'en pensez vous ?}
-    \comment[CC]{d'acc.}
-    \comment[MLT]{Du coup, dans nos échanges par mails on disait qu'on était ok pour changer, mais finalement cette solution vous convient?
-Je comprends pourquoi il/elle a tiqué sur le terme, mais c'est pas non plus hors de propos vu l'application qu'on en fait...
-bref, je suis partagée!}[(JR) moi ca me va de virer le meta]
 
 	% 5. Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index
-	\item \comment[JR]{je pensais qu'on prenait la moyenne ? je verifie... ligne 175 on écrit la médiane en effet}
-	\comment[CC]{ça doit être moi parce que je pensais qu'on prenait effectivement la médiane, qui est plus robuste aux extrêmes, mais si c'est la moyenne et que ça reste proche entre différentes réplications why not.}[(JR) il me semble que les distribs sont assez centrées et que ca revient quasiment au même ; par contre Clem si tu as pris median pour les stats et moi mean pour les plots c'est pas fou ; tu as encore le code des analyses stats ?
+    %\comment[JR]{je pensais qu'on prenait la moyenne ? je verifie... ligne 175 on écrit la médiane en effet}
+	%\comment[CC]{ça doit être moi parce que je pensais qu'on prenait effectivement la médiane, qui est plus robuste aux extrêmes, mais si c'est la moyenne et que ça reste proche entre différentes réplications why not.}[(JR) il me semble que les distribs sont assez centrées et que ca revient quasiment au même ; par contre Clem si tu as pris median pour les stats et moi mean pour les plots c'est pas fou ; tu as encore le code des analyses stats ?
 (c'est sur le git ?)]
-	\comment[JR]{c'est la merde, on prend certaines fois la mediane et certaines fois la moyenne...}
+	%\comment[JR]{c'est la merde, on prend certaines fois la mediane et certaines fois la moyenne...}
+    %\comment[FL]{certes c'est pas fou mais ca peut se défendre en disant que la médiane sert pour communiquer de façon accessible, sur une dimension pour laquelle c'était pertinent, mais que de façon générique on a préféré travailler sur la loi de Gauss car tout le monde à en tête ces indicateurs moyenne / ecartype c'est plus "passe partout"?
+}
 
@@
 Would we get qualitatively different results from different indices?
-	\item \comment[JR]{on avait teste avec le Monge ca donnait a peut pres pareil ; donc les normes alpha surement encore plus. je regarde.}
+	\item The approach proposed in the paper is indeed rather exploratory.
+This explains why we have chosen to focus on a single index (L2) as this index is relatively well-known in the community.
+Previous exploratory analyses have shown that the results were rather stable when using the Monge method.
+Adapting the mathematical aggregator to the different indices would constitute an interesting perspective, but is out of the scope of the paper which already focuses on two models and proposes a new framework for analysing the impact of initial pattern parameters.
 
@@
 	\item In the case of Schelling, this is the scale at which most empirical segregation indices are computed.
-In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty....), or to a generic landscape where a resource is grown otherwise.
+In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty, 2005), or to a generic landscape where a resource is grown otherwise.
 In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density, and that not acknowledging this diversity leads to potentially false results.
-Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most harmful one (e.g.
-the least realistic, although this is the case of most Schelling models).
-	%\comment[CC]{Ref du papier de batty à ajouter (je ne l'ai pas trouvée}[(c'est le Batty, 2005)]
+Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most common simplification of the heterogeneity of space, with consequences on the model outputs often not evaluated.
 
@@
 The satisfaction and segregation indices are computed with regard to the people in the cell and the people present in neighbouring cells.
-	%\comment[JR]{il a mal lu ou on est pas assez explicites, a detailler}
-	%\comment[CC]{ça va comme réponse? }[(JR) parfait]
-	%\comment[CC]{argumentaire à ajouter dans le papier}[(cc)done]
 
@@
 We chose to use actual distribution of European cities to constraint our density generation.
-%\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
 
 	% 12. Line 641: NetLogo supports grids, continuous space, and GIS elements. It does not restrict the user in any way.
-    % \comment[JR]{plutot d'accord avec lui mais parce que proficient en NetLogo, peut etre que des debutants sont plus contraints ? quelquun aurait une ref là dessus ?}
-    %\comment[MLT]{En fait, c'est surtout le cas des pre-built simulations proposées dans la librairie de NetLogo.
-J'ai un peu cherché, je trouve pas de références qui montrent explicitement la surabondance des grilles dans les modèles hors librairie...}[(JR) ok yes en effet !
-cela devrait suffire comme argument ?]
 	\item We agree that NetLogo does not restrict the use on what implementation to use.
@@
 We clarified this point in the introduction.
-    %\comment[JR]{il s'agirait de nuancer ici}
-	%\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
 
@@
 I am thinking of work in the line of disease diffusion, crime etc.
-	\item \comment[JR]{il a un point car ce sont des classes de modèles qu'on a pas trop cité ?
-par contre la sensibilite a ces distribs precises n'est pas testée, ou pas systeématiquement : il faut qu'on clarifie ou reformule}
-    \comment[MLT]{En fait, c'est là aussi une question de reformulation.
-Je crois qu'il faudrait répondre quelque chose comme : "We acknowledge the fact that many ABM applications use "real" data and spatial features for their simulations.
+	\item We acknowledge the fact that many ABM applications use "real" data and spatial features for their simulations.
 However, as explained by (Thomas et al.
-2017) in the case of LUTI models for instance, most simulations do not evaluate the impact of city delineations nor MAUP on the output of the models, which constitutes a potential limit in the interpretation of results.}{(JR) ca me va comme reformulation}
+2017) in the case of LUTI models for instance, most simulations do not evaluate the impact of city delineations nor MAUP on the output of the models, which constitutes a potential limit in the interpretation of results.
 
@@
 Many agent-based models are also not valid for other areas.
-	\item \comment[JR]{idem la on parle pas exactement du meme type de modeles, mais son exemple prouve notre propos : le modele tourne sur une situation unique, et on teste pas l'effet de l'espace.}
-    \comment[MLT]{Bon, c'est clairement quelqu'un qui fait de l'appliqué et qui veut pas entendre notre argument ;) On peut aller dans son sens et modifier notre propos en écrivant : "However, this step is rarely explicit and authors are therefore not able to generalize the conclusion of the model to other spatial entities."}
+	\item What we meant was actually that this step is rarely explicit and authors are therefore not able to generalize the conclusion of the simulations to other spatial context.
+We rephrased the paragraph accordingly.
 
@@
 I think a clear differentiation between the two groups and the positioning of the two case studies in the second group would make the paper more balanced.
-	\item Concerning the remarks on synthetic population generation and the different types of models, we clearly stated our positioning regarding models built from theory, for which the proposed sensitivity analysis workflow is more suited and that our two case studies illustrate.\comment[JR]{oui du coup en disant ça ça repond à ses deux points précédents}\comment[JR]{positionement a plus souligner dans le papier}
+	\item Concerning the remarks on synthetic population generation and the different types of models, we clearly stated our positioning regarding models built from theory, for which the proposed sensitivity analysis workflow is more suited and that our two case studies illustrate.
 
@@
 	\item Indeed, we have tried to review articles where some kind of sensitivity to initial spatial conditions are tackled.
-But first, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.
-Second, the references cited mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but none provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper.
+However we would like to emphasize that the references cited in this section are few compared to the mass of geosimulation articles published with hardly any mention of dependence of output on initial spatial condition.
+Moreover, even if such sensitivity is analysed, it mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but to the best of our knowledge no article provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper.
 We have tried to argue this point at the end of the section.
-    %\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]
 
@@
 I wonder if the Euclidean distance is the best measure to determine the difference between the phase vectors of different spatial inputs.
-	% -> 
 	\item The comment on other possible techniques to compare phase diagrams echoes one other reviewer comment, and we indeed expanded the discussion on this point, showing first that other distance do not change much qualitative results and with more literature for possible extensions.
@@
 En termes thematique c'est plus coherent pour nous, lui doit avoir en tete des indices de sensibilite globaux type Saltelli.
-$\rightarrow$ a voir comment justifier ca bien}{(JR) inclure en mise en contexte ce lien avec les Saltelli et indexes globaux}
+$\rightarrow$ a voir comment justifier ca bien}{(JR) inclure en mise en contexte ce lien avec les Saltelli et indexes globaux}\comment[FL]{oui je trouve qu'on a le droit de rester limiter à cela c'est déjà pas mal!}
 
 	% Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.
-
-	\item Color scale for distance was reversed.
-    \comment[JR]{Clem tu es d'accord là dessus ?}[(CC) Non mais il chipote complètement à ce niveau!
-Vu la figure pour moi c'est totalement équivalent dans un sens et dans l'autre, le tout c'était qu'il y'a une double descente chromatique et y'a même pas de densité = rouge ou quoi puisque c'est un rapport.
-Si tu veux lui faire plaisir on change mais je vois pas trop la nécessité.][(JR) bon je vais refaire la figure pour caresser dans le sens du poil - TODO]
+    \item The double color progression in Fig 4 and 5 is used to highlight the opposition between ratio values above/under 1 but is not meant to introduce any hierarchy in the interpretation.
+We would therefore prefer to keep the figures as they are if this is ok for you?
 
@@
 	\item We have tried to make the methodology and result section flow better in the corresponding sections.
-    \comment[CC]{à faire en derniere relecture}
 
@@
 	\item We have agreed with the reviewer on ``validation'' and removed the concept of validation altogether, replacing it with model evaluation or sensitivity analysis.
-    %\comment[JR]{enlever le concept de validation}
 
@@
 	\item We have worked towards improving the English along with the overall argument.
-      \comment[CC]{ok pour cette phrase + derniere relecture a faire}
 
@@
 	\item 
-    \comment[JR]{cf remarques other reviewers sur ce point}[(MLT) Sa remarque fait sens, mais j'avoue ne pas être trop au point sur cette partie de la littérature...
+%    \comment[JR]{cf remarques other reviewers sur ce point}[(MLT) Sa remarque fait sens, mais j'avoue ne pas être trop au point sur cette partie de la littérature...
 vous avez des références?][(JR)concernant la spatial sensitivity : justement c'est pas trop fait non plus dans les autres champs !
@@
 We have clarified this point in section 2.3
-    %\comment[JR]{on regarde des indicateurs aggreges qui ne dependent pas de la trajectoire de chaque agent, si le placement initial est random il na aucune influence sur ceux-ci (cf les distribs) -> ajouter des graphes de comportement stat des modeles en Appendix ?}
-	%\comment[CC]{à modifier dans le papier section 2.3}[(cc) done]
 
@@
 We have clarified this in section 3.2.
-    %\comment[CC]{à modifier dans le papier section 3.2}[(cc) done]
 
@@
 You touch upon shape when talking about NetLogo's use of raster as opposed to vector in the Discussion - could you expand on this?
-	\item We have not addressed the shape of cells and the uncertainty of data in this paper, only the diversity of shapes of urban morphologies (e.g. heterogeneity).
+	\item We thank the reviewer for this good remark.
+We provide references to papers addressing these issues.
+However, we have not addressed the shape of cells and the uncertainty of data in this paper, only the diversity of shapes of urban morphologies (e.g.
+heterogeneity).
 
@@
 by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).
-	\item To the best of our knowledge, there does not exist a study of urban morphologies similar to the one we used but at a global scale (this should be linked to the absence of a raster of population with a resolution fine enough).
+	\item To the best of our knowledge, there is no study of urban morphologies similar to the one we used but at a global scale (this should be linked to the absence of a raster of population with a resolution fine enough).
 Working on other regional systems of cities such as the US could be possible but remains out of the scope of this paper.
 We added this interesting suggestion to the discussion.
-	%\comment[JR]{to the best of my knowledge there is no full-world morphological classification of cities (and surely because global 1km density dataset is not robust enough for that - already with the European grid it is not perfect)}[(MLT) Peut-être qu'on peut intégrer sa remarque dans la partie discussion?
-Je ne connais pas non plus de classification morphologique des villes à l'échelle du monde.
-Les cas européens sont privilégiés du fait de la disponibilité de données comme Corine Land Cover et Urban Atlas (pour l'usage du sol) ou Geostat (grilles de densité de pop de résolution 1km), mais ça doit aussi exister pour les US au moins?
-Il y a aussi le projet gridded population of the world de l'université de Columbia mais la résolution n'est pas assez fine (2.5 arcminute) pour une analyse de la morpho urbaine.]
 
 	% Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?
-	\item \comment[JR]{no justification for the difference in number of runs (just experimental setup contingence\ldots ) $\rightarrow$ either new runs or redo computations with 50 for Schelling}\comment[JR]{ca va etre chaud patate pour les new computations ; je vais regarder les ratio de sharpe, avec un peu de chance il y a besoin de plus de repets pour schelling...}
+	\item 
+    %\comment[JR]{no justification for the difference in number of runs (just experimental setup contingence\ldots ) $\rightarrow$ either new runs or redo computations with 50 for Schelling}\comment[JR]{ca va etre chaud patate pour les new computations ; je vais regarder les ratio de sharpe, avec un peu de chance il y a besoin de plus de repets pour schelling...}
 

### 2018-09-28 paper.tex
Update on Overleaf.

@@
 Considerations of such issues in the \replaced{geosimulation}{geographical modelling} literature are rather scarce.
-\replaced{However,}{Despite the opportunity of simulation in the representation of space and its potential observation biases,} there have been some noticeable attempts \replace{at analysing the impact of three types of initial spatial characteristics on model outcomes}{to evaluate the effects of the same set of mechanisms run from different initial spatial conditions, which have focused on the impact of}:
+\replaced{However,}{Despite the opportunity of simulation in the representation of space and its potential observation biases,} there have been some noticeable attempts \replaced{at analysing the impact of three types of initial spatial characteristics on model outcomes}{to evaluate the effects of the same set of mechanisms run from different initial spatial conditions, which have focused on the impact of}:
 \begin{itemize}
@@
 The sensitivity analysis relates the \added{variations in the} model's \replaced{outcomes}{parameters} to how the density spatial distribution was generated and \added{to} the patterns of density generated.
-\deleted{in order to produce qualitative knowledge from the model and its behaviour.
-In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space.
+\deleted{in order to produce qualitative knowledge from the model and its behaviour.} In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space.
 In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
@@
 We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment).
-A solution is to use systematic quantitative procedures to compare them.
-Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
+A solution is to use systematic quantitative procedures to compare them \added{to a reference case}.
+Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median \added{?} of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
 
-To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.
+To our knowledge there exists no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.
 We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$.
@@
 where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given \replaced{generator}{meta-} parameters $\vec{\alpha}_i$.
-$d$ is a functional distance that we take simply as Euclidian distance.
+$d$ is a functional distance that we take simply as Euclidean distance.
 The internal variabilities are estimated as the variance within each phase diagram $f_{\vec{\alpha_i}}$.
@@
 
-The last methodological point which we need to emphasize is the relation between the present workflow and model exploration workflows in general.
-The ideas of multi-modeling and extensive model exploration are nothing from new as ~\cite{openshaw1983data} already advocated for ``model-crunching'', but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities.
-The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write modulable exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
+The last methodological point which we need to emphasize is the relationship between the present workflow and model exploration workflows in general.
+The ideas of multi-modelling and extensive model exploration are nothing from new - ~\cite{openshaw1983data} already advocated for ``model-crunching'' \added{in 1983 -}, but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities.
+The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write \replaced{flexible}{modulable} exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
 In our case, this tool is a powerful way to embed both the sensitivity analysis and the \deleted{meta-}sensitivity analysis \added{to initial spatial conditions}, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file.
@@
 The spatial generator applies an urban morphogenesis model \citep{Batty2007} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}.
-An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily in our workflow.
-Grids are generated through an iterative process which, starting from a void grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
+An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily into our workflow.
+Grids are generated through an iterative process which, starting from an empty grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
 More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
-At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (8 surrounding patches).
+At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (\added{i.e.
+its} 8 surrounding patches).
 To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside.
@@
 \cite{LeNechet2015}).
-We sample the \replaced{generator}{meta-} parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a point cloud with high discrepancy.
+We sample the \replaced{generator}{meta-} parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a \replaced{scatter}{point cloud} with high discrepancy.
 We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}.
@@
 \added{The spatial generator and its resulting grids are relevant to the case study models we have picked (Schelling and Sugarscape) because it produces density grids at a ``metropolitan scale'', which is the scale at which both model were initially intended to be.
-In the case of Schelling's segregation model for example, this scale is the one at which most empirical segregation indices are computed and compared to the model outputs.
+In the case of Schelling's segregation model for example, this scale is the one at which most empirical segregation indexes are computed and compared to the model outputs.
 In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources \citep{batty2005agents}, or to a generic landscape where a resource is grown otherwise.
@@
 It has been shown by \cite{Schelling1969} that even tolerant agents tend to produce segregated patterns because of the complexity of their local interactions and the snowball effect of individual moves on the global distribution of agents in the city.
-The main parameters of this model are the tolerance level (maximum \% of agents different to {\it ego} accepted in the neighbourhood), the scope of sensing, the minority split and the percentage of vacant spaces in the housing market.
+The main parameters of this model are the tolerance level (maximum \% of agents different to {\it ego} accepted in the neighbourhood), the scope of sensing, the \added{global majority/}minority split and the percentage of vacant spaces in the housing market.
 In addition, we are interested in testing the impact of the initial spatial distribution of housing capacity in this project, using the generated grids.
-The outcome of the model is measured by a combination of three segregation indices: Dissimilarity, Moran's Index, Entropy~\citep{brown2006spatial}.
+The outcome of the model is measured by a combination of three segregation indices: Dissimilarity, Moran's I and Entropy~\citep{brown2006spatial}.
 \replaced{We use a ad-hoc implementation of the Schelling model, both in Scala for performance reasons and in NetLogo to ensure visualization of model dynamics.
@@
 \added{In general, the implementations of Schelling models allow only one agent per cell, and their initial distribution is random, therefore following a uniform distribution across the modelled city.
-In this experiment, we allow more than one agent to be in a given same cell.
+In this experiment, we allow more than one agent to be in a given cell.
 The potential density of a cell is defined by the density grid generated.
@@
 Empirical distributions of density in cities are important in our framework because we want to test models with realistic ranges of initial patterns of density distribution.
-Therefore we cannot limit ourselves to a single 2-peak mountain in the classical Sugarscape model and an isotropic square city in Schelling.
-We chose to use actual distribution of European cities to constraint our density generation.}
+Therefore we cannot limit ourselves an isotropic square city in Schelling.
+We chose to use the actual distributions of European cities to constraint our density generation.}
 
@@
 \added{In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants.
-Following \cite{batty2005agents}, it is a good metaphor of an urban system.} We extend the implementation with agents wealth distribution of \cite{li2009netlogo}.
+Following \cite{batty2005agents}, it can be considered as a metaphor of an urban system.} We extend the implementation with agents wealth distribution of \cite{li2009netlogo}.
 The outcome of the model is measured by a Gini index of inequality for resource distribution.
@@
 We choose different experiment designs, both for \replaced{generator parameters}{meta-parameters} and for the phase diagram, to demonstrate the robustness of the method to technical choices.
-In principle, our workflow applies with to manner to generate a spatial configuration (even taking real configurations) and any manner to establish phase diagrams.
+In principle, our workflow applies \replaced{regardless of the way we}{with to manner to }generate a spatial configuration (even taking real configurations) and \replaced{the way we}{any manner to} establish phase diagrams.
 
@@
 
-\added{The implementations of the models were done in NetLogo.
+\added{The implementations of the models were done from NetLogo.
 We modified the sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model.
@@
 For each density grid, we obtain the average squared distance between corresponding points of the phase diagrams, i.e.
-the median value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated).
+the median\added{?} value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated).
 This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry).
@@
 
-Another way of quantifying the density grids, instead of looking at the \replaced{generator parameters}{meta-parameters of their generator}, is to look at the resulting indicators of urban form, such as Moran index, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and context).
+Another way of quantifying the density grids, instead of looking at the \replaced{generator parameters}{meta-parameters of their generator}, is to look at the resulting indicators of urban form, such as Moran's I, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and context).
 This 4-dimensional space defined a morphological space.
@@
 In sprawled spaces, inequalities are thus fostered by a lack of minimal local resources, whereas population will drive inequality in concentrated spaces.
-The process is thus completely reversed depending on the grid chosen to run the model, which would have significant impacts if one tried to deduce policies from this model.
+The process is thus completely reversed depending on the grid chosen to run the model \added{on}, which would have significant impacts if one tried to deduce policy from this model.
 %This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
@@
 The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\citep{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
-Map comparison methods, popular in environmental sciences, provide numeral tools to compare two dimensional fields~\citep{visser2006map,kuhnert2005comparing}.
+Map comparison methods, popular in environmental sciences, provide numerous tools to compare two dimensional fields~\citep{visser2006map,kuhnert2005comparing}.
 To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\citep{10.1371/journal.pone.0178165}.
 \added{The transfer of methods used to compare sequences \citep{kruskal1983overview} or time-series \citep{liao2005clustering} is a possible way to develop measures between phase diagrams.
-The higher dimension of the phase diagram we study must however be considered with caution when transferring method, in a way analog to the application of global sensitivity indices to spatial data \citep{lilburne2009sensitivity}.} The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.
+The higher dimension of the phase diagrams we study must however be considered with caution when transferring methods, in a way analog to the application of global sensitivity indices to spatial data \citep{lilburne2009sensitivity}.} The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.
 
@@
 An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
-For example, \replaced{spatial structure may be easier to implement as a raster rather a vector in NetLogo models, what could influence the implementation choices of some non-experienced modelers}{the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements}.
+For example, \replaced{spatial structure may be easier to implement as a raster rather a vector in NetLogo models, which could influence the implementation choices of some non-experienced modelers}{the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements}.
 Its toroidal default setting might also have influenced the work of many modellers who did not question explicitly the representation of space.
@@
 As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions.
-Geography and the social sciences in general have been strongly influenced by physics in the last decades, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space.
+Geography and the social sciences in general have been strongly influenced by physics in the last decades, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space \added{in basic models}.
 We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity the for social sciences to build their own stream of methodologies in the modelling domain.
@@
 %% Schelling
+% Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%  1.053  12.368  19.434  25.268  24.328 468.563 
+%> summary(sres$entropySharpe)
+%    Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
+%  0.5458   8.2591  14.5112  17.0100  17.9372 244.3230 
+%> summary(sres$moranSharpe)
+%    Min.  1st Qu.   Median     Mean  3rd Qu.     Max. 
+%  0.0001   0.5554   0.7782   3.2724   3.8979 121.6829
 
-For both models we estimate the Sharpe ratios for each indicator by
+For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\Eb{X}}/\hat{\sigma}(X)$ with standard estimators for average and standard deviation.
+The summary statistics computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}.
+
+%%%%%%%%%%%%%%
+\begin{table*}
+\caption{\added{\textbf{Summary statistics of Sharpe ratio estimated on repetitions for each parameter point.}}\label{tab:summarysharpes}}
+	\begin{tabular}{|cccccc|}
+Model/Indicator & Min. & 1st Qu. &  Median &  Mean & 3rd Qu. & Max. \\\hline
+Sugarscape/Gini & 2.343 & 14.413 & 20.993 & 20.098 & 26.208 & 49.808\\\hline 
+Schelling/Dissimilarity & 1.053 & 12.368 & 19.434 & 25.268 & 24.328 & 468.563\\
+Schelling/Entropy & 0.5458 & 8.2591 & 14.5112 & 17.0100 & 17.9372 & 244.3230\\
+Schelling/Moran & 0.0001 & 0.5554 & 0.7782 & 3.2724 & 3.8979 & 121.6829\\\hline
+    \end{tabular}
+\end{table*}
+%%%%%%%%%%%%%%
 
@@
 \added{We describe here the tests done with other distances to compare phase diagrams.
-We tested normalized Minkovski distances, defined by $d(x,y) = \left(\frac{1}{N}\cdot \sum_i \left|x_i - y_i\right|^{\alpha}\right)^{\frac{1}{\alpha}}$, for varying values of $\alpha$ from $\alpha = 1$ (Manhattan distance) to $\alpha = 1000$ (close to the infinite norm), including $\alpha = 2$ (Euclidian distance) which is}
+We tested normalized Minkovski distances, defined by $d(x,y) = \left(\frac{1}{N}\cdot \sum_i \left|x_i - y_i\right|^{q}\right)^{\frac{1}{q}}$, for varying values of $q$ from $q = 1$ (Manhattan distance) to $q = 10$, including $q = 2$ (Euclidian distance) which is used in main text.}
 
@@
 %  0.000   2.083   2.431   2.380   2.713   3.664
-%[1] "alpha = 1000"
+
+% Sugarscape
+%[1] "alpha = 1"
 %   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
-%  0.000   3.481   3.868   3.960   4.262   6.781
+%0.07184 0.13996 0.39816 0.42452 0.64751 1.13169 
+%[1] "alpha = 2"
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+%0.08909 0.19794 1.52272 1.29594 2.16371 2.98273 
+%[1] "alpha = 3"
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+% 0.1025  0.2288  3.1495  2.5074  4.2718  5.5378 
+%[1] "alpha = 10"
+%   Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
+% 0.1540  0.4466  9.8757  7.6246 13.0550 16.8001
+
+\begin{table*}
+	\caption{\added{\textbf{Summary statistics of different distances for schelling.}}}
+	\begin{tabular}{|ccccccc|}
+    	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
+        1 & 0.00000 & 0.02838 & 0.06263 & 0.10608 & 0.16818 & 0.37018 \\
+        2 & 0.0000 & 0.1520 & 0.2517 & 0.3107 & 0.4359 & 0.8155 \\
+        3 & 0.0000 & 0.3709 & 0.5133 & 0.5860 & 0.7435 & 1.2930 \\
+        10 & 0.000 & 2.083 & 2.431 & 2.380 & 2.713 & 3.664 \\\hline
+    \end{tabular}
+\end{table*}
+
+\begin{table*}
+	\caption{\added{\textbf{Summary statistics of different distances for sugarscape.}}}
+	\begin{tabular}{|ccccccc|}
+    	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
+        1 & 0.07184 & 0.13996 & 0.39816 & 0.42452 & 0.64751 & 1.13169\\
+        2 & 0.08909 & 0.19794 & 1.52272 & 1.29594 & 2.16371 & 2.98273\\
+        3 & 0.1025 & 0.2288 & 3.1495 & 2.5074 & 4.2718 & 5.5378\\
+        10 & 0.1540 & 0.4466 & 9.8757 & 7.6246 & 13.0550 & 16.8001\\\hline
+    \end{tabular}
+\end{table*}
 

### 2018-09-28 Response_to_Reviewers.tex
Update on Overleaf.

@@
     \item In order to conciliate as much as possible the different points of view about the comparison of phase diagram, we added some developments and comments about how we defined, built, justified and compared phase diagrams in the paper.
-    %\comment[FL]{je ne sais pas si dire "conflict" sera productif?, de même que de parler de "parties"? ci-dessous en commentaires la réponse précédente, j'ai juste reformulé}[(JR) ok nickel]
-    % \item There was a conflict between the comments of the third referee and the other referees about the comparison of phase diagram.
-In order to conciliate as much as possible the different parties, we added some developments and comments about how we defined, built, justified and compared phase diagrams in the paper
     \end{itemize}
@@
 
-\begin{figure*}[htbp] 
+%%%%%%%%%%%%%%
+\begin{minipage}{\textwidth}
 \begin{center} 
@@
 	\includegraphics{figures/SpaceMatters2.bmp}
-} \caption{\textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (implementation on Netlogo).}} \label{fig:Fig1}
+} \textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (implementation on Netlogo).} \label{fig:Fig1}
 \end{center}
-\end{figure*}
+\end{minipage}
+%%%%%%%%%%%%%%
 
@@
 This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.
-% It does not reach a segregated state for high values of intolerance, but it does not cause a problem in our case as indicators can be computed on any model state (and actually reflect the fact that it is not segregated).
-	 %\comment[CC]{le modèle ne produit pas de ségrégation quand les individus sont très intolérants?
-Comment ça?
-Bien sûr qu'il le fait (sous contrainte du taux de vacance).
-Cela dit y'a bien un point sur lequel l'implem Netlogo n'est pas bonne: c'est que la relocalisation des mécontents est random alors que dans le bouquin de Schelling, apparemment, c'était dans un voisinage proche.}[(JR) ah oui cest selon la vacance ; mais il y a des regimes ou ca converge jamais sous forme de cluster et ca reste un etat dynamique sans structure ; oui par contre on a du reprendre la random reloc, c'est ce que font les physiciens aussi]
 

### 2018-09-28 Response_to_Reviewers.tex
Update on Overleaf.

@@
 
-In order to get a first idea of the potential impact of spatial initial conditions on to the results of the simulation, we implemented on Segregation model a geographical constraint: some cells cannot be inhabited (yellow cells in Figure \ref{Fig1}).
+In order to get a first idea of the potential impact of spatial initial conditions on to the results of the simulation, we implemented on Segregation model a geographical constraint: some cells cannot be inhabited (yellow cells in Figure 1).
 The way those “forbidden” cells are distributed throughout space will indeed influence the steps of the model.
@@
 A value of 0 would indicate a spatial distribution of yellow cells so as to minimize spatial autocorrelation.
-A value of 1 indicates on the contrary a concentration process that tends towards a high value of autocorrelation (see below Figure \ref{Fig1}).
+A value of 1 indicates on the contrary a concentration process that tends towards a high value of autocorrelation (see below Figure 1).
+% note : no fig envt in moderncv
 
@@
 \begin{center} 
-\resizebox{0.9\textwidth}{!}{ 
-	\includegraphics{figures/SpaceMatters2.bmp}
-} \textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (implementation on Netlogo).} \label{fig:Fig1}
+	\includegraphics[width=\textwidth]{figures/SpaceMatters2.png}
+
+\textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (Netlogo implementation).}
 \end{center}
@@
     \item The double color progression in Fig 4 and 5 is used to highlight the opposition between ratio values above/under 1 but is not meant to introduce any hierarchy in the interpretation.
-We would therefore prefer to keep the figures as they are if this is ok for you?
+We reversed the order of color according to your remark.
+    %We would therefore prefer to keep the figures as they are if this is ok for you?
 

### 2018-09-28 Response_to_Reviewers.tex
Update on Overleaf.

@@
 \begin{minipage}{\textwidth}
+\bigskip\bigskip
 \begin{center} 
@@
 
-\textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (Netlogo implementation).}
+Fig. 1: \textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (Netlogo implementation).}
 \end{center}
+\bigskip\bigskip
 \end{minipage}
@@
 I wonder if the Euclidean distance is the best measure to determine the difference between the phase vectors of different spatial inputs.
-	\item The comment on other possible techniques to compare phase diagrams echoes one other reviewer comment, and we indeed expanded the discussion on this point, showing first that other distance do not change much qualitative results and with more literature for possible extensions.
-	%\comment[JR]{analyse à ajouter}
+	\item The comment on other possible techniques to compare phase diagrams echoes one other reviewer comment, and we indeed expanded the discussion on this point, showing first that other distances do not change much qualitative results (analysis added in supplementary material) and with more literature for possible extensions.
 
@@
     \item The double color progression in Fig 4 and 5 is used to highlight the opposition between ratio values above/under 1 but is not meant to introduce any hierarchy in the interpretation.
-We reversed the order of color according to your remark.
-    %We would therefore prefer to keep the figures as they are if this is ok for you?
+We reversed the order of colors according to your remark.
 
@@
 	% Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?
-	\item 
-    %\comment[JR]{no justification for the difference in number of runs (just experimental setup contingence\ldots ) $\rightarrow$ either new runs or redo computations with 50 for Schelling}\comment[JR]{ca va etre chaud patate pour les new computations ; je vais regarder les ratio de sharpe, avec un peu de chance il y a besoin de plus de repets pour schelling...}
+	\item Concerning the number of repetitions, statistical analysis show that Schelling indicators need more repetitions for an equivalent size of confidence interval on the estimated mean, as the Sharpe ratios are mostly lower than for Sugarscape.
+This remark was added in the paper and the statistical analysis as a supplementary material.
 
@@
 Similarly, the large simulation datasets used for results presented in the paper are openly available on a dataverse repository (provisorily anonymized also).
-	%\comment[JR]{TODO dataverse. / statement open code and data}
 

### 2018-09-28 paper.tex
.

@@
 However, this step is rarely explicit.
-\deleted{For example
-
-and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one.
+\deleted{For example and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one.
 }A meaningful way of representing \added{initial} space might \deleted{thus }be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour \replaced{to different}{in} plausible spatial arrangements.
@@
 
-\newpage
+%\newpage
 
@@
 
-For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\Eb{X}}/\hat{\sigma}(X)$ with standard estimators for average and standard deviation.
-The summary statistics computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}.
+\added{For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\mathbb{E}{X}}/\hat{\sigma}(X)$ with standard estimators for average and standard deviation.
+The summary statistics computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}.}
 
@@
 \caption{\added{\textbf{Summary statistics of Sharpe ratio estimated on repetitions for each parameter point.}}\label{tab:summarysharpes}}
-	\begin{tabular}{|cccccc|}
+\centering
+	\begin{tabular}{|ccccccc|}
+	\hline
 Model/Indicator & Min. & 1st Qu. &  Median &  Mean & 3rd Qu. & Max. \\\hline
@@
 	\caption{\added{\textbf{Summary statistics of different distances for schelling.}}}
+	\centering
 	\begin{tabular}{|ccccccc|}
+	\hline
     	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
@@
 	\caption{\added{\textbf{Summary statistics of different distances for sugarscape.}}}
+	\centering
 	\begin{tabular}{|ccccccc|}
+	\hline
     	$q$ & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\\hline
@@
 \begin{figure*}
-	\caption{\added{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.}}}
+\caption{\added{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}}
 \begin{mdframed}
@@
 		let wantedcolor color
-    		let stranger-rate count neighboragents with [color = wantedcolor] / count neighboragents
+		let stranger-rate count neighboragents with [color = wantedcolor] /
+			count neighboragents
 		if stranger-rate < similar-wanted [
@@
 % Minimal netlogo code for the sugarscape model
-\caption{\added{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}}
+\caption{\added{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} Conventions are the same than for Schelling pseudo-code.}}
 \begin{mdframed}
@@
 while [time < final-time] [
-	; regrow the patch resources
+;regrow the patch resources
     ask patches [set psugar min (list max-psugar (psugar + 1))]

### 2018-09-28 Response_to_Reviewers.tex
Update on Overleaf.

@@
 	% 5. Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index
-    %\comment[JR]{je pensais qu'on prenait la moyenne ? je verifie... ligne 175 on écrit la médiane en effet}
-	%\comment[CC]{ça doit être moi parce que je pensais qu'on prenait effectivement la médiane, qui est plus robuste aux extrêmes, mais si c'est la moyenne et que ça reste proche entre différentes réplications why not.}[(JR) il me semble que les distribs sont assez centrées et que ca revient quasiment au même ; par contre Clem si tu as pris median pour les stats et moi mean pour les plots c'est pas fou ; tu as encore le code des analyses stats ?
-(c'est sur le git ?)]
-	%\comment[JR]{c'est la merde, on prend certaines fois la mediane et certaines fois la moyenne...}
-    %\comment[FL]{certes c'est pas fou mais ca peut se défendre en disant que la médiane sert pour communiquer de façon accessible, sur une dimension pour laquelle c'était pertinent, mais que de façon générique on a préféré travailler sur la loi de Gauss car tout le monde à en tête ces indicateurs moyenne / ecartype c'est plus "passe partout"?
-}
+    \item There was a mistake in the writing, as we never use the median but always the mean, which as you recall is consistent with the choice of distance we made. This was corrected in the paper.
 
 	% 6. Line 232: comparing phase diagrams is related to the topic of comparing patterns. It would make sense to mention the methods that are usually used to compare patterns
-	\item \comment[JR]{biblio a rajouter}
+	\item Some general references on comparing patterns were added.
 
@@
 Adapting the mathematical aggregator to the different indices would constitute an interesting perspective, but is out of the scope of the paper which already focuses on two models and proposes a new framework for analysing the impact of initial pattern parameters.
+We added a supplementary material showing the distances obtained with other distances, which are similar to the results in main text.
 
@@
 Perhaps this should be explained earlier in the paper.
-	\item \comment[JR]{ce qu'il suggère c'est un indice intégré sur l'ensemble des phases diagramme, donc un ``meta-meta-indice'' en quelque sorte, nous on reste à un niveau de meta seulement car on regarde juste pour chaque s'il bouge par rapport à un point de reference.
-En termes thematique c'est plus coherent pour nous, lui doit avoir en tete des indices de sensibilite globaux type Saltelli.
-$\rightarrow$ a voir comment justifier ca bien}{(JR) inclure en mise en contexte ce lien avec les Saltelli et indexes globaux}\comment[FL]{oui je trouve qu'on a le droit de rester limiter à cela c'est déjà pas mal!}
+	\item Concerning the measure of the distance of all phase diagram to one reference phase diagram, this choice was made as we preferred having a local sensitivity index for which the distribution can be studied, rather than a global completely integrated sensitivity index (similar to Saltelli or Morris indexes for example).
+This was explained better in the paper.
 
@@
 	% You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields?
-	\item 
-%    \comment[JR]{cf remarques other reviewers sur ce point}[(MLT) Sa remarque fait sens, mais j'avoue ne pas être trop au point sur cette partie de la littérature...
-vous avez des références?][(JR)concernant la spatial sensitivity : justement c'est pas trop fait non plus dans les autres champs !
-je parlais avec un ecologue, leur espace est jamais varié autre que donnees reelles ; j'ai l'impressions que geosciences, hydrologie etc idem.
-Pour comparer phase diagrams j'ai pas en tete de domaine ou ca se ferait (faute d'avoir cherché, justement quand on cherchait une methode) - le plus proche c'est les patterns comparison etc.
-comme l'on suggere les reviewers.]
+	\item We do not have knowledge of examples in which the comparison of patterns or fields is called ``phase diagram comparison'', and we neither have knowledge of such work in physics for example, from which the term originated.
+There is naturally all the literature we cited on similar issues in other domains, and the extension of the discussion gives elements of answer to this point.
 

### 2018-09-28 paper.tex
Update on Overleaf.

@@
-\documentclass[preprint,5p,times,twocolumn,authoryear]{elsarticle}
+\documentclass[preprint,5p,times,twocolumn,authoryear,draft]{elsarticle}
 
@@
 A solution is to use systematic quantitative procedures to compare them \added{to a reference case}.
-Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the median \added{?} of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
+Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the \replaced{mean}{median} of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
 
@@
 For a multi-dimensional phase diagram, we \replaced{average}{sum} these relative distances over the components.
+\added{Given a set of phase diagram to be compared, we will study the distribution of this distance to an arbitrary phase diagram for all diagrams, rather than an aggregated measure which would be similar to global sensitivity methods \citep{saltelli2008global}.}
 
@@
 
+\added{As detailed in supplementary material~\ref{app:convergence}, more repetitions are needed for Schelling indicators than for Sugarscape, in order to obtain a similar relative confidence in the estimation of averages.
+We ran for this reason a different number of replications for each model.}
+
 We choose different experiment designs, both for \replaced{generator parameters}{meta-parameters} and for the phase diagram, to demonstrate the robustness of the method to technical choices.
@@
 For each density grid, we obtain the average squared distance between corresponding points of the phase diagrams, i.e.
-the median\added{?} value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated).
+the \replaced{mean}{median} value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated).
 This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry).
 Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.
+\added{We tested the sensitivity to the type of distance, using a Minkovski distance with a varying exponent.
+The results are presented in supplementary material~\ref{app:distances}, and show a similar sensitivity to spatial initial conditions.}
 
@@
 
+\label{app:convergence}
+
 %%%
@@
 \added{For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\mathbb{E}{X}}/\hat{\sigma}(X)$ with standard estimators for average and standard deviation.
-The summary statistics computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}.}
+The summary statistics of this ratio computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}.
+Under the assumption of a normal distribution, the width of the confidence interval at level $c$ is given by $\left|\mu_+ - \mu_-\right| = 4\cdot\sigma \sigma z_{c} \sqrt{n}$ where $\sigma$ is the standard deviation, $z_{c}$ is the quantile at which $c$ is attained by the cumulative distribution, which is around $1.96$ for a 95\% confidence interval.
+This means that to obtain a confidence interval of width $\kappa \cdot \sigma$, one needs a number $n \simeq (8 / \kappa )^2$ of repetitions.
+This gives 64 repetitions for $\kappa = 1$.
+As the Sharpes ratios are in general smaller for Schelling indicators than for Sugarscape, we take $n = 50$ for Sugarscape and $n = 100$ for Schelling to have s similar confidence in estimations.}
 
@@
 \centering
-	\begin{tabular}{|ccccccc|}
+	\begin{tabular}{|c|cccccc|}
 	\hline
@@
 
+\label{app:distances}
+
 \added{We describe here the tests done with other distances to compare phase diagrams.
-We tested normalized Minkovski distances, defined by $d(x,y) = \left(\frac{1}{N}\cdot \sum_i \left|x_i - y_i\right|^{q}\right)^{\frac{1}{q}}$, for varying values of $q$ from $q = 1$ (Manhattan distance) to $q = 10$, including $q = 2$ (Euclidian distance) which is used in main text.}
+We tested normalized Minkovski distances, defined by $d(x,y) = \left(\frac{1}{N}\cdot \sum_i \left|x_i - y_i\right|^{q}\right)^{\frac{1}{q}}$, for varying values of $q$ from $q = 1$ (Manhattan distance) to $q = 10$, including $q = 2$ (Euclidian distance) which is used in main text.
+The Table~\ref{tab:schellingdistances} gives the summary statistics of each distance computed on all initial configurations for the Schelling model.
+The Table~\ref{tab:sugarscapedistances} gives the same statistics for the sugarscape model.
+We naturally obtain smaller difference with the Manhattan distance but which remain significant (averages of 10\% for the Schelling model and 40 \% for Sugarscape), and variabilities with higher values of the Minkovski exponent are much higher.
+This results confirm the high variability observed in main text with the Euclidian distance.}
 
@@
 \begin{table*}
-	\caption{\added{\textbf{Summary statistics of different distances for schelling.}}}
+	\caption{\added{\textbf{Summary statistics of different distances for schelling.}}}\label{tab:schellingdistances}
 	\centering
@@
 \begin{table*}
-	\caption{\added{\textbf{Summary statistics of different distances for sugarscape.}}}
+	\caption{\added{\textbf{Summary statistics of different distances for sugarscape.}}}\label{tab:sugarscapedistances}
 	\centering
@@
 We recall that source code is openly available at} \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
+\added{The pseudo-code is in the style of NetLogo code, which is already easily readable.}
 

### 2018-09-28 paper.tex
.

@@
-\documentclass[preprint,5p,times,twocolumn,authoryear,draft]{elsarticle}
+\documentclass[preprint,5p,times,twocolumn,authoryear]{elsarticle}
 
@@
 	\centering
-	\begin{tabular}{|ccccccc|}
+	\begin{tabular}{|c|cccccc|}
 	\hline
@@
 while [time < final-time] [
-;regrow the patch resources
     ask patches [set psugar min (list max-psugar (psugar + 1))]
-
-	; make agents exploit resources
 	ask turtles [

### 2018-09-28 paper.tex
Update on Overleaf.

--- 2018-09-28 paper.tex
@@
 \added{The transfer of methods used to compare sequences \citep{kruskal1983overview} or time-series \citep{liao2005clustering} is a possible way to develop measures between phase diagrams.
-The higher dimension of the phase diagrams we study must however be considered with caution when transferring methods, in a way analog to the application of global sensitivity indices to spatial data \citep{lilburne2009sensitivity}.} The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.
+The higher dimension of the phase diagrams we study must however be considered with caution when transferring methods, in a way analog to the application of global sensitivity indexes to spatial data \citep{lilburne2009sensitivity}.
+We can also note than more generally, this problem of comparing phase diagrams is a particular instance of the more generic issue of comparing patterns, which for example include unsupervised learning techniques \citep{hastie2009unsupervised}.} The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.
 
@@
 \caption{\added{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}}
+\bigskip
 \begin{mdframed}
@@
 \caption{\added{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} Conventions are the same than for Schelling pseudo-code.}}
+\bigskip
 \begin{mdframed}

### 2018-09-28 Response_to_Reviewers.tex
final

@@
 
-In order to get a first idea of the potential impact of spatial initial conditions on to the results of the simulation, we implemented on Segregation model a geographical constraint: some cells cannot be inhabited (yellow cells in Figure 1).
+In order to get a first idea of the potential impact of spatial initial conditions on to the results of the simulation, we implemented on the Schelling Segregation model a geographical constraint: some cells cannot be inhabited (yellow cells in Figure 1).
 The way those “forbidden” cells are distributed throughout space will indeed influence the steps of the model.
@@
 Adapting the mathematical aggregator to the different indices would constitute an interesting perspective, but is out of the scope of the paper which already focuses on two models and proposes a new framework for analysing the impact of initial pattern parameters.
-We added a supplementary material showing the distances obtained with other distances, which are similar to the results in main text.
+We however added a supplementary material showing the distances obtained with other distances, which are similar to the results in main text.
 
@@
 
-	\item In the case of Schelling, this is the scale at which most empirical segregation indices are computed.
+	\item More particularly concerning the choice of this generator in the case of Schelling, this is the scale at which most empirical segregation indices are computed.
 In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty, 2005), or to a generic landscape where a resource is grown otherwise.
@@
 	% 10. Explain how the density grids are used in the Schelling model, do you allow more than one agent to be in the same cell? 
-	\item Yes, we allow more agents to be in the same cell.
+	\item Regarding the use of the density grid in the Schelling model, yes we allow more agents to be in the same cell.
 The potential density is defined by the density grid generated, and its population with the same number of agents in all grids.
@@
 	% 11. Line 297: I am not sure why empirical density destitutions of cities are important in the context of this study. 
-	%\item cf infra. 
 	\item Density distributions of empirical cities are important because we want to test models with realistic ranges of initial patterns of density distribution.
@@
 	\item The term ``mythical'' line 671 was a language mistake and corrected, ``textbook'' ABM was meant.
-    %\comment[MLT]{Use "textbook ABM instead?}[(JR) yes, better]
 \end{itemize}
@@
 Moreover, even if such sensitivity is analysed, it mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but to the best of our knowledge no article provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper.
-We have tried to argue this point at the end of the section.
+We have tried to argue this point at the end of the first section.
 
@@
 	\item Concerning the measure of the distance of all phase diagram to one reference phase diagram, this choice was made as we preferred having a local sensitivity index for which the distribution can be studied, rather than a global completely integrated sensitivity index (similar to Saltelli or Morris indexes for example).
-This was explained better in the paper.
+This was better explained in the paper.
 
@@
 Maybe this was explained in the paper but I missed it.
-	\item Yes, classically, the initial distribution of agents is random in both models and, most importantly for our point, it follows a uniform distribution of density.
+	\item Regarding the way these models are used in the literature, yes, the initial distribution of agents is classically random in both models and, most importantly for our point, it follows a uniform distribution of density.
 In our implementation, the initial location of agents is also random in Schelling, but it follows a distribution of density given by the generated grid.
 In Sugarscape, the agents are initialised at random but the resources follow a distribution of density given by the generated grid.
-We have clarified this point in section 2.3
+We have clarified this point in section 2.3.
 
@@
 You touch upon shape when talking about NetLogo's use of raster as opposed to vector in the Discussion - could you expand on this?
-	\item We thank the reviewer for this good remark.
+	\item We thank the reviewer for this good remark regarding connected issues such as heterogeneity or the choice raster/vector.
 We provide references to papers addressing these issues.

### 2018-09-28 paper.tex
final

@@
 
-%\usepackage[final]{changes}
-\usepackage{changes}
+\usepackage[final]{changes}
+%\usepackage{changes}
 
@@
 \address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
-\address[f]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilité Transport, Marne-la-Vallée, France}
+\address[f]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilit{\'e} Transport, Marne-la-Vallée, France}
 
@@
 
-\added{As detailed in supplementary material~\ref{app:convergence}, more repetitions are needed for Schelling indicators than for Sugarscape, in order to obtain a similar relative confidence in the estimation of averages.
+\added{As detailed in~\ref{app:convergence}, more repetitions are needed for Schelling indicators than for Sugarscape, in order to obtain a similar relative confidence in the estimation of averages.
 We ran for this reason a different number of replications for each model.}
@@
 \added{We tested the sensitivity to the type of distance, using a Minkovski distance with a varying exponent.
-The results are presented in supplementary material~\ref{app:distances}, and show a similar sensitivity to spatial initial conditions.}
+The results are presented in~\ref{app:distances}, and show a similar sensitivity to spatial initial conditions.}
 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams}
+\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams.png}
 \caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by \replaced{their generator parameters}{the meta-parameters of their generator}.} Relative distance as a function of \replaced{generator parameters}{meta-parameters} $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/relativedistance_morphspace}
+\includegraphics[width=0.8\textwidth]{figures/relativedistance_morphspace.png}
 \caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text).
@@
 Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents, \added{transportation networks generators \citep{raimbault2018multi}, or coupled road network and population raster generators \citep{raimbault2018urban}.}
+% Generators of abstract urban systems at a macroscopic scale could also be useful tools to better understand urban growth models as sketched by \cite{raimbault2018unveiling}.}
 
@@
 Low values of $\alpha$ give diffuse patterns, with however clear centers for a high diffusion.
-We do not discuss it the role of other parameters, but according to \cite{raimbault2018calibration}, diffusion steps give smoother forms, and the rate between total population and the population increment at each step (which is equivalent to the total number of steps) is crucial to select non-stationary distributions that are closer to real configurations.
+We do not discuss here the role of other parameters, but according to \cite{raimbault2018calibration}, diffusion steps give smoother forms, and the rate between total population and the population increment at each step (which is equivalent to the total number of steps) is crucial to select non-stationary distributions that are closer to real configurations.
 \cite{raimbault2018calibration} also shows the existence of non-linear behaviors in some regions of the parameter space, so the description we gave here shall not be interpreted as a linear link between generator parameters and the morphological properties of the generated grids.}
@@
 
-\added{For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\mathbb{E}{X}}/\hat{\sigma}(X)$ with standard estimators for average and standard deviation.
+\added{For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\mathbb{E}}\left[X\right]/\hat{\sigma}(X)$ with standard estimators for average and standard deviation.
 The summary statistics of this ratio computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}.
-Under the assumption of a normal distribution, the width of the confidence interval at level $c$ is given by $\left|\mu_+ - \mu_-\right| = 4\cdot\sigma \sigma z_{c} \sqrt{n}$ where $\sigma$ is the standard deviation, $z_{c}$ is the quantile at which $c$ is attained by the cumulative distribution, which is around $1.96$ for a 95\% confidence interval.
+Under the assumption of a normal distribution, the width of the confidence interval at level $c$ is given by $\left|\mu_+ - \mu_-\right| = 4\cdot \sigma \cdot z_{c} / \sqrt{n}$ where $\sigma$ is the standard deviation, $z_{c}$ is the quantile at which $c$ is attained by the cumulative distribution, which is around $1.96$ for a 95\% confidence interval.
 This means that to obtain a confidence interval of width $\kappa \cdot \sigma$, one needs a number $n \simeq (8 / \kappa )^2$ of repetitions.
 This gives 64 repetitions for $\kappa = 1$.
-As the Sharpes ratios are in general smaller for Schelling indicators than for Sugarscape, we take $n = 50$ for Sugarscape and $n = 100$ for Schelling to have s similar confidence in estimations.}
+As the Sharpe ratios are in general smaller for Schelling indicators than for Sugarscape, we take $n = 50$ for Sugarscape and $n = 100$ for Schelling to have a similar confidence in estimations.}
 
@@
 The Table~\ref{tab:sugarscapedistances} gives the same statistics for the sugarscape model.
-We naturally obtain smaller difference with the Manhattan distance but which remain significant (averages of 10\% for the Schelling model and 40 \% for Sugarscape), and variabilities with higher values of the Minkovski exponent are much higher.
+We naturally obtain smaller difference with the Manhattan distance but which remain significant (averages of 10\% for the Schelling model and 40 \% for sugarscape), and variabilities with higher values of the Minkovski exponent are much higher.
 This results confirm the high variability observed in main text with the Euclidian distance.}
@@
 
-\added{We give below the pseudo-code for the implementations we used of both sugarscape and schelling models.
+\added{We give below the pseudo-code for the implementations we used of both sugarscape (Fig.~\ref{frame:sugarscapecode}) and Schelling (Fig.~\ref{frame:schellingcode}) models.
 We recall that source code is openly available at} \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
@@
 \begin{figure*}
-\caption{\added{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}}
+\caption{\added{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}}\label{frame:schellingcode}
 \bigskip
@@
 % Minimal netlogo code for the sugarscape model
-\caption{\added{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} Conventions are the same than for Schelling pseudo-code.}}
+\caption{\added{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} Conventions are the same than for Schelling pseudo-code.}}\label{frame:sugarscapecode}
 \bigskip

### 2018-09-28 paper.tex
.

@@
 Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes.
-However, initial spatial conditions are usually \replaced{not modified systematically}{taken for granted} in geographical models, thus leaving \deleted{completely} unexplored the effect of \added{initial} spatial arrangements on the interactions of agents \replaced{with one another as well as}{and of that of agents} with their environment.
-In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic \added{spatial configuration} generator \deleted{controlled by meta-parameters} in order to create density grids with which spatial simulation models are initialised.
-We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant \replaced{on}{in} the two models.
+However, initial spatial conditions are usually not modified systematically in geographical models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment.
+In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic spatial configuration generator in order to create density grids with which spatial simulation models are initialised.
+We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models.
 Furthermore, this effect is sometimes larger than the effect of parameters' value change.

### 2018-12-09 paper.tex
resppreviewers
jasss_template

@@
-\documentclass[preprint,5p,times,twocolumn,authoryear]{elsarticle}
+\documentclass{JASSS}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% Editorial fields (to be set in case of publication)	
+% Please leave this section untouched
+%\doinum{10.18564/jasss.xxxx}
+%\volume{xx}
+%\issue{x}
+%\article{x}
+%\pubyear{20xx}
+%\received{dd-mmm-yyyy}
+%\accepted{dd-mmm-yyyy}
+%\published{dd-mmm-yyyy}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% title, authors and affiliations
+
+\title{Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models}
+
+%All authors should be included in the submission.
+To anonymise the submission, set to 'true' the \reviewcopy command below.
+However, before submitting  you can check that all the informations you entered are correct by temporarily setting it to 'false'.
+Please remember to set it back to 'true' before the submission.
+\reviewcopy{true}
+
+\author[1,2]{Juste Raimbault}
+\affil[1]{UPS CNRS 3611 ISC-PIF, Paris, France}
+\affil[2]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
+
+\author[3,4]{Cl{\' e}mentine Cottineau}
+\affil[3]{Centre for Advanced Spatial Analysis, University College London, UK}
+\affil[4]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
+
+\author[5]{Marion Le Texier}
+\affil[5]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
+
+\author[6]{Florent Le N{\' e}chet}
+\affil[6]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilit{\'e} Transport, Marne-la-Vallée, France}
+
+\author[1]{Romain Reuillon}
+\affil[1]{UPS CNRS 3611 ISC-PIF, Paris, France}
+%\author{}
+
+\email{juste.raimbault@polytechnique.edu}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% REFERENCES should be included using the \citep, \citet, etc. commands provided by the natbib package
+
+\usepackage{natbib}
+	\setcitestyle{authoryear,round,aysep={}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+% EXTRA PACKAGES. Please place here any extra package you need along with your own command definitions
 
@@
 
-\usepackage{soul}
-\soulregister\cite7
-\soulregister\citep7
-\soulregister\ref7
+%\usepackage{soul}
+%\soulregister\cite7
+%\soulregister\citep7
+%\soulregister\ref7
 
-\usepackage[final]{changes}
+%\usepackage[final]{changes}
 %\usepackage{changes}
+%\setaddedmarkup{\textcolor{black}{\hl{#1}}}
+%\setdeletedmarkup{\textcolor{red}{\sout{#1}}}
 
-\setaddedmarkup{\textcolor{black}{\hl{#1}}}
-\setdeletedmarkup{\textcolor{red}{\sout{#1}}}
-
-\journal{Computers, Environment and Urban Systems}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \begin{document}
+\maketitle
 
-\begin{frontmatter}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\title{Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models}
-
-\author[a,b]{Juste Raimbault\corref{cor1}}
-\author[c,d]{Cl{\' e}mentine Cottineau}
-\author[e]{Marion Le Texier}
-\author[f]{Florent Le N{\' e}chet}
-\author[a]{Romain Reuillon}
-%\author{}
-
-\address[a]{UPS CNRS 3611 ISC-PIF, Paris, France}
-\address[b]{UMR CNRS 8504 G{\'e}ographie-cit{\'e}s, Paris, France}
-\address[c]{Centre for Advanced Spatial Analysis, University College London, UK}
-\address[d]{UMR CNRS 8097 Centre Maurice Halbwachs, Paris, France}
-\address[e]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
-\address[f]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilit{\'e} Transport, Marne-la-Vallée, France}
-
-\cortext[cor1]{Corresponding author: juste.raimbault@polytechnique.edu}
+% Abstract and keywords
 
@@
 
-\begin{keyword}
-Space \sep Initial conditions \sep Sensitivity \sep ABM
-\end{keyword}
+\begin{keywords}
+Space, Initial conditions, Sensitivity, ABM
+\end{keywords}
 
-\end{frontmatter}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Start of  paragraph numbering. Please leave this untouched
+\parano{}
 
@@
 
-Computer simulation has been recognised and \added{is }increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}\added{.
-It has been identified as an emerging field and coined under the term geosimulation by \cite{benenson2004geosimulation}.} Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
-The specificity of geographical models compared to other social science models is that space and spatial interactions are given a prime role, \deleted{as }geographers \replaced{being}{are} driven by an explicit interest in studying the way space influences the outcomes of \replaced{social processes modelled.}{the model in understanding the role of space on social interactions and environmental processes, in a city for example}.
-We think \replaced{that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions}{simulation can become a very good tool to achieve this}, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of \replaced{the model outputs}{ variations} to the way space is represented.
+Computer simulation has been recognised and is increasingly used by geographers as an efficient tool to explore geographical processes, hypotheses and predictive scenarios within virtual laboratories \citep{batty1971modelling, batty2007model, carley1999generating, Quesneletal2009}.
+It has been identified as an emerging field and coined under the term geosimulation by \cite{benenson2004geosimulation}.
+Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
+The specificity of geographical models compared to other social science models is that space and spatial interactions are given a prime role, geographers being driven by an explicit interest in studying the way space influences the outcomes of social processes modelled.
+We think that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of the model outputs to the way space is represented.
 Unfortunately, the first condition is not always met, and the second is seldom even mentioned.
@@
 This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}.
-Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex systems generally, a given configuration contains clues \replaced{about}{on} past bifurcations, that can have had dramatic effects on the state of the system.
+Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex systems generally, a given configuration contains clues about past bifurcations, that can have had dramatic effects on the state of the system.
 Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models.
-\replaced{Ironically, these very models tend to be the models most frequently used}{..
-which are the main models to be found} in geosimulation.
+Ironically, these very models tend to be the models most frequently used in geosimulation.
 
@@
 Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures.
-In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as \added{any }other initial conditions, but of more interest to the geographer.
+In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as any other initial conditions, but of more interest to the geographer.
 \\
 
-Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in \deleted{the most }sprawled cities.
+Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in sprawled cities.
 On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities.
@@
 However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic density.
-Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents \replaced{tend to be}{is still} approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}.
-The isotropic simplification is potentially harmful to the representation of urban processes \deleted{under inquiry }because density and accessibility have environmental, economic and social consequences.
+Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents tend to be approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}.
+The isotropic simplification is potentially harmful to the representation of urban processes because density and accessibility have environmental, economic and social consequences.
 Additionally, we expect the initial spatial distribution of agents to influence simulation results in the long run \citep{Castellanoetal2009}, because the agents' rule of action itself may depend on the spatial structure of the environment.
-For example, households can have different preferences with respect to the built-environment they \added{might} want to live in \citep{SpielmanHarrison2014}, or agents \replaced{moving around will sense a different set of objects within the same fixed radius depending on}{can move around and sense their environment up to a certain distance, so that the resulting number of sensed objects will depend on} the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}.
-The way modellers represent \added{initial} space is therefore a central element of geographical simulation models.
+For example, households can have different preferences with respect to the built-environment they might want to live in \citep{SpielmanHarrison2014}, or agents moving around will sense a different set of objects within the same fixed radius depending on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}.
+The way modellers represent initial space is therefore a central element of geographical simulation models.
 However, this step is rarely explicit.
-\deleted{For example and authors tend to generalise the conclusion of the model from a single spatial case study (be it a theoretical pattern or a singular empirical one.
-}A meaningful way of representing \added{initial} space might \deleted{thus }be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour \replaced{to different}{in} plausible spatial arrangements.
+A meaningful way of representing initial space might be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour to different plausible spatial arrangements.
 
@@
 In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape).
-In no way do we \replaced{pretend}{want} to provide a full exploration of these two particular models, their attractors and/or potential policy implications.
-Instead, we present a way of performing a sensitivity analysis to \replaced{initial}{the} spatial conditions of models by using a spatial generator to produce a variety of density grids, which are \replaced{taken as input by}{then fed into} simulation models at initialisation.
-The generator being controlled by \replaced{its own}{meta-} parameters, we can then relate the parameters used to generate initial spatial conditions to the variation \replaced{simulation outcomes}{in outcome of the simulations}.
-The purpose is two-fold: (i) to test the robustness of simulation results to small variations of \replaced{generator}{meta-} parameters and (ii) to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
-Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies \added{to our knowledge}.
+In no way do we pretend to provide a full exploration of these two particular models, their attractors and/or potential policy implications.
+Instead, we present a way of performing a sensitivity analysis to initial spatial conditions of models by using a spatial generator to produce a variety of density grids, which are taken as input by simulation models at initialisation.
+The generator being controlled by its own parameters, we can then relate the parameters used to generate initial spatial conditions to the variation simulation outcomes.
+The purpose is two-fold: (i) to test the robustness of simulation results to small variations of generator parameters and (ii) to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
+Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies to our knowledge.
 In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.
@@
 For example, \citet{wheeler2006urban} shows that, in the US, sprawling cities are more unequal than their compact counterparts with respect to income.
-Dynamically, sprawl \added{in American cities} consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}.
+Dynamically, sprawl in American cities consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}.
 Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}.
@@
 
-Considerations of such issues in the \replaced{geosimulation}{geographical modelling} literature are rather scarce.
-\replaced{However,}{Despite the opportunity of simulation in the representation of space and its potential observation biases,} there have been some noticeable attempts \replaced{at analysing the impact of three types of initial spatial characteristics on model outcomes}{to evaluate the effects of the same set of mechanisms run from different initial spatial conditions, which have focused on the impact of}:
-\begin{itemize}
+Considerations of such issues in the geosimulation literature are rather scarce.
+However, there have been some noticeable attempts at analysing the impact of three types of initial spatial characteristics on model outcomes:
+\begin{itemize*}
 \item The accuracy of geo-localised input data;
@@
 \item The degree of spatial heterogeneity modelled.
-\end{itemize}
+\end{itemize*}
 
-\paragraph{Geo-localised input data accuracy} \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and \added{the} scale of analysis.
+\subsubsection{Geo-localised input data accuracy}
+
+\citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and the scale of analysis.
 They provide a few examples on how the use of Exploratory Spatial Data Analysis (ESDA) prior to simulation runs can help avoiding measurement errors of model behaviour and outcomes.
@@
 
-\paragraph{Spatial system shape, precision, and boundaries} \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e.
+\subsubsection{Spatial system shape, precision, and boundaries}
+
+\citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e.
 made equivalent to) to Axelrod Culture Model.
@@
 \citet{LeTexierCaruso2017}, using a set of different theoretical spatial systems, demonstrate the impact of the regularity and aggregation levels, or centrality/periphery effects, on spatial diffusion dynamics of euro coins.
-Similar issues were also dealt with in physical sciences: for example, \cite{horritt2001effects} studies the effects of grid cell size on the behaviour of a raster flood model and shows that increasing resolution does not increase model prediction performance \replaced{below a certain level}{under a certain limit}.
+Similar issues were also dealt with in physical sciences: for example, \cite{horritt2001effects} studies the effects of grid cell size on the behaviour of a raster flood model and shows that increasing resolution does not increase model prediction performance below a certain level.
 Similar conclusions are obtained by \cite{vazquez2002effect}, unveiling an intermediate optimal spatial resolution regarding model performance and computation time.
@@
 
-\paragraph{Spatial heterogeneity} \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
+\subsubsection{Spatial heterogeneity}
+
+\citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
 They reveal conjoint and non-linear effects on the vacancy rates and tolerance levels on segregation patterns.
@@
 
-Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.) \replaced{or}{either} on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
-\added{More so, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.}
+Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.) or on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
+More so, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.
 
@@
 The general method workflow of our method is illustrated in Figure \ref{fig:method}.
-In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on \replaced{its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
-We call these parameters ``generator parameters'' to distinguish them from the standard parameters of the models and to highlight their higher position in the simulation process}{meta-parameters and feeds the model $\mu$ at initialisation (lower branch)}.
+In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
+We call these parameters ``generator parameters'' to distinguish them from the standard parameters of the models and to highlight their higher position in the simulation process.
 The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.
-The sensitivity analysis relates the \added{variations in the} model's \replaced{outcomes}{parameters} to how the density spatial distribution was generated and \added{to} the patterns of density generated.
-\deleted{in order to produce qualitative knowledge from the model and its behaviour.} In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space.
+The sensitivity analysis relates the variations in the model's outcomes to how the density spatial distribution was generated and to the patterns of density generated.
+In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space.
 In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
 %%%%%%%%%%%%%%%%%%%%%%
-\begin{figure*}[htbp] 
-\begin{center} 
-\resizebox{0.9\textwidth}{!}{ 
-	\includegraphics{figures/SchemaMeta_1.png}
-} \caption{\textbf{General workflow.}} \label{fig:method}
-\end{center}
-\end{figure*} %
+\begin{figure}[!t]
+\centering
+\includegraphics[width=0.9\textwidth]{figures/SchemaMeta_1.png}
+\caption{\textbf{General workflow.}} \label{fig:method}
+\end{figure} %
 %%%%%%%%%%%%%%%%%%%%%%
@@
 We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment).
-A solution is to use systematic quantitative procedures to compare them \added{to a reference case}.
-Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the \replaced{mean}{median} of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
+A solution is to use systematic quantitative procedures to compare them to a reference case.
+Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the mean of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
 
@@
 We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$.
-Phase diagrams are denoted by \replaced{the same}{a common} function $\mu$ indexed by \replaced{the generator parameters}{what we call \emph{meta-parameters}} $\vec{\alpha}$, \replaced{which}{that} capture the spatial configuration \added{(in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself)}.
+Phase diagrams are denoted by the same function $\mu$ indexed by the generator parameters $\vec{\alpha}$, which capture the spatial configuration (in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself).
 The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by
@@
 
-where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given \replaced{generator}{meta-} parameters $\vec{\alpha}_i$.
+where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given generator parameters $\vec{\alpha}_i$.
 $d$ is a functional distance that we take simply as Euclidean distance.
 The internal variabilities are estimated as the variance within each phase diagram $f_{\vec{\alpha_i}}$.
-For a multi-dimensional phase diagram, we \replaced{average}{sum} these relative distances over the components.
-\added{Given a set of phase diagram to be compared, we will study the distribution of this distance to an arbitrary phase diagram for all diagrams, rather than an aggregated measure which would be similar to global sensitivity methods \citep{saltelli2008global}.}
+For a multi-dimensional phase diagram, we average these relative distances over the components.
+Given a set of phase diagram to be compared, we will study the distribution of this distance to an arbitrary phase diagram for all diagrams, rather than an aggregated measure which would be similar to global sensitivity methods \citep{saltelli2008global}.
 
 The last methodological point which we need to emphasize is the relationship between the present workflow and model exploration workflows in general.
-The ideas of multi-modelling and extensive model exploration are nothing from new - ~\cite{openshaw1983data} already advocated for ``model-crunching'' \added{in 1983 -}, but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities.
-The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write \replaced{flexible}{modulable} exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
-In our case, this tool is a powerful way to embed both the sensitivity analysis and the \deleted{meta-}sensitivity analysis \added{to initial spatial conditions}, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file.
+The ideas of multi-modelling and extensive model exploration are nothing from new - ~\cite{openshaw1983data} already advocated for ``model-crunching'' in 1983 -, but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities.
+The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write flexible exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
+In our case, this tool is a powerful way to embed both the sensitivity analysis and the sensitivity analysis to initial spatial conditions, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file.
 In this paper, we use the OpenMOLE platform for the spatial environment and the model coupling, placing ourselves in the framework of multi-modelling \citep{cottineau2015modular}.
-\added{We use therefore OpenMOLE's functionalities for model embedding through workflow, design of experiments (parameter sampling) and high performance environment access.}
+We use therefore OpenMOLE's functionalities for model embedding through workflow, design of experiments (parameter sampling) and high performance environment access.
 
@@
 More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
-At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (\added{i.e.
-its} 8 surrounding patches).
+At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (i.e.
+its 8 surrounding patches).
 To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside.
@@
 The grid then has a population of $t_f \cdot N$ (the population lost due to diffusion process to the outside is reallocated through a normalization procedure at the end of the steps).
-Grids are thus generated from the combination of the values of these four \replaced{generator}{meta-} parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed.
+Grids are thus generated from the combination of the values of these four generator parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed.
 To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment.
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\begin{figure*}[htbp] \begin{center} 
-	\includegraphics[width=\linewidth]{figures/spatialGen.png}
+\begin{figure}[!t]
+\centering
+	\includegraphics[width=\textwidth]{figures/spatialGen.png}
  \caption{\textbf{Four examples of grids produced by the spatial generator.} The lighter the red, the denser the area.
 Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$.
-\added{Parameter values used here are for the first grid $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 971)$, the second $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 176)$, the third $(\alpha = 1.4,\beta=0.045, N = 102, n_d = 2, t_f = 618)$, and the fourth $(\alpha = 1.8,\beta=0.114, N = 108, n_d = 1, t_f = 227)$.}}
+Parameter values used here are for the first grid $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 971)$, the second $(\alpha = 0.4,\beta=0.006, N = 25, n_d = 1, t_f = 176)$, the third $(\alpha = 1.4,\beta=0.045, N = 102, n_d = 2, t_f = 618)$, and the fourth $(\alpha = 1.8,\beta=0.114, N = 108, n_d = 1, t_f = 227)$.}
 \label{fig:spatialGen}
-\end{center}
-\end{figure*} %
+\end{figure} %
 %%%%%%%%%%%%%%%%%%%%%%
@@
 \cite{LeNechet2015}).
-We sample the \replaced{generator}{meta-} parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a \replaced{scatter}{point cloud} with high discrepancy.
+We sample the generator parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a scatter with high discrepancy.
 We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}.
-It yields a subset of 170 \deleted{interesting} grids matching empirical densities, which constituted our set of different initial spatial conditions.
+It yields a subset of 170 grids matching empirical densities, which constituted our set of different initial spatial conditions.
 These are further clustered into three classes of morphology (figure \ref{fig:densityTypes}): compact (e.g.
@@
 
-\added{The spatial generator and its resulting grids are relevant to the case study models we have picked (Schelling and Sugarscape) because it produces density grids at a ``metropolitan scale'', which is the scale at which both model were initially intended to be.
+The spatial generator and its resulting grids are relevant to the case study models we have picked (Schelling and Sugarscape) because it produces density grids at a ``metropolitan scale'', which is the scale at which both model were initially intended to be.
 In the case of Schelling's segregation model for example, this scale is the one at which most empirical segregation indexes are computed and compared to the model outputs.
@@
 In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density and that acknowledging this diversity might leads to a variation in the model outputs.
-Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most unrealistic one, although unfortunately the most common one in Schelling implementations.}
+Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most unrealistic one, although unfortunately the most common one in Schelling implementations.
 
 %%%%%%%%%%%%%%%%%%%%%%
-\begin{figure*}[htbp] \begin{center} 
-\resizebox{0.9\textwidth}{!}{ 
-	\includegraphics{figures/GridPresentation.png}
-} \caption{Correspondence between European urban density structures and grids produced with the spatial generator.} 
-\label{fig:densityTypes} \end{center} \end{figure*} %
+\begin{figure}[!t]
+\centering
+	\includegraphics[width=0.9\textwidth]{figures/GridPresentation.png}
+ \caption{Correspondence between European urban density structures and grids produced with the spatial generator.} 
+\label{fig:densityTypes}
+\end{figure} %
 %%%%%%%%%%%%%%%%%%%%%%
@@
 It has been shown by \cite{Schelling1969} that even tolerant agents tend to produce segregated patterns because of the complexity of their local interactions and the snowball effect of individual moves on the global distribution of agents in the city.
-The main parameters of this model are the tolerance level (maximum \% of agents different to {\it ego} accepted in the neighbourhood), the scope of sensing, the \added{global majority/}minority split and the percentage of vacant spaces in the housing market.
+The main parameters of this model are the tolerance level (maximum \% of agents different to {\it ego} accepted in the neighbourhood), the scope of sensing, the global majority/minority split and the percentage of vacant spaces in the housing market.
 In addition, we are interested in testing the impact of the initial spatial distribution of housing capacity in this project, using the generated grids.
 The outcome of the model is measured by a combination of three segregation indices: Dissimilarity, Moran's I and Entropy~\citep{brown2006spatial}.
-\replaced{We use a ad-hoc implementation of the Schelling model, both in Scala for performance reasons and in NetLogo to ensure visualization of model dynamics.
-The pseudo-code of the implemented model is available in~\ref{app:code} and source code are available on the repository of the project.}{We translated the Netlogo implementation of Schelling into Scala for this project.}
+We use a ad-hoc implementation of the Schelling model, both in Scala for performance reasons and in NetLogo to ensure visualization of model dynamics.
+The pseudo-code of the implemented model is available in~\ref{app:code} and source code are available on the repository of the project.
 
-\added{In general, the implementations of Schelling models allow only one agent per cell, and their initial distribution is random, therefore following a uniform distribution across the modelled city.
+In general, the implementations of Schelling models allow only one agent per cell, and their initial distribution is random, therefore following a uniform distribution across the modelled city.
 In this experiment, we allow more than one agent to be in a given cell.
@@
 Therefore we cannot limit ourselves an isotropic square city in Schelling.
-We chose to use the actual distributions of European cities to constraint our density generation.}
+We chose to use the actual distributions of European cities to constraint our density generation.
 
@@
 The main parameters of this model are the number of agents, their minimal and maximal resource levels.
-\added{In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants.
-Following \cite{batty2005agents}, it can be considered as a metaphor of an urban system.} We extend the implementation with agents wealth distribution of \cite{li2009netlogo}.
+In an urban environment, Sugarscape can be used to model how the spatial distribution of any type of goods or services can influence the spread of wealth among inhabitants.
+Following \cite{batty2005agents}, it can be considered as a metaphor of an urban system.
+We extend the implementation with agents wealth distribution of \cite{li2009netlogo}.
 The outcome of the model is measured by a Gini index of inequality for resource distribution.
-\added{We are interested in testing the impact of the spatial distribution of the resource, using the generated grids.}
+We are interested in testing the impact of the spatial distribution of the resource, using the generated grids.
 
@@
 We run 50 repetitions for each configuration, which yields reasonable convergence properties.
-The initial spatial configuration varies across 50 different grids, generated by sampling \replaced{generator parameters}{meta-parameters for the generator} in a LHS.
+The initial spatial configuration varies across 50 different grids, generated by sampling generator parameters in a LHS.
 We did not use the clustered grids to test the flexibility of our framework, which is demonstrated in this case by a direct sequential coupling of the generator and the model.
@@
 
-\added{As detailed in~\ref{app:convergence}, more repetitions are needed for Schelling indicators than for Sugarscape, in order to obtain a similar relative confidence in the estimation of averages.
-We ran for this reason a different number of replications for each model.}
+As detailed in~\ref{app:convergence}, more repetitions are needed for Schelling indicators than for Sugarscape, in order to obtain a similar relative confidence in the estimation of averages.
+We ran for this reason a different number of replications for each model.
 
-We choose different experiment designs, both for \replaced{generator parameters}{meta-parameters} and for the phase diagram, to demonstrate the robustness of the method to technical choices.
-In principle, our workflow applies \replaced{regardless of the way we}{with to manner to }generate a spatial configuration (even taking real configurations) and \replaced{the way we}{any manner to} establish phase diagrams.
+We choose different experiment designs, both for generator parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices.
+In principle, our workflow applies regardless of the way we generate a spatial configuration (even taking real configurations) and the way we establish phase diagrams.
 
@@
 
-\added{The implementations of the models were done from NetLogo.
+The implementations of the models were done from NetLogo.
 We modified the sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model.
-Both pseudo-codes are available in~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at} \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
-\added{Density grids are also available at this address.
-Simulation data are available for reproducibility on the dataverse repository at }\url{https://doi.org/10.7910/DVN/A0N5GV}.
+Both pseudo-codes are available in~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
+Density grids are also available at this address.
+Simulation data are available for reproducibility on the dataverse repository at \url{https://doi.org/10.7910/DVN/A0N5GV}.
 
@@
 For each density grid, we obtain the average squared distance between corresponding points of the phase diagrams, i.e.
-the \replaced{mean}{median} value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated).
+the mean value of the final output measure, such as segregation or inequality, for a given value of parameters in the two setups (isotropic and generated).
 This average squared distance for each point is then related to the average variance of each of the phase diagram (the reference one and the one for the grid under inquiry).
 Therefore, values greater than 1 will mean that inter-diagram variability is more important than intra-diagram variability.
-\added{We tested the sensitivity to the type of distance, using a Minkovski distance with a varying exponent.
-The results are presented in~\ref{app:distances}, and show a similar sensitivity to spatial initial conditions.}
+We tested the sensitivity to the type of distance, using a Minkovski distance with a varying exponent.
+The results are presented in~\ref{app:distances}, and show a similar sensitivity to spatial initial conditions.
 
-\paragraph{Sugarscape}
+\subsubsection{Sugarscape}
 
@@
 Indeed, the relative distance between the phase diagrams of different density grids and the phase diagram of the reference case ranges from 0.09 to 2.98 with a median of 1.52 and an average value of 1.30.
-The mean distance above 1 means that, on average, the model is more sensitive to \replaced{the generator parameters}{meta-parameters of grid generation} than to its own parameters (population and sugar endowment) in the reference model.
+The mean distance above 1 means that, on average, the model is more sensitive to the generator parameters than to its own parameters (population and sugar endowment) in the reference model.
 Moreover, the maximum distance of 2.98 means that the variation due to the change of grid can be up to three times bigger than the variation due to the model parameters.
-We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in the \replaced{generator parameter}{meta-parameter} space.
+We plot in Fig.~\ref{fig:sugarscape-distance-meta} the distribution of these distances in the generator parameter space.
 Each point represent one of the 50 different density grids used to initialise the distribution of sugar in the model.
-The points are projected with respect to the \replaced{generator parameters}{meta-parameters used to generate the grid}, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case.
+The points are projected with respect to the generator parameters, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case.
 Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e.
@@
 %%%%%%%%%%%%%
-\begin{figure*}
+\begin{figure}[!t]
 \centering
 \includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams.png}
-\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by \replaced{their generator parameters}{the meta-parameters of their generator}.} Relative distance as a function of \replaced{generator parameters}{meta-parameters} $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
+\caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by their generator parameters.} Relative distance as a function of generator parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
 \label{fig:sugarscape-distance-meta}
-\end{figure*}
+\end{figure}
 %%%%%%%%%%%%%
 
-Another way of quantifying the density grids, instead of looking at the \replaced{generator parameters}{meta-parameters of their generator}, is to look at the resulting indicators of urban form, such as Moran's I, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and context).
+Another way of quantifying the density grids, instead of looking at the generator parameters, is to look at the resulting indicators of urban form, such as Moran's I, average distance, rank-size slope and entropy (see~\cite{LeNechet2015} for precise definition and context).
 This 4-dimensional space defined a morphological space.
@@
 The first component defines a ``level of sprawl'' and of scattering, whereas the second one represents the level aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation (top left of figure \ref{fig:sugarscape-distance-pca}).
-It is confirmed by the behaviour as a function of \replaced{generator parameters}{meta-parameters}, as high values of $\alpha$ also yield high distance.
+It is confirmed by the behaviour as a function of generator parameters, as high values of $\alpha$ also yield high distance.
 In terms of model processes, it shows that congestion mechanisms in the gathering of the resource induces fast increases of inequality.
@@
 %%%%%%%%%%%%%
-\begin{figure*}
+\begin{figure}[!t]
 \centering
@@
 \label{fig:sugarscape-distance-pca}
-\end{figure*}
+\end{figure}
 %%%%%%%%%%%%%
 
-\paragraph{Schelling} 
+\subsubsection{Schelling} 
 Within a standard Schelling model (i.e.
@@
 
-\paragraph{Schelling}
+\subsubsection{Schelling}
 
@@
 \textbf{Density Grid = Discontinuous}       &            & 0.111 ***                        &                   & 0.068 ***         &                      & \textit{0.00}              \\
-\textbf{Attraction \replaced{generator }{meta-}parameter $\alpha$} &            & -0.083 ***                       &                   & -0.053 ***        &                      & 0.014 ***                \\
-\textbf{Diffusion \replaced{generator }{meta-}parameter $\beta$}   &            & 0.323 ***                        &                   & 0.218 ***         &                      & 0.017 ***           \\ \hline
+\textbf{Attraction generator parameter $\alpha$} &            & -0.083 ***                       &                   & -0.053 ***        &                      & 0.014 ***                \\ 
+\textbf{Diffusion generator parameter $\beta$}   &            & 0.323 ***                        &                   & 0.218 ***         &                      & 0.017 ***           \\ \hline
 \textbf{R2 (\%)}                            & 30.6       & 34.7                             & 24.1              & 25.6              & 23.9                 & 24.0                    \\ 
@@
 In table \ref{tab:regressionSchelling}, we see that the type of density grid with which the model is initialised correlates to a certain extent with the level of segregation measured at the end of the simulation run.
-Indeed, compared to \added{the reference case of} compact \added{(monocentric)} density patterns, polycentric grids produce more dissimilarity and entropy between the location of green and red agents.
+Indeed, compared to the reference case of compact (monocentric) density patterns, polycentric grids produce more dissimilarity and entropy between the location of green and red agents.
 Discontinuous grids have the same effect, although attenuated.
@@
 
-\paragraph{Sugarscape}
+\subsubsection{Sugarscape}
 
@@
 In sprawled spaces, inequalities are thus fostered by a lack of minimal local resources, whereas population will drive inequality in concentrated spaces.
-The process is thus completely reversed depending on the grid chosen to run the model \added{on}, which would have significant impacts if one tried to deduce policy from this model.
+The process is thus completely reversed depending on the grid chosen to run the model on, which would have significant impacts if one tried to deduce policy from this model.
 %This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
@@
 \label{fig:sugarscape-phasediagrams}
-\end{figure*}
+\end{figure}
 %%%%%%%%%%%%%
@@
 
-\paragraph{Comparing phase diagrams} Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point.
-For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \deleted{\emph{meta phase transitions}} phase transitions in the space of \replaced{generator}{meta} parameters.
+\subsubsection{Comparing phase diagrams}
+
+Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point.
+For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong phase transitions in the space of generator parameters.
 The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\citep{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools.
@@
 To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\citep{10.1371/journal.pone.0178165}.
-\added{The transfer of methods used to compare sequences \citep{kruskal1983overview} or time-series \citep{liao2005clustering} is a possible way to develop measures between phase diagrams.
+The transfer of methods used to compare sequences \citep{kruskal1983overview} or time-series \citep{liao2005clustering} is a possible way to develop measures between phase diagrams.
 The higher dimension of the phase diagrams we study must however be considered with caution when transferring methods, in a way analog to the application of global sensitivity indexes to spatial data \citep{lilburne2009sensitivity}.
-We can also note than more generally, this problem of comparing phase diagrams is a particular instance of the more generic issue of comparing patterns, which for example include unsupervised learning techniques \citep{hastie2009unsupervised}.} The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.
+We can also note than more generally, this problem of comparing phase diagrams is a particular instance of the more generic issue of comparing patterns, which for example include unsupervised learning techniques \citep{hastie2009unsupervised}.
+The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.
 
-\paragraph{Platform constraints and docking challenges}
+\subsubsection{Platform constraints and docking challenges}
 
 An aspect that we have not touched upon in the article with respect to the sensitivity to initial spatial conditions is the importance of the modelling platform as a constraint in the formalisation of space.
-For example, \replaced{spatial structure may be easier to implement as a raster rather a vector in NetLogo models, which could influence the implementation choices of some non-experienced modelers}{the use of Netlogo fosters the use of raster/grid spatial inputs rather than vector elements}.
+For example, spatial structure may be easier to implement as a raster rather a vector in NetLogo models, which could influence the implementation choices of some non-experienced modelers.
 Its toroidal default setting might also have influenced the work of many modellers who did not question explicitly the representation of space.
@@
 
-\paragraph{Reproducibility and Applicability}
+\subsubsection{Reproducibility and Applicability}
 
@@
 They are available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
-\added{Future work could be done to compare these or generate grids with a larger morphological span, covering other typical urban forms that can be found in the world.}
+Future work could be done to compare these or generate grids with a larger morphological span, covering other typical urban forms that can be found in the world.
 
-Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents, \added{transportation networks generators \citep{raimbault2018multi}, or coupled road network and population raster generators \citep{raimbault2018urban}.}
+Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents, transportation networks generators \citep{raimbault2018multi}, or coupled road network and population raster generators \citep{raimbault2018urban}.
 % Generators of abstract urban systems at a macroscopic scale could also be useful tools to better understand urban growth models as sketched by \cite{raimbault2018unveiling}.}
 
-\paragraph{An emancipation opportunity for social sciences.}
+\subsubsection{An emancipation opportunity for social sciences.}
 
 As~\citet{pumain2003approche} points out in an overview of complexity approaches in geography, transfer of models and concepts between disciplines may induce a transfer of corresponding assumptions.
-Geography and the social sciences in general have been strongly influenced by physics in the last decades, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space \added{in basic models}.
+Geography and the social sciences in general have been strongly influenced by physics in the last decades, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space in basic models.
 We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity the for social sciences to build their own stream of methodologies in the modelling domain.
@@
 We did so by implementing a spatial generator whose output is used as input for the simulation model.
-We applied this approach to two \replaced{textbook}{mythical} ABMs: Schelling and Sugarscape.
+We applied this approach to two textbook ABMs: Schelling and Sugarscape.
 With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
@@
 
-%%%%%%%%%%%%
-%% References
-%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% End of  paragraph numbering. Please leave this untouched
+\endparano
 
-\bibliographystyle{elsarticle-harv}
-\bibliography{spacematters}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\appendix
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%\newpage
+% APPENDICES
+% Put your appendices here. Please use the normal sectioning command, e.g.,
+% \section{Appendix A: <title of the appendix>}
+% \section{Appendix B: <title of the appendix>}
+% ...
 
-\section{\added{Behavior of the density grid generator}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\added{We give in Fig.~\ref{fig:stylized-generator} plots roughly summarizing the behavior of the density grid generator, according to parameters $(\alpha,\beta)$.
+\section{Appendix A: Behavior of the density grid generator}
+
+We give in Fig.~\ref{fig:stylized-generator} plots roughly summarizing the behavior of the density grid generator, according to parameters $(\alpha,\beta)$.
 To put it in a simple way, high values of $\alpha$ give highly hierarchical configurations, and diminishing $\beta$ increase the number of centers.
@@
 We do not discuss here the role of other parameters, but according to \cite{raimbault2018calibration}, diffusion steps give smoother forms, and the rate between total population and the population increment at each step (which is equivalent to the total number of steps) is crucial to select non-stationary distributions that are closer to real configurations.
-\cite{raimbault2018calibration} also shows the existence of non-linear behaviors in some regions of the parameter space, so the description we gave here shall not be interpreted as a linear link between generator parameters and the morphological properties of the generated grids.}
+\cite{raimbault2018calibration} also shows the existence of non-linear behaviors in some regions of the parameter space, so the description we gave here shall not be interpreted as a linear link between generator parameters and the morphological properties of the generated grids.
 
 %%%%%%%%%%%%%%
-\begin{figure*}
+\begin{figure}[!t]
+	\centering
 	\includegraphics[width=0.49\linewidth]{figures/alpha_beta_examples.png}
     \includegraphics[width=0.47\linewidth]{figures/alpha_beta_schemas.png}
-    \caption{\added{\textbf{Stylized behavior of the density grid generator.} (Left) Examples of grids with their position in the $(\alpha,\beta)$ parameter space; (Right) Stylized interpretation of the forms obtained as a function of $\alpha$ and $\beta$.}\label{fig:stylized-generator}}
-\end{figure*}
+    \caption{\textbf{Stylized behavior of the density grid generator.} (Left) Examples of grids with their position in the $(\alpha,\beta)$ parameter space; (Right) Stylized interpretation of the forms obtained as a function of $\alpha$ and $\beta$.}\label{fig:stylized-generator}
+\end{figure}
 %%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%
-\section{\added{Additional statistical analysis}}
+\section{Appendix B: Additional statistical analysis}
 
@@
 
-\added{For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\mathbb{E}}\left[X\right]/\hat{\sigma}(X)$ with standard estimators for average and standard deviation.
+For both models we estimate the Sharpe ratios for each indicator by $S(X) = \hat{\mathbb{E}}\left[X\right]/\hat{\sigma}(X)$ with standard estimators for average and standard deviation.
 The summary statistics of this ratio computed on repetitions for all parameter points are given in Table~\ref{tab:summarysharpes}.
@@
 This gives 64 repetitions for $\kappa = 1$.
-As the Sharpe ratios are in general smaller for Schelling indicators than for Sugarscape, we take $n = 50$ for Sugarscape and $n = 100$ for Schelling to have a similar confidence in estimations.}
+As the Sharpe ratios are in general smaller for Schelling indicators than for Sugarscape, we take $n = 50$ for Sugarscape and $n = 100$ for Schelling to have a similar confidence in estimations.
 
 %%%%%%%%%%%%%%
-\begin{table*}
-\caption{\added{\textbf{Summary statistics of Sharpe ratio estimated on repetitions for each parameter point.}}\label{tab:summarysharpes}}
+\begin{table}[!t]
 \centering
@@
     \end{tabular}
-\end{table*}
+    \caption{\textbf{Summary statistics of Sharpe ratio estimated on repetitions for each parameter point.}}
+\label{tab:summarysharpes}
+\end{table}
 %%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%
-\section{\added{Additional figures for the Schelling model}}
+\section{Appendix C: Additional figures for the Schelling model}
 
-\added{The Fig.~\ref{fig:schelling-distance-meta} gives the phase diagrams distances as a function of generator parameters and morphological components, similarly to the Sugarscape model in main text.
+The Fig.~\ref{fig:schelling-distance-meta} gives the phase diagrams distances as a function of generator parameters and morphological components, similarly to the Sugarscape model in main text.
 In absolute, this version of the Schelling model seems less sensitive to density grids than the sugarscape model, as we do not obtain a high range of values here.
-We however obtain measures ranging from 0 to 0.85 with the euclidian distance, what is however characteristic of a significant sensitivity to space.}
+We however obtain measures ranging from 0 to 0.85 with the euclidian distance, what is however characteristic of a significant sensitivity to space.
 
@@
 \caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams.
-We show them in the \replaced{generator }{meta-}parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
-\end{figure*}
+We show them in the generator parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
+\end{figure}
 %%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%
-\section{\added{Comparison of phase diagram with other distances}}
+\section{Appendix D: Comparison of phase diagram with other distances}
 
@@
 
-\added{We describe here the tests done with other distances to compare phase diagrams.
+We describe here the tests done with other distances to compare phase diagrams.
 We tested normalized Minkovski distances, defined by $d(x,y) = \left(\frac{1}{N}\cdot \sum_i \left|x_i - y_i\right|^{q}\right)^{\frac{1}{q}}$, for varying values of $q$ from $q = 1$ (Manhattan distance) to $q = 10$, including $q = 2$ (Euclidian distance) which is used in main text.
@@
 We naturally obtain smaller difference with the Manhattan distance but which remain significant (averages of 10\% for the Schelling model and 40 \% for sugarscape), and variabilities with higher values of the Minkovski exponent are much higher.
-This results confirm the high variability observed in main text with the Euclidian distance.}
+This results confirm the high variability observed in main text with the Euclidian distance.
 
@@
 
-\begin{table*}
-	\caption{\added{\textbf{Summary statistics of different distances for schelling.}}}\label{tab:schellingdistances}
+\begin{table}[!t]
+
 	\centering
@@
     \end{tabular}
-\end{table*}
+    \caption{\textbf{Summary statistics of different distances for schelling.}}\label{tab:schellingdistances}
+\end{table}
 
-\begin{table*}
-	\caption{\added{\textbf{Summary statistics of different distances for sugarscape.}}}\label{tab:sugarscapedistances}
+\begin{table}[!t]
+
 	\centering
@@
     \end{tabular}
-\end{table*}
+    \caption{\textbf{Summary statistics of different distances for sugarscape.}}\label{tab:sugarscapedistances}
+\end{table}
 
 %%%%%%%%%%%%%%%
-\section{\added{Pseudo-code for models}}
+\section{Appendix E: Pseudo-code for models}
 \label{app:code}
 
-\added{We give below the pseudo-code for the implementations we used of both sugarscape (Fig.~\ref{frame:sugarscapecode}) and Schelling (Fig.~\ref{frame:schellingcode}) models.
-We recall that source code is openly available at} \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
-\added{The pseudo-code is in the style of NetLogo code, which is already easily readable.}
+We give below the pseudo-code for the implementations we used of both sugarscape (Fig.~\ref{frame:sugarscapecode}) and Schelling (Fig.~\ref{frame:schellingcode}) models.
+We recall that source code is openly available at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
+The pseudo-code is in the style of NetLogo code, which is already easily readable.
 
 %%%%%%%%%%%%%%%
-\begin{figure*}
-\caption{\added{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}}\label{frame:schellingcode}
+\begin{figure}[!t]
+\caption{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}\label{frame:schellingcode}
 \bigskip
@@
 		let wantedcolor color
-		let stranger-rate count neighboragents with [color = wantedcolor] /
-			count neighboragents
+		let stranger-rate count neighboragents
+		   with [color = wantedcolor] / count neighboragents
 		if stranger-rate < similar-wanted [
@@
     \end{mdframed}
-\end{figure*}
+\end{figure}
 %%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%
-\begin{figure*}
+\begin{figure}[!t]
 % Minimal netlogo code for the sugarscape model
-\caption{\added{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} Conventions are the same than for Schelling pseudo-code.}}\label{frame:sugarscapecode}
+\caption{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} Conventions are the same than for Schelling pseudo-code.}\label{frame:sugarscapecode}
 \bigskip
@@
 \end{mdframed}
-\end{figure*}
+\end{figure}
 %%%%%%%%%%%%%%%
 
+\bibliographystyle{jasss}
+\bibliography{spacematters}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \end{document}

### 2018-12-13 coverletter.tex
Update on Overleaf.

@@
 % recipient data
-\recipient{Editors CEUS}{}
-\date{June 13, 2018}
+\recipient{Editors JASSS}{}
+\date{December 13, 2018}
 \opening{Dear Editors,}
@@
 
+% the authors' postal and email addresses and the authors' home pages on the web
+
+% authors bio
+% Juste Raimbault is a PhD student at Université Paris 7, UMR CNRS 8504 Géographie-cités and UMR-T 9403 IFSTTAR LVMT, under the supervision of Arnaud Banos and Florent Le Néchet.
+He holds an engineer degree from Ecole Polytechnique and Ecole des Ponts et Chaussées, and a Master in Complex Systems Science from Ecole Polytechnique.
+His research focuses on complexity approaches to the relations between networks and territories.
+% UPS CNRS 3611 ISC-PIF, 113 rue Nationale 75013 Paris
+% juste.raimbault@polytechnique.edu
+% http://parisgeo.cnrs.fr/spip.php?article6835&lang=en
+
+% Dr.
+Clémentine Cottineau is a research associate at the Centre for Advanced Spatial Analysis of University College London.
+She studied geography and economics at University Paris 1 Panthéon-Sorbonne and holds a PhD in geography from this university.
+Her thesis aimed at modelling urbanisation in the post-Soviet space with modular agent-based models.
+Since 2014, she has been working at the Centre for Advanced Spatial Analysis on urban economics and parametric scaling in the context of European systems of cities, using tools from complexity science to investigate inequality in cities and economic disparities between cities.
+% UMR CNRS 8097 Centre Maurice Halbwachs, 48 boulevard Jourdan 75014 Paris
+% clementine.cottineau@ens.fr
+% http://clementinecttn.github.io/
+
+% Dr.
+Marion Le Texier is associate professor at University of Rouen Normandie.
+She holds a PhD in geography from both Paris Diderot University and Luxembourg University.
+Her thesis aimed at modelling the circulation of euro coins through European borders with statistical and agent-based models.
+Previous to joining Rouen Normandie University in 2016, she has been a Jean Monet fellow at the European University Institute in Florence and a post-doc at Luxembourg University.
+Her work deals with data and modelling issues in different thematic contexts: access to nature in cities, international online media coverage of earthquakes, to name but two.
+% UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France
+% marion.le-texier@univ-rouen.fr
+% http://umr-idees.fr/user/marion-le-texier/
+
+% Dr.
+Florent Le Néchet is associate professor at  Paris-Est Marne-la-Vallée University (UPEM).
+He holds a PhD in planning from Paris-Est University (UPE).
+His thesis aimed at modelling the relationships between urban form and mobility patterns at metropolitan level, comparing several European cities, especially in France and Germany.
+His work deals with charactering and measuring mobility in cities, in relation to the built environment, and involves statistical techniques as well as multi-agent modelling.
+% Universit{\'e} Paris-Est, Laboratoire Ville Mobilit{\'e} Transport, Marne-la-Vallée, France
+% Florent.Lenechet@u-pem.fr 
+% http://www.lvmt.fr/equipe/florent-le-nechet/
+
+% Dr.
+Romain Reuillon is a research fellow at UMR CNRS 8504 Géographie-cités and at the Complex System Institute Paris Ile-de-France.
+He holds a PhD in Computer Science from Université Clermont II.
+His work deals with intensive computation model exploration, focusing on new techniques and methods to extract knowledge from models of simulation.
+He is the creator of the OpenMole platform for model exploration.
+% UPS CNRS 3611 ISC-PIF, 113 rue Nationale 75013 Paris
+% romain.reuillon@iscpif.fr
+%
+
 \justify
-My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in Computer, Environment and Urban Systems.
+My coauthors and I are pleased to submit an original research article entitled ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models'' for consideration for publication in the Journal of Artificial Societies and Social Simulation.
 
@@
 the spatial configuration of agents and their environment), and more precisely their spatial initial conditions.
-We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of CEUS, a journal dealing with computerised models and urban systems.
+We think the subject in general and the novelty of the paper in particular will meet the interest of the readers of JASSS, a journal dealing with computerised models and social systems.
 
 Despite some preliminary results being presented as at ECTQG 2015, Bari, and at Geocomputation Conference 2017, Leeds, we confirm that neither the manuscript nor any parts of its content are currently under consideration or published in another journal.
-We did not have any previous interaction with CEUS regarding this manuscript.
+We did not have any previous interaction with JASSS regarding this manuscript.
 We have no conflicts of interest to disclose.
-We do not have any opposed reviewer.
-\justify
+
+%We do not have any opposed reviewer.
+%We would like to oppose the following reviewer: Pr.
+I.
+Benenson, University of Tel-Aviv.
+As an potential editor for a previous version of this manuscript which was finally rejected from the journal Computer, Environments and Urban Systems, he tried to make us change the purpose and message of the paper, wanting us to rewrite the paper as a targeted study of the Schelling model, ignoring the majority of reviewers comments and getting out of his role as an editor.
+This non objective relation between him and us had already began in 2015 at ECTQG when it was presented.
+For this reason, we do not think to would be scientific nor productive that he would be a reviewer of this paper.
 

### 2019-06-11 paper.tex
Update on Overleaf.
clean_figs

@@
 \centering
-\includegraphics[width=0.9\textwidth]{figures/SchemaMeta_1.png}
+%\includegraphics[width=0.9\textwidth]{figuresraw/SchemaMeta_1.png}
+\includegraphics[width=0.9\textwidth]{figures/Fig1.png}
 \caption{\textbf{General workflow.}} \label{fig:method}
@@
 $d$ is a functional distance that we take simply as Euclidean distance.
-The internal variabilities are estimated as the variance within each phase diagram $f_{\vec{\alpha_i}}$.
+The internal variabilities are estimated as the variance within each phase diagram $\mu_{\vec{\alpha_i}}$.
 For a multi-dimensional phase diagram, we average these relative distances over the components.
@@
 \centering
-	\includegraphics[width=\textwidth]{figures/spatialGen.png}
+	%\includegraphics[width=\textwidth]{figuresraw/spatialGen.png}
+	\includegraphics[width=\textwidth]{figures/Fig2.png}
  \caption{\textbf{Four examples of grids produced by the spatial generator.} The lighter the red, the denser the area.
@@
 \centering
-	\includegraphics[width=0.9\textwidth]{figures/GridPresentation.png}
+	%\includegraphics[width=0.9\textwidth]{figuresraw/GridPresentation.png}
+	\includegraphics[width=0.9\textwidth]{figures/Fig3.png}
  \caption{Correspondence between European urban density structures and grids produced with the spatial generator.} 
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/relativedistance_metaparams.png}
+	%\includegraphics[width=0.8\textwidth]{figuresraw/relativedistance_metaparams.png}
+	\includegraphics[width=0.8\textwidth]{figures/Fig4.png}
 \caption{\textbf{Relative distances of phase diagrams by initial spatial grids described by their generator parameters.} Relative distance as a function of generator parameters $\alpha$ (strength of preferential attachment) and $\beta$ (strength of diffusion process).}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/relativedistance_morphspace.png}
+	%\includegraphics[width=0.8\textwidth]{figuresraw/relativedistance_morphspace.png}
+	\includegraphics[width=0.8\textwidth]{figures/Fig5.png}
 \caption{\textbf{Relative distances of phase diagrams to the reference across grids.} Relative distance as a function of two first principal components of the morphological space (see text).
@@
 \centering
-\includegraphics[width=0.48\textwidth]{figures/phasediagram_id27_maxSugar110}
-\includegraphics[width=0.48\textwidth]{figures/phasediagram_id0_maxSugar110}
+%\includegraphics[width=0.48\textwidth]{figuresraw/phasediagram_id27_maxSugar110}
+%\includegraphics[width=0.48\textwidth]{figuresraw/phasediagram_id0_maxSugar110}
+	\includegraphics[width=\textwidth]{figures/Fig6.png}
 \caption{\textbf{Examples of phase diagrams.} We show two dimensional phase diagrams on $(P,s_-)$, both at fixed $s_+ = 110$.
@@
 	\centering
-	\includegraphics[width=0.49\linewidth]{figures/alpha_beta_examples.png}
-    \includegraphics[width=0.47\linewidth]{figures/alpha_beta_schemas.png}
+	%\includegraphics[width=0.49\linewidth]{figuresraw/alpha_beta_examples.png}
+    %\includegraphics[width=0.47\linewidth]{figuresraw/alpha_beta_schemas.png}
+    \includegraphics[width=\textwidth]{figures/Fig7.png}
     \caption{\textbf{Stylized behavior of the density grid generator.} (Left) Examples of grids with their position in the $(\alpha,\beta)$ parameter space; (Right) Stylized interpretation of the forms obtained as a function of $\alpha$ and $\beta$.}\label{fig:stylized-generator}
@@
 \centering
-\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams_mean.png}
-\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace_mean.png}
+%\includegraphics[width=0.48\textwidth]{figuresraw/schelling-relativedistance_metaparams_mean.png}
+%\includegraphics[width=0.48\textwidth]{figuresraw/schelling-relativedistance_morphspace_mean.png}
+\includegraphics[width=\textwidth]{figures/Fig8.png}
 \caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams.

### 2019-06-11 figures.sh

@@ -1,0 +1,24 @@
+
+# fig 1
+cp figuresraw/SchemaMeta_1.png figures/Fig1.png
+
+# fig 2
+cp figuresraw/spatialGen.png figures/Fig2.png
+
+# fig 3
+cp figuresraw/GridPresentation.png figures/Fig3.png
+
+# fig 4
+cp figuresraw/relativedistance_metaparams.png figures/Fig4.png
+
+# fig 5
+cp figuresraw/relativedistance_morphspace.png figures/Fig5.png
+
+# fig 6
+montage figuresraw/phasediagram_id27_maxSugar110.png figuresraw/phasediagram_id0_maxSugar110.png -tile 2x1 -geometry +5+0 figures/Fig6.png
+
+# fig 7
+montage figuresraw/alpha_beta_examples.png figuresraw/alpha_beta_schemas.png -tile 2x1 -geometry +50+0 figures/Fig7.png
+
+# fig 8
+montage figuresraw/schelling-relativedistance_metaparams_mean.png figuresraw/schelling-relativedistance_morphspace_mean.png -tile 2x1 -geometry +5+0 figures/Fig8.png

### 2019-06-11 Response_to_Reviewers.tex

--- 2018-09-28 Response_to_Reviewers.tex
@@
 % recipient data
-\recipient{Editor CEUS}{}
+\recipient{Editor JASSS}{}
 \date{\today}
@@
 
-\makelettertitle
+%\makelettertitle
 
@@
 
-Thank you for the feedback on the draft paper ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models''.
+%Thank you for the feedback on the draft paper ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models''.
 The suggestions and comments will undoubtedly be of great value to the paper.
@@
 
-% Thank you for submitting your manuscript to Computers, Environment and Urban Systems.
-I have completed the review of your manuscript and a summary is appended below.
-The reviewers recommend reconsideration of your paper following major revision.
-I invite you to resubmit your manuscript after addressing all reviewer comments.
+\textbf{Response to reviewers}
 
-% When revising your paper, pay special attention to the explicit presentation of every model that is investigated in the paper.
-Present each model as an algorithm/unambiguous pseudocode and supply the Netlogo code for downloading (all this can be in appendix).
-The Schelling model, for example, has several Netlogo implementation, and it is critically important to know which of them is investigated.
-The same is true for the Sugarscape model.
-See comments of the reviewer #1 in this respect.
+\bigskip
 
-% When resubmitting your manuscript, please carefully consider all issues mentioned in the reviewers' comments, outline every change made point by point, and provide suitable rebuttals for any comments not addressed.
+%We now turn to point-by-point responses to referees comments.
 
-As a response to several reviewers, we made the following general changes:
-
-\begin{itemize}
-	\item Several comments, from the first and third referees especially, suggested that we clarify and explicit the description of our two models: we clarified the descriptions and justified our choices more thoroughly, and also provided pseudo-codes for each model in the Appendix so as to help the reader navigate the implementation chosen compared to other articles.
-Finally, we mentioned more clearly in the text the fact that source codes are openly available in a github repository.
-    \item In order to conciliate as much as possible the different points of view about the comparison of phase diagram, we added some developments and comments about how we defined, built, justified and compared phase diagrams in the paper.
-    \end{itemize}
+\textbf{Point-by-point response to referees comments:}
 
@@
 
-We now turn to point-by-point responses to referees comments.
+\textbf{First referee:}
 
@@
 
-Concerning specifically the comments made by the first referee:
+%The paper describes an approach that takes spatial sensitivity analysis explicitly and thoroughly into account.
+Based on the criticism that most agent-based-modelling and simulation approaches do neglect a careful recognition of the spatial impact on social phenomena, it introduces a “spatial generator” whose task is to generate different possible spatial configurations that are then used as a comparative measurement in socio-spatial simulations to prove the influence of spatial structures.
+To evaluate the strengths and weaknesses of the use of the spatial generator tool the author(s) applies two classical approaches – Schelling’s segregation model and the Sugarscape model – as benchmark models.
 
-\begin{itemize}
-	% 1.
-I think it would be better if the author would focus on a single model.
-This way they could describe the model and the analyses in more detail.
-As it stands, the models are treated as black boxes.
-	\item The suggestion to focus on a single model is a qualified comment.
-However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies rather than with a single model.
-In that perspective, the ``black-box'' is seen as an advantage as it reinforces genericity.
-We acknowledge the fact that this choice led us to shorten the model presentations.
-Nevertheless, we have taken notice of your remark by improving the description of the models and their results through the addition of pseudo-codes in Appendix, result descriptions in the text and by recalling more clearly the link to the model source we used.
-Furthermore, we have focused on some of the most classical models of the ABM literature so as to lower the reader's cost of entry into the methodology.
+\begin{enumerate}
+ % The methodology described is very convincing as it allows the utilisation of a tool that takes the spatial conditions – as causes or effects or intermediates – actually explicitly into account.
+With this approach, the paper convincingly outlines a pragmatic application by creating typical spatial compositions in urban contexts that is monocentric, polycentric or discontinuous cities.
+From this methodological perspective of reviewing, it would be helpful to add some more information to approaches/techniques used within the tool of the spatial generator.
+For example, paragraph 2.8 mentions four dimensions of spatial structure but leave the reader alone with just the terms.
+Either describe these or omit it.
+   \item \comment[JR]{will add some description on urban form indicators etc.}
 
-   % I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis
-   \item We agree that illustrative examples would provide a relevant starting point to  the analysis but we were afraid that their addition to the paper would make the introduction less straightforward and have the opposite effect to the one we were seeking.
-As we were nevertheless very interested by your remark, we provide a short answer to your proposition.
+  %This also applies to the three segregation indexes: what exactly do they measure?
+  \item
 
-In order to get a first idea of the potential impact of spatial initial conditions on to the results of the simulation, we implemented on Segregation model a geographical constraint: some cells cannot be inhabited (yellow cells in Figure 1).
-The way those “forbidden” cells are distributed throughout space will indeed influence the steps of the model.
-We introduced various spatial initial conditions in the model that we controlled by one clustering parameter on the forbidden cells.
-A value of 0 would indicate a spatial distribution of yellow cells so as to minimize spatial autocorrelation.
-A value of 1 indicates on the contrary a concentration process that tends towards a high value of autocorrelation (see below Figure 1).
-% note : no fig envt in moderncv
+  % The description of the spatial generator’s procedure, outlined in paragraph 2.7, would also benefit if more information is added, for example, to delineate the terms “strength of attrac!
+tion” (attraction in which respect?) and “strength of the diffusion process” (is strength equal to intensity?
+If so, is it then in space and/or time?).
+  \item \comment[JR]{will add more precision in the description of the generator}
 
-We then compute the number of communities arising when the model converge, hence the number of different related components of similar neighbors.
-Using 270 replications of 11 values of the clustering parameter (from 0 to 1 by step 0.1), we obtained a significant growth in terms of number of communities when the constraints on residential locations are concentrated in space, which might be interpreted in such example as a bigger segregation when density of housing is heterogeneously distributed through space.
-%In the rest of the article, we wish to further investigate such spatial effects in a much more systematic way.
+  %While the methodological part of the paper is convincing the theoretical and epistemological background is less so.
+In short, the paper would improve significantly, if most of the geographical and empirical associations would be left out.
+Even claimed by the author(s), the approach has not much to do with the empirical growth or development of cities, just because the approach reduces cities’ growth to the growth itself, omitting issues such as historical genesis, cultural differentiation, economic background political and legal circumstances, etc.
+  \item \comment[JR]{most of this empirical literature was added because of one CEUS reviewer, no ?}
 
-%%%%%%%%%%%%%%
-\begin{minipage}{\textwidth}
-\bigskip\bigskip
-\begin{center} 
-	\includegraphics[width=\textwidth]{figures/SpaceMatters2.png}
+  % The problem, from a geographical point of view, is that the paper switches between “geographical” and “spatial” without making any differentiation.
+In so doing, the author(s) runs immediately into the problem of adequate definitions.
+In paragraph 1.2 the definition of geographical systems as “social agents interacting with one another, within a limited portion of space” is misleading.
+First, it is a tautology (geographical systems in space); secondly, geographical cannot be described with social agents, because geographical systems are composed of spatial units, their relations, and a spatially defined system border.
+Social systems, in turn, might be described with humans as elements, their relations, and again a socially defined system border.
+Moreover, with respect to sociological system theory (Luhmann 1993), space does not add any knowledge to the nature of social systems.
+This approach can be (and I would add: has to be) criticised, however, not by interming!
+ling fuzzily the social with the geographical, as happened with this definition.
+  \item \comment[JR]{by spatial we indeed mean some kind of spatial container for agents right ? which is far from geographical in the sense he puts it indeed}
 
-Fig. 1: \textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (Netlogo implementation).}
-\end{center}
-\bigskip\bigskip
-\end{minipage}
-%%%%%%%%%%%%%%
+  % Then, in paragraph 1.3, the complexity of systems with respect to segregation is also problematically described.
+Emergent behaviour that may lead to segregation is first and foremost a social process, taking social interactions at the local scale (income, housing costs, etc.) and/or at the macro scale (laws, rules, economic system, cultural attitudes, etc.) into account.
+These interactions can then be associated with the material and relational dimension of geographical spaces.
+The question remains: what is the role of geographical systems?
+In this respect, the paper offers some relevant clues such as the MAUP problem or the “uncertain geographic context problem”.
+  \item
 
-	% 2.
-It is very important to describe the models.
-This is especially relevant for the Schelling model, which has many variants; each has different dependencies on the initial pattern.
-In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.
-	\item There was a misunderstanding for the Schelling implementation actually used, as we used our own implementation which pseudo-code has been added in Appendix of the paper and source code is openly available on the git repository of the project.
-This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.
+  % The problem of setting space and geography as synonymous continues in paragraphs 1.9 and following.
+What is described here as “spatial system shape” is the geometry of spaces but not the geography of spaces.
+  % -> yes indeed
+  \item
 
-    \item More generally concerning the choice of these two models, we aim at helping the readers to distance themselves from the case studies and focus on the genericity of the approach.
-This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs.
-This is again an argument in favor of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.
+  % Even more critically is the arbitrary conflation of social and spatial units as happened in paragraph 2.1, as geographical entities are described with “people, housing, networks, etc.” One is inevitably asking oneself: do geographical entities comprise of simply everything?
+Obviously not, but the paper seems to argue this way.
+  \item
 
-	% 3. I am not sure why the Sugarscape model has been chosen given that the paper seems to be focused on cities.
-	\item Furthermore concerning the choice of the Sugarscape model, we think it is a good example of a stylized model of a spatialized socio-economic system, and according to (Batty, 2005) a good metaphor of an urban system.
-Our choice was accordingly clarified in the paper.
+  % Due to this, attempts to explain the growth of existing cities should be avoided.
+The approach is reductionist and has its strength in reflecting growth at an abstract level nevertheless.
+The concluding remark in paragraph 5.1 “With the Schelling experiment, we found … that polycentric and discontinuous cities appear systematically more segregated than compact cities …” is elusive.
+Either it is a trivial fact, simply because polycentric cities have more centres and thus, by nature, more opportunities from which segregation can originate.
+Or, the results can be used as a starting point to dig deeper in geographical reasons, taking social, economic, cultural, and political, but also topographical issues into consideration.
+This geographical reasoning, however, is outside of the scope of this paper.
+  \item \comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate}
 
-	% 4. Consider replacing “meta parameters” with something like “generator parameters” or “initial pattern parameters.”
-	\item We agree that the term of meta-parameters may have been confusing, and following the suggestion we replaced it with ``generator parameters''.
+   % Formal issues that have to be solved: the numbering of paragraphs is not coherent. The reference of Wheeler (2006) is missing the source of the article.
+   \item
 
-	% 5. Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index
-    \item There was a mistake in the writing, as we never use the median but always the mean, which as you recall is consistent with the choice of distance we made. This was corrected in the paper.
+\end{enumerate}
 
-	% 6. Line 232: comparing phase diagrams is related to the topic of comparing patterns. It would make sense to mention the methods that are usually used to compare patterns
-	\item Some general references on comparing patterns were added.
+\bigskip
+\bigskip
 
-	% 7.
-Line 238: Please explain why this particular index was chosen and discuss its properties.
-Connect the discussion to other indices like L1, and Linf norms, etc.
-Would we get qualitatively different results from different indices?
-	\item The approach proposed in the paper is indeed rather exploratory.
-This explains why we have chosen to focus on a single index (L2) as this index is relatively well-known in the community.
-Previous exploratory analyses have shown that the results were rather stable when using the Monge method.
-Adapting the mathematical aggregator to the different indices would constitute an interesting perspective, but is out of the scope of the paper which already focuses on two models and proposes a new framework for analysing the impact of initial pattern parameters.
-We added a supplementary material showing the distances obtained with other distances, which are similar to the results in main text.
+\textbf{Second referee:}
 
-	% 8. Regarding the spatial generator: the authors should explain why this generator is relevant to these two models? 
-	\item This spatial generator is relevant to both model as it produces density grids at a ``metropolitan scale'' (see detailed discussion added in Appendix and the original paper from which we used the generator), which is the scale at which both model were initially intended to be.
-We made this aspect more explicit.
+%This manuscript presents a methodology to explore the effects of spatial parameters in ABM.
+They basically explore what they called “generator parameters”, to distinguish them from other statistical ones, which include density spatial distribution and patterns.
+To do so, they compared phase diagrams and calculated the mean of final values.
+%As case-studies, they used the actual distributions of three European cities as a means to constraint the density generation in two archetypical model (Schelling's segregation model and Epstein and Axtell's Sugarscape).
 
-	\item In the case of Schelling, this is the scale at which most empirical segregation indices are computed.
-In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty, 2005), or to a generic landscape where a resource is grown otherwise.
-In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density, and that not acknowledging this diversity leads to potentially false results.
-Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most common simplification of the heterogeneity of space, with consequences on the model outputs often not evaluated.
+\medskip
 
-	% 9. The authors should illustrate the contribution of each parameter of the generator to the generated initial pattern using a figure. Details of the generators should be included in an appendix
-	\item The behavior of the spatial generator was described with more details in Appendix.
+\begin{enumerate}
 
-	% 10. Explain how the density grids are used in the Schelling model, do you allow more than one agent to be in the same cell? 
-	\item Yes, we allow more agents to be in the same cell.
-The potential density is defined by the density grid generated, and its population with the same number of agents in all grids.
-This distribution is the one at initialisation.
-If the potential density is not reached, more agents can move in the cell, otherwise it is deemed full and unavailable for further moves.
-The satisfaction and segregation indices are computed with regard to the people in the cell and the people present in neighbouring cells.
+  % While the different outcomes of the two implementation exercises are not the main point, I would suggest authors to provide some more elaborated explanations.
+The two models have different behavioural assumptions and represent a different 'interaction' context.
+How can be extrapolate and isolate spatial effects when models have different logics and theoretical purposes?
+  \item
 
-	% 11. Line 297: I am not sure why empirical density destitutions of cities are important in the context of this study. 
-	%\item cf infra. 
-	\item Density distributions of empirical cities are important because we want to test models with realistic ranges of initial patterns of density distribution.
-Therefore we cannot limit ourselves to a single 2-peak mountain in the classical Sugarscape model and an isotropic square city in Schelling.
-We chose to use actual distribution of European cities to constraint our density generation.
+  % In 4.1, the authors suggest potential developments and point to certain limitations.
+While limitations are well-discussed, readers who are not specialist of spatial models could benefit from a more comprehensive discussion on potential applications and extensions.
+For instance, it would be useful to have suggestions on how to use this methodology to run empirical models.
+Suppose to have a nice dataset and a case of a urban context to be modelled.
+Could the exploration of different spatial configurations even be used to run 'counter-factual' or 'what if' tests on actual spatial configurations?
+  \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}
 
-	% 12. Line 641: NetLogo supports grids, continuous space, and GIS elements. It does not restrict the user in any way.
-	\item We agree that NetLogo does not restrict the use on what implementation to use.
-However, some features are much more easier as raster or the contrary, which could influence the choice of beginners.
-Furthermore, models in the built-in NetLogo model library mostly rely on rasters.
-We clarified that statement in the paper.
+  % Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), whcih could be interesting to see implemented in ABM.
+Spatial econometrics based on empirical dataset is developing fast.
+It would be useful to understand how these models can enrich ABM and vice-versa.
+  \item
 
-	% 13. Line 301: I am not sure what the authors mean by “interesting grids”
-	\item The term ``interesting grids'' was a typo and corrected, ``grids'' only was meant.
+  % Furthermore, even among limitations, perhaps authors could discuss data limitations and the fact that social processes (networks, mobility) cannot conflate with spatial dynamics.
+Here, it is a bit misleading to call "generator parameters" ONLY the spatial ones!
+  \item
 
-	% 14. Line 671: The same comment regarding “mythical ABMs” 
-	\item The term ``mythical'' line 671 was a language mistake and corrected, ``textbook'' ABM was meant.
-    %\comment[MLT]{Use "textbook ABM instead?}[(JR) yes, better]
-\end{itemize}
+  % In this respect, I would encourage authors to further develop their closing section, which now does not stimulate readers to imagine implications of authors' findings and message to their own research.
+I would expand this section to provide a more detailed discussion on implications for model makers, experts on social geography and urban scholars who work with ABM, non-specialist in geography who would like to base their models in more proper geographical constraints.
+And perhaps, authors could suggest future line of research that could show-case their argument.
+  \item
 
-Concerning the comments made by the second referee:
+%I appreciate that models are fully available to readers.
 
-%This is an interesting article that I read with true interest. The paper fits the profile of this journal and can reach a wider audience.
+\end{enumerate}
 
-%I also have some concerns:
-
-\begin{itemize}
-	% In the abstract the authors claim that initial spatial conditions are often taken for granted.
-I do not agree to this statement.
-Initial spatial condition can have a large impact on the results of a model, however I do not think that the initial spatial conditions are often taken for granted.
-	\item We forcefully agree with the referee in that initial spatial conditions can have a large impact on the results of the model.
-Maybe not all papers take them for granted, but what we meant is that compared to other parameters, initial spatial conditions are much less often explored using sensitivity analysis.
-Actually, we know only a few papers which have explored the impact of initial spatial conditions as a systematic criterion in a sensitivity analysis of model results.
-We clarified this point in the introduction.
-
-	% Page 2 line 88: the authors state that “Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density”, I do not think this is true.
-Many models generate a heterogeneous synthetic population that is based on local data (replicating the actual population as closely as possible) and in far more detail then what is done in this paper.
-I am thinking of work in the line of disease diffusion, crime etc.
-	\item We acknowledge the fact that many ABM applications use "real" data and spatial features for their simulations.
-However, as explained by (Thomas et al.
-2017) in the case of LUTI models for instance, most simulations do not evaluate the impact of city delineations nor MAUP on the output of the models, which constitutes a potential limit in the interpretation of results.
-
-	% Page 2 line 98: “However, this step is rarely explicit and authors tend to generalize the conclusion of the model from a single, spatial case study”.
-I do not understand the first part of the this statement.
-The steps taken are often very explicit.
-Due to the amount of time it takes to correctly represent all spatial data, including population densities and the variation of other population variables over space, many authors indeed restrict themselves to a single case study due to the time intensity of this task.
-Many agent-based models are also not valid for other areas.
-	\item What we meant was actually that this step is rarely explicit and authors are therefore not able to generalize the conclusion of the simulations to other spatial context.
-We rephrased the paragraph accordingly.
-
-	% The authors do not differentiate between data driven empirical models versus theory-driven models.
-I think a clear differentiation between the two groups and the positioning of the two case studies in the second group would make the paper more balanced.
-	\item Concerning the remarks on synthetic population generation and the different types of models, we clearly stated our positioning regarding models built from theory, for which the proposed sensitivity analysis workflow is more suited and that our two case studies illustrate.
-
-	% In section 1.3 (Previous considerations on the effects of the spatial configuration in simulation models) authors provide a lot of evidence on previous work on the impact of initial spatial conditions on the output of models.
-This is contrary to the statement in the abstract that initial conditions are often taken for granted.
-	\item Indeed, we have tried to review articles where some kind of sensitivity to initial spatial conditions are tackled.
-However we would like to emphasize that the references cited in this section are few compared to the mass of geosimulation articles published with hardly any mention of dependence of output on initial spatial condition.
-Moreover, even if such sensitivity is analysed, it mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but to the best of our knowledge no article provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper.
-We have tried to argue this point at the end of the section.
-
-	% The most interesting part of the research to me, is not the way the density layers are generated but the use of the phase diagrams, defined as a vector of final aggregated model outputs.
-As these are vectors, they are not so different from time series and many techniques exist to compare vectors in this domain.
-I wonder if the Euclidean distance is the best measure to determine the difference between the phase vectors of different spatial inputs.
-	\item The comment on other possible techniques to compare phase diagrams echoes one other reviewer comment, and we indeed expanded the discussion on this point, showing first that other distances do not change much qualitative results (analysis added in supplementary material) and with more literature for possible extensions.
-
-	% Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.
-	\item The functionalities of OpenMOLE we used were detailed in the article; we did not use genetic algorithms but sampling methods and high performance computing environments.
-
-	% In section 3.1, page 8, line 404-405 the authors mention that the distance of the phase diagrams was measured with respect to the reference phase diagram computed on the default initial spatial conditions.
-Actually this came as a surprise to me.
-I had the impression that all phase diagrams would be evaluated against all other diagrams.
-Perhaps this should be explained earlier in the paper.
-	\item Concerning the measure of the distance of all phase diagram to one reference phase diagram, this choice was made as we preferred having a local sensitivity index for which the distribution can be studied, rather than a global completely integrated sensitivity index (similar to Saltelli or Morris indexes for example).
-This was explained better in the paper.
-
-	% Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.
-    \item The double color progression in Fig 4 and 5 is used to highlight the opposition between ratio values above/under 1 but is not meant to introduce any hierarchy in the interpretation.
-We reversed the order of colors according to your remark.
-
-	% I appreciate the fact that the authors in section 4 clearly indicate the limitations of their work (and I share the limitations listed under “comparing phase diagrams”).
-Personally I do not think that releasing the density grids is an important opportunity.
-As mentioned before I think that the core of the work is in the comparison of the phase vectors.
-I am not so sure the second opportunity listed in 4.2 is very valid.
-I do not think that the CEUS journal targets social sciences exactly.
-	\item Thanks for the remarks on discussion, in particular on releasing density grids.
-We have tried to shorten the considerations which you have found not too useful and will let the readers decide if the grids could be of any use to them.
-\end{itemize}
-
-Concerning the comments made by the third referee:
-
-\begin{itemize}
-	% I found the research question posed by the author(s) interesting and I agree with the author(s) that many modelers don’t consider how spatial arrangements impact the outcomes of the model and tend to focus only on the parameters.
-The paper makes good use of the complexity and spatial literature in defining the problem and demonstrates how urban form impacts the results.
-However, while the paper is innovative, I found it really difficult to interpret and follow the results.
-I would suggest to the author to rewrite the results and methodology so that they flow better and also provide more context with respect to how to interpret the results.
-	\item We have tried to make the methodology and result section flow better in the corresponding sections.
-
-	% On a side note, one of your key words you use “validation” but in what sense?
-You don’t really come back to this in the paper, also you write about “the validity of generative models is uncertain until their results are proven robust and representative of ’real-world’ conditions.” In the abstract but this is it.
-What do you consider real world conditions to be?
-You seem to allude that you need to have the spatial generator to generate realistic (styles) city shapes for testing on the Schelling model but I am not convinced.
-After the abstract all notion of validation is missing.
-	\item We have agreed with the reviewer on ``validation'' and removed the concept of validation altogether, replacing it with model evaluation or sensitivity analysis.
-
-	% Geosimulation is introduced but not defined or referenced.  This should be clarified as the term has several meanings etc.
-	\item For the definition of geosimulation, we meant the one coined by (Benenson and Torrens, 2004), this was made explicit and the reference was added.
-
-	% In places the English along with the overall argument could be better articulated.
-For example, on page 1 you write: “We think simulation can become a very good tool to achieve this,”  Could this be rewritten or better argued.
-You are not the only one to believe this but you don’t show this in your paper.
-	\item We have worked towards improving the English along with the overall argument.
-
-	% Page 4: you write “Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time” This makes little sense.
-What it should be is “Generally, sensitivity analyses have focused either on the spatial context or the heterogeneity of space” or something like that.
-Basically you need the “or”.
-	\item Thank you for the suggestion on page 4. We have modified the paragraph accordingly.
-
-	% You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields?
-	\item We do not have knowledge of examples in which the comparison of patterns or fields is called ``phase diagram comparison'', and we neither have knowledge of such work in physics for example, from which the term originated.
-There is naturally all the literature we cited on similar issues in other domains, and the extension of the discussion gives elements of answer to this point.
-
-	% In the Schelling and sugarscape models how are the initial population of agents distributed?
-Is it random?
-I ask as this is unclear in the current version of the paper but it is vital if your purpose is to explore how the initial spatial configuration impacts the results of the model.
-Should the agents not be all located at the same location for each run to rule out the impact of initial placement of agents on the resulting patterns?
-This is something that is not addressed in the paper.
-I know for example, the Schelling model in NetLogo normally places the agents at random locations but in your reimplementation how is this done?
-The same goes for the sugarscape model in NetLogo.
-I believe a clearer description of the model initialization would be good and how its not the initial random placement of agents but the actual environment that impacts the outcomes.
-Maybe this was explained in the paper but I missed it.
-	\item Yes, classically, the initial distribution of agents is random in both models and, most importantly for our point, it follows a uniform distribution of density.
-In our implementation, the initial location of agents is also random in Schelling, but it follows a distribution of density given by the generated grid.
-In Sugarscape, the agents are initialised at random but the resources follow a distribution of density given by the generated grid.
-We have clarified this point in section 2.3
-
-	% In Section 1.2 talk about “(monocentric vs. polycentric for example)” but in the results you don’t discuss the monocentric versions explicitly using these terms.
-	\item We believe that the monocentric versions are explicitly analysed in table 1: the term "monocentric" itself does not appear in the regression results but it is only because it is taken as the reference case under the term "compact' in the regression.
-We have clarified this in section 3.2.
-
-	% Also I appreciate you are trying to show rather complex results but reading the results several times I found the presentation of them (both written and visual) really hard to follow and understand.
-I was wondering if maybe you could find a better way to show / discuss them.
-Maybe a  worked example of how each set of parameters leads to a point in say Figure 4.
-Also it is confusing why the results from both models are not presented using the same styles of figures.
-For example, for the sugar scape model you different styles of figures than for the segregation model.
-	\item Additional figures for the Schelling model presenting the results in a similar way were provided and commented in a way that tries to make it easier to grasp for the reader.
-Unfortunately, the figures showing grid estimates of phase diagrams could not be realized for the Schelling model because of the different sampling of the parameter space used.
-
-\end{itemize}
-
-Concerning the comments made by the fourth referee:
-
-\begin{itemize}
-	% In sensitivity analysis of spatial models, stochastic effects and small parameter changes are normally applied.
-What about initial conditions?
-The authors test the effects of different arrangements (e.g.
-monocentric vs.
-polycentric density patterns) on agent-based Schelling segregation models and Sugarscape's unequal societies model.
-Overall, I find this to be a very good paper.
-Very well written in general.
-Good review of the problem, clearly outlining (with references) the need for focus on initial conditions in geosimulation.
-The chief factors (data uncertainty, nature of spatial units (e.g.
-shape) and spatial heterogeneity) on initial spatial conditions are covered: -how many of these do you test?
-I see heterogeneity addressed, but what about shape and uncertainty?
-You touch upon shape when talking about NetLogo's use of raster as opposed to vector in the Discussion - could you expand on this?
-	\item We thank the reviewer for this good remark.
-We provide references to papers addressing these issues.
-However, we have not addressed the shape of cells and the uncertainty of data in this paper, only the diversity of shapes of urban morphologies (e.g.
-heterogeneity).
-
-	% Rigorous experimental method described.
-In the matching of European metropolitan areas to your generated outputs, I wonder what you are missing by not using cities from other continents.
-i.e.
-by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).
-	\item To the best of our knowledge, there is no study of urban morphologies similar to the one we used but at a global scale (this should be linked to the absence of a raster of population with a resolution fine enough).
-Working on other regional systems of cities such as the US could be possible but remains out of the scope of this paper.
-We added this interesting suggestion to the discussion.
-
-	% Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?
-	\item Concerning the number of repetitions, statistical analysis show that Schelling indicators need more repetitions for an equivalent size of confidence interval on the estimated mean, as the Sharpe ratios are mostly lower than for Sugarscape.
-This remark was added in the paper and the statistical analysis as a supplementary material.
-
-	% You say your models are open source - do you give details of that source?  
-	\item Concerning the source code of the models, an open public repository was put into place and the link provided in the paper (to an anonymized version to ensure blind review).
-Similarly, the large simulation datasets used for results presented in the paper are openly available on a dataverse repository (provisorily anonymized also).
-
-	% Small edits:
-	% X line 68: "...a different system"
-	% figure 1: modify font size so that labels do not go to the edge of their respective objects
-	% TODO Clem
-	% figure 2: can you add the parameters that led to each of the grids
-	% TODO Juste
-	% figure 3: check spelling of "Augburg" (should be Augsburg)
-	% TODO Clem
-	% line 350: "...consists of..."
-	% X line 455: "sa,e" = "same"? OK
-	%figure 4 and figure 5 captions: make it explicit that we're seeing the Sugarscape results here.
-	%line 515: "analyse" = "analysis"
-	%line 593: "at reach" = "within reach"
-	%line 671: would "classic" or "hypothetical" be better words (choice dependent on your meaning) than "mythical", which implies these are models of centuries ago and somehow out of reach..
-	\item Spelling errors and approximations in the vocabulary chosen were corrected.
-\end{itemize}
-
-\justify
-
-\makeletterclosing
+%\makeletterclosing
 

### 2019-06-11 old/Response_to_reviewers_ceus.tex

%% start of file `template.tex'.
%% Copyright 2006-2013 Xavier Danaux (xdanaux@gmail.com).
%
% This work may be distributed and/or modified under the
% conditions of the LaTeX Project Public License version 1.3c,
% available at http://www.latex-project.org/lppl/.

\documentclass[11pt,a4paper,sans]{moderncv}        % possible options include font size ('10pt', '11pt' and '12pt'), paper size ('a4paper', 'letterpaper', 'a5paper', 'legalpaper', 'executivepaper' and 'landscape') and font family ('sans' and 'roman')

\usepackage[document]{ragged2e}
% pour justifier

% moderncv themes
\moderncvstyle{banking}                            % style options are 'casual' (default), 'classic', 'oldstyle' and 'banking'
\moderncvcolor{red}                                % color options 'blue' (default), 'orange', 'green', 'red', 'purple', 'grey' and 'black'
\renewcommand{\familydefault}{\rmdefault} 
%\nopagenumbers{} 
\usepackage[utf8]{inputenc} 
\usepackage[scale=0.9]{geometry}

\def \draft {1}
\usepackage{xparse}
\usepackage{ifthen}
\DeclareDocumentCommand{\comment}{o m o o o o}
{\ifthenelse{\draft=1}{
  \IfValueT{#1}{
      \textcolor{red}{\textbf{C (#1) : }#2}
      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
      \IfValueT{#4}{\textcolor{green}{\textbf{A2 : }#4}}
      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
      \IfValueT{#6}{\textcolor{blue}{\textbf{A4 : }#6}}
    }
    \IfNoValueT{#1}{
      \textcolor{red}{\textbf{C : }#2}
      \IfValueT{#3}{\textcolor{blue}{\textbf{A1 : }#3}}
      \IfValueT{#4}{\textcolor{green}{\textbf{A2 : }#4}}
      \IfValueT{#5}{\textcolor{red!50!blue}{\textbf{A3 : }#5}}
      \IfValueT{#6}{\textcolor{blue}{\textbf{A4 : }#6}}
    }
 }{}
}

\firstname{}
\lastname{}
\begin{document}

% recipient data
\recipient{Editor CEUS}{}
\date{\today}
\opening{Dear Editor,}
\closing{Yours faithfully,\\
Juste Raimbault and co-authors%\\
%Université Paris 7 - UMR CNRS 8504 Géographie-cités
}
         % use an optional argument to use a string other than "Enclosure", or redefine \enclname

%\makelettertitle

%Dear Editor,

\justify

%Thank you for the feedback on the draft paper ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models''. The suggestions and comments will undoubtedly be of great value to the paper. It was updated accordingly.

\medskip

% Thank you for submitting your manuscript to Computers, Environment and Urban Systems. I have completed the review of your manuscript and a summary is appended below. The reviewers recommend reconsideration of your paper following major revision. I invite you to resubmit your manuscript after addressing all reviewer comments.

% When revising your paper, pay special attention to the explicit presentation of every model that is investigated in the paper. Present each model as an algorithm/unambiguous pseudocode and supply the Netlogo code for downloading (all this can be in appendix). The Schelling model, for example, has several Netlogo implementation, and it is critically important to know which of them is investigated. The same is true for the Sugarscape model. See comments of the reviewer #1 in this respect.

% When resubmitting your manuscript, please carefully consider all issues mentioned in the reviewers' comments, outline every change made point by point, and provide suitable rebuttals for any comments not addressed.

%Comments from the editors and reviewers:
%Please reformat your reply to the reviewers according to the journal demands.
%1. Your reply should consist of the comment-reply pairs.
%2. For each review, copy every comment, including positive ones, and reply to each directly and in full. If necessary, you may cite  in your reply textual changes that are made in the manuscript. 
%3. If you disagree with the reviewer's comment, state this clearly in your reply and explain your rebuttal.
%4. Separate between your replies to the comments of the editor (if any) and each of the reviewers. 
%5. Number comments of each reviewer and reply to them in exactly the same order as they are listed in the review.
%6. If different reviewers point to the same flaws, repeat your replies, while mentioning that this is the same comment as already replied. 
%7. Use italics when you cite reviewer's comment and normal font for your reply. 
%You may preserve your general reply to the reviewers' comments, as it is now, while it should be located after the above list of the direct replies in your reply letter.

\textbf{Response to reviewers}

\bigskip

%We now turn to point-by-point responses to referees comments.

\textbf{Point-by-point response to referees comments:}

\medskip

%Concerning specifically the comments made by the first referee:
\textbf{First referee :}

\medskip

\begin{enumerate}
	\item (a) \textit{I think it would be better if the author would focus on a single model. This way they could describe the model and the analyses in more detail. As it stands, the models are treated as black boxes.}

	\medskip

	$\rightarrow$ The suggestion to focus on a single model is a qualified comment. However, after presenting our research to several audiences, we came to the conclusion that the genericity of our approach would be better served by two case studies rather than with a single model. In that perspective, the ``black-box'' is seen as an advantage as it reinforces genericity. We acknowledge the fact that this choice led us to shorten the model presentations. Nevertheless, we have taken notice of your remark by improving the description of the models and their results through the addition of pseudo-codes in Appendix, result descriptions in the text and by recalling more clearly the link to the model source we used. Furthermore, we have focused on some of the most classical models of the ABM literature so as to lower the reader's cost of entry into the methodology.
	\bigskip

    \setcounter{enumi}{0}
   \item (b) \textit{I believe it would be better to begin with an illustrative example on how two or three initial patterns influence the model results and then use the suggested method for a more extensive analysis.}

   \medskip

   $\rightarrow$ We agree that illustrative examples would provide a relevant starting point to  the analysis but we were afraid that their addition to the paper would make the introduction less straightforward and have the opposite effect to the one we were seeking. As we were nevertheless very interested by your remark, we provide a short answer to your proposition. In order to get a first idea of the potential impact of spatial initial conditions on to the results of the simulation, we implemented on the Schelling Segregation model a geographical constraint: some cells cannot be inhabited (yellow cells in Figure 1). The way those “forbidden” cells are distributed throughout space will indeed influence the steps of the model. We introduced various spatial initial conditions in the model that we controlled by one clustering parameter on the forbidden cells. A value of 0 would indicate a spatial distribution of yellow cells so as to minimize spatial autocorrelation. A value of 1 indicates on the contrary a concentration process that tends towards a high value of autocorrelation (see below Figure 1). We then compute the number of communities arising when the model converge, hence the number of different related components of similar neighbors. Using 270 replications of 11 values of the clustering parameter (from 0 to 1 by step 0.1), we obtained a significant growth in terms of number of communities when the constraints on residential locations are concentrated in space, which might be interpreted in such example as a bigger segregation when density of housing is heterogeneously distributed through space.

%%%%%%%%%%%%%%
\begin{minipage}{\textwidth}
\bigskip\bigskip
\begin{center} 
	\includegraphics[width=\textwidth]{figures/SpaceMatters2.png}

Fig. 1: \textbf{Effects of clustering of forbidden cells on the number of communities emerging from an adapted Schelling model (Netlogo implementation).}
\end{center}
\bigskip\bigskip
\end{minipage}
%%%%%%%%%%%%%%

	\item \textit{It is very important to describe the models. This is especially relevant for the Schelling model, which has many variants; each has different dependencies on the initial pattern. In my opinion, the NetLogo implementation is not the best choice because the model does not reach a state of segregation when agents are highly intolerant.}

	\medskip

	$\rightarrow$ There was a misunderstanding for the Schelling implementation actually used, as we used our own implementation which pseudo-code has been added in Appendix of the paper and source code is openly available on the git repository of the project. This version implements the one used by (Gauvin et al., 2009), which also influenced us for the use of the term ``phase diagram''.

	\bigskip

	\item \textit{I am not sure why the Sugarscape model has been chosen given that the paper seems to be focused on cities.}

	\medskip

	$\rightarrow$ Furthermore concerning the choice of the Sugarscape model, we think it is a good example of a stylized model of a spatialized socio-economic system, and according to (Batty, 2005) a good metaphor of an urban system. Our choice was accordingly clarified in the paper. More generally concerning the choice of these two models, we aim at helping the readers to distance themselves from the case studies and focus on the genericity of the approach. This also explains why we have chosen Sugarscape as a second example: these two models are often used as toy models due to the simplicity of their mechanisms and therefore do not require a high level of expertise nor technicality to apprehend the impacts of the initial spatial conditions on the model outputs. This is again an argument in favor of our choice: space-dependent variations are to be expected even with the most simple sociospatial models.
	\bigskip

	\item \textit{Consider replacing “meta parameters” with something like “generator parameters” or “initial pattern parameters.”}

	\medskip

	$\rightarrow$ We agree that the term of meta-parameters may have been confusing, and following the suggestion we replaced it with ``generator parameters''.
	\bigskip

	\item \textit{Line 228: Please explain why the median was chosen. Especially given that you use an L2 norm and variances in the index}

    \medskip

    $\rightarrow$ There was a mistake in the writing, as we never use the median but always the mean, which as you recall is consistent with the choice of distance we made. This was corrected in the paper.

	\bigskip

	\item \textit{Line 232: comparing phase diagrams is related to the topic of comparing patterns. It would make sense to mention the methods that are usually used to compare patterns.}

	\medskip

	$\rightarrow$ Some general references on comparing patterns were added (section 4.1).
	\bigskip

	\item \textit{Line 238: Please explain why this particular index was chosen and discuss its properties. Connect the discussion to other indices like L1, and Linf norms, etc. Would we get qualitatively different results from different indices? }
	\medskip

	$\rightarrow$ The approach proposed in the paper is indeed rather exploratory. This explains why we have chosen to focus on a single index (L2) as this index is relatively well-known in the community. Previous exploratory analyses have shown that the results were rather stable when using the Monge method. Adapting the mathematical aggregator to the different indices would constitute an interesting perspective, but is out of the scope of the paper which already focuses on two models and proposes a new framework for analysing the impact of initial pattern parameters. We however added a supplementary material showing the distances obtained with other distances, which are similar to the results in main text.
	\bigskip

	\item \textit{Regarding the spatial generator: the authors should explain why this generator is relevant to these two models?}

	\medskip

	$\rightarrow$ This spatial generator is relevant to both model as it produces density grids at a ``metropolitan scale'' (see detailed discussion added in Appendix and the original paper from which we used the generator), which is the scale at which both model were initially intended to be. We made this aspect more explicit. More particularly concerning the choice of this generator in the case of Schelling, this is the scale at which most empirical segregation indices are computed. In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty, 2005), or to a generic landscape where a resource is grown otherwise. In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density, and that not acknowledging this diversity leads to potentially false results. Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most common simplification of the heterogeneity of space, with consequences on the model outputs often not evaluated.
	\bigskip

	\item \textit{The authors should illustrate the contribution of each parameter of the generator to the generated initial pattern using a figure. Details of the generators should be included in an appendix.}

	\medskip

	$\rightarrow$ The behavior of the spatial generator was described with more details in Appendix.

	\bigskip

	\item \textit{Explain how the density grids are used in the Schelling model, do you allow more than one agent to be in the same cell? }

	\medskip

	$\rightarrow$ Regarding the use of the density grid in the Schelling model, yes we allow more agents to be in the same cell. The potential density is defined by the density grid generated, and its population with the same number of agents in all grids. This distribution is the one at initialisation. If the potential density is not reached, more agents can move in the cell, otherwise it is deemed full and unavailable for further moves. The satisfaction and segregation indices are computed with regard to the people in the cell and the people present in neighbouring cells.

	\bigskip

	\item \textit{Line 297: I am not sure why empirical density destitutions of cities are important in the context of this study.}

	\medskip

	$\rightarrow$ Density distributions of empirical cities are important because we want to test models with realistic ranges of initial patterns of density distribution. Therefore we cannot limit ourselves to a single 2-peak mountain in the classical Sugarscape model and an isotropic square city in Schelling. We chose to use actual distribution of European cities to constraint our density generation.
	\bigskip

	\item \textit{Line 641: NetLogo supports grids, continuous space, and GIS elements. It does not restrict the user in any way.}

	\medskip

	$\rightarrow$ We agree that NetLogo does not restrict the use on what implementation to use. However, some features are much more easier as raster or the contrary, which could influence the choice of beginners. Furthermore, models in the built-in NetLogo model library mostly rely on rasters. We clarified that statement in the paper.

	\bigskip

	\item \textit{Line 301: I am not sure what the authors mean by “interesting grids”}

	\medskip

	$\rightarrow$ The term ``interesting grids'' was a typo and corrected, ``grids'' only was meant.
	\bigskip

	\item \textit{Line 671: The same comment regarding “mythical ABMs”}

	\medskip

	$\rightarrow$ The term ``mythical'' line 671 was a language mistake and corrected, ``textbook'' ABM was meant.
\end{enumerate}

\bigskip
\bigskip

%Concerning the comments made by the second referee:

\textbf{Second referee: }

\bigskip
\bigskip

%This is an interesting article that I read with true interest. The paper fits the profile of this journal and can reach a wider audience.

%I also have some concerns:

\begin{enumerate}
	\item \textit{In the abstract the authors claim that initial spatial conditions are often taken for granted. I do not agree to this statement. Initial spatial condition can have a large impact on the results of a model, however I do not think that the initial spatial conditions are often taken for granted.}

	\medskip

	$\rightarrow$ We forcefully agree with the referee in that initial spatial conditions can have a large impact on the results of the model. Maybe not all papers take them for granted, but what we meant is that compared to other parameters, initial spatial conditions are much less often explored using sensitivity analysis. Actually, we know only a few papers which have explored the impact of initial spatial conditions as a systematic criterion in a sensitivity analysis of model results. We clarified this point in the introduction.
	\bigskip

	\item \textit{Page 2 line 88: the authors state that “Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density”, I do not think this is true. Many models generate a heterogeneous synthetic population that is based on local data (replicating the actual population as closely as possible) and in far more detail then what is done in this paper. I am thinking of work in the line of disease diffusion, crime etc.}

	\medskip

	$\rightarrow$ We acknowledge the fact that many ABM applications use "real" data and spatial features for their simulations. However, as explained by (Thomas et al. 2017) in the case of LUTI models for instance, most simulations do not evaluate the impact of city delineations nor MAUP on the output of the models, which constitutes a potential limit in the interpretation of results.
	\bigskip

	\item \textit{Page 2 line 98: “However, this step is rarely explicit and authors tend to generalize the conclusion of the model from a single, spatial case study”. I do not understand the first part of the this statement. The steps taken are often very explicit. Due to the amount of time it takes to correctly represent all spatial data, including population densities and the variation of other population variables over space, many authors indeed restrict themselves to a single case study due to the time intensity of this task. Many agent-based models are also not valid for other areas.}

	\medskip

	$\rightarrow$ What we meant was actually that this step is rarely explicit and authors are therefore not able to generalize the conclusion of the simulations to other spatial context. We rephrased the paragraph accordingly.
	\bigskip

	\item \textit{The authors do not differentiate between data driven empirical models versus theory-driven models. I think a clear differentiation between the two groups and the positioning of the two case studies in the second group would make the paper more balanced.}

	\medskip

	$\rightarrow$ Concerning the remarks on synthetic population generation and the different types of models, we clearly stated our positioning regarding models built from theory, for which the proposed sensitivity analysis workflow is more suited and that our two case studies illustrate.
	\bigskip

	\item \textit{In section 1.3 (Previous considerations on the effects of the spatial configuration in simulation models) authors provide a lot of evidence on previous work on the impact of initial spatial conditions on the output of models. This is contrary to the statement in the abstract that initial conditions are often taken for granted.}

	\medskip

	$\rightarrow$ Indeed, we have tried to review articles where some kind of sensitivity to initial spatial conditions are tackled. However we would like to emphasize that the references cited in this section are few compared to the mass of geosimulation articles published with hardly any mention of dependence of output on initial spatial condition. Moreover, even if such sensitivity is analysed, it mostly provide some hints of impact of initial spatial conditions on the output of models (due to the accuracy of data, shape of boundaries and spatial heterogeneity) but to the best of our knowledge no article provide a systematic inquiry into the impact of systematically defined spatial input on the model output in the way we try to provide later on in the paper. We have tried to argue this point at the end of the first section.
	\bigskip

	\item \textit{The most interesting part of the research to me, is not the way the density layers are generated but the use of the phase diagrams, defined as a vector of final aggregated model outputs. As these are vectors, they are not so different from time series and many techniques exist to compare vectors in this domain. I wonder if the Euclidean distance is the best measure to determine the difference between the phase vectors of different spatial inputs.}

	\medskip

	$\rightarrow$ The comment on other possible techniques to compare phase diagrams echoes one other reviewer comment, and we indeed expanded the discussion on this point, showing first that other distances do not change much qualitative results (analysis added in supplementary material) and with more literature for possible extensions.
	\bigskip

	\item \textit{Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.}

	\medskip

	$\rightarrow$ The functionalities of OpenMOLE we used were detailed in the article; we did not use genetic algorithms but sampling methods and high performance computing environments.
	\bigskip

	\item \textit{In section 3.1, page 8, line 404-405 the authors mention that the distance of the phase diagrams was measured with respect to the reference phase diagram computed on the default initial spatial conditions. Actually this came as a surprise to me. I had the impression that all phase diagrams would be evaluated against all other diagrams. Perhaps this should be explained earlier in the paper.}

	\medskip

	$\rightarrow$ Concerning the measure of the distance of all phase diagram to one reference phase diagram, this choice was made as we preferred having a local sensitivity index for which the distribution can be studied, rather than a global completely integrated sensitivity index (similar to Saltelli or Morris indexes for example). This was better explained in the paper.
	\bigskip

	\item \textit{Figure 4 and 5: I find it confusing that the green color indicates higher distance values and the red color indicates lower values. I think this should be reversed.}

	\medskip

    $\rightarrow$ The double color progression in Fig 4 and 5 is used to highlight the opposition between ratio values above/under 1 but is not meant to introduce any hierarchy in the interpretation. We reversed the order of colors according to your remark.
   	\bigskip

	\item \textit{I appreciate the fact that the authors in section 4 clearly indicate the limitations of their work (and I share the limitations listed under “comparing phase diagrams”). Personally I do not think that releasing the density grids is an important opportunity. As mentioned before I think that the core of the work is in the comparison of the phase vectors. I am not so sure the second opportunity listed in 4.2 is very valid. I do not think that the CEUS journal targets social sciences exactly.}

	\medskip

	$\rightarrow$ Thanks for the remarks on discussion, in particular on releasing density grids. We have tried to shorten the considerations which you have found not too useful and will let the readers decide if the grids could be of any use to them.
\end{enumerate}

%Concerning the comments made by the third referee:

\bigskip
\bigskip

\textbf{Third referee :}

\bigskip

\begin{enumerate}
	\item \textit{I found the research question posed by the author(s) interesting and I agree with the author(s) that many modelers don’t consider how spatial arrangements impact the outcomes of the model and tend to focus only on the parameters. The paper makes good use of the complexity and spatial literature in defining the problem and demonstrates how urban form impacts the results. However, while the paper is innovative, I found it really difficult to interpret and follow the results. I would suggest to the author to rewrite the results and methodology so that they flow better and also provide more context with respect to how to interpret the results.}

	\medskip

	$\rightarrow$ We have tried to make the methodology and result section flow better in the corresponding sections.
	\bigskip

	\item \textit{On a side note, one of your key words you use “validation” but in what sense?  You don’t really come back to this in the paper, also you write about “the validity of generative models is uncertain until their results are proven robust and representative of ’real-world’ conditions.” In the abstract but this is it.  What do you consider real world conditions to be? You seem to allude that you need to have the spatial generator to generate realistic (styles) city shapes for testing on the Schelling model but I am not convinced.  After the abstract all notion of validation is missing.}

	\medskip

	$\rightarrow$ We have agreed with the reviewer on ``validation'' and removed the concept of validation altogether, replacing it with model evaluation or sensitivity analysis.
	\bigskip

	\item \textit{Geosimulation is introduced but not defined or referenced.  This should be clarified as the term has several meanings etc.}

	\medskip

	$\rightarrow$ For the definition of geosimulation, we meant the one coined by (Benenson and Torrens, 2004), this was made explicit and the reference was added. 
    \bigskip

	\item \textit{In places the English along with the overall argument could be better articulated. For example, on page 1 you write: “We think simulation can become a very good tool to achieve this,”  Could this be rewritten or better argued. You are not the only one to believe this but you don’t show this in your paper.}

	\medskip

	$\rightarrow$ We have worked towards improving the English along with the overall argument. 
	\bigskip

	\item \textit{Page 4: you write “Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time” This makes little sense. What it should be is “Generally, sensitivity analyses have focused either on the spatial context or the heterogeneity of space” or something like that. Basically you need the “or”.}

	\medskip

	$\rightarrow$ Thank you for the suggestion on page 4. We have modified the paragraph accordingly.
	\bigskip

	\item \textit{You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields? }

	\medskip

	$\rightarrow$ We do not have knowledge of examples in which the comparison of patterns or fields is called ``phase diagram comparison'', and we neither have knowledge of such work in physics for example, from which the term originated. There is naturally all the literature we cited on similar issues in other domains, and the extension of the discussion gives elements of answer to this point.
	\bigskip

	\item \textit{In the Schelling and sugarscape models how are the initial population of agents distributed? Is it random? I ask as this is unclear in the current version of the paper but it is vital if your purpose is to explore how the initial spatial configuration impacts the results of the model. Should the agents not be all located at the same location for each run to rule out the impact of initial placement of agents on the resulting patterns? This is something that is not addressed in the paper. I know for example, the Schelling model in NetLogo normally places the agents at random locations but in your reimplementation how is this done? The same goes for the sugarscape model in NetLogo. I believe a clearer description of the model initialization would be good and how its not the initial random placement of agents but the actual environment that impacts the outcomes. Maybe this was explained in the paper but I missed it.}

	\medskip

	$\rightarrow$ Regarding the way these models are used in the literature, yes, the initial distribution of agents is classically random in both models and, most importantly for our point, it follows a uniform distribution of density. In our implementation, the initial location of agents is also random in Schelling, but it follows a distribution of density given by the generated grid. In Sugarscape, the agents are initialised at random but the resources follow a distribution of density given by the generated grid. We have clarified this point in section 2.3.
         \bigskip

	\item  \textit{In Section 1.2 talk about “(monocentric vs. polycentric for example)” but in the results you don’t discuss the monocentric versions explicitly using these terms.}

	\medskip

	$\rightarrow$ We believe that the monocentric versions are explicitly analysed in table 1: the term "monocentric" itself does not appear in the regression results but it is only because it is taken as the reference case under the term "compact' in the regression. We have clarified this in section 3.2.
	\bigskip

	\item \textit{Also I appreciate you are trying to show rather complex results but reading the results several times I found the presentation of them (both written and visual) really hard to follow and understand. I was wondering if maybe you could find a better way to show / discuss them. Maybe a  worked example of how each set of parameters leads to a point in say Figure 4. Also it is confusing why the results from both models are not presented using the same styles of figures. For example, for the sugar scape model you different styles of figures than for the segregation model.}

	\medskip

	$\rightarrow$ Additional figures for the Schelling model presenting the results in a similar way were provided and commented in a way that tries to make it easier to grasp for the reader. Unfortunately, the figures showing grid estimates of phase diagrams could not be realized for the Schelling model because of the different sampling of the parameter space used.

\end{enumerate}

\bigskip
\bigskip

\textbf{Fourth referee :}

%Concerning the comments made by the fourth referee:

\bigskip

\begin{enumerate}
	\item \textit{In sensitivity analysis of spatial models, stochastic effects and small parameter changes are normally applied. What about initial conditions? The authors test the effects of different arrangements (e.g. monocentric vs. polycentric density patterns) on agent-based Schelling segregation models and Sugarscape's unequal societies model. Overall, I find this to be a very good paper. Very well written in general. Good review of the problem, clearly outlining (with references) the need for focus on initial conditions in geosimulation. The chief factors (data uncertainty, nature of spatial units (e.g. shape) and spatial heterogeneity) on initial spatial conditions are covered: -how many of these do you test? I see heterogeneity addressed, but what about shape and uncertainty? You touch upon shape when talking about NetLogo's use of raster as opposed to vector in the Discussion - could you expand on this?}

	\medskip

	$\rightarrow$ We thank the reviewer for this good remark regarding connected issues such as heterogeneity or the choice raster/vector. We provide references to papers addressing these issues. However, we have not addressed the shape of cells and the uncertainty of data in this paper, only the diversity of shapes of urban morphologies (e.g. heterogeneity). 
	\bigskip

	\item \textit{Rigorous experimental method described. In the matching of European metropolitan areas to your generated outputs, I wonder what you are missing by not using cities from other continents. i.e. by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).}

	\medskip

	$\rightarrow$ To the best of our knowledge, there is no study of urban morphologies similar to the one we used but at a global scale (this should be linked to the absence of a raster of population with a resolution fine enough). Working on other regional systems of cities such as the US could be possible but remains out of the scope of this paper. We added this interesting suggestion to the discussion.
	\bigskip

	\item \textit{Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?}

	\medskip

	$\rightarrow$ Concerning the number of repetitions, statistical analysis show that Schelling indicators need more repetitions for an equivalent size of confidence interval on the estimated mean, as the Sharpe ratios are mostly lower than for Sugarscape. This remark was added in the paper and the statistical analysis as a supplementary material.
	\bigskip

	\item \textit{You say your models are open source - do you give details of that source?}

	\medskip

	$\rightarrow$ Concerning the source code of the models, an open public repository was put into place and the link provided in the paper (to an anonymized version to ensure blind review). Similarly, the large simulation datasets used for results presented in the paper are openly available on a dataverse repository (provisorily anonymized also).
	\bigskip

	\item \textit{Small edits: line 68: "...a different system"; figure 1: modify font size so that labels do not go to the edge of their respective objects; figure 2: can you add the parameters that led to each of the grids; figure 3: check spelling of "Augburg" (should be Augsburg); line 350: "...consists of..."; line 455: "sa,e" = "same"?; figure 4 and figure 5 captions: make it explicit that we're seeing the Sugarscape results here; line 515: "analyse" = "analysis"; line 593: "at reach" = "within reach"; line 671: would "classic" or "hypothetical" be better words (choice dependent on your meaning) than "mythical", which implies these are models of centuries ago and somehow out of reach..}

	\medskip

	$\rightarrow$ Spelling errors and approximations in the vocabulary chosen were corrected.
\end{enumerate}

\bigskip
\bigskip

%As a response to several reviewers, we made the following general changes:

\textbf{General comments to all reviewers and the editor: }

\bigskip

\begin{enumerate}
	\item Several comments, from the first and third referees especially, suggested that we clarify and explicit the description of our two models: we clarified the descriptions and justified our choices more thoroughly, and also provided pseudo-codes for each model in the Appendix so as to help the reader navigate the implementation chosen compared to other articles. Finally, we mentioned more clearly in the text the fact that source codes are openly available in a github repository.
    \item In order to conciliate as much as possible the different points of view about the comparison of phase diagram, we added some developments and comments about how we defined, built, justified and compared phase diagrams in the paper.
    \end{enumerate}

\justify

%\makeletterclosing

\end{document}

%% end of file `template.tex'.

### 2019-07-25 Response_to_Reviewers.tex
Update on Overleaf.

@@
 
+Resubmit 15 Aug. 2019\\
+
+We want to thank the two reviewers for their thorough evaluation and discussion of our paper, as well as the challenging questions they asked.
+We have revised the manuscript in a way that addresses most of the comments.
+In this response, we first want to address the two main questions that arose for both reviews.
+The first one (A) relates to the original mis-specification/ambiguities between the social and spatial components of the modelling process.
+The second one (B) relates to the contribution of the approach in terms of potential use for model makers, for domain experts and in operational contexts.
+
+ \medskip
+A. In terms of the ambiguity between the social and spatial component 
+\begin{itemize}
+\item A.1.
+We have precised and clarified the argument of the paper (mainly in introduction) by qualifying our contribution, which is to provide an operational framework for studying the influence of geometric structures of space on the results of a geosimulation model.
+We have thus distinguished two effects of geographical space in simulation modelling: the ones that are part of the model, reflecting for example the role played by distance on decreasing agents' perception or potential social interactions; and the ones which act as geometric structures "on which" the social elements simulated interact, i.e.
+geometric constraints of the simulation.
+The first type of effects is addressed by traditional modelling.
+Our original contribution is to tackle the second type.
+\comment[CC]{done in 1.5 and 1.8}
+
+\item A.2.
+We have clarified the terms used in our paper introduction to distinguish between: {\bf sociospatial systems} (i.e.
+groups of social agents whose behaviour is constrained by their position in geographical space.
+This term refers to empirical systems, which are the targets of the modelling process), {\bf interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and {\bf geometric structure}, a term we reserve for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model.
+When we talk about the generator we actually used in the paper, we call it a "density grid generator" because the is the only geometric structure it generates, whereas a "spatial generator" could potentially generate graph or point structures.
+\comment[CC]{done in 1.5}
+
+ \item A.3.
+We have modified figure 1 to express the clarified approach visually.
+Naming the generator parameter "gamma" has helped reduced the confusion that the term 'meta-parameter' was inducing.
+We have also replaced the expressions "quantitative results" by "sensitivity analysis" and "qualitative results" by "variations by types of geometric structure", in the figure as in the subsections of the paper.
+\comment[CC]{done}
+\end{itemize}
+
+ \medskip
+B. To highlight limitations as well as promising perspectives of improvement, answer the concern about the genericity of the method and expose its links with the conceptual and empirical domains:
+
+ \begin{itemize}
+
+  \item B.1.
+We have added a discussion of how segregation indicators can themselves be influenced by geometric structures (because they give the same weight to cells of different density).
+This is a limitation of the method as it is but also an avenue for promoting stronger indicators in empirical research.
+\comment{il faut reprendre ces elements dans l'article}
+
+  \item B.2.
+To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, We added a paragraph in conclusion, discussing how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences).
+\comment{il faut reprendre ces elements dans l'article}
+
+  \item B.3.
+As a perspective to reduce this bias, we could compute ego-neighbourhoods rather than a regular grid (another paper).
+We could also in another paper vary the size of grid and its cell shape (cf.
+MAUP), but also generate other types of geometric structures to feed into geosimulation models (for example networks).
+\comment{il faut reprendre ces elements dans l'article}
+\end{itemize}
+
 \bigskip
 
+\textbf{to do list:}
+
+ {\bf Sociospatial system} pour parler des systèmes empiriques. \comment[MLT]{Done}\\
+ {\bf interactions} pour parler de ce qui se passe dans le modele.\\
+  {\bf geometric structures} pour parler des outputs issus du générateur.\\
+  + parler de "density grid generator" quand on développe le nôtre. \\
+  caser" Jessop (2008) points in particular to the interdependence between social phenomena and the spatial environment where they take place.
+We model the social system as the interaction dynamics of agents and spatial units.
+The spatial generator produces the geometric support for these interactions." in intro 1.4 \\
+
 %We now turn to point-by-point responses to referees comments.
-
+\medskip
 \textbf{Point-by-point response to referees comments:}
@@
    \item \comment[JR]{will add some description on urban form indicators etc.}
+   \item To describe more the spatial generator we used in a few sentences (JR) + situate this one among the many possible ways of making a spatial generator as Young2002 (FLN/MLT).
+   %This also applies to the three segregation indexes: what exactly do they measure?
+  \item Describe segregation index in a few sentences \comment[CC]{done in section "case study models"}
 
-  %This also applies to the three segregation indexes: what exactly do they measure?
-  \item
-
-  % The description of the spatial generator’s procedure, outlined in paragraph 2.7, would also benefit if more information is added, for example, to delineate the terms “strength of attrac!
-tion” (attraction in which respect?) and “strength of the diffusion process” (is strength equal to intensity?
+  % The description of the spatial generator’s procedure, outlined in paragraph 2.7, would also benefit if more information is added, for example, to delineate the terms “strength of attraction” (attraction in which respect?) and “strength of the diffusion process” (is strength equal to intensity?
 If so, is it then in space and/or time?).
@@
 Even claimed by the author(s), the approach has not much to do with the empirical growth or development of cities, just because the approach reduces cities’ growth to the growth itself, omitting issues such as historical genesis, cultural differentiation, economic background political and legal circumstances, etc.
-  \item \comment[JR]{most of this empirical literature was added because of one CEUS reviewer, no ?}
+  \item \comment[JR]{most of this empirical literature was added because of one CEUS reviewer, no ?}\comment[MLT]{If we want to keep this part to please everyone (like our former reviewer), we could explicitly mention that we model the growth of a city (in terms of population density) which is the results of a bunch of dynamics ...
+and not the dynamics themselves?}\comment[CC] We have clarified the semantic terms used in sections 1.5 to 1.8 and in the figure 1.
+\comment[FLN]{ qui montre que la démarche méta est ici appliquée à un domaine empirique qu'il aide à explorer mais d'autres domaines peuvent s'emparer de la démarche méta.}
+\item show disagreement with omitting empirical/geographical but demonstrate helps understand geographical question but the is generic and valuable for it.
 
@@
 ling fuzzily the social with the geographical, as happened with this definition.
-  \item \comment[JR]{by spatial we indeed mean some kind of spatial container for agents right ? which is far from geographical in the sense he puts it indeed}
+  \item \comment[JR]{by spatial we indeed mean some kind of spatial container for agents right ?
+which is far from geographical in the sense he puts it indeed}\comment[MLT]{use sociospatial system instead?
+The term is broadly used in the literature to describe the interdependence between social phenomena and their spatial environment.
+We could also quote \hyperlink{https://journals.sagepub.com/doi/abs/10.1068/d9107}{Jessop et al.
+2008}}
+\item cf. comment A
+\item reprendre tout le papier pour éliminer les termes indéfinis "space", "pattern" etc.
++ en conclu mentionner que ça s'applique aux automates cellulaires (sans agents) et aux modèles écologiques (agents-arbres).
+  \item in conclusion, we need to clarify the answer to the question "do geometric structure influence sociospatial systems dynamics" by saying it depends on the model.
+≠ black\&white and we can demonstrate it.
+The nature of social systems is a given (in the model mu, which is not the contribution of this paper) and we explore the consequence of geometric structures on the spatial expression of social systems.
 
@@
 In this respect, the paper offers some relevant clues such as the MAUP problem or the “uncertain geographic context problem”.
-  \item
+  \item see comment A. cite Banos, 2012
 
@@
   % -> yes indeed
-  \item
+  \item see comment A.
 
@@
 Obviously not, but the paper seems to argue this way.
-  \item
+  \item see comment A.
 
   % Due to this, attempts to explain the growth of existing cities should be avoided.
-The approach is reductionist and has its strength in reflecting growth at an abstract level nevertheless.
+  \item we agree.
+it's not the point of the paper, only the metaphore of the spatial generator.
++ supprimer la ref à "urban growth" dans le texte et dire d'inspiration urban science mais dont seules les grilles finales nous intéressent.
+On aurait pu utiliser des generator issus de graphes, de percolation etc.
+  The core of the paper is not this, it's the approach of including gamma and applying to mu to understand systems better, whatever gamma and whatever mu in whatever knowledge domain.
+
+ \item in discussion, mention that it's one isolated example of geometry but other cases could be thought of as well, using networks and graphs, but also empirical shape files.
+
+  %The approach is reductionist and has its strength in reflecting growth at an abstract level nevertheless.
 The concluding remark in paragraph 5.1 “With the Schelling experiment, we found … that polycentric and discontinuous cities appear systematically more segregated than compact cities …” is elusive.
@@
 This geographical reasoning, however, is outside of the scope of this paper.
-  \item \comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate}
+  \item \comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate} \comment[MLT]{I kind of disagree with the idea that it is straightforward that centralised cities produce more segregation as it is not a direct parameter of the model but an output.
+Plus: what argument would he or she have to state that a big center would be automatically less segregated that several smaller ones?
+And I believe that explaining why polycentric cities are more segregated than others is out of the scope of the paper.
+But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?}
+
+  \item in discussion, say that indeed the indicators can themselves be influenced by geometric structures (cf.
+comment B) but on the processes, there is nothing obvious about the influence of structures on behavioural spatial patterns (if you imagine a model with no interactions between agents for example) + it's a difference between the model and its potential addition to the real systems.
+cf.
+B
+
+  \item rewrite 1.11 \& 1.12 which is misleading
 
    % Formal issues that have to be solved: the numbering of paragraphs is not coherent. The reference of Wheeler (2006) is missing the source of the article.
-   \item
+    \item check jasss style for paragraph numbering (Tous)
 
@@
 How can be extrapolate and isolate spatial effects when models have different logics and theoretical purposes?
-  \item
+  \item see above comment A, discussion has been amended and details added.
 
@@
 While limitations are well-discussed, readers who are not specialist of spatial models could benefit from a more comprehensive discussion on potential applications and extensions.
-For instance, it would be useful to have suggestions on how to use this methodology to run empirical models.
+  \item see comment B
+
+  %For instance, it would be useful to have suggestions on how to use this methodology to run empirical models.
 Suppose to have a nice dataset and a case of a urban context to be modelled.
 Could the exploration of different spatial configurations even be used to run 'counter-factual' or 'what if' tests on actual spatial configurations?
-  \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}
+  \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model} cf.
+comment B.3
 
-  % Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), whcih could be interesting to see implemented in ABM.
-Spatial econometrics based on empirical dataset is developing fast.
-It would be useful to understand how these models can enrich ABM and vice-versa.
-  \item
+  \item it as not explicit enough and we revised the presentation of our approach (which is generic and not empirical) in intro (especially fig.
+1).
+We thank reviewer 2 for highlighting this limitation.
+
+  % Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), which could be interesting to see implemented in ABM.
+  \item cf. discussion of egohoods in conclusion (comment B)
+
+  %Spatial econometrics based on empirical dataset is developing fast. It would be useful to understand how these models can enrich ABM and vice-versa. 
+   \item this is very interesting to inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010.
+We need to add a paragraph about this discussion in conclusion and how our approach allows to go back to questioning empirical cases from the model and questioning our concepts ('neighbourhood').
+cf.
+comment B
 
@@
 Here, it is a bit misleading to call "generator parameters" ONLY the spatial ones!
-  \item
-
-  % In this respect, I would encourage authors to further develop their closing section, which now does not stimulate readers to imagine implications of authors' findings and message to their own research.
+ \item we have clarified this issue in intro but referring to different terms and figure 1 which differentiates between the mu model and gama parameters. (comment A)
+  %In this respect, I would encourage authors to further develop their closing section, which now does not stimulate readers to imagine implications of authors' findings and message to their own research.
 I would expand this section to provide a more detailed discussion on implications for model makers, experts on social geography and urban scholars who work with ABM, non-specialist in geography who would like to base their models in more proper geographical constraints.
 And perhaps, authors could suggest future line of research that could show-case their argument.
-  \item
+  \item cf. new discussion/conclusion and comment B
 
 %I appreciate that models are fully available to readers. 
+\item We do too!
 

### 2019-07-25 paper.tex
Update on Overleaf.

@@
 \begin{abstract}
-Although simulation models of geographical systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions.
+Although simulation models of socio-spatial systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions.
 Sensitivity analysis usually includes the analysis of the effect of stochasticity on the variability of results, as well as the effects of small parameter changes.
-However, initial spatial conditions are usually not modified systematically in geographical models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment.
-In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic spatial configuration generator in order to create density grids with which spatial simulation models are initialised.
+However, initial spatial conditions are usually not modified systematically in socio-spatial models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment.
+In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic spatial configuration generator in order to create density grids with which socio-spatial simulation models are initialised.
 We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models.
@@
 It has been identified as an emerging field and coined under the term geosimulation by \cite{benenson2004geosimulation}.
-Simulation also appears as a way to overcome the difficult analytic resolution of many spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
+Simulation also appears as a way to overcome the difficult analytic resolution of many socio-spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
 The specificity of geographical models compared to other social science models is that space and spatial interactions are given a prime role, geographers being driven by an explicit interest in studying the way space influences the outcomes of social processes modelled.
@@
 
-Geographical systems can be crudely described as social agents interacting with one another, within a limited portion of space.
+Socio-spatial systems can be crudely described as social agents interacting with one another, within a limited portion of space.
 Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependency \citep{arthur1994increasing}.
@@
 Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex systems generally, a given configuration contains clues about past bifurcations, that can have had dramatic effects on the state of the system.
-Therefore, strong spatio-temporal path-dependencies in the trajectory of individual cities and changing social environments over time prohibits the use of ergodic models.
+Therefore, strong spatio-temporal path-dependencies in the trajectory of individual territories and changing social environments over time prohibits the use of ergodic models.
 Ironically, these very models tend to be the models most frequently used in geosimulation.
 
-Self-organization has been shown to be a central feature of geographical systems in general and of cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
+Self-organization has been shown to be a central feature of socio-spatial systems in general and of cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
 In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}.
-Complexity is partially due to bifurcations, which are determinant in spatial systems~\citep{Wilson1981, Wilson2002}.
+Complexity is partially due to bifurcations, which are determinant in socio-spatial systems~\citep{Wilson1981, Wilson2002}.
 Indeed, in spatially explicit simulation models, the non-linearity of local interactions is very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures.
@@
 In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions.
-However, most geographical models, especially cellular automata, still represent cities as uniform grids of isotropic density.
+However, most socio-spatial models, especially cellular automata, still represent cities as uniform grids of isotropic density.
 Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents tend to be approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}.
@@
 For example, households can have different preferences with respect to the built-environment they might want to live in \citep{SpielmanHarrison2014}, or agents moving around will sense a different set of objects within the same fixed radius depending on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}.
-The way modellers represent initial space is therefore a central element of geographical simulation models.
+The way modellers represent initial space is therefore a central element of socio-spatial simulation models.
 However, this step is rarely explicit.
@@
 
-In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape).
-In no way do we pretend to provide a full exploration of these two particular models, their attractors and/or potential policy implications.
-Instead, we present a way of performing a sensitivity analysis to initial spatial conditions of models by using a spatial generator to produce a variety of density grids, which are taken as input by simulation models at initialisation.
+%In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape).
+
+In this paper, we aim to provide an operational framework for studying the influence of geometric structures of space on the results of a simulation model representing a sociospatial system.
+{\bf Sociospatial systems} are understood here as groups of social agents whose behaviour is constrained by their position in geographical space, in the empirical world.
+The aim of the geosimulation models we are interested in is to represent, simulate and explain the dynamics of these systems.
+In the sociospatial systems as well as in the models representing them, agents are in {\bf interaction} among themselves and with their geographical environment.
+The {\bf geometric structure} of the environment defined at the initialisation of the model and its influence on the model outputs is our object in this paper.
+We present how to generate the geometric outputs to are used as initial spatial conditions in the model.
+To this end, we use the example of a density grid generator and feed its outputs to two application models (Schelling and Sugarscape).
+In no way do we pretend to provide a full exploration of these two particular models, of their attractors and/or potential policy implications.
+Instead, we present a way of performing a sensitivity analysis to initial spatial conditions of models generated systematically.
 The generator being controlled by its own parameters, we can then relate the parameters used to generate initial spatial conditions to the variation simulation outcomes.
-The purpose is two-fold: (i) to test the robustness of simulation results to small variations of generator parameters and (ii) to study the non-trivial effects of typical categories of spatial distribution (monocentric \textit{vs.} polycentric for example) on the results of a given model.
+The purpose is two-fold: (i) to test the robustness of simulation results to small variations of generator parameters and (ii) to study the non-trivial effects of typical categories of geometric structures (monocentric \textit{vs.} polycentric for example) on the results of a given model.
 Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies to our knowledge.
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{Previous considerations on the effects of the spatial configuration in simulation models}
+\subsection{The effects of the space in simulation models}
 
+\subsubsection{Spatial processes}
 In 'real-life', different spatial configurations of people in a city tend to be associated with different distributions of income, carbon emissions, educational outcomes, etc.
@@
 Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}.
-The discussion of these effects has also been associated with the field of geostatistics since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984, FotheringhamWong1991}.
+These effects of geographical space on social systems correspond to processes to be included in geosimulation models.\\
+
+\subsubsection{Spatial representation}
+A discussion of the effects of spatial encoding and representation has also been associated with the field of geostatistics since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984, FotheringhamWong1991}.
 For example, \citet{Kwan2012} has argued for a careful examination of what she coins the 'uncertain geographic context problem' (UGCoP), i.e.
-of the spatial configuration of geographical units even when the size and delineation of the area are the same.\\
-
+of the spatial configuration of geographical units even when the size and delineation of the area are the same.
 Considerations of such issues in the geosimulation literature are rather scarce.
@@
 \begin{itemize*}
-\item The accuracy of geo-localised input data;
-\item The shape, precision and boundaries of the modelled spatial system;
-\item The degree of spatial heterogeneity modelled.
-\end{itemize*}
-
-\subsubsection{Geo-localised input data accuracy}
-
+\item The accuracy of geo-localised input data.
 \citet{Thomasetal2017} show that data selection in LUTI model is inter-related with the delineation of the spatial system boundaries and the scale of analysis.
@@
 In the context of spatial interaction models, \citet{hagen2012new} acknowledge the dilemma between spatial resolution and the computational burden, and suggest a method of adaptive zoning (where the size of destination zones depends on the distance to origin) to solve it.
-
-\subsubsection{Spatial system shape, precision, and boundaries}
-
+\item The shape, precision and boundaries of the modelled spatial system.
 \citet{Axtelletal1996} highlight the sensitivity of the average number of stable cultural regions generated to the effect of the territory width implemented in a version of the Sugarscape model which is docked (i.e.
@@
 Spatial resolution also plays a role in the Schelling model: \citet{Singhetal2009} show that the segregation patterns for certain tolerance values are strictly a small city phenomenon (8x8 city-lattice) and do not work for a larger spatial lattice (100x100), where segregation appears only for certain combinations of tolerance threshold and vacancy density values.
-
-\subsubsection{Spatial heterogeneity}
-
+\item The degree of spatial heterogeneity modelled.
 \citet{StaufferSolomon2007} introduce asymetric interactions and empty residences in Schelling's model run on a large and regular lattice.
@@
 In a spatial epidemics model for an infectious disease, parameterised on real data, \citet{smith2002predicting} finds that the physical landscape heterogeneity, in particular the presence of rivers, locally influence the propagation speed.
+\end{itemize*}
 
-Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.) or on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
+\subsubsection{Spatial structures}
+We can distinguish a last category of spatial effects in geosimulation, which are the geometric constraints of the environment modelled at initialisation on the course and output of the simulation.
+Our original contribution is to tackle this type of spatial effects.
+In this paper, we present an operation framework which allows to systematically measure the impact of the initial geometric structures on the aggregate behaviour of simulation models.
+We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.
+
+%Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.) or on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time.
 More so, the references cited in this section are few compared to the mass of geosimulation articles published without any mention of dependence of output on initial spatial condition.
@@
 %\cite{ligmann2013spatially}
-
-In this paper, we build on these contributions and go further by systematically measuring the impact of the city density distribution aggregate model behaviours (e.g.
-the final segregation level for Schelling).
-We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.
 
@@
 
-The general method workflow of our method is illustrated in Figure \ref{fig:method}.
+The general workflow of our method is illustrated in Figure \ref{fig:method}.
 In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
@@
 The sensitivity analysis relates the variations in the model's outcomes to how the density spatial distribution was generated and to the patterns of density generated.
-In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the distribution of geographical entities (people, housing, networks, etc.) in space.
+In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the spatial distribution of socio-spatial entities (people, housing, networks, etc.).
 In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
@@
 %\includegraphics[width=0.9\textwidth]{figuresraw/SchemaMeta_1.png}
-\includegraphics[width=0.9\textwidth]{figures/Fig1.png}
+\includegraphics[width=0.9\textwidth]{figures/Fig1_v2.png}
 \caption{\textbf{General workflow.}} \label{fig:method}
@@
 
+\subsubsection{Schelling}
 Schelling's model consists in an abstract urban housing market where agents of different attributes (for example: red or green) sense their environment, evaluate their satisfaction in terms of neighbourhood composition (how many reds and greens?), and relocate if unsatisfied.
@@
 In addition, we are interested in testing the impact of the initial spatial distribution of housing capacity in this project, using the generated grids.
-The outcome of the model is measured by a combination of three segregation indices: Dissimilarity, Moran's I and Entropy~\citep{brown2006spatial}.
+\\
+The outcome of the model is measured by a combination of three segregation indices: Dissimilarity, Moran's I and Entropy.
+The dissimilarity index (or Duncan's D) is a global measure of segregation and the most widely used, although it does not account for local variations~\citep{white1986segregation, brown2006spatial}.
+It corresponds to the minimum percentage of the population who would have to move to another location so that the global area exhibits a uniform distribution of groups across its constituent areas (each small area would then display the same distribution as the area as a whole).
+Moran's I is a spatial segregation index which denotes the overall spatial autocorrelation of a group.
+It takes positive values when a group is clustered in space (people of the same group as oneself are over-represented in the neighbourhood) and negative values when a group is dispersed (people of the same group as oneself are under-represented in the neighbourhood).
+Entropy can be used as an indicator of spatial segregation because it measures the evenness of distribution of groups in space, although the metaphor of thermodynamics is not straightforward \citep{barner2017multiscale}.
+In our case, areas with higher values of entropy are considered more segregated because the mix of groups is uneven across small areas, whereas low entropy denotes more uniform distributions.\\
+
 We use a ad-hoc implementation of the Schelling model, both in Scala for performance reasons and in NetLogo to ensure visualization of model dynamics.
-The pseudo-code of the implemented model is available in~\ref{app:code} and source code are available on the repository of the project.
+The pseudo-code of the implemented model is available in~\ref{app:code} and source code are available on the repository of the project.\\
 
@@
 
+\subsubsection{Sugarscape}
 Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \cite{EpsteinAxtell1996}.
@@
 
-\subsection{Quantitative variation across grids}
+\subsection{Sensitivity analysis}
 
@@
 
-\subsection{Qualitative variations}
+\subsection{Variation by type of structure}
 \label{sec:qualResults}
@@
 
-We consider that the method presented in this paper holds great potential for strengthening geographical models' exploration.
+We consider that the method presented in this paper holds great potential for strengthening socio-spatial models' exploration.
 However, two limits and two areas of opportunities have still not been tackled.
@@
 With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space.
-We think that this method has the potential to increase the arsenal of evaluation of geographical models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
+We think that this method has the potential to increase the arsenal of evaluation of socio-spatial models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
 %%%%%%%%%%%%%

### 2019-08-06 Response_to_Reviewers.tex
.

@@
 
-%Dear Editor,
-
 \justify
-
-%Thank you for the feedback on the draft paper ``Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models''.
-The suggestions and comments will undoubtedly be of great value to the paper.
-It was updated accordingly.
-
-\medskip
 
@@
 
-Resubmit 15 Aug. 2019\\
+%Resubmit 15 Aug. 2019\\
+
+\bigskip
 
@@
 Our original contribution is to tackle the second type.
-\comment[CC]{done in 1.5 and 1.8}
+%\comment[CC]{done in 1.5 and 1.8}
 
 \item A.2.
-We have clarified the terms used in our paper introduction to distinguish between: {\bf sociospatial systems} (i.e.
+We have clarified the terms used in our paper introduction to distinguish between: \textbf{sociospatial systems} (i.e.
 groups of social agents whose behaviour is constrained by their position in geographical space.
-This term refers to empirical systems, which are the targets of the modelling process), {\bf interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and {\bf geometric structure}, a term we reserve for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model.
+This term refers to empirical systems, which are the targets of the modelling process), \textbf{interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and \textbf{geometric structure}, a term we reserve for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model.
 When we talk about the generator we actually used in the paper, we call it a "density grid generator" because the is the only geometric structure it generates, whereas a "spatial generator" could potentially generate graph or point structures.
-\comment[CC]{done in 1.5}
+%\comment[CC]{done in 1.5}
 
@@
 We have also replaced the expressions "quantitative results" by "sensitivity analysis" and "qualitative results" by "variations by types of geometric structure", in the figure as in the subsections of the paper.
-\comment[CC]{done}
+%\comment[CC]{done}
 \end{itemize}
@@
 
- {\bf Sociospatial system} pour parler des systèmes empiriques. \comment[MLT]{Done}\\
- {\bf interactions} pour parler de ce qui se passe dans le modele.\\
-  {\bf geometric structures} pour parler des outputs issus du générateur.\\
+ \textbf{Sociospatial system} pour parler des systèmes empiriques. \comment[MLT]{Done}\\
+ \textbf{interactions} pour parler de ce qui se passe dans le modele.\\
+  \textbf{geometric structures} pour parler des outputs issus du générateur.\\
   + parler de "density grid generator" quand on développe le nôtre. \\
@@
 With this approach, the paper convincingly outlines a pragmatic application by creating typical spatial compositions in urban contexts that is monocentric, polycentric or discontinuous cities.
-From this methodological perspective of reviewing, it would be helpful to add some more information to approaches/techniques used within the tool of the spatial generator.
+ 	\item \textit{From this methodological perspective of reviewing, it would be helpful to add some more information to approaches/techniques used within the tool of the spatial generator.
 For example, paragraph 2.8 mentions four dimensions of spatial structure but leave the reader alone with just the terms.
-Either describe these or omit it.
-   \item \comment[JR]{will add some description on urban form indicators etc.}
-   \item To describe more the spatial generator we used in a few sentences (JR) + situate this one among the many possible ways of making a spatial generator as Young2002 (FLN/MLT).
-   %This also applies to the three segregation indexes: what exactly do they measure?
-  \item Describe segregation index in a few sentences \comment[CC]{done in section "case study models"}
+Either describe these or omit it.}
 
-  % The description of the spatial generator’s procedure, outlined in paragraph 2.7, would also benefit if more information is added, for example, to delineate the terms “strength of attraction” (attraction in which respect?) and “strength of the diffusion process” (is strength equal to intensity?
+   $\rightarrow$ To describe more the spatial generator we added a few sentences; and the generator was furthermore better situated among the many possible others in the discussion.
+   %ways of making a spatial generator as Young2002 (FLN/MLT).
+ 	% item \comment[JR]{will add some description on urban form indicators etc.}
+
+  \medskip
+
+   \item \textit{This also applies to the three segregation indexes: what exactly do they measure?}
+
+  $\rightarrow$ segregation indexes were better described in the section describing the models.
+  %\comment[CC]{done in section "case study models"}
+
+  \medskip
+
+  \item \textit{The description of the spatial generator’s procedure, outlined in paragraph 2.7, would also benefit if more information is added} \ldots
+  %, for example, to delineate the terms “strength of attraction” (attraction in which respect?) and “strength of the diffusion process” (is strength equal to intensity?
 If so, is it then in space and/or time?).
-  \item \comment[JR]{will add more precision in the description of the generator}
 
-  %While the methodological part of the paper is convincing the theoretical and epistemological background is less so.
-In short, the paper would improve significantly, if most of the geographical and empirical associations would be left out.
-Even claimed by the author(s), the approach has not much to do with the empirical growth or development of cities, just because the approach reduces cities’ growth to the growth itself, omitting issues such as historical genesis, cultural differentiation, economic background political and legal circumstances, etc.
-  \item \comment[JR]{most of this empirical literature was added because of one CEUS reviewer, no ?}\comment[MLT]{If we want to keep this part to please everyone (like our former reviewer), we could explicitly mention that we model the growth of a city (in terms of population density) which is the results of a bunch of dynamics ...
-and not the dynamics themselves?}\comment[CC] We have clarified the semantic terms used in sections 1.5 to 1.8 and in the figure 1.
-\comment[FLN]{ qui montre que la démarche méta est ici appliquée à un domaine empirique qu'il aide à explorer mais d'autres domaines peuvent s'emparer de la démarche méta.}
-\item show disagreement with omitting empirical/geographical but demonstrate helps understand geographical question but the is generic and valuable for it.
+  $\rightarrow$ \comment[JR]{will add more precision in the description of the generator}
 
-  % The problem, from a geographical point of view, is that the paper switches between “geographical” and “spatial” without making any differentiation.
-In so doing, the author(s) runs immediately into the problem of adequate definitions.
+  \medskip
+
+  \item \textit{While the methodological part of the paper is convincing the theoretical and epistemological background is less so.
+In short, the paper would improve significantly, if most of the geographical and empirical associations would be left out.}
+  %Even claimed by the author(s), the approach has not much to do with the empirical growth or development of cities, just because the approach reduces cities’ growth to the growth itself, omitting issues such as historical genesis, cultural differentiation, economic background political and legal circumstances, etc.
+
+   $\rightarrow$ We have clarified the semantic terms used in sections 1.5 to 1.8 and in the figure 1.
+
+   %\comment[JR]{most of this empirical literature was added because of one CEUS reviewer, no ?}\comment[MLT]{If we want to keep this part to please everyone (like our former reviewer), we could explicitly mention that we model the growth of a city (in terms of population density) which is the results of a bunch of dynamics ...
+and not the dynamics themselves?}
+  %\comment[FLN]{ qui montre que la démarche méta est ici appliquée à un domaine empirique qu'il aide à explorer mais d'autres domaines peuvent s'emparer de la démarche méta.}
+
+   % show disagreement with omitting empirical/geographical but demonstrate helps understand geographical question but the is generic and valuable for it.
+
+\medskip
+
+  \item \textit{The problem, from a geographical point of view, is that the paper switches between “geographical” and “spatial” without making any differentiation. \ldots}
+  %In so doing, the author(s) runs immediately into the problem of adequate definitions.
 In paragraph 1.2 the definition of geographical systems as “social agents interacting with one another, within a limited portion of space” is misleading.
@@
 ling fuzzily the social with the geographical, as happened with this definition.
-  \item \comment[JR]{by spatial we indeed mean some kind of spatial container for agents right ?
+  %\item \comment[JR]{by spatial we indeed mean some kind of spatial container for agents right ?
 which is far from geographical in the sense he puts it indeed}\comment[MLT]{use sociospatial system instead?
@@
 2008}}
-\item cf. comment A
-\item reprendre tout le papier pour éliminer les termes indéfinis "space", "pattern" etc.
+
+ $\rightarrow$ cf. comment A
+
+ \medskip
+
+  %\item reprendre tout le papier pour éliminer les termes indéfinis "space", "pattern" etc.
 + en conclu mentionner que ça s'applique aux automates cellulaires (sans agents) et aux modèles écologiques (agents-arbres).
-  \item in conclusion, we need to clarify the answer to the question "do geometric structure influence sociospatial systems dynamics" by saying it depends on the model.
+  %\item in conclusion, we need to clarify the answer to the question "do geometric structure influence sociospatial systems dynamics" by saying it depends on the model.
 ≠ black\&white and we can demonstrate it.
@@
 
-  % Then, in paragraph 1.3, the complexity of systems with respect to segregation is also problematically described.
-Emergent behaviour that may lead to segregation is first and foremost a social process, taking social interactions at the local scale (income, housing costs, etc.) and/or at the macro scale (laws, rules, economic system, cultural attitudes, etc.) into account.
+  \item \textit{Then, in paragraph 1.3, the complexity of systems with respect to segregation is also problematically described.} \ldots
+  % Emergent behaviour that may lead to segregation is first and foremost a social process, taking social interactions at the local scale (income, housing costs, etc.) and/or at the macro scale (laws, rules, economic system, cultural attitudes, etc.) into account.
 These interactions can then be associated with the material and relational dimension of geographical spaces.
@@
 In this respect, the paper offers some relevant clues such as the MAUP problem or the “uncertain geographic context problem”.
-  \item see comment A. cite Banos, 2012
 
-  % The problem of setting space and geography as synonymous continues in paragraphs 1.9 and following.
-What is described here as “spatial system shape” is the geometry of spaces but not the geography of spaces.
+  $\rightarrow$ see comment A. cite Banos, 2012
+
+  \medskip
+
+  \textit{The problem of setting space and geography as synonymous continues in paragraphs 1.9 and following.
+What is described here as "spatial system shape" is the geometry of spaces but not the geography of spaces.}
   % -> yes indeed
-  \item see comment A.
 
-  % Even more critically is the arbitrary conflation of social and spatial units as happened in paragraph 2.1, as geographical entities are described with “people, housing, networks, etc.” One is inevitably asking oneself: do geographical entities comprise of simply everything?
-Obviously not, but the paper seems to argue this way.
-  \item see comment A.
+  $\rightarrow$ see comment A.
 
-  % Due to this, attempts to explain the growth of existing cities should be avoided.
-  \item we agree.
-it's not the point of the paper, only the metaphore of the spatial generator.
+  \medskip
+
+  \item \textit{Even more critically is the arbitrary conflation of social and spatial units as happened in paragraph 2.1, as geographical entities are described with “people, housing, networks, etc.” One is inevitably asking oneself: do geographical entities comprise of simply everything?
+Obviously not, but the paper seems to argue this way.}
+
+  $\rightarrow$ All these terms were clarified (see comment A).
+
+  \medskip
+
+  \item \textit{Due to this, attempts to explain the growth of existing cities should be avoided.}
+
+  $\rightarrow$ We agree with this comment and recall that urban growth processes are not the main focus of the paper, they are here an instantiation of a given $\gamma$ generator, which could have been replaced by others such as network models, procedural modeling, etc.
+The position and discussion of the paper was amended to precise that (see comment A).
+
+  %it's not the point of the paper, only the metaphore of the spatial generator.
 + supprimer la ref à "urban growth" dans le texte et dire d'inspiration urban science mais dont seules les grilles finales nous intéressent.
 On aurait pu utiliser des generator issus de graphes, de percolation etc.
-  The core of the paper is not this, it's the approach of including gamma and applying to mu to understand systems better, whatever gamma and whatever mu in whatever knowledge domain.
+  %The core of the paper is not this, it's the approach of including gamma and applying to mu to understand systems better, whatever gamma and whatever mu in whatever knowledge domain.
+  %\item in discussion, mention that it's one isolated example of geometry but other cases could be thought of as well, using networks and graphs, but also empirical shape files.
 
- \item in discussion, mention that it's one isolated example of geometry but other cases could be thought of as well, using networks and graphs, but also empirical shape files.
+  \medskip
 
@@
 
-  % While the different outcomes of the two implementation exercises are not the main point, I would suggest authors to provide some more elaborated explanations.
+  \item \textit{While the different outcomes of the two implementation exercises are not the main point, I would suggest authors to provide some more elaborated explanations.
 The two models have different behavioural assumptions and represent a different 'interaction' context.
-How can be extrapolate and isolate spatial effects when models have different logics and theoretical purposes?
-  \item see above comment A, discussion has been amended and details added.
+How can be extrapolate and isolate spatial effects when models have different logics and theoretical purposes?}
 
-  % In 4.1, the authors suggest potential developments and point to certain limitations.
-While limitations are well-discussed, readers who are not specialist of spatial models could benefit from a more comprehensive discussion on potential applications and extensions.
-  \item see comment B
+  $\rightarrow$ see above comment A, discussion has been amended and details added.
+  \medskip
 
-  %For instance, it would be useful to have suggestions on how to use this methodology to run empirical models.
+  \item \textit{In 4.1, the authors suggest potential developments and point to certain limitations.
+While limitations are well-discussed, readers who are not specialist of spatial models could benefit from a more comprehensive discussion on potential applications and extensions.}
+
+  $\rightarrow$ see comment B
+
+  \medskip
+
+  \textit{For instance, it would be useful to have suggestions on how to use this methodology to run empirical models.
 Suppose to have a nice dataset and a case of a urban context to be modelled.
-Could the exploration of different spatial configurations even be used to run 'counter-factual' or 'what if' tests on actual spatial configurations?
-  \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model} cf.
-comment B.3
+Could the exploration of different spatial configurations even be used to run 'counter-factual' or 'what if' tests on actual spatial configurations?}
+
+  $\rightarrow$ cf. comment B.3
+  % \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}
 
@@
 
-  % Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), which could be interesting to see implemented in ABM.
-  \item cf. discussion of egohoods in conclusion (comment B)
+  \medskip
+
+  \textit{Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), which could be interesting to see implemented in ABM.}
+
+  $\rightarrow$ cf. discussion of egohoods in conclusion (comment B)
+
+  \medskip
 
@@
 
-  % Furthermore, even among limitations, perhaps authors could discuss data limitations and the fact that social processes (networks, mobility) cannot conflate with spatial dynamics.
-Here, it is a bit misleading to call "generator parameters" ONLY the spatial ones!
- \item we have clarified this issue in intro but referring to different terms and figure 1 which differentiates between the mu model and gama parameters. (comment A)
-  %In this respect, I would encourage authors to further develop their closing section, which now does not stimulate readers to imagine implications of authors' findings and message to their own research.
-I would expand this section to provide a more detailed discussion on implications for model makers, experts on social geography and urban scholars who work with ABM, non-specialist in geography who would like to base their models in more proper geographical constraints.
+  \medskip
+
+  \item\textit{Furthermore, even among limitations, perhaps authors could discuss data limitations and the fact that social processes (networks, mobility) cannot conflate with spatial dynamics.
+Here, it is a bit misleading to call "generator parameters" ONLY the spatial ones!}
+
+  $\rightarrow$ we have clarified this issue in introduction but referring to different terms and figure 1 which differentiates between the mu model and gama parameters. (see comment A)
+
+ \medskip
+
+  \item\textit{In this respect, I would encourage authors to further develop their closing section, which now does not stimulate readers to imagine implications of authors' findings and message to their own research.} \ldots
+  %I would expand this section to provide a more detailed discussion on implications for model makers, experts on social geography and urban scholars who work with ABM, non-specialist in geography who would like to base their models in more proper geographical constraints.
 And perhaps, authors could suggest future line of research that could show-case their argument.
-  \item cf. new discussion/conclusion and comment B
+
+  $\rightarrow$ cf. new discussion/conclusion and comment B
 
 %I appreciate that models are fully available to readers. 
-\item We do too!
+%\item We do too!
 

### 2019-08-06 paper.tex
.

@@
 
+\subsubsection{Sensitivity of indicators to the geometry}
+
+While we investigated the sensitivity to geometry of models together with their indicators, a potential research direction is to study the dependance of indicators themselves to the spatial configuration.
+% TODO finish
+
 \subsubsection{Platform constraints and docking challenges}
@@
 
-Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents, transportation networks generators \citep{raimbault2018multi}, or coupled road network and population raster generators \citep{raimbault2018urban}.
-% Generators of abstract urban systems at a macroscopic scale could also be useful tools to better understand urban growth models as sketched by \cite{raimbault2018unveiling}.}
+Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents, transportation networks generators \citep{raimbault2018multi}, or coupled road network and population raster generators \citep{raimbault2019urban}.
+The particular density grid generator we used here is an example among possible others, and our methodological contribution is generic as it does not depend on which generator is chosen.
 
@@
 We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity the for social sciences to build their own stream of methodologies in the modelling domain.
+
+This relates to relations between empirical, conceptual and modeling dimensions of quantitative research in social sciences, as coined by \cite{livet2010ontology}. % TODO finish
 

### 2019-08-11 Response_to_Reviewers.tex
Update on Overleaf.

@@
 This term refers to empirical systems, which are the targets of the modelling process), \textbf{interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and \textbf{geometric structure}, a term we reserve for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model.
-When we talk about the generator we actually used in the paper, we call it a "density grid generator" because the is the only geometric structure it generates, whereas a "spatial generator" could potentially generate graph or point structures.
+When we talk about the generator we actually used in the paper, we call it a "density grid generator" because this is the only geometric structure it generates, whereas a "spatial generator" could potentially generate graph or point structures.
 %\comment[CC]{done in 1.5}
@@
 This is a limitation of the method as it is but also an avenue for promoting stronger indicators in empirical research.
-\comment{il faut reprendre ces elements dans l'article}
+% \comment[MLT]{j'ai formulé un paragraphe court sur ce point en discussion, en lien avec le point B.3}
 
   \item B.2.
-To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, We added a paragraph in conclusion, discussing how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences).
+To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, We added a paragraph in discussion, explaining how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences).
 \comment{il faut reprendre ces elements dans l'article}
@@
 We could also in another paper vary the size of grid and its cell shape (cf.
-MAUP), but also generate other types of geometric structures to feed into geosimulation models (for example networks).
-\comment{il faut reprendre ces elements dans l'article}
+MAUP), but also generate other types of geometric structures to feed into geosimulation models (for example points or networks).
+% \comment{j'ai formulé un paragraphe court sur ce point en discussion, en lien avec le point B.1}
 \end{itemize}
@@
 
-\textbf{to do list:}
+%\textbf{to do list:}
 
- \textbf{Sociospatial system} pour parler des systèmes empiriques. \comment[MLT]{Done}\\
- \textbf{interactions} pour parler de ce qui se passe dans le modele.\\
-  \textbf{geometric structures} pour parler des outputs issus du générateur.\\
-  + parler de "density grid generator" quand on développe le nôtre. \\
-  caser" Jessop (2008) points in particular to the interdependence between social phenomena and the spatial environment where they take place.
+ % \textbf{Sociospatial system} pour parler des systèmes empiriques, \textbf{interactions} pour parler de ce qui se passe dans le modele,  \textbf{geometric structures} pour parler des outputs issus du générateur et  parler de \textbf{density grid generator} quand on développe le nôtre.
+\comment{done}
+
+ % caser" Jessop (2008) points in particular to the interdependence between social phenomena and the spatial environment where they take place.
 We model the social system as the interaction dynamics of agents and spatial units.
-The spatial generator produces the geometric support for these interactions." in intro 1.4 \\
+The spatial generator produces the geometric support for these interactions." in intro 1.4 \\ \comment[MLT]{To avoid redundancies, I have simply added the reference to Jessop when we describe Figure 1.}
 
@@
 
-  $\rightarrow$ cf. discussion of egohoods in conclusion (comment B)
+  $\rightarrow$ we added a paragraph on the limits of our density grid approach in discussion, in which we mention the possibility to use egohoods to better measure segregation.
 

### 2019-08-11 paper.tex
Update on Overleaf.

@@
 \author[5]{Marion Le Texier}
-\affil[5]{UMR 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
+\affil[5]{UMR CNRS 6266 IDEES, Universit{\'e} de Rouen Normandie, France}
 
@@
 However, initial spatial conditions are usually not modified systematically in socio-spatial models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment.
-In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic spatial configuration generator in order to create density grids with which socio-spatial simulation models are initialised.
+In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic geometric structures generator in order to create density grids with which socio-spatial simulation models are initialised.
 We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models.
@@
 Self-organization has been shown to be a central feature of socio-spatial systems in general and of cities in particular ~\citep{AllenSanglier1981,saint1989villes, Portugali2000}.
-In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, Batty2007}.
+In the vocabulary of complex systems, cities also exhibit emergent properties at macroscopic scales~\citep{pumain2006hierarchy, AzizAlaouiBertelle2009}, which can be simulated through microscopic interactions between agents~\citep{Wu2002, batty2007cities}.
 Complexity is partially due to bifurcations, which are determinant in socio-spatial systems~\citep{Wilson1981, Wilson2002}.
@@
 In this paper, we aim to provide an operational framework for studying the influence of geometric structures of space on the results of a simulation model representing a sociospatial system.
-{\bf Sociospatial systems} are understood here as groups of social agents whose behaviour is constrained by their position in geographical space, in the empirical world.
+Following \citep{Jessop2008} {\bf Sociospatial systems} are understood here as groups of social agents whose behaviour is constrained by their position in geographical space, in the empirical world.
 The aim of the geosimulation models we are interested in is to represent, simulate and explain the dynamics of these systems.
@@
 The {\bf geometric structure} of the environment defined at the initialisation of the model and its influence on the model outputs is our object in this paper.
-We present how to generate the geometric outputs to are used as initial spatial conditions in the model.
+We present how to generate the geometric outputs which are used as initial spatial conditions in the model.
 To this end, we use the example of a density grid generator and feed its outputs to two application models (Schelling and Sugarscape).
@@
 
-The spatial generator applies an urban morphogenesis model \citep{Batty2007} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}.
+The spatial generator applies an urban morphogenesis model \citep{batty2007cities} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}.
 An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily into our workflow.
@@
 The implementations of the models were done from NetLogo.
-We modified the sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model.
+We modified the Sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model.
 Both pseudo-codes are available in~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
@@
 We tested the sensitivity to the type of distance, using a Minkovski distance with a varying exponent.
-The results are presented in~\ref{app:distances}, and show a similar sensitivity to spatial initial conditions.
+The results are presented in~\ref{app:distances}, and show a similar sensitivity to geometric structures.
 
@@
 
-We obtain a very strong sensitivity to initial spatial conditions for the Sugarscape model.
+We obtain a very strong sensitivity to geometric structures for the Sugarscape model.
 Indeed, the relative distance between the phase diagrams of different density grids and the phase diagram of the reference case ranges from 0.09 to 2.98 with a median of 1.52 and an average value of 1.30.
@@
 We consider that the method presented in this paper holds great potential for strengthening socio-spatial models' exploration.
-However, two limits and two areas of opportunities have still not been tackled.
+However, three limits and two areas of opportunities have still not been tackled.
 
@@
 
-While we investigated the sensitivity to geometry of models together with their indicators, a potential research direction is to study the dependance of indicators themselves to the spatial configuration.
-% TODO finish
+While we investigated the sensitivity to geometry of models together with their indicators, a potential research direction is to study the dependence of indicators themselves to the spatial configuration.
+Indeed, the segregation indices are likely to be zoning, size or population density dependant \citep{Wong1997,ReardonOSullivan2004}.
+Generating point data rather than an grid could help overcoming these issues, and appears as a promising direction for future research.
+This methodological claim is also interesting from theoretical and empirical views as point data may be used to construct \textit{egohoods} \citep{Hippetal2013}, i.e.
+a smooth definition of neighbourhoods as opposed to the commonly used non-overlapping units.
 
@@
 This relates to relations between empirical, conceptual and modeling dimensions of quantitative research in social sciences, as coined by \cite{livet2010ontology}. % TODO finish
+% to inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, we added a paragraph in conclusion, discussing how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how ’neighbourhood’ is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences)
 
@@
 We think that this method has the potential to increase the arsenal of evaluation of socio-spatial models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
+
 %%%%%%%%%%%%%

### 2019-08-12 Response_to_Reviewers.tex
Update on Overleaf.

@@
 We want to thank the two reviewers for their thorough evaluation and discussion of our paper, as well as the challenging questions they asked.
-We have revised the manuscript in a way that addresses most of the comments.
+We have revised the manuscript in a way that we believe addresses most of the comments.
 In this response, we first want to address the two main questions that arose for both reviews.
@@
 We have clarified the terms used in our paper introduction to distinguish between: \textbf{sociospatial systems} (i.e.
-groups of social agents whose behaviour is constrained by their position in geographical space.
-This term refers to empirical systems, which are the targets of the modelling process), \textbf{interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and \textbf{geometric structure}, a term we reserve for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model.
-When we talk about the generator we actually used in the paper, we call it a "density grid generator" because this is the only geometric structure it generates, whereas a "spatial generator" could potentially generate graph or point structures.
-%\comment[CC]{done in 1.5}
+groups of social agents whose behaviour is constrained by their position in geographical space)%.
+This term refers to empirical systems, which are the targets of the modelling process)% Mise en commentaire par FL 190812 : je trouve que cela alourdi inutilement la phrase
+, \textbf{interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and \textbf{geometric structure}, a term we use for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model.
+When we talk about the generator we actually used in the paper, we call it a "density grid generator" because it generates a square grid with heterogeneous density, whereas in all generality a "spatial generator" could potentially generate graph or point structures.
+%\comment[CC]{done in 1.5} %Changement par FL 190812 Because this is the only geometric structure it generates
 
@@
  \medskip
-B. To highlight limitations as well as promising perspectives of improvement, answer the concern about the genericity of the method and expose its links with the conceptual and empirical domains:
+B. To highlight limitations as well as promising perspectives of improvement, we answer the concern about the genericity of the method and expose its links with the conceptual and empirical domains:
 
@@
   \item B.1.
-We have added a discussion of how segregation indicators can themselves be influenced by geometric structures (because they give the same weight to cells of different density).
+We have added a discussion of\comment[FL]{on?} how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities).
 This is a limitation of the method as it is but also an avenue for promoting stronger indicators in empirical research.
@@
   \item B.2.
-To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, We added a paragraph in discussion, explaining how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences).
+To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, We added a paragraph in discussion, explaining how our approach allows us to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences).
 \comment{il faut reprendre ces elements dans l'article}
@@
   \item B.3.
-As a perspective to reduce this bias, we could compute ego-neighbourhoods rather than a regular grid (another paper).
-We could also in another paper vary the size of grid and its cell shape (cf.
+As a perspective to reduce this bias, we could compute ego-neighbourhoods rather than a regular grid (another paper)\comment[FL]{pas clair : pour calculer les indicateurs?
+pour les interactions?
+à préciser comment sont utilisés ces ego neighborhoods}.
+We could also\comment[FL]{plutôt : In further researches, we plan to extend the analysis...} in another paper vary the size of grid and its cell shape (cf.
 MAUP), but also generate other types of geometric structures to feed into geosimulation models (for example points or networks).
@@
 
-   $\rightarrow$ To describe more the spatial generator we added a few sentences; and the generator was furthermore better situated among the many possible others in the discussion.
+   $\rightarrow$ To describe in more details the density grid generator we added a few sentences; and the generator was furthermore better situated among the many possible others in the discussion.
    %ways of making a spatial generator as Young2002 (FLN/MLT).
@@
 
-  $\rightarrow$ see comment A. cite Banos, 2012
+  $\rightarrow$ see comment A. cite Banos, 2012 \comment[FL]{quel était le point exactement avec Banos?}
 
@@
 
-  %The approach is reductionist and has its strength in reflecting growth at an abstract level nevertheless.
+  \item \textit{The approach is reductionist and has its strength in reflecting growth at an abstract level nevertheless.
 The concluding remark in paragraph 5.1 “With the Schelling experiment, we found … that polycentric and discontinuous cities appear systematically more segregated than compact cities …” is elusive.
@@
 This geographical reasoning, however, is outside of the scope of this paper.
-  \item \comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate} \comment[MLT]{I kind of disagree with the idea that it is straightforward that centralised cities produce more segregation as it is not a direct parameter of the model but an output.
+}\comment[FL]{j'ai pas compris pourquoi la phrase du reviewer avait été mise en commentaire, j'ai remis}
+ \comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate} \comment[MLT]{I kind of disagree with the idea that it is straightforward that centralised cities produce more segregation as it is not a direct parameter of the model but an output.
 Plus: what argument would he or she have to state that a big center would be automatically less segregated that several smaller ones?
 And I believe that explaining why polycentric cities are more segregated than others is out of the scope of the paper.
-But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?}
+But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?} \comment[FK]{du coup je fais une proposition de réponse ci-dessous en reprenant ces différents éléments}
+  We understand this statements, however we wish to underline that our objective in the discussion of the Schelling results was not to establish a general result on the links between polycentricity and segregation, which are indeed very complex and multidimensional as advocated by the reviewer.
+Instead, we wished to show what kind of results can be obtained using the framework we suggest.
+If we do not prove that in general, and in empirical context, polycentricity creates more segregation, we obtained, in the modelling context, under Schelling assumptions, that polycentric geometric structures are more often that not associated with segregated patterns, a result which we do not believe to be solely resulting from the fact that more centers leads to more opportunities of segregation (which is true), because it seems also true that monocentric structures, with higher densities, can more easily accommodate heterogeneity of agents within cells, and hence a more patchy type of segregation which can for instance being observed in London.
+We included a comment in the discussion section, saying that we hope we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?
+\comment[FL]{à la relecture de l'article s'assurer que cette prise de recul est bien explicitée}
 
+  \item \comment[FL]{ne pas oublier les trois points ci-dessous!!!}
   \item in discussion, say that indeed the indicators can themselves be influenced by geometric structures (cf.
@@
   % \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}
-
-  \item it as not explicit enough and we revised the presentation of our approach (which is generic and not empirical) in intro (especially fig.
-1).
-We thank reviewer 2 for highlighting this limitation.
+  it as not explicit enough and we revised the presentation of our approach (which is generic and not empirical) in intro (especially fig. 1). We thank reviewer 2 for highlighting this limitation.
 
@@
 
-  \textit{Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), which could be interesting to see implemented in ABM.}
+  \item \textit{Furthermore, there are now examples of spatial models that go beyond administrative neighborhoods and overimposed spatial boundaries, i.e., egohoods (e.g., https://onlinelibrary.wiley.com/doi/abs/10.1111/1745-9125.12006), which could be interesting to see implemented in ABM.}
 
@@
 We need to add a paragraph about this discussion in conclusion and how our approach allows to go back to questioning empirical cases from the model and questioning our concepts ('neighbourhood').
-cf.
-comment B
+
+   $\rightarrow$  cf. comment B
 

### 2019-08-12 paper.tex
Update on Overleaf.

@@
 We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models.
-Furthermore, this effect is sometimes larger than the effect of parameters' value change.
+We wish to illustrate the potential  add spatial sensitivity analysis during the exploration of their model and that specialist of thethematics would see the opportunity to dig deeper into our results?%Furthermore, this effect is sometimes larger than the effect of parameters' value change.%Comment FL : j'ai supprimé cette dernière phrase qui ne me semblait pas essentielle et rajouté à la place une phrase d'ouverture de Marion.
 \end{abstract}

### 2019-08-12 paper.tex
Update on Overleaf.

--- 2019-08-11 paper.tex
@@
 We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models.
-Furthermore, this effect is sometimes larger than the effect of parameters' value change.
+We wish to illustrate the potential interest of adding spatial sensitivity analysis during the exploration of models for both modellers and thematic specialists.%Furthermore, this effect is sometimes larger than the effect of parameters' value change.%Comment FL : j'ai supprimé cette dernière phrase qui ne me semblait pas essentielle et rajouté à la place une phrase d'ouverture de Marion.
 \end{abstract}
@@
 \begin{keywords}
-Space, Initial conditions, Sensitivity, ABM
+Space, Initial conditions, Sensitivity, Agent-Based Models
 \end{keywords}
@@
 The process is thus completely reversed depending on the grid chosen to run the model on, which would have significant impacts if one tried to deduce policy from this model.
-%This second example confirms thus the importance of sensitivity of simulation models to the initial spatial conditions.
-
-% phase diagrams -> ok well different qualitatively
-%          spAlpha spDiffsteps spDiffusion spGrowth spPopulation
-% id=27 : 0.7913103    2.376837   0.1440293 157.4147 4852.746
-% id=0 : 2.562398    3.753032   0.1316788 128.4632 4753.983
-% maxSugar = 110
 
@@
 
-This relates to relations between empirical, conceptual and modeling dimensions of quantitative research in social sciences, as coined by \cite{livet2010ontology}. % TODO finish
+This relates to relations between empirical, conceptual and modeling dimensions of quantitative research in social sciences, as coined by \cite{livet2010ontology}.
+In our case, a contribution in the modeling domain aiming at extracting further knowledge on model behavior (sensitivity to geometrical context), naturally leads to questioning theoretical concepts and empirical definition on which the model was based, such as for example the meaning of neighborhood.
 % to inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, we added a paragraph in conclusion, discussing how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how ’neighbourhood’ is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences)

### 2019-08-12 Response_to_Reviewers.tex
.

@@
 We have also replaced the expressions "quantitative results" by "sensitivity analysis" and "qualitative results" by "variations by types of geometric structure", in the figure as in the subsections of the paper.
-%\comment[CC]{done}
 \end{itemize}
@@
   \item B.2.
-To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, We added a paragraph in discussion, explaining how our approach allows us to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences).
+To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, we added a paragraph in discussion, explaining how our approach allows us to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences).
 \comment{il faut reprendre ces elements dans l'article}

### 2019-08-12 paper.tex
Update on Overleaf.

@@
 Simulation also appears as a way to overcome the difficult analytic resolution of many socio-spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
-The specificity of geographical models compared to other social science models is that space and spatial interactions are given a prime role, geographers being driven by an explicit interest in studying the way space influences the outcomes of social processes modelled.
+The specificity of geographical models compared to other social science models is that space and spatial interactions are given a prime role, geographers being driven by an explicit interest in studying the way "space" in broad sense influences the outcomes of social processes modelled.
 We think that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of the model outputs to the way space is represented.
@@
 This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions.
+\comment[FL]{il faut ici citer des approches du même type}.
 To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\
@@
 \subsection{Definition of the problem}
-
-Socio-spatial systems can be crudely described as social agents interacting with one another, within a limited portion of space.
-Social agents thus constitute the microscopic level of the system, and they are framed by a system-time that evolves irreversibly, creating temporal and cumulative effects, also known as path-dependency \citep{arthur1994increasing}.
-Therefore, observing a system at different points in time does not equate to observing different systems at a single point in time.
+%change1 : within a limited portion of space
+%change2 : irreversibly, creating temporal and 
+Socio-spatial systems can be crudely described as social agents interacting with one another via the geometric structure of space \comment[FL]{change2}.
+Social agents thus constitute the microscopic level of the system, and they are framed by a system-time\comment[FL]{manque un mot} that evolves\comment[FL]{change2} with potential cumulative effects, also known as path-dependency \citep{arthur1994increasing}.
+Therefore, observing one system at different points in time does not equate to observing different systems at a single point in time.
 This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}.
@@
 Therefore, strong spatio-temporal path-dependencies in the trajectory of individual territories and changing social environments over time prohibits the use of ergodic models.
-Ironically, these very models tend to be the models most frequently used in geosimulation.
+Ironically, these very models tend to be the models most frequently used in geosimulation.\comment[FL]{on ne revient pas du tout là dessus à la fin : est ce que les cas d'étude qu'on mobilise sont si path dependant que ca?
+Si c'est pertinent : revenir dessus.
+Sinon, enlever cette phrase}
 
@@
 Dynamically, sprawl in American cities consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}.
-Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}.
+Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}\comment[FL]{je ne comprends pas ce point : est-ce lié à un modèle de simulation?}.
 These effects of geographical space on social systems correspond to processes to be included in geosimulation models.\\

### 2019-08-12 Response_to_Reviewers.tex
.

@@
 
-  $\rightarrow$ segregation indexes were better described in the section describing the models.
-  %\comment[CC]{done in section "case study models"}
+  $\rightarrow$ Segregation indexes were better described in the section describing the models.
 
@@
 
-  $\rightarrow$ \comment[JR]{will add more precision in the description of the generator}
+  $\rightarrow$ The description of the spatial generator was partly reformulated with more details.
 
@@
 
- $\rightarrow$ cf. comment A
+ $\rightarrow$ This issue is tackled by comment A.
 
@@
 
-  $\rightarrow$ see comment A.
+  $\rightarrow$ We agree and corrected the semantics (see comment A).
 
@@
 This geographical reasoning, however, is outside of the scope of this paper.
-}\comment[FL]{j'ai pas compris pourquoi la phrase du reviewer avait été mise en commentaire, j'ai remis}
- \comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate} \comment[MLT]{I kind of disagree with the idea that it is straightforward that centralised cities produce more segregation as it is not a direct parameter of the model but an output.
+}
+  %\comment[FL]{j'ai pas compris pourquoi la phrase du reviewer avait été mise en commentaire, j'ai remis}
+ %\comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate} \comment[MLT]{I kind of disagree with the idea that it is straightforward that centralised cities produce more segregation as it is not a direct parameter of the model but an output.
 Plus: what argument would he or she have to state that a big center would be automatically less segregated that several smaller ones?
@@
 But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?} \comment[FK]{du coup je fais une proposition de réponse ci-dessous en reprenant ces différents éléments}
-  We understand this statements, however we wish to underline that our objective in the discussion of the Schelling results was not to establish a general result on the links between polycentricity and segregation, which are indeed very complex and multidimensional as advocated by the reviewer.
-Instead, we wished to show what kind of results can be obtained using the framework we suggest.
+  We understand this statement, however we wish to underline that our objective in the discussion of the Schelling results was not to establish a general result on the links between polycentricity and segregation, which are indeed very complex and multidimensional as advocated by the reviewer.
+Instead, we aimed at showing what kind of results could be obtained using the framework we suggest.
 If we do not prove that in general, and in empirical context, polycentricity creates more segregation, we obtained, in the modelling context, under Schelling assumptions, that polycentric geometric structures are more often that not associated with segregated patterns, a result which we do not believe to be solely resulting from the fact that more centers leads to more opportunities of segregation (which is true), because it seems also true that monocentric structures, with higher densities, can more easily accommodate heterogeneity of agents within cells, and hence a more patchy type of segregation which can for instance being observed in London.
@@
 
-  \item \comment[FL]{ne pas oublier les trois points ci-dessous!!!}
+  \comment[FL]{ne pas oublier les trois points ci-dessous!!!}
   \item in discussion, say that indeed the indicators can themselves be influenced by geometric structures (cf.
@@
 
-  $\rightarrow$ see above comment A, discussion has been amended and details added.
+  $\rightarrow$ See above comment A, discussion has been amended and details added.
   \medskip
@@
 
-  $\rightarrow$ see comment B
+  $\rightarrow$ See comment B above.
 
@@
 
-  $\rightarrow$ cf. comment B.3
+  $\rightarrow$ This point was not explicit enough and we revised the presentation of our approach (which is generic and not empirical) in introduction (especially fig.
+1), as already stated in comment B.3 above.
+We thank reviewer 2 for highlighting this limitation.
   % \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}
-  it as not explicit enough and we revised the presentation of our approach (which is generic and not empirical) in intro (especially fig. 1). We thank reviewer 2 for highlighting this limitation.
 
@@
 
-  $\rightarrow$ we added a paragraph on the limits of our density grid approach in discussion, in which we mention the possibility to use egohoods to better measure segregation.
+  $\rightarrow$ We added a paragraph on the limits of our density grid approach in discussion, in which we mention the possibility to use egohoods to better measure segregation.
 
@@
 
-  %Spatial econometrics based on empirical dataset is developing fast. It would be useful to understand how these models can enrich ABM and vice-versa. 
-   \item this is very interesting to inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010.
-We need to add a paragraph about this discussion in conclusion and how our approach allows to go back to questioning empirical cases from the model and questioning our concepts ('neighbourhood').
+   \item \textit{Spatial econometrics based on empirical dataset is developing fast. It would be useful to understand how these models can enrich ABM and vice-versa.}
 
-   $\rightarrow$  cf. comment B
+     $\rightarrow$ This points suggest that our work can inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010.
+We added a paragraph on this in the discussion discussion, and suggested how our approach allows to go back to questioning empirical cases from the model and questioning our concepts ('neighbourhood').
 
@@
 
-  $\rightarrow$ we have clarified this issue in introduction but referring to different terms and figure 1 which differentiates between the mu model and gama parameters. (see comment A)
+  $\rightarrow$ We have clarified this issue in introduction but referring to different terms and figure 1 which differentiates between the $\mu$ model and $\gamma$ parameters (see comment A).
 
@@
 
-  $\rightarrow$ cf. new discussion/conclusion and comment B
+  $\rightarrow$ The discussion and conclusion was amended in this direction (see comment B).
 

### 2019-08-12 paper.tex
.

--- 2019-08-12 paper.tex
@@
 An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily into our workflow.
-Grids are generated through an iterative process which, starting from an empty grid, adds a quantity $N$ (population) at each time step $t$, allocating it through preferential attachment on population density, characterised by its strength of attraction $\alpha$.
-More precisely, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
-At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process of strength $\beta$: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (i.e.
+Population density grids, at the typical scale of a metropolitan area, are generated by combining the opposite processes of urban dispersal (negative externalities) with urban concentration (positive externalities).
+More precisely, grids are generated through an iterative process which simulates successive time steps with a fixed population gain at each time step.
+Starting from an empty grid, the model adds a quantity $N$ (population) at each time step $t$.
+The new population is allocated through preferential attachment on previous population density.
+Formally, each added unit has a probability equal to $P_i^{\alpha}/\sum_k P_k^{\alpha}$ to be added to a patch $i$ with population $P_i$, all $N$ units being added independently and in parallel.
+The attachment parameter $\alpha$ can thus be interpreted as a ``strength of attraction'', in the sense that increasing it will lead to a higher instantaneous concentration.
+At the end of each time step, this growth process is smoothed $n_d$ times using a diffusion process: each patch transmits an equal share of $\beta\cdot P_i$ to its Moore neighborhood (i.e.
 its 8 surrounding patches).
+The parameter $\beta$ can be interpreted as a strength of diffusion: increasing it will lead to larger share of population being diffused in space.
 To avoid border effects such as a reflexion on the border of the world, border patches diffuse to the outside.
@@
 \centering
-	%\includegraphics[width=\textwidth]{figuresraw/spatialGen.png}
 	\includegraphics[width=\textwidth]{figures/Fig2.png}
@@
 
-% Generator parameters values:
-%  1) ex_sp-diffusion=0.0060_sp-growth-rate=25_sp-diffusion-steps=1_sp-alpha-localization=0.4_ticks=971_sp-population=24275.000000000102
-%  2) ex_sp-diffusion=0.0060_sp-growth-rate=25_sp-diffusion-steps=1_sp-alpha-localization=0.4_ticks=176_sp-population=4400.000000000003
-%  3) ex_sp-max-pop=80260_sp-diffusion=0.045_sp-growth-rate=102_sp-diffusion-steps=2_sp-alpha-localization=1.4_ticks=618
-%  4) ex_sp-max-pop=382170_sp-diffusion=0.114_sp-growth-rate=108_sp-diffusion-steps=1_sp-alpha-localization=1.8_ticks=227
+In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure: their concentration index, hierarchy index, centrality index and homogeneity index (cf.
+\cite{LeNechet2015}).
+These morphological dimensions are in practice captured using the rank-size slope to capture hierarchy, the average distance between individuals to capture concentration, distribution entropy to capture homogeneity, and Moran spatial autocorrelation index to measure centrality.
 
-In order to generate density grids which correspond to empirical density distributions, we select among the generated grids using an objective function which matches the point cloud of 110 metropolitan areas in Europe described by four dimensions of spatial structure: their concentration index, hierarchy index, centrality index and continuity index (cf.
-\cite{LeNechet2015}).
-We sample the generator parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a scatter with high discrepancy.
+We sample the generator parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a scatter with low discrepancy in high dimensions.
 We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}.

### 2019-08-12 paper.tex
Update on Overleaf.

@@
 %%%%%%%%%%%%%%%%%%%%%%
-
+%change1 ``generator parameters''
 The general workflow of our method is illustrated in Figure \ref{fig:method}.
-In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a spatial generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
-We call these parameters ``generator parameters'' to distinguish them from the standard parameters of the models and to highlight their higher position in the simulation process.
+In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a density grid generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
+We call these parameters \comment[FL]{change1} $\gamma$ parameters to distinguish them from the standard parameters of the models (called $\mu$ parameters) and to highlight their higher position in the simulation process.
 The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.
@@
 In particular, we want to emphasize that spatial effects derive not only from grid size or shape effects, but also from the heterogeneity of the spatial distribution of socio-spatial entities (people, housing, networks, etc.).
-In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling (generation of synthetic population for instance \citep{bhat1999activity}).
+In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling - generation of synthetic population for instance \citep{bhat1999activity}.
 %%%%%%%%%%%%%%%%%%%%%%
@@
 A solution is to use systematic quantitative procedures to compare them to a reference case.
-Technically, because of stochasticity, we represent the output of the model for a given combination of parameter values $\vec{P}$ as the mean of the final values of an output indicator $O$ obtained for the replications of the model initialized with $\vec{P}$.\\
-
-To our knowledge there exists no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.
-We introduce a measure of the relative distance $d_r$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$.
+Technically, because of stochasticity, we represent the output of the model for a given set of parameter values as the mean of the final values of an output indicator obtained for the replications of the model initialized with the set of parameter values \\
+%removed : $\vec{P}$  $O$ $\vec{P}$.
+To our knowledge there exists no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature\comment[FL]{signaler tout de même des tentatives on ne part pas de rien!}.
+We introduce a measure of the relative distance $d_r(\mu_{\vec{\alpha}_1},\mu_{\vec{\alpha}_2})$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$.
 Phase diagrams are denoted by the same function $\mu$ indexed by the generator parameters $\vec{\alpha}$, which capture the spatial configuration (in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself).
@@
 
-where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given generator parameters $\vec{\alpha}_i$.
+%where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given generator parameters $\vec{\alpha}_i$.
+% pour le coup cette précision me semble superflue
+\comment[FL]{dire plutôt pourquoi le choix de cette fonction qui n'a rien de trivial : d'où cela sort-il? Quels avantages de cette fonction mathématique? Comment est calculée la Variance?}
 $d$ is a functional distance that we take simply as Euclidean distance.
@@
 The ideas of multi-modelling and extensive model exploration are nothing from new - ~\cite{openshaw1983data} already advocated for ``model-crunching'' in 1983 -, but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities.
-The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write flexible exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
+The model exploration platform OpenMOLE~\citep{reuillon2013openmole}\comment[FL]{citer d'autres plateformes -StopPub et d'autre part : justifier en quoi c'est nécessaire?
+A mon sens : ici très utile, mais pas nécessairement utile si moins de paramètres ; mettre donc les questions de calculabilités au coeur de cette discussion} allows to embed any model as a blackbox, to write flexible exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
 In our case, this tool is a powerful way to embed both the sensitivity analysis and the sensitivity analysis to initial spatial conditions, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file.
@@
 
-The spatial generator applies an urban morphogenesis model \citep{batty2007cities} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}.
+The density grid generator applies an urban morphogenesis model \citep{batty2007cities} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}\comment[FL]{manque ici la référence à d'autres techniques : par exemple le modèle multimonocentrique d'Anas, certes non génératif, mais pourquoi serait-ce à exclure de cette approche?}.
 An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily into our workflow.
@@
 	\includegraphics[width=\textwidth]{figures/Fig2.png}
- \caption{\textbf{Four examples of grids produced by the spatial generator.} The lighter the red, the denser the area.
+ \caption{\textbf{Four examples of grids produced by the density grid generator.} The lighter the red, the denser the area.
 Changing the growth rate $N$ allows to have more or less chaotic shapes (two first compared to the two last grids for example) corresponding to different levels of convergence of the model, whereas local radius can be tuned with the interplay of aggregation strength $\alpha$ and diffusion strength $\beta$.
@@
 \cite{LeNechet2015}).
+These four dimensions were chosen as complementary descriptors of spatial organisation at urban or metropolitain level - see also \cite{tsai2005quantifying} and \cite{schwarz2010urban}.
+They account for : , the extend to which population is clustered in a central city, at macro scale (centrality) and micro scale (concentration), the extent to which density grid values are similar or contrasted, regardless of location, at macro scale (hierarchy) and micro scale (homogeneity)\comment[FL]{check with Juste quels indicateurs ont été utlisés}.
 These morphological dimensions are in practice captured using the rank-size slope to capture hierarchy, the average distance between individuals to capture concentration, distribution entropy to capture homogeneity, and Moran spatial autocorrelation index to measure centrality.
@@
 section \ref{sec:qualResults}).
-
+\comment[FL]{à discuter}
 The spatial generator and its resulting grids are relevant to the case study models we have picked (Schelling and Sugarscape) because it produces density grids at a ``metropolitan scale'', which is the scale at which both model were initially intended to be.
@@
 %%%%%%%%%%%%%%%%%%%%%%
+\comment[FL]{à discuter, rappeler sens des paramètres alpha, beta, etc. ; pourquoi appeler cela des "meta paramètres"? maintenant on appele cela des gamma parametres?}
 
@@
 
-Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point.
+Comparing phase diagrams is as we saw not straightforward, and further developments of our method imply testing alternative methods for this particular point.
 For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong phase transitions in the space of generator parameters.

### 2019-08-13 Response_to_Reviewers.tex
Update on Overleaf.

@@
 But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?} \comment[FK]{du coup je fais une proposition de réponse ci-dessous en reprenant ces différents éléments}
-  We understand this statement, however we wish to underline that our objective in the discussion of the Schelling results was not to establish a general result on the links between polycentricity and segregation, which are indeed very complex and multidimensional as advocated by the reviewer.
+
+  $\rightarrow$ We understand this statement, however we wish to underline that our objective in the discussion of the Schelling results was not to establish a general result on the links between polycentricity and segregation, which are indeed very complex and multidimensional as advocated by the reviewer.
 Instead, we aimed at showing what kind of results could be obtained using the framework we suggest.
 If we do not prove that in general, and in empirical context, polycentricity creates more segregation, we obtained, in the modelling context, under Schelling assumptions, that polycentric geometric structures are more often that not associated with segregated patterns, a result which we do not believe to be solely resulting from the fact that more centers leads to more opportunities of segregation (which is true), because it seems also true that monocentric structures, with higher densities, can more easily accommodate heterogeneity of agents within cells, and hence a more patchy type of segregation which can for instance being observed in London.
-We included a comment in the discussion section, saying that we hope we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?
-\comment[FL]{à la relecture de l'article s'assurer que cette prise de recul est bien explicitée}
+We included a comment in the discussion section, saying that we hope we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results \comment[FL]{à la relecture de l'article s'assurer que cette prise de recul est bien explicitée}
 

### 2019-08-13 paper.tex
Update on Overleaf.

@@
 Dynamically, sprawl in American cities consists in the addition of new developments which have been occupied by different groups of population, resulting in a concentration of the wealthy in suburban pockets and of pockets of poverty in the inner city area \citep{jargowsky2002sprawl}.
-Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}\comment[FL]{je ne comprends pas ce point : est-ce lié à un modèle de simulation?}.
+Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}.
 These effects of geographical space on social systems correspond to processes to be included in geosimulation models.\\

### 2019-08-13 Response_to_Reviewers.tex
.

--- 2019-08-12 Response_to_Reviewers.tex
@@
 
-  \item \textit{The problem, from a geographical point of view, is that the paper switches between “geographical” and “spatial” without making any differentiation. \ldots}
-  %In so doing, the author(s) runs immediately into the problem of adequate definitions.
+  \item \textit{The problem, from a geographical point of view, is that the paper switches between “geographical” and “spatial” without making any differentiation.
+In so doing, the author(s) runs immediately into the problem of adequate definitions.
 In paragraph 1.2 the definition of geographical systems as “social agents interacting with one another, within a limited portion of space” is misleading.
@@
 This approach can be (and I would add: has to be) criticised, however, not by interming!
-ling fuzzily the social with the geographical, as happened with this definition.
+ling fuzzily the social with the geographical, as happened with this definition.}
   %\item \comment[JR]{by spatial we indeed mean some kind of spatial container for agents right ?

### 2019-08-14 Response_to_Reviewers.tex
Update on Overleaf.

@@
 
-  $\rightarrow$ see comment A. cite Banos, 2012 \comment[FL]{quel était le point exactement avec Banos?}
+  $\rightarrow$ see comment A.
+As for the role of geometric structures in the production of segregation, Banos (2012) has shown that with the same Schelling mechanisms and the same parameter values but different network structures of neighbourhoods (scale-free, regular, etC.) the segregation observed at the end of simulation varied significantly between network structures.
+This is the kind of approach we systematise here, although with a spatial grid instead of network structures.
 
@@
   $\rightarrow$ We added a paragraph on the limits of our density grid approach in discussion, in which we mention the possibility to use egohoods to better measure segregation.
+Indeed, at present, the use of a grid of varying density means that in dense areas, agents have many neighbours and thus a higher probability of mixing, which affects segregation measurement.
+With egohoods, each agents would have the same number of neighbours taken into account for the calcultation, but their spatial scope would vary (for some agents, it would be for example 100 neighbours from the same cell whereas for others, it would be the nearest 100 neighbours taken from 25 surrounding cells).
 

### 2019-08-14 paper.tex
.
Update on Overleaf.

@@
 \author[6]{Florent Le N{\' e}chet}
-\affil[6]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilit{\'e} Transport, Marne-la-Vallée, France}
+\affil[6]{Universit{\'e} Paris-Est, Laboratoire Ville Mobilit{\'e} Transport, UMR-T 9403, Université Paris-Est Marne-la-Vallée, France}
 
@@
 \begin{abstract}
+\comment[FL]{le titre de l'article ce serait presque "geometry" matters? non?}
 Although simulation models of socio-spatial systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions.
@@
 Simulation also appears as a way to overcome the difficult analytic resolution of many socio-spatial models which were developed in the past, as well as to explore the possible (alternative) trajectories of path-dependent social and ecological systems.
-The specificity of geographical models compared to other social science models is that space and spatial interactions are given a prime role, geographers being driven by an explicit interest in studying the way "space" in broad sense influences the outcomes of social processes modelled.
-We think that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions, provided that models include relevant spatial descriptions and behavioural rules which take space into account, and provided that the model evaluation includes a sensitivity analysis of the model outputs to the way space is represented.
+The specificity of geographical models compared to other social science models is that space and spatial interactions are given a prime role, geographers being driven by an explicit interest in studying the way "space" in broad sense, and more specifically geometry or topology of space influences the outcomes of social processes modelled.
+We think that simulation approaches are uniquely positioned to represent the complexity of socio-spatial interactions, provided that models include relevant spatial descriptions and behavioural rules which take spatial proximity into account, and provided that the model evaluation includes a sensitivity analysis of the model outputs to the way space is represented.
 Unfortunately, the first condition is not always met, and the second is seldom even mentioned.
+Various contribution have been made on ad hoc models to test for the influence of geometry on simulation outputs \citep{lilburne2009sensitivity, sun2007traffic}, but without providing a generic framework fitted for most agent-based models.
 This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions.
-\comment[FL]{il faut ici citer des approches du même type}.
 To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\
@@
 For example, households can have different preferences with respect to the built-environment they might want to live in \citep{SpielmanHarrison2014}, or agents moving around will sense a different set of objects within the same fixed radius depending on the topology \citep{Banos2012} and distribution of density of the sensed environment \citep{LauriJaggi2003, FossettDietrich2009}.
-The way modellers represent initial space is therefore a central element of socio-spatial simulation models.
+The way modellers represent initial geometry of space is therefore a central element of socio-spatial simulation models.
 However, this step is rarely explicit.
-A meaningful way of representing initial space might be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour to different plausible spatial arrangements.
+A meaningful way to address it might be to consider, not necessarily the peculiarities of every city, but at least their broad density structures so as to estimate the variability of the model behaviour to different plausible spatial arrangements.
 
@@
 These four dimensions were chosen as complementary descriptors of spatial organisation at urban or metropolitain level - see also \cite{tsai2005quantifying} and \cite{schwarz2010urban}.
-They account for : , the extend to which population is clustered in a central city, at macro scale (centrality) and micro scale (concentration), the extent to which density grid values are similar or contrasted, regardless of location, at macro scale (hierarchy) and micro scale (homogeneity)\comment[FL]{check with Juste quels indicateurs ont été utlisés}.
-These morphological dimensions are in practice captured using the rank-size slope to capture hierarchy, the average distance between individuals to capture concentration, distribution entropy to capture homogeneity, and Moran spatial autocorrelation index to measure centrality.
+They account for : the extend to which population is clustered in a central city, with two complementary indicators : Moran index (called "centrality") and distance between individuals ("concentration), the extent to which density grid values are similar or contrasted, regardless of location, with two complementary indicators : entropy of distribution ("homogeneity") and rank-size slope ("hierarchy").
 
-We sample the generator parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a scatter with low discrepancy in high dimensions.
+We sample the $\gamma$ parameter space using a Latin Hypercube Sampling, which is a convenient technique to have a scatter with low discrepancy in high dimensions.
 We sample 2000 points in the 4-dimensional space of parameters {$\alpha$, $\beta$, $n_d$, $N$}.
@@
  \caption{Correspondence between European urban density structures and grids produced with the spatial generator.}
+ \comment[FL]{à discuter, rappeler sens des paramètres alpha, beta, etc.
+; pourquoi appeler cela des "meta paramètres"?
+maintenant on appele cela des gamma parametres?} \comment[JR]{+ correc nombre de chiffres significatifs}
 \label{fig:densityTypes}
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\comment[FL]{à discuter, rappeler sens des paramètres alpha, beta, etc. ; pourquoi appeler cela des "meta paramètres"? maintenant on appele cela des gamma parametres?}
 
@@
 
+\section{Conclusion}
+\comment[FL]{je déplace cette section, remettez là à la fin si vous pensez que c'était mieux}
+After reviewing the extensive literature on spatial biases in statistical and simulation models, we presented a method to analyse the sensitivity of a simulation's results to the initial spatial configuration.
+We did so by implementing a spatial generator whose output is used as input for the simulation model.
+We applied this approach to two textbook ABMs: Schelling and Sugarscape.
+With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
+With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space.
+We think that this method has the potential to increase the arsenal of evaluation of socio-spatial models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
-Although the applications we present here are limited by the simplicity of the models, we think that the method could (and should) be applied to larger models including domain mechanisms and more empirical initialisation data, for example synthetic populations.
+We wish to underline the interest of adding sensitivity analysis of the geometry of space during the exploration of models.
+As we have observed for simple agent-based models, uncertainties in the initial distribution of space on which agents will interact play a role on the variability in model outputs.
+This would certainly be even more important for more sophisticated model, with more type of agents and type of interactions, insomuch as agent-based model tend to be used in realistic, diverse settings.
+Therefore, we believe that for many applied problems, studying the specific variability of model outputs regarding the geometry of space would be important to assess the transferability of results in other urban settings.
+
+This is especially important in the context of the increasing recognition of the complexity of urban space and of the role cities plays in various aspects of sustainability : modelling cities with agent-based models might become more frequent \citep{perez2016agent} and we argue it is important to account for the diversity of urban configurations when disseminating the results of any model in the planning community.
+
+we think that the method could (and should) be applied to larger models including domain mechanisms and more empirical initialisation data, for example synthetic populations.
 The sensitivity analysis to initial spatial conditions could then be either a replication on the spatial allocation of the synthetic population, or a series of spatial permutations of the empirical spatial inputs.
@@
 This relates to relations between empirical, conceptual and modeling dimensions of quantitative research in social sciences, as coined by \cite{livet2010ontology}.
-In our case, a contribution in the modeling domain aiming at extracting further knowledge on model behavior (sensitivity to geometrical context), naturally leads to questioning theoretical concepts and empirical definition on which the model was based, such as for example the meaning of neighborhood.
+In our case, a contribution in the modeling domain aiming at extracting further knowledge on model behavior (sensitivity to geometrical context), may lead to questioning theoretical concepts and empirical definition on which the model was based, such as for example the meaning of neighborhood, hence geometry of space!
+
 % to inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, we added a paragraph in conclusion, discussing how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how ’neighbourhood’ is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences)
-
-\section{Conclusion}
-
-After reviewing the extensive literature on spatial biases in statistical and simulation models, we presented a method to analyse the sensitivity of a simulation's results to the initial spatial configuration.
-We did so by implementing a spatial generator whose output is used as input for the simulation model.
-We applied this approach to two textbook ABMs: Schelling and Sugarscape.
-With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
-With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space.
-We think that this method has the potential to increase the arsenal of evaluation of socio-spatial models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
 

### 2019-08-14 Response_to_Reviewers.tex
Update on Overleaf.

@@
   \item B.1.
-We have added a discussion of\comment[FL]{on?} how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities).
+We have added a discussion on how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities).
 This is a limitation of the method as it is but also an avenue for promoting stronger indicators in empirical research.
@@
 And I believe that explaining why polycentric cities are more segregated than others is out of the scope of the paper.
-But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?} \comment[FK]{du coup je fais une proposition de réponse ci-dessous en reprenant ces différents éléments}
+But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?} \comment[FL]{du coup je fais une proposition de réponse ci-dessous en reprenant ces différents éléments}
 

### 2019-08-14 Response_to_Reviewers.tex
rev_ok

@@
 Our original contribution is to tackle the second type.
-%\comment[CC]{done in 1.5 and 1.8}
 
@@
 We have clarified the terms used in our paper introduction to distinguish between: \textbf{sociospatial systems} (i.e.
-groups of social agents whose behaviour is constrained by their position in geographical space)%.
-This term refers to empirical systems, which are the targets of the modelling process)% Mise en commentaire par FL 190812 : je trouve que cela alourdi inutilement la phrase
-, \textbf{interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and \textbf{geometric structure}, a term we use for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model.
+groups of social agents whose behaviour is constrained by their position in geographical space), \textbf{interactions} (what happens between agents and between agents and their environment, in the model as well as in the empirical system) and \textbf{geometric structure}, a term we use for qualifying the geometric outputs of the spatial generator which are used as initial spatial conditions in the model.
 When we talk about the generator we actually used in the paper, we call it a "density grid generator" because it generates a square grid with heterogeneous density, whereas in all generality a "spatial generator" could potentially generate graph or point structures.
-%\comment[CC]{done in 1.5} %Changement par FL 190812 Because this is the only geometric structure it generates
 
@@
   \item B.1.
-We have added a discussion on how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities).
+We have added a discussion of on how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities).
 This is a limitation of the method as it is but also an avenue for promoting stronger indicators in empirical research.
-% \comment[MLT]{j'ai formulé un paragraphe court sur ce point en discussion, en lien avec le point B.3}
 
@@
 To inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, we added a paragraph in discussion, explaining how our approach allows us to go back to questioning empirical cases from the model and questioning our concepts (for example how 'neighbourhood' is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences).
-\comment{il faut reprendre ces elements dans l'article}
 
   \item B.3.
-As a perspective to reduce this bias, we could compute ego-neighbourhoods rather than a regular grid (another paper)\comment[FL]{pas clair : pour calculer les indicateurs?
-pour les interactions?
-à préciser comment sont utilisés ces ego neighborhoods}.
-We could also\comment[FL]{plutôt : In further researches, we plan to extend the analysis...} in another paper vary the size of grid and its cell shape (cf.
+As a perspective to reduce this bias, we could compute ego-neighbourhoods rather than a regular grid, and make these systematically vary and study the effect on emergent behavior and indicators.
+In further researches, we plan to extend the analysis to vary the size of grid and its cell shape (cf.
 MAUP), but also generate other types of geometric structures to feed into geosimulation models (for example points or networks).
-% \comment{j'ai formulé un paragraphe court sur ce point en discussion, en lien avec le point B.1}
 \end{itemize}
@@
 
-%\textbf{to do list:}
-
- % \textbf{Sociospatial system} pour parler des systèmes empiriques, \textbf{interactions} pour parler de ce qui se passe dans le modele,  \textbf{geometric structures} pour parler des outputs issus du générateur et  parler de \textbf{density grid generator} quand on développe le nôtre.
-\comment{done}
-
- % caser" Jessop (2008) points in particular to the interdependence between social phenomena and the spatial environment where they take place.
-We model the social system as the interaction dynamics of agents and spatial units.
-The spatial generator produces the geometric support for these interactions." in intro 1.4 \\ \comment[MLT]{To avoid redundancies, I have simply added the reference to Jessop when we describe Figure 1.}
-
-%We now turn to point-by-point responses to referees comments.
 \medskip
@@
 
-   %\comment[JR]{most of this empirical literature was added because of one CEUS reviewer, no ?}\comment[MLT]{If we want to keep this part to please everyone (like our former reviewer), we could explicitly mention that we model the growth of a city (in terms of population density) which is the results of a bunch of dynamics ...
-and not the dynamics themselves?}
-  %\comment[FLN]{ qui montre que la démarche méta est ici appliquée à un domaine empirique qu'il aide à explorer mais d'autres domaines peuvent s'emparer de la démarche méta.}
-
-   % show disagreement with omitting empirical/geographical but demonstrate helps understand geographical question but the is generic and valuable for it.
-
 \medskip
@@
 ling fuzzily the social with the geographical, as happened with this definition.}
-  %\item \comment[JR]{by spatial we indeed mean some kind of spatial container for agents right ?
-which is far from geographical in the sense he puts it indeed}\comment[MLT]{use sociospatial system instead?
-The term is broadly used in the literature to describe the interdependence between social phenomena and their spatial environment.
-We could also quote \hyperlink{https://journals.sagepub.com/doi/abs/10.1068/d9107}{Jessop et al.
-2008}}
 
@@
  \medskip
-
-  %\item reprendre tout le papier pour éliminer les termes indéfinis "space", "pattern" etc.
-+ en conclu mentionner que ça s'applique aux automates cellulaires (sans agents) et aux modèles écologiques (agents-arbres).
-  %\item in conclusion, we need to clarify the answer to the question "do geometric structure influence sociospatial systems dynamics" by saying it depends on the model.
-≠ black\&white and we can demonstrate it.
-The nature of social systems is a given (in the model mu, which is not the contribution of this paper) and we explore the consequence of geometric structures on the spatial expression of social systems.
 
@@
 
-  \textit{The problem of setting space and geography as synonymous continues in paragraphs 1.9 and following.
+  \item \textit{The problem of setting space and geography as synonymous continues in paragraphs 1.9 and following.
 What is described here as "spatial system shape" is the geometry of spaces but not the geography of spaces.}
-  % -> yes indeed
 
@@
 
-  %it's not the point of the paper, only the metaphore of the spatial generator.
-+ supprimer la ref à "urban growth" dans le texte et dire d'inspiration urban science mais dont seules les grilles finales nous intéressent.
-On aurait pu utiliser des generator issus de graphes, de percolation etc.
-  %The core of the paper is not this, it's the approach of including gamma and applying to mu to understand systems better, whatever gamma and whatever mu in whatever knowledge domain.
-  %\item in discussion, mention that it's one isolated example of geometry but other cases could be thought of as well, using networks and graphs, but also empirical shape files.
-
   \medskip
@@
 }
-  %\comment[FL]{j'ai pas compris pourquoi la phrase du reviewer avait été mise en commentaire, j'ai remis}
- %\comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate} \comment[MLT]{I kind of disagree with the idea that it is straightforward that centralised cities produce more segregation as it is not a direct parameter of the model but an output.
-Plus: what argument would he or she have to state that a big center would be automatically less segregated that several smaller ones?
-And I believe that explaining why polycentric cities are more segregated than others is out of the scope of the paper.
-But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?} \comment[FL]{du coup je fais une proposition de réponse ci-dessous en reprenant ces différents éléments}
 
@@
 If we do not prove that in general, and in empirical context, polycentricity creates more segregation, we obtained, in the modelling context, under Schelling assumptions, that polycentric geometric structures are more often that not associated with segregated patterns, a result which we do not believe to be solely resulting from the fact that more centers leads to more opportunities of segregation (which is true), because it seems also true that monocentric structures, with higher densities, can more easily accommodate heterogeneity of agents within cells, and hence a more patchy type of segregation which can for instance being observed in London.
-We included a comment in the discussion section, saying that we hope we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results \comment[FL]{à la relecture de l'article s'assurer que cette prise de recul est bien explicitée}
+We included a comment in the discussion section, saying that we hope we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results.
 
-  \comment[FL]{ne pas oublier les trois points ci-dessous!!!}
-  \item in discussion, say that indeed the indicators can themselves be influenced by geometric structures (cf.
+ % \comment{in discussion, say that indeed the indicators can themselves be influenced by geometric structures (cf.
 comment B) but on the processes, there is nothing obvious about the influence of structures on behavioural spatial patterns (if you imagine a model with no interactions between agents for example) + it's a difference between the model and its potential addition to the real systems.
 cf.
-B
+B}
 
-  \item rewrite 1.11 \& 1.12 which is misleading
+	\medskip
 
-   % Formal issues that have to be solved: the numbering of paragraphs is not coherent. The reference of Wheeler (2006) is missing the source of the article.
-    \item check jasss style for paragraph numbering (Tous)
+   \item \textit{Formal issues that have to be solved: the numbering of paragraphs is not coherent. The reference of Wheeler (2006) is missing the source of the article.}
+
+   $\rightarrow$ These issues were corrected.
 

### 2019-08-14 paper.tex
rev_ok

@@
 
+%\comment[FL]{le titre de l'article ce serait presque "geometry" matters? non?} bof ca parle moins
+
 \begin{abstract}
-\comment[FL]{le titre de l'article ce serait presque "geometry" matters? non?}
 Although simulation models of socio-spatial systems in general and agent-based models in particular represent a fantastic opportunity to explore socio-spatial behaviours and to test a variety of scenarios for public policy, the validity of generative models is uncertain unless their results are proven robust and representative of 'real-world' conditions.
@@
 We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models.
-We wish to illustrate the potential interest of adding spatial sensitivity analysis during the exploration of models for both modellers and thematic specialists.%Furthermore, this effect is sometimes larger than the effect of parameters' value change.%Comment FL : j'ai supprimé cette dernière phrase qui ne me semblait pas essentielle et rajouté à la place une phrase d'ouverture de Marion.
+We wish to illustrate the potential interest of adding spatial sensitivity analysis during the exploration of models for both modellers and thematic specialists.
 \end{abstract}
@@
 This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions.
-To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.\\
+To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.
 
 \subsection{Definition of the problem}
-%change1 : within a limited portion of space
-%change2 : irreversibly, creating temporal and 
-Socio-spatial systems can be crudely described as social agents interacting with one another via the geometric structure of space \comment[FL]{change2}.
-Social agents thus constitute the microscopic level of the system, and they are framed by a system-time\comment[FL]{manque un mot} that evolves\comment[FL]{change2} with potential cumulative effects, also known as path-dependency \citep{arthur1994increasing}.
+
+Socio-spatial systems can be crudely described as social agents interacting with one another via the geometric structure of space.
+Social agents thus constitute the microscopic level of the system, and they are contained within a spatio-temporal structure that evolves with potential cumulative effects, also known as path-dependency \citep{arthur1994increasing}.
 Therefore, observing one system at different points in time does not equate to observing different systems at a single point in time.
-This general property of ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}.
+This general property of non-ergodicity applies to geographical elements such as road networks or built-up areas \citep{pumain2003approche}.
 Similarly to what \citet{gell1995quark} calls \emph{frozen accidents} in complex systems generally, a given configuration contains clues about past bifurcations, that can have had dramatic effects on the state of the system.
 Therefore, strong spatio-temporal path-dependencies in the trajectory of individual territories and changing social environments over time prohibits the use of ergodic models.
-Ironically, these very models tend to be the models most frequently used in geosimulation.\comment[FL]{on ne revient pas du tout là dessus à la fin : est ce que les cas d'étude qu'on mobilise sont si path dependant que ca?
-Si c'est pertinent : revenir dessus.
-Sinon, enlever cette phrase}
+Ironically, these very models tend to be the models most frequently used in geosimulation.
+With these kind of models, the influence of the geometric structure of space will be even more important than in the case without path-dependency.
 
@@
 In that sense, the impact of initial \emph{spatial} settings on final outcomes is assumed to be significant just as any other initial conditions, but of more interest to the geographer.
-\\
 
@@
 \subsection{Objective}
-
-%In this paper, we offer a methodological solution to the problem of sensitivity of simulation models to initial spatial conditions with two application cases (Schelling and Sugarscape).
 
@@
 %%%%%%%%%%%%%%%%%%%%%%
-\subsection{The effects of the space in simulation models}
+\section{The effects of the space in simulation models}
 
-\subsubsection{Spatial processes}
+\subsection{Spatial processes}
 In 'real-life', different spatial configurations of people in a city tend to be associated with different distributions of income, carbon emissions, educational outcomes, etc.
@@
 Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}.
-These effects of geographical space on social systems correspond to processes to be included in geosimulation models.\\
+These effects of geographical space on social systems correspond to processes to be included in geosimulation models.
 
-\subsubsection{Spatial representation}
+\subsection{Spatial representation}
 A discussion of the effects of spatial encoding and representation has also been associated with the field of geostatistics since the exposure of the Modifiable Areal Unit Problem (MAUP) \citep{Openshaw1984, FotheringhamWong1991}.
@@
 
-\subsubsection{Spatial structures}
+\subsection{Spatial structures}
 We can distinguish a last category of spatial effects in geosimulation, which are the geometric constraints of the environment modelled at initialisation on the course and output of the simulation.
@@
 %%%%%%%%%%%%%%%%%%%%%%
-%change1 ``generator parameters''
+
 The general workflow of our method is illustrated in Figure \ref{fig:method}.
 In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a density grid generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
-We call these parameters \comment[FL]{change1} $\gamma$ parameters to distinguish them from the standard parameters of the models (called $\mu$ parameters) and to highlight their higher position in the simulation process.
+We call these parameters $\gamma$ parameters to distinguish them from the standard parameters of the models (called $\mu$ parameters) and to highlight their higher position in the simulation process.
 The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.
@@
 In the models we used as examples, the initial spatial configurations can be either flat or heterogeneous, monocentric or polycentric, based on external databases and on internal modelling - generation of synthetic population for instance \citep{bhat1999activity}.
+
 %%%%%%%%%%%%%%%%%%%%%%
@@
 \centering
-%\includegraphics[width=0.9\textwidth]{figuresraw/SchemaMeta_1.png}
 \includegraphics[width=0.9\textwidth]{figures/Fig1_v2.png}
@@
 A solution is to use systematic quantitative procedures to compare them to a reference case.
-Technically, because of stochasticity, we represent the output of the model for a given set of parameter values as the mean of the final values of an output indicator obtained for the replications of the model initialized with the set of parameter values \\
-%removed : $\vec{P}$  $O$ $\vec{P}$.
-To our knowledge there exists no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature\comment[FL]{signaler tout de même des tentatives on ne part pas de rien!}.
-We introduce a measure of the relative distance $d_r(\mu_{\vec{\alpha}_1},\mu_{\vec{\alpha}_2})$ between two phase diagrams $\mu_{\vec{\alpha}_1}$ and $\mu_{\vec{\alpha}_2}$.
-Phase diagrams are denoted by the same function $\mu$ indexed by the generator parameters $\vec{\alpha}$, which capture the spatial configuration (in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself).
-The $d_r$ measure is taken as the share of between-diagrams variability relative to their internal variability, given formally in the case of a one-dimensional phase diagram by
-
+Technically, because of stochasticity, we represent the output of the model for a given set of parameter values as the mean of the final values of an output indicator obtained for the replications of the model initialized with the set of parameter values.
+To our knowledge there exists no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature (see discussion below).
+We introduce a measure of the relative distance $d_r(\mu_{\vec{\gamma}_1},\mu_{\vec{\gamma}_2})$ between two phase diagrams $\mu_{\vec{\gamma}_1}$ and $\mu_{\vec{\gamma}_2}$.
+Phase diagrams are denoted by the same function $\mu$ indexed by the generator parameters $\vec{\gamma}$, which capture the spatial configuration (in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself).
+We choose to compare the variability of each phase diagram to the variability between them.
+We take therefore a basic adimentional ratio measure, given formally in the case of a one-dimensional phase diagram by
 \begin{equation}\label{eq:phase-distance}
-d_r\left(\mu_{\vec{\alpha}_1},\mu_{\vec{\alpha}_2}\right) = 2 \cdot \frac{d(\mu_{\vec{\alpha}_1},\mu_{\vec{\alpha}_2})^2}{Var\left[\mu_{\vec{\alpha}_1}\right] + Var\left[\mu_{\vec{\alpha}_2}\right]}
+d_r\left(\mu_{\vec{\gamma}_1},\mu_{\vec{\gamma}_2}\right) = 2 \cdot \frac{d(\mu_{\vec{\gamma}_1},\mu_{\vec{\gamma}_2})^2}{Var\left[\mu_{\vec{\gamma}_1}\right] + Var\left[\mu_{\vec{\gamma}_2}\right]}
 \end{equation}
-
-%where $\mu_{\vec{\alpha}_i}$ are the phase diagrams at given generator parameters $\vec{\alpha}_i$.
-% pour le coup cette précision me semble superflue
-\comment[FL]{dire plutôt pourquoi le choix de cette fonction qui n'a rien de trivial : d'où cela sort-il? Quels avantages de cette fonction mathématique? Comment est calculée la Variance?}
-$d$ is a functional distance that we take simply as Euclidean distance.
-The internal variabilities are estimated as the variance within each phase diagram $\mu_{\vec{\alpha_i}}$.
+where $d$ is a functional distance.
+We test in Appendix D different distances and show that results are qualitatively robust to this choice.
+We thus discuss results with the Euclidean distance in the following.
+The internal variabilities are estimated as the variance within each phase diagram $\mu_{\vec{\gamma_i}}$ (in practice computed with the unbiased variance estimator).
 For a multi-dimensional phase diagram, we average these relative distances over the components.
@@
 The last methodological point which we need to emphasize is the relationship between the present workflow and model exploration workflows in general.
-The ideas of multi-modelling and extensive model exploration are nothing from new - ~\cite{openshaw1983data} already advocated for ``model-crunching'' in 1983 -, but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities.
-The model exploration platform OpenMOLE~\citep{reuillon2013openmole}\comment[FL]{citer d'autres plateformes -StopPub et d'autre part : justifier en quoi c'est nécessaire?
-A mon sens : ici très utile, mais pas nécessairement utile si moins de paramètres ; mettre donc les questions de calculabilités au coeur de cette discussion} allows to embed any model as a blackbox, to write flexible exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
+The ideas of multi-modelling and extensive model exploration are nothing from new - \cite{openshaw1983data} already advocated for ``model-crunching'' in 1983 -, but their effective use only begins to emerge thanks to the development of new methods and tools together with an explosion of computation capabilities.
+The model exploration platform OpenMOLE~\citep{reuillon2013openmole} allows to embed any model as a blackbox, to write flexible exploration workflow using advanced methodologies such as genetic algorithms and to distribute transparently the computation on large scale computation infrastructures such as clusters or computation grids.
+While tools and platforms providing similar functionalities exist, such as for example Behavior Space of NetLogo~\citep{tisue2004netlogo} for model exploration, \cite{bakker2016scripting} for interactive model development, or interfaces to access High Performance Computing services~\citep{vecchiola2009high}, none provide the three aspects simultaneously in an integrated manner.
 In our case, this tool is a powerful way to embed both the sensitivity analysis and the sensitivity analysis to initial spatial conditions, and to allow the coupling of any spatial generator with any model in a straightforward way as long as the model can take its spatial configuration as an input or from an input file.
@@
 We use therefore OpenMOLE's functionalities for model embedding through workflow, design of experiments (parameter sampling) and high performance environment access.
+As our method quickly increases the amount of computation needed (we ran models approximatively $7\cdot 10^6$ times with a total computation time of around 2 years equivalent CPU), the use of OpenMOLE was crucial in our work.
 
@@
 
-The density grid generator applies an urban morphogenesis model \citep{batty2007cities} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}\comment[FL]{manque ici la référence à d'autres techniques : par exemple le modèle multimonocentrique d'Anas, certes non génératif, mais pourquoi serait-ce à exclure de cette approche?}.
+The density grid generator applies an urban morphogenesis model \citep{batty2007cities} which has been generalised, explored and calibrated by~\citet{raimbault2018calibration}.
+To generate population density distributions, other models such as other morphogenesis models~\citep{rybski2013distance}, kernel mixtures~\citep{anas1998urban} or more operational cellular automaton models of urban growth~\citep{herold2003spatiotemporal} could be used, as our general method is proposed to be independent of the generators and models chosen.
+The model of \citet{raimbault2018calibration} has the advantage of producing a broad range of existing urban forms with reasonable level of complexity.
 An open implementation and a characterisation of the urban forms which the model can produce allow us to integrate it easily into our workflow.
@@
 Grids are thus generated from the combination of the values of these four generator parameters $\alpha$, $\beta$, $n_d$ and $N$, in addition to the random seed.
-To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment.
+To ease our exploration, only the distribution of density is allowed to vary rather than the size of the grid, which we fix to a 50x50 square environment (this size provides a good compromise between accuracy of the model to reproduce forms and computational complexity, and furthermore corresponds to the order of magnitude of raster grids for metropolitan areas).
 We furthermore fix the total population at $t_f\cdot N = 100,000$, and determine therein the number of steps needed at a given $N$.
@@
 This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results.
-We select 15 grids of each type to work with in the computation of sensitivity analysis of an intra-urban model (cf.
-section \ref{sec:qualResults}).
-\comment[FL]{à discuter}
+We select 15 grids of each type to capture possible variabilities within and between different types of grid.
 The spatial generator and its resulting grids are relevant to the case study models we have picked (Schelling and Sugarscape) because it produces density grids at a ``metropolitan scale'', which is the scale at which both model were initially intended to be.
@@
 \centering
-	%\includegraphics[width=0.9\textwidth]{figuresraw/GridPresentation.png}
-	\includegraphics[width=0.9\textwidth]{figures/Fig3.png}
- \caption{Correspondence between European urban density structures and grids produced with the spatial generator.}
- \comment[FL]{à discuter, rappeler sens des paramètres alpha, beta, etc.
-; pourquoi appeler cela des "meta paramètres"?
-maintenant on appele cela des gamma parametres?} \comment[JR]{+ correc nombre de chiffres significatifs}
+	\includegraphics[width=0.9\textwidth]{figures/Fig3_v2.png}
+ \caption{Correspondence between European urban density structures and grids produced with the spatial generator.
+We also give for each the generator parameters leading to the density grid (we recall that $\alpha$ is a level of aggregation, $\beta$ a level of diffusion, $n$ the number of diffusions per time steps, and $N$ the population growth rate at each time step.}
 \label{fig:densityTypes}
@@
 In addition, we are interested in testing the impact of the initial spatial distribution of housing capacity in this project, using the generated grids.
-\\
+
 The outcome of the model is measured by a combination of three segregation indices: Dissimilarity, Moran's I and Entropy.
@@
 Entropy can be used as an indicator of spatial segregation because it measures the evenness of distribution of groups in space, although the metaphor of thermodynamics is not straightforward \citep{barner2017multiscale}.
-In our case, areas with higher values of entropy are considered more segregated because the mix of groups is uneven across small areas, whereas low entropy denotes more uniform distributions.\\
+In our case, areas with higher values of entropy are considered more segregated because the mix of groups is uneven across small areas, whereas low entropy denotes more uniform distributions.
 
 We use a ad-hoc implementation of the Schelling model, both in Scala for performance reasons and in NetLogo to ensure visualization of model dynamics.
-The pseudo-code of the implemented model is available in~\ref{app:code} and source code are available on the repository of the project.\\
-
+The pseudo-code of the implemented model is available in Appendix E and source code are available on the repository of the project.
 In general, the implementations of Schelling models allow only one agent per cell, and their initial distribution is random, therefore following a uniform distribution across the modelled city.
@@
 \subsubsection{Sugarscape}
+
 Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \cite{EpsteinAxtell1996}.
@@
 \subsection{Experiment design}
+
 For Sugarscape, we explore three dimensions of the parameter space: the total population of agents $P\in \left[10;510\right]$, the minimal initial agent resource $s_{-}\in \left[10;100\right]$ and the maximal initial agent resource $s_{+}\in \left[110;200\right]$.
@@
 
-As detailed in~\ref{app:convergence}, more repetitions are needed for Schelling indicators than for Sugarscape, in order to obtain a similar relative confidence in the estimation of averages.
+As detailed in Appendix B, more repetitions are needed for Schelling indicators than for Sugarscape, in order to obtain a similar relative confidence in the estimation of averages.
 We ran for this reason a different number of replications for each model.
-
 We choose different experiment designs, both for generator parameters and for the phase diagram, to demonstrate the robustness of the method to technical choices.
@@
 We modified the Sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model.
-Both pseudo-codes are available in~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
+Both pseudo-codes are available in Appendix E, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}.
 Density grids are also available at this address.
@@
 We tested the sensitivity to the type of distance, using a Minkovski distance with a varying exponent.
-The results are presented in~\ref{app:distances}, and show a similar sensitivity to geometric structures.
+The results are presented in Appendix D, and show a similar sensitivity to geometric structures.
 
@@
 The points are projected with respect to the generator parameters, and coloured according to the relative distance of the phase diagram of the simulations using this grid to the phase diagram of the reference case.
-Therefore, the figure ~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e.
+Therefore, Fig.~\ref{fig:sugarscape-distance-meta} shows that the grids generated with a high $\alpha$ (i.e.
 with a small number of very high density cells) produce simulation results that vary more between the reference case and the generated grid with the same values of parameters than within the reference case because of parameter variations.
@@
 \centering
-	%\includegraphics[width=0.8\textwidth]{figuresraw/relativedistance_metaparams.png}
 	\includegraphics[width=0.8\textwidth]{figures/Fig4.png}
@@
 The first two components represent 92\% of cumulated variance.
-The first component defines a ``level of sprawl'' and of scattering, whereas the second one represents the level aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation (top left of figure \ref{fig:sugarscape-distance-pca}).
+The first component defines a ``level of sprawl'' and of scattering, whereas the second one represents the level aggregation.\footnote{We have $PC1 = 0.76\cdot distance + 0.60\cdot entropy + 0.03\cdot moran + 0.24\cdot slope$ and $PC2 = -0.26\cdot distance + 0.18\cdot entropy + 0.91\cdot moran + 0.26\cdot slope$.} We find that grids producing the highest deviations are the ones with a low level of sprawl and a high aggregation (top left of Fig.~\ref{fig:sugarscape-distance-pca}).
 It is confirmed by the behaviour as a function of generator parameters, as high values of $\alpha$ also yield high distance.
@@
 \centering
-	%\includegraphics[width=0.8\textwidth]{figuresraw/relativedistance_morphspace.png}
 	\includegraphics[width=0.8\textwidth]{figures/Fig5.png}
@@
 The objective of this quantitative experiment is to evaluate to which extent this phase diagram is modified when different density grids are applied.
-We show in Fig.\ref{fig:schelling-distance-meta} (Supplementary Material) the values of the relative distance as a function of meta-parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape.
+We show in Fig.\ref{fig:schelling-distance-meta} (Supplementary Material) the values of the relative distance as a function of generator parameters and in the reduced morphological space, in a way similar to the analyse done with Sugarscape.
 Variations are less considerable than for Sugarscape across phase diagrams, but values close to 1 show that several configurations are as sensitive to initial spatial conditions than to their parameters.
@@
 
-\section{Conclusion}
-\comment[FL]{je déplace cette section, remettez là à la fin si vous pensez que c'était mieux}
-After reviewing the extensive literature on spatial biases in statistical and simulation models, we presented a method to analyse the sensitivity of a simulation's results to the initial spatial configuration.
-We did so by implementing a spatial generator whose output is used as input for the simulation model.
-We applied this approach to two textbook ABMs: Schelling and Sugarscape.
-With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
-With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space.
-We think that this method has the potential to increase the arsenal of evaluation of socio-spatial models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
-
 %%%%%%%%%%%%%%%%%%%%%%
@@
 This would certainly be even more important for more sophisticated model, with more type of agents and type of interactions, insomuch as agent-based model tend to be used in realistic, diverse settings.
+Indeed, regarding path-dependency which was part of issues motivating our approach, the models we studied are not path-dependent for the aggregated indicators (while the final agent configuration naturally is).
+We postulate that intrinsically non-ergodic models such as the one studied by~\cite{coupe2017modeliser} should be more sensitive to geometry.
 Therefore, we believe that for many applied problems, studying the specific variability of model outputs regarding the geometry of space would be important to assess the transferability of results in other urban settings.
@@
 
-we think that the method could (and should) be applied to larger models including domain mechanisms and more empirical initialisation data, for example synthetic populations.
+We think that the method could (and should) be applied to larger models including domain mechanisms and more empirical initialisation data, for example synthetic populations.
 The sensitivity analysis to initial spatial conditions could then be either a replication on the spatial allocation of the synthetic population, or a series of spatial permutations of the empirical spatial inputs.
@@
 This relates to relations between empirical, conceptual and modeling dimensions of quantitative research in social sciences, as coined by \cite{livet2010ontology}.
-In our case, a contribution in the modeling domain aiming at extracting further knowledge on model behavior (sensitivity to geometrical context), may lead to questioning theoretical concepts and empirical definition on which the model was based, such as for example the meaning of neighborhood, hence geometry of space!
+In our case, a contribution in the modeling domain aiming at extracting further knowledge on model behavior (sensitivity to geometrical context), may lead to questioning theoretical concepts and empirical definition on which the model was based, such as for example the meaning of neighborhood, hence geometry of space.
 
 % to inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, we added a paragraph in conclusion, discussing how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how ’neighbourhood’ is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences)
+
+\section{Conclusion}
+
+%\comment[FL]{je déplace cette section, remettez là à la fin si vous pensez que c'était mieux} % c'est une conclusion generale, ca parait mieux a la fin non ?
+
+After reviewing the extensive literature on spatial biases in statistical and simulation models, we presented a method to analyse the sensitivity of a simulation's results to the initial spatial configuration.
+We did so by implementing a spatial generator whose output is used as input for the simulation model.
+We applied this approach to two textbook ABMs: Schelling and Sugarscape.
+With the Schelling experiment, we found that the different urban morphologies impact the interaction patterns, and that polycentric and discontinuous cities appear systematically more segregated than compact cities in terms of dissimilarity and entropy index.
+With Sugarscape, we show that the model is more sensitive to space than to its other parameters in the reference Netlogo implementation, both qualitatively and quantitatively: the amplitude of variations across density grids is larger than the amplitude in each phase diagram, and the behaviour of the phase diagram is qualitatively different in different regions of the morphological space.
+We think that this method has the potential to increase the arsenal of evaluation of socio-spatial models, in order to assess the sensitivity of models to their initial spatial conditions but also to learn about the impact of the urban form on social mechanisms.
 
@@
 \centering
-%\includegraphics[width=0.48\textwidth]{figuresraw/schelling-relativedistance_metaparams_mean.png}
-%\includegraphics[width=0.48\textwidth]{figuresraw/schelling-relativedistance_morphspace_mean.png}
 \includegraphics[width=\textwidth]{figures/Fig8.png}

### 2019-08-14 Response_to_Reviewers.tex

@@
   \item B.1.
-We have added a discussion of on how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities).
+We have added a discussion on how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities).
 This is a limitation of the method as it is but also an avenue for promoting stronger indicators in empirical research.
+% \comment[MLT]{j'ai formulé un paragraphe court sur ce point en discussion, en lien avec le point B.3}
 
@@
 
+  %\comment[FL]{j'ai pas compris pourquoi la phrase du reviewer avait été mise en commentaire, j'ai remis}
+ %\comment[JR]{I rather agree that it is quite reductionist, and that the discussion may be valid only at very abstract levels (on which we were not explicit maybe ?) - maybe make these conclusions more moderate} \comment[MLT]{I kind of disagree with the idea that it is straightforward that centralised cities produce more segregation as it is not a direct parameter of the model but an output.
+Plus: what argument would he or she have to state that a big center would be automatically less segregated that several smaller ones?
+And I believe that explaining why polycentric cities are more segregated than others is out of the scope of the paper.
+But we could add his or her comment in the discussion section, saying that we hove we have convinced people to add spatial sensitivity analysis during the exploration of their model and that specialist of the thematics would see the opportunity to dig deeper into our results?} \comment[FL]{du coup je fais une proposition de réponse ci-dessous en reprenant ces différents éléments}
+
   $\rightarrow$ We understand this statement, however we wish to underline that our objective in the discussion of the Schelling results was not to establish a general result on the links between polycentricity and segregation, which are indeed very complex and multidimensional as advocated by the reviewer.

### 2019-08-31 Response_to_Reviewers.tex
Update on Overleaf.

--- 2019-08-14 Response_to_Reviewers.tex
@@
   \item B.1.
-We have added a discussion of on how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities).
+We have added a discussion on how segregation indicators can themselves be linked with geometric structure (which can typically happens if an indicator give the same weight to cells of different densities).
 This is a limitation of the method as it is but also an avenue for promoting stronger indicators in empirical research.
@@
 
- % \comment{in discussion, say that indeed the indicators can themselves be influenced by geometric structures (cf.
-comment B) but on the processes, there is nothing obvious about the influence of structures on behavioural spatial patterns (if you imagine a model with no interactions between agents for example) + it's a difference between the model and its potential addition to the real systems.
-cf.
-B}
-
 	\medskip
@@
      $\rightarrow$ This points suggest that our work can inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010.
-We added a paragraph on this in the discussion discussion, and suggested how our approach allows to go back to questioning empirical cases from the model and questioning our concepts ('neighbourhood').
+We added a paragraph on this in the discussion, and suggested how our approach allows to go back to questioning empirical cases from the model and questioning our concepts ('neighbourhood').
 

### 2019-08-31 paper.tex
Update on Overleaf.

@@
 However, initial spatial conditions are usually not modified systematically in socio-spatial models, thus leaving unexplored the effect of initial spatial arrangements on the interactions of agents with one another as well as with their environment.
-In this paper, we present a method to assess the effect of some initial spatial conditions on simulation models, using a systematic geometric structures generator in order to create density grids with which socio-spatial simulation models are initialised.
+In this article, we present a method to assess the effect of variation of some initial spatial conditions on simulation models, using a systematic geometric structures generator in order to create density grids with which socio-spatial simulation models are initialised.
 We show, with the example of two classical agent-based models (Schelling's models of segregation and Sugarscape's model of unequal societies) and a straightforward open-source work-flow using high performance computing, that the effect of initial spatial arrangements is significant on the two models.
@@
 Unfortunately, the first condition is not always met, and the second is seldom even mentioned.
-Various contribution have been made on ad hoc models to test for the influence of geometry on simulation outputs \citep{lilburne2009sensitivity, sun2007traffic}, but without providing a generic framework fitted for most agent-based models.
+Various contributions have been made on ad hoc models to test for the influence of geometry on simulation outputs \citep{lilburne2009sensitivity, sun2007traffic}, but without providing a generic framework fitted for most agent-based models.
 This paper aims to fill a methodological and conceptual gap, which is a systematic testing of the sensitivity of a model's outcomes to its initial spatial conditions.
-To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model.
+To demonstrate the genericity of our approach, we develop two applications with classic simulation models commonly used as case studies for comparing and aligning simulation models \citep{Axtelletal1996, wilensky2007making}: \citet{schelling1971dynamic}'s model of segregation and \citet{EpsteinAxtell1996}'s Sugarscape model of unequal societies.
 
@@
 Finally, although this may seem obvious, cities are not regular grids, and the distribution of density (of jobs, residents, buildings, etc.) is far from isotropic, even in sprawled cities.
-On the contrary, there is a significant diversity in the way people, activities and structures are distributed in cities.
+On the contrary, there is a significant diversity in the way people, activities and structures are distributed within cities.
 In Europe for example, \citet{LeNechet2015} quantifies and classifies six broad types of residential density distributions.
-However, most socio-spatial models, especially cellular automata, still represent cities as uniform grids of isotropic density.
+However, most socio-spatial models, especially cellular automata, still represent cities, hence geometric support of spatial interactions, as uniform grids of isotropic density.
 Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents tend to be approximated by a constant density \citep{arribas2014diverse}, although previous research shows that it is computationally and methodologically feasible to use accurate locations in a simple model such as Schelling's \citep{benenson2002entity}.
@@
 
-In this paper, we aim to provide an operational framework for studying the influence of geometric structures of space on the results of a simulation model representing a sociospatial system.
-Following \citep{Jessop2008} {\bf Sociospatial systems} are understood here as groups of social agents whose behaviour is constrained by their position in geographical space, in the empirical world.
+In this article, we aim to provide an operational framework for studying the influence of geometric structures of space on the results of a simulation model representing a sociospatial system.
+Following \citep{Jessop2008} {\bf Sociospatial systems} are understood here as groups of social agents whose behaviour is constrained by their position in geographical space.
 The aim of the geosimulation models we are interested in is to represent, simulate and explain the dynamics of these systems.
-In the sociospatial systems as well as in the models representing them, agents are in {\bf interaction} among themselves and with their geographical environment.
+In the sociospatial systems as well as in the models representing them, agents are in {\bf interaction} with other agents and with their geographical environment.
 The {\bf geometric structure} of the environment defined at the initialisation of the model and its influence on the model outputs is our object in this paper.
@@
 The purpose is two-fold: (i) to test the robustness of simulation results to small variations of generator parameters and (ii) to study the non-trivial effects of typical categories of geometric structures (monocentric \textit{vs.} polycentric for example) on the results of a given model.
-Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but hardly implemented and achieved in previous studies to our knowledge.
-In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape as a second one.
+Our approach allows for a systematic comparison of several aspects of the spatial configuration problem, which have been suggested by \citet{filatova2013spatial}, but to the best of our knowledge hardly implemented and achieved in previous studies.
+In particular, it is applied to the effects of urban form on simulation results, using Schelling's model as a first case study and Sugarscape model as a second one.
 
@@
 \subsection{Spatial processes}
-In 'real-life', different spatial configurations of people in a city tend to be associated with different distributions of income, carbon emissions, educational outcomes, etc.
+Several empirical studies emphasize the statistical correlations between spatial configurations of people in a city and different distributions of income, carbon emissions, educational outcomes, etc.
 For example, \citet{wheeler2006urban} shows that, in the US, sprawling cities are more unequal than their compact counterparts with respect to income.
@@
 Similarly, in terms of pollution for example, \citet[p.173]{schwanen2001travel} show that \textit{``deconcentration of urban land uses encourages driving and discourages the use of public transport as well as cycling and walking''}.
-These effects of geographical space on social systems correspond to processes to be included in geosimulation models.
+These effects of geographical space on social systems correspond to processes to be included in geosimulation models, as a way to disentangle sources of variation arising from socio-spatial processes and from initial configuration of the geometry of space.
 
@@
 Our original contribution is to tackle this type of spatial effects.
-In this paper, we present an operation framework which allows to systematically measure the impact of the initial geometric structures on the aggregate behaviour of simulation models.
-We illustrate the potential generalisation of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape.
+In this paper, we present an operational framework which allows us to systematically measure the impact of the initial geometric structures on the aggregate behaviour of simulation models.
+We illustrate the potential generecity of our approach by running two distinctive agent-based models: Schelling's model and Sugarscape model.
 
@@
 In addition to the usual protocol (upper branch of the figure), which consists of running a model $\mu$ with different values of its parameters, we introduce a density grid generator which depends on its very own set of parameters and feeds the model $\mu$ at initialisation (lower branch).
-We call these parameters $\gamma$ parameters to distinguish them from the standard parameters of the models (called $\mu$ parameters) and to highlight their higher position in the simulation process.
+We call these parameters $\gamma$ parameters to distinguish them from the standard parameters of the models (called $\mu$ parameters).
 The resulting configurations \emph{can be} clustered into qualitative types of spatial patterns.
@@
 Following \citet{Gauvinetal2009}, we define a phase diagram as the vector of final aggregated model outputs considered as a function of model parameters.
-We have as many phase diagrams than we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment).
+We have as many phase diagrams as  we have spatial grids, which makes a qualitative visual comparison not realistic (with around 50 different spatial configurations for each model experiment).
 A solution is to use systematic quantitative procedures to compare them to a reference case.
@@
 Phase diagrams are denoted by the same function $\mu$ indexed by the generator parameters $\vec{\gamma}$, which capture the spatial configuration (in practice these can be parameters of an upstream model to generate the configuration, or a description of the configuration itself).
-We choose to compare the variability of each phase diagram to the variability between them.
-We take therefore a basic adimentional ratio measure, given formally in the case of a one-dimensional phase diagram by
+We choose to compare the inner variability of each phase diagram to the variability between them.
+We take therefore a simple adimentional ratio measure, given formally in the case of a one-dimensional phase diagram by
 \begin{equation}\label{eq:phase-distance}
@@
 These four dimensions were chosen as complementary descriptors of spatial organisation at urban or metropolitain level - see also \cite{tsai2005quantifying} and \cite{schwarz2010urban}.
-They account for : the extend to which population is clustered in a central city, with two complementary indicators : Moran index (called "centrality") and distance between individuals ("concentration), the extent to which density grid values are similar or contrasted, regardless of location, with two complementary indicators : entropy of distribution ("homogeneity") and rank-size slope ("hierarchy").
+They account for : the extend to which population is clustered in a central city, with two complementary indicators : Moran spatial autocorrelation index (called "centrality") and distance between individuals ("concentration), the extent to which density grid values are similar or contrasted, regardless of location, with two complementary indicators : entropy of distribution ("homogeneity") and rank-size slope ("hierarchy").
 
@@
 It yields a subset of 170 grids matching empirical densities, which constituted our set of different initial spatial conditions.
-These are further clustered into three classes of morphology (figure \ref{fig:densityTypes}): compact (e.g.
-Vienna), polycentric (Liege) and discontinuous (Augsburg).
+These are further clustered into three classes of morphology (figure \ref{fig:densityTypes}) that we label 'compact' (e.g.
+Vienna), 'polycentric' (Liege) and 'discontinuous' (Augsburg) after \cite{LeNechet2015}.
 This clustering allows to evaluate the non-trivial effects of a meaningful urban form on simulation results.
@@
 In sprawled spaces, inequalities are thus fostered by a lack of minimal local resources, whereas population will drive inequality in concentrated spaces.
-The process is thus completely reversed depending on the grid chosen to run the model on, which would have significant impacts if one tried to deduce policy from this model.
+The process is thus completely reversed depending on the grid chosen to run the model on, which would have significant impacts if one tried to draw policy recommendations from this model.
 
@@
 Indeed, the segregation indices are likely to be zoning, size or population density dependant \citep{Wong1997,ReardonOSullivan2004}.
-Generating point data rather than an grid could help overcoming these issues, and appears as a promising direction for future research.
+Generating point data rather than a grid could help overcoming these issues, and appears as a promising direction for future research.
 This methodological claim is also interesting from theoretical and empirical views as point data may be used to construct \textit{egohoods} \citep{Hippetal2013}, i.e.
@@
 Geography and the social sciences in general have been strongly influenced by physics in the last decades, that beside their highly enriching impact~\citep{o2015physicists}, may have softly imposed strong assumptions such as homogeneity and isotropy of space in basic models.
-We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity the for social sciences to build their own stream of methodologies in the modelling domain.
+We believe that a renewed approach on the role of space as we proposed, in other terms insisting that \emph{space matters}, is an opportunity for social sciences to build their own stream of methodologies in the modelling domain.
 
 This relates to relations between empirical, conceptual and modeling dimensions of quantitative research in social sciences, as coined by \cite{livet2010ontology}.
-In our case, a contribution in the modeling domain aiming at extracting further knowledge on model behavior (sensitivity to geometrical context), may lead to questioning theoretical concepts and empirical definition on which the model was based, such as for example the meaning of neighborhood, hence geometry of space.
-
-% to inform the interactions between empirics, concepts and models as discussed by Livet et al., 2010, we added a paragraph in conclusion, discussing how our approach allows to go back to questioning empirical cases from the model and questioning our concepts (for example how ’neighbourhood’ is modelled since Schelling compared to the meaning of neighbourhood in empirical/theoretical social sciences)
+In our case, a contribution in the modeling domain aiming at extracting further knowledge on model behavior (sensitivity to geometrical context), may lead to questioning theoretical concepts and empirical definitions on which the model was based, such as for example the meaning of neighborhood, hence geometry of space.
 

### 2019-09-02 paper.tex
submit_v2

@@
 
-%\comment[FL]{le titre de l'article ce serait presque "geometry" matters? non?} bof ca parle moins
-
 \begin{abstract}
@@
 \parano{}
-
-%%%%%%%
-%% Highlights
-%
-% - We investigate the sensitivity of geosimulation models to initial spatial conditions.
-% - We generate stylized population density grids.
-% - We assess the sensitivity of two classical agent-based models, Schelling's model and Sugarscape's model.
-% - Strong quantitative and qualitative effects of spatial initial conditions are shown for these models.
 
@@
 
-Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \cite{EpsteinAxtell1996}.
+Sugarscape is a model of resource extraction which simulates the unequal distribution of wealth within a heterogenous population \citep{EpsteinAxtell1996}.
 Although it "is designed to study the interaction of many plausible social mechanisms" \cite[p.125]{Axtelletal1996}, we refer in this paper to the first (and simplest) version of the model, where "processes allow its agents to look for, move to, and eat a resource ("sugar") which grows on its $[...]$ array of cells".

### 2019-09-02 Response_to_Reviewers.tex
.

@@
   \item \textit{Then, in paragraph 1.3, the complexity of systems with respect to segregation is also problematically described.} \ldots
-  % Emergent behaviour that may lead to segregation is first and foremost a social process, taking social interactions at the local scale (income, housing costs, etc.) and/or at the macro scale (laws, rules, economic system, cultural attitudes, etc.) into account.
-These interactions can then be associated with the material and relational dimension of geographical spaces.
-The question remains: what is the role of geographical systems?
-In this respect, the paper offers some relevant clues such as the MAUP problem or the “uncertain geographic context problem”.
 
   $\rightarrow$ see comment A.
-As for the role of geometric structures in the production of segregation, Banos (2012) has shown that with the same Schelling mechanisms and the same parameter values but different network structures of neighbourhoods (scale-free, regular, etC.) the segregation observed at the end of simulation varied significantly between network structures.
+As for the role of geometric structures in the production of segregation, Banos (2012) has shown that with the same Schelling mechanisms and the same parameter values but different network structures of neighbourhoods (scale-free, regular, etc.) the segregation observed at the end of simulation varied significantly between network structures.
 This is the kind of approach we systematise here, although with a spatial grid instead of network structures.
@@
   \item \textit{The approach is reductionist and has its strength in reflecting growth at an abstract level nevertheless.
-The concluding remark in paragraph 5.1 “With the Schelling experiment, we found … that polycentric and discontinuous cities appear systematically more segregated than compact cities …” is elusive.
+The concluding remark in paragraph 5.1 ``With the Schelling experiment, we found \ldots that polycentric and discontinuous cities appear systematically more segregated than compact cities\ldots'' is elusive.
 Either it is a trivial fact, simply because polycentric cities have more centres and thus, by nature, more opportunities from which segregation can originate.
@@
 We thank reviewer 2 for highlighting this limitation.
-  % \item \comment[JR]{this is a relevant remark but far beyond the scope of the paper, as an extension of what we did but with perturbation of real datasets / sensitivity to noise - Julien wanted to begin working on that actually and it should be integrated into spatialdata at some point - we also are supposed to implement such a perturbator to test the 4city model}
 
@@
   \item\textit{In this respect, I would encourage authors to further develop their closing section, which now does not stimulate readers to imagine implications of authors' findings and message to their own research.} \ldots
-  %I would expand this section to provide a more detailed discussion on implications for model makers, experts on social geography and urban scholars who work with ABM, non-specialist in geography who would like to base their models in more proper geographical constraints.
-And perhaps, authors could suggest future line of research that could show-case their argument.
 
@@
 
-%I appreciate that models are fully available to readers. 
-%\item We do too!
-
 \end{enumerate}
-
-%\makeletterclosing
 
